### 2020-01-18 18-whitepaper/article.tex
Init

\documentclass[a4paper, twocolumn]{article}
\newcommand{\papertitle}{Whitepaper: Smarter Workflow Processing with the Next Generation Data-Centric Interfaces}

\usepackage[a4paper, margin=2cm]{geometry}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage[english]{babel}
\usepackage[colorlinks=true,urlcolor=red]{hyperref}
\usepackage{url}
\usepackage{cleveref}
\usepackage{todonotes}

\usepackage{multicol}
\setlength{\columnsep}{1cm}

\usepackage{fancyhdr}
\fancyhead{}
\fancyfoot{}
\fancyhead[l]{\papertitle}
\fancyhead[r]{\includegraphics[height=1em]{ngi-logo}}
\fancyfoot[r]{\thepage}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{1pt}
\renewcommand{\footrulewidth}{1pt}

\usepackage{titling}
\pretitle{\begin{center}\Large\bfseries}
\posttitle{\end{center}\vskip 0.5em}

\graphicspath{{./assets/}}

\title{\papertitle}

\author{Julian M. Kunkel
  \textit{University of Reading}
	\and
  Chris Hoffman
  \textit{National Center for Atmospheric Research}
  \and
  Andy Poling
  \textit{Kove}
  \and
  Jay Lofstead \textit{Sandia National Laboratories}
}
\date{\today}

\begin{document}
\maketitle
\thispagestyle{fancy}

\section*{Abstract}
The efficient, convenient, and robust execution of data-driven workflows and enhanced data management are key for productivity in scientific computing and computer-aided Research Development and Engineering (RD\&E).
Traditionally in HPC the concerns of storage and compute are separated and optimized independently from each other, and, moreover, independently from the needs of end-to-end user workflows that ultimately lead to insight.
While there are good reasons for this strategy, there are workflows that run also efficiently with alternative organizations.
For example, Big data tools integrate compute and storage capabilities into a holistic solution demonstrating the benefit of tight integrating for data-driven workflows.
As workflows become increasingly complex and blend beyond data centers while at the same time storage hierarchies becomes deeper, the community investigates means to re-organize storage access to fully utilize such environments.

Even within a single data center, utilizing homogeneous storage and compute infrastructure efficiently is complex for experts and needs a manual definition.
The efficient management of data and compute capabilities in a heterogeneous environment, however, is an unresolved question as the execution of individual tasks from workflows may benefit from alternative hardware architectures and infrastructures.

In this white paper, we describe the vision for the Next-Generation Data-Centric Interface (NGI) initiative that aims to tackle the aforementioned storage and compute challenges in a holistic approach.
Within the NGI consortium, we fuse state-of-the-art concepts into a revolutionary interface by rethinking how data-driven I/O should be done.
Basically, we believe workflows composed of data, compute and communication intensive tasks should drive the interfaces and hardware configurations to best support these programming models.
From the user perspective, NGI aims to provide an abstraction to data-driven processing and data management.
The NGI interface ultimately will increase the opportunity of implementations for smarter scheduling of compute and storage in heterogeneous environments considering characteristics of workflows and data.
From the system perspective, the vision of NGI aims to derive an execution plan that utilizes the available hardware and software infrastructure within and across data centers but also on small-scale systems.

\section{Introduction}

High-Performance Computing harnesses the fastest hardware components to enable
the execution of tightly coupled applications from science and industry.
Typical use-cases cover the numerical simulation of physical systems or the analysis of large-scale observational data.
While the hardware of HPC systems is on the bleeding edge, the software landscape of the infrastructure is lacking behind modern concepts found in domains like cloud computing.
There are various reasons for the lack of innovation, e.g., the market capitalization is limited; many HPC applications have a life time of sometimes several decades\footnote{For example, CTH at Sandia is more than 30 years old, but it has evolved over the years as hardware has changed.
It is not uncommon that codes take 10 years to write and receive a first optimized setup.
Then they run for another 30 years with constant evolution based on platform changes and functionality expansions.}; the goal in HPC is to get science done and not to have modern and user-friendly systems.
Therefore, software developed to target the HPC market is rarely used outside data centers; while there is no reason to use developed applications outside this scope, software to aid developers or for managing HPC infrastructure suffers the same cause.
Nowadays, HPC adapts infrastructure software from the enterprise and cloud market as much as possible; at best, adapting (stretching) already existing software packages to the needs of HPC.

From the perspective of big data and cloud solutions, the state-of-the-art in the HPC software ecosystem is perceived as conservative and outdated at least; for example, HPC still uses batch processing and dedicates nodes to the execution of individual applications which were the first concepts implemented at the dawn of computing.
Similar to writing code in assembler, often it is argued that HPC treats user-friendliness and manageability to obtain the best possible performance.
However, as the complexity of systems increases and various alternative storage and compute technologies provide unique characteristics, it is barely possible to define optimal data-driven workflows using low-level constructs for the following reasons:

\begin{itemize}
  \item Compute focus: HPC considers compute, storage, and network as separate resources and focuses on the compute aspects, e.g., I/O requires to retrieve/store data from a compute node.
  While this provides a wonderful separation of concerns, it disregards the opportunity to process data closer to storage or between applications.
  \item Complexity: a data-driven workflow is highly complex and may benefit from using a heterogeneous set of compute and storage technology at the same time.
  Knowing the capabilities, interfaces, and performance characteristics of individual components is mandatory to make best use of them.
  \item Portability: moving from one infrastructure to another or even when modifying the infrastructure requires to rethink and manually adapt the workflow as the characteristics of the overall environment changes.
  \item Cost-efficiency: it is time consuming to utilize low-level constructs often only increasing performance slightly; also, the cost of labor must be taken into account.
  From the perspective of scientists and engineers, the time-to-solution from experimental design to insight matters and not the cost-savings for the public data center.
  \item Conservative users: users are reluctant to change their workflows for the reasons that older workflows are proven to be correct and it is labor intensive to manually adapt workflows.
\end{itemize}

For these reasons, workflows are typically encoded manually and run suboptimally on HPC systems.
Additionally, users of public data center services supply limited information about their experiments, typically the level of detail is like "I need X CPU hours and Y storage capacity" and they have not to proof their ability to use these systems efficiently.
Still, the systems these workflows run on cost up to 100 million of dollars.
Compare this to experiments conducted in physics like the Large Hadron Collider (LHC) at CERN or the Square Kilometer Array (SKA); execution of complex experiments may need years of planning involving the scientists.
While it is important to keep some freedom to conduct ad-hoc experiments, we raise the question what could be achieved by systems and operators knowing workflows in more detail.
We argue that knowledge about workflow execution and overall experimental design helps to
optimize execution of workflows and additionally increase the performance, throughput and  cost-efficiency of the environments, therewith, providing an incentive to users and data-centers that cannot be neglected.

\medskip

With the Next Generation Interfaces initiative, we aim to define a high-level interfaces that, firstly, meets the needs of users and data centers, and, secondly, that allows implementations to exploit the capabilities of current and future storage and compute landscapes.
The remainder of this article is structured as follows: \\
\todo[inline]{TODO}

In \Cref{sec:background}, we provide relevant background information for workflows.
\Cref{sec:challenges}
\Cref{sec:ngiVision}
\Cref{sec:ngWorkflows}
\Cref{sec:community}
\Cref{sec:conclusions}

\section{Workflows}
\label{sec:background}
This section provides an introduction to workflows that solve computational problems, the mandatory steps to encode and setup tasks and describe data, and the relevance of  heterogeneous infrastructure to run workflows.
We focus on high-level considerations that illustrate mandatory concepts.

%Kubernetes

\subsection{Workflow Description}

To solve a computational problem, we can consider to describe the necessary steps from the start to the insight as a workflow.
Such a workflow description must comprise of the individual data processing steps, the expected  products, together with dependencies between the steps and to input.
The nodes can be annotated with characteristics of the individual steps;
data can be annotated with metadata describing its content, the expected data life cycle information, e.g., the value of data and how long should it be kept; tasks can be annotated with the information needed to start processing or with performance models that bound the  expectations for resources (time, memory, storage capacity).

An example execution of a workflow is illustrated in \Cref{fig:workflow} as a graph (we have not shown characteristics and annotations).
Arrows indicate dependencies between nodes that represent tasks and data.
In the example, Task\,1 needs two datasets to perform its work, it directly communicates with Task\,2 and produces Product\,1.
Most of the workflow can run automatically except the manual quality control of the products and the final data usage of Product\,3 at the end.
The last step represents the typical uncertainty of data reuse, when this workflow is created it is unclear how Product\,3 will be used further, it is apparent that it will be.
Further workflows that use the data may be created similarly and use the product as input.

Note that a realistic workflow will be much more complex and involves not a single parallel job execution but many to generate a robust answer to the research question.

\begin{figure}[b]
  \centering
  \includegraphics[width=0.75\columnwidth]{workflow}
  \caption{Example High-Level Workflow without Characteristics}
  \label{fig:workflow}
\end{figure}

\subsection{Task Definition}

The high-level interaction of a task with its dependencies is generally specified in the workflow.
A compute task is encoded in a programming language by a developer\footnote{We often treat developer and HPC user as synonyms as a fraction of HPC users are developers for parallel applications to some extent.}.
A team of developers encodes the algorithm that solves the problem -- e.g., in many cases,  scientists that generate a numerical simulation implement the model programmaticially.
Often they use existing libraries that provide a piece of functionality or infrastructure.
In the design phase of the program, a parallel programming paradigm and suitable data structures that allow efficient data processing are chosen.
In the world of parallel computing, the developer must also consider strategies to parallelize the code and data processing; this often involves to use APIs and runtime systems that allow the efficient usage of supercomputers.

While most parts of the algorithm can be a tightly coupled simulation, as part of the algorithm, input data must be fetched and data products are generated which are processes that are executed outside of the main simulation.
Typically, before the input data can be used some kind of preprocessing must be applied, similarly, in the output several known post-processing steps may be applied.
Also, the programmer is responsible for making the optimal decisions on how to map data structures from memory to I/O calls and place these files onto storage systems.

We consider that pre/post-processing are part of the data-driven scope of NGI.
Indeed, we claim the programmer should be able to define the input/output on an abstract fashion and connect processing steps as part of the general workflow.

\subsection{Heterogeneous Infrastructure}

The variety of tasks executed by a single workflow may benefit from a heterogeneous storage and compute infrastructure and potentially even span multiple locations.
\Cref{fig:heterogeneous} shows such an environment but focuses on the computation and storage.
A variety of accelerators (GPU, TPU, FPGAs), active storage, in-memory, and in-network computing technologies provide processing capabilities across and beyond an HPC center optimized for certain workloads.
Data centers typically provide more than one storage and compute class with different characteristics to optimize cost and efficiency.
Likewise, storage is deployed data center wide, locally available to individual nodes, or subsets of nodes (e.g., racks); depending on the need, the characteristics range from predictable low-latency (in-memory storage, NVMe) to online storage (SSD, HDD), to cheap storage for long-term archival (tape).

The processing across geographic locations is nowadays mandatory to tame the data volumes of observational or simulation data.
Also, the lines between cloud computing and HPC are blurry, attempts are made to augment processing by flexibly migrate bursts of computation to cloud environments, ingest data from cloud sources or provided to customers.
Consider a simple scenario where sensor data captured is pre-processed (cleaned) locally, then enriched with an additional data source in a small-scale HPC facility or by using fog computing, then transferred to a data center that performs the number crunching and generates data products.
These data products may be made available on cloud storage to customers for individual processing.

\begin{figure}[b]
  \centering
  \includegraphics[width=\columnwidth]{system}
  \caption{Example Heterogeneous HPC Landscape}
  \label{fig:heterogeneous}
\end{figure}

\section{Challenges and Limitations}
\label{sec:challenges}
The current HPC environment faces several challenges and limitations in respect to the execution of data-driven workflows.

\paragraph{Low-level interfaces:}
In a nutshell, a stream of bytes isn't a workable data model.
HPC data management and data-flow processing utilizes low-level interfaces.
Storage relies still on traditional file system objects of directories and files -- provided by the POSIX interface or by object storage.
Underlying storage systems are receiving or delivering streams of bytes, and it cannot make any intelligent decisions about how to optimize data layout for performance, or resiliency.
It deals with data blocks and files and requires instructions on how to select and control available optimization strategies.

In practice, applications build up layers so that the application itself is usually working with the data mediated by a library which provides application specific semantics --- but at the lowest level such a library is limited by the underlying storage APIs (POSIX).
This has performance implications due to synchronization in distributed environments but also severely limits the ability of the underlying storage systems to optimize I/O operations.

Data-driven processing suffers similarly, from the high-level view of processes that are executed on a given compute node.

\paragraph{Explicit workflows:}
Typically, as part of shell scripts, users explicitly encode data processing workflows as a sequence of steps consisting of input, compute, and output.
The compute-intense parts (jobs) and their dependencies are managed by a cluster resource manager that dispatches them.
These loosely coupled workflows require that intermediate results are stored on persistent storage to be used by the next task.
Data transfer, however, needs a dominant fraction of energy compared to computation favoring offloading of computation to storage.

\paragraph{Suboptimal Storage Tiering:}
Organizing the data placement on storage tiers is typically performed manually by the developer/user or via policies.
This leads to suboptimal optimizations, is tedious and error prone.

Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in form of files) on a particular storage system and typically moving data between storage by using scripts.
One limitation here is that decisions about how data structures are mapped and packaged into files are made once by the producing application, and cannot be changed without manual intervention by a downstream application.

A policy system (e.g. burst buffer) aims to simplify the data movement for the user but  typically migrates objects in the coarse granularity of files.
However, the semantical information that can be used by a policy system to make decisions is limited, e.g.,  data location, file extension, age of the file.

Any hard-coded decision has significant implications for performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
In any case, data migration needs resources on the source and the target storage system to copy data.
Thus, it is preferable to direct data to locations where it can be stored and directly used by subsequent workflows; at best coupling the data source producing data with a data consumer to prevent storing intermediate products on persistent storage.

\paragraph{Task and data equality:}

All tasks and data products are handled identically by an HPC system meaning there is no differentiation between individual jobs and data products.
This is problematic as different use cases may not only differ in priority and deadline but as the speed they proceed depends on the critical path of the execution.

Fair enough, batch systems understand a priority for jobs.
However, once a set of nodes is allocated explicitly to a compute job, the job is run exclusively and may utilize shared resources such as storage systems and network as any other job. %-- quality of service features available in hardware are rarely used

This is more sever for data products: a storage system provides a single level of performance and resilience and performance for all use cases.
Theoretically, different fault-tolerance and performance policies can be realized with different storage systems and using management policies.
However, these policies would be coarse-grained and would not account for the priority of the workflow and most importantly the value of data;
the value of data depends on aspects like costs to reproduce the data (can it be re-produced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow.
While value and priority should influence fault-tolerance strategies and imply quality of service for performance and availability, in practice this is not the case as the storage just sees blocks and files.

\paragraph{Manual optimization:}
Utilizing the compute and storage characteristics of a heterogeneous environment as depicted in \Cref{fig:heterogeneous} is challenging at least.
The optimization of a single parallel job already is quite challenging.
To make optimal use of available features, programmers and users are forced to think about technological aspects and terms, e.g., setting low-level MPI hints for file striping, while they understand their workflow.
Similarly, an efficient data placement is complex.

\medskip

An explicit workflows combined with manual optimization is the root-cause of subsequent problems.

\paragraph{Access pattern intransparency:}
As access patterns are primarily implicitly defined in application code and embedded in workflows, the understanding and debugability of behavior is limited.
Firstly, there are many alternatives ways of programming data-intensive computation to achieve the identical data products.
Secondly, many storage and compute interfaces are low-level.
Thus, data centers have often limited insight into the access patterns as well.
Moreover, some users are abusing the storage system in non-obvious ways, for example to exchange notifications.
Together with the possibility to observe access patterns at low-level, these issues lead to the situation that understanding and improving of access patterns become difficult.
Limited insight into behavior leads to suboptimal tuning.

Manual and hard-coded workflows cannot handle changes in the environment and are error prone.
Therefore, users must be able to express their workflows in an abstract fashion that allows the system to generate (near-)optimal execution plans and monitor their execution.

\section{NGI Vision}
\label{sec:ngiVision}

Our vision for Next Generation Interfaces is a new approach for data management and data-driven computation.
%\paragraph{Data Model}
NGI tackles the challenge of storage and data-flow computation in an holistic way accepting that data storage is only aiding the goal of generating insight.
Besides providing a typical abstract data model, NGI treats domain metadata\footnote{Therewith, we mean non-technical information about the data, for instance, the name of an experiment.}, workflows, and information lifecycle management (ILM) as first-class citizens allowing the storage to understand these concepts.
This user information is exploited by smart components of NGI which aim to utilize capabilities of  heterogeneous system landscapes.

To describe the features and benefits of the approach further, we sketch the user and system perspective in the following, as this highlights how we believe we should work in the 21st century.
The vision covers not only HPC but spans cloud, enterprise, and personalized IT.

\subsection{User Perspective}

\newcommand{\bnf}[1]{\textless #1\textgreater}

From the user's perspective, the interface and system matches the core requirements and expectations including reliability, availability, and usability.

\paragraph{Data model:} the data model is based on state-of-the-art solutions like HDF5+X where X indicates an extension for ILM, workflows and processing.
However, it abstracts from data localization and low-level concepts like files and directories and puts its focus on data organization and data structures.
The integration into applications is shown in \Cref{fig:ngilayers}.
Applications may utilize the NGI API directly, or utilize existing middleware\footnote{Which won't be able to utilize the full set of features if not extending its interface.}.
From the perspective of a developer of middleware or an application like web publishing, NGI provides a common feature set that reduces the amount of code replication inside domain-specific solutions.

\paragraph{Namespace:} data is accessed using information from the domain metadata, i.e., similar to using methods of data indexing services.
The system won't relying on a hierarchical namespace but a dynamic mapping to the POSIX namespace can be provided to enable compatibility for POSIX applications.
For example, a user may decide to see a hierarchy like an experiments \bnf{Date}/\bnf{Name}/\bnf{Timestep}/\bnf{Variable}.

\paragraph{Workflow specification:} the user defines a workflow describing the intentions of data manipulation and the data lifecycle including manual (user) involvement in the data processing.
This also includes the definition of constraints, for instance, deadlines for the generation of the final data products (insight).
Workflows can be defined using arbitrary tools including Python.
Note that a user does not provide a concrete mapping to system or infrastructure but expects that the system derives an optimized \textit{execution plan} that covers \textit{heterogeneous infrastructures}.
It does not matter where data is processed and how and where data is stored but only that the desired result is computed with all constraints met.
Workflows and data lineage of data products are recorded and can be inspected by the user upon demand allowing to reproduce and validate correct execution of experiments.

\paragraph{Behavioral contracts:}
Besides the core characteristics, users expect the execution of their workflows under constraints (including cost and time constraints).
The system is expected to always show the expected characteristics and shield the user from system-specifics and undesired events (like faults or data loss).
The characteristics may improve over time, for instance, by expanding hardware capabilities, but the user won't have to change their workflow specification.
Still, for expert users, interfaces for introspecting system state and decision making are provided.

\paragraph{Prescriptive analysis:} a toolset for users and administrators allows to explore scenarios.
Similarly, to an SQL Explain statement, this provides an unprecedented level of introspection, for instance, to predict the runtime or execution costs of a workflow on a given system.
Sharing the workflow description with vendors allows them to conduct predictions for runtime and costs for their solutions.
The ecosystem may also prescribe alternative systems to use in an open environment, e.g., this part of the workflow should be run on data center A and this part on cloud provider B.

\subsection{System Perspective}

An implementation of the NGI API exploits the user-supplied information and system characteristics intelligently to meet the user expectations and smartly use heterogeneous compute and storage landscapes.
It is expected that machine learning techniques will be applied to serve the individual aspects and evolve the systems over time.

\paragraph{Awareness:}
the system knows and understands user metadata and the overarching workflows (with the user intentions).
It is self-aware, knowing the nominal characteristics of the hardware and software components, constantly comparing the system and workflow state with the expectations to perform mitigating actions to heal deviations.
At best, each individual hardware component would ship with a behavioral model that can be integrated hierarchically up to a global system model.
With behavioral models we not only mean performance models but reliability and cost-models as well.

\paragraph{Resource management:}
the system is in charge to manage and provision the available hardware resources autonomously.
Ultimately, it aims to utilize the individual characteristics of heterogeneous compute and storage landscapes by utilizing intelligent scheduling.
Data placement and localization is beyond mere tiering strategies, parts of data may be placed on certain storage technology or replicated in different representations across storage characteristics.
Therewith, serving different access patterns and workflows at the same time.
Naturally, the system will replicate (expand) upon the availability of storage space while reducing redundancy under shortage of space.

\paragraph{Liquid computing:}
the system is in charge to manage and provision the available hardware resources autonomously and to assign compute and storage resources to tasks.
Therewith, it may use, e.g., GPU, CPU, and in-network computing capabilities to serve pieces of a single workflow.
Moreover, as workflows stretch across data centers and over cloud and edge, an environment that schedules and dispatches work and storage efficiently is key.
We call this concept \textbf{liquid computing} as the compute resources are dynamically provisioned and assigned; while data flows through the system, it is processed.
Naturally, processing closer to the data sources is likely to improve system throughput.
The programming of the processing will be based on a high-level functional language providing primitives for data-flow computing and compiled into a virtual machine or machine code when the code is to be dispatched.
The capabilities of different hardware components may be limited realizing only a subset of the language which must be taken into account by the scheduler.

\paragraph{Smart scheduling:}
supplied with user workflows and their characteristics, the system derives execution plans for the computation and make data placement decisions that aim to utilize the available hardware resources.
As NGI is in charge for both, it may decide to trade computation vs. storage capacity or other characteristics (e.g., costs, energy consumption); for example, enabling automatic recomputation of intermediate states utilizing virtualization and container technologies.
Similarly, the scheduler may decide to couple subsequent steps of a workflow directly without storing intermediate data products on persistent storage.
Data might be replicated but enables the system to rerun parts of the workflow in case of a data loss.

The automatic scheduling is key to efficient processing and will allow constant improvement without forcing user to adjust their workflows.
In most cases, individual resources are not used exclusively by a single task but QoS methods ensure that constraints of workflows are met.
Note that this significant change for HPC environments, as compute resources are typically dedicated for individual jobs.

\paragraph{Extensible ecosystem:}
the hardware and software landscape will be open allowing to integrate solutions from different hardware and software vendors.
Systems connected may use different interfaces including traditional POSIX interfaces, NoSQL, or distributed hash tables.
Imagine that a certain class of workflows performs inefficiently on a given infrastructure indicating a crucial processing or storage capability is missing.
Adding new capabilities or resources to the system will allow the scheduler to use them in subsequent user-workflows speeding up the processing; this should be as easy as adding an appliance to a business environment.

\begin{figure}[b]
  \centering
  \includegraphics[width=0.75\columnwidth]{layers-ngi}
  \caption{Layers with NGI}
  \label{fig:ngilayers}
\end{figure}

\section{Conclusions}
\label{sec:conclusions}

Cloud definition of desired state and contracts.

Indeed many research prototypes address subproblems
* But not all aspects together
* Competing approaches; the standardization we describe here does not compete!

In this paper, we purposely do not mention too many of the ongoing RD\&E in storage and compute.

To wrap up, NGI provides a high-level view on data and processing on heterogeneous platforms while shielding the user from execution details.

Open ...

\includegraphics[width=2cm]{ngi-logo}

\noindent\url{https://ngi.vi4io.org}

\end{document}

### 2020-01-18 18-whitepaper/use-cases1.tex
Init

\documentclass[a4paper, twocolumn]{article}
\newcommand{\pversion}{0.5}
\newcommand{\papertitle}{NGI Use Cases: Smarter Workflow Processing in Climate/Weather}

\usepackage[a4paper, margin=2cm]{geometry}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage[english]{babel}
\usepackage[colorlinks=true,urlcolor=red]{hyperref}
\usepackage{url}
\usepackage{cleveref}
\usepackage{todonotes}

\usepackage{multicol}
\setlength{\columnsep}{1cm}

\usepackage{fancyhdr}
\fancyhead{}
\fancyfoot{}
\fancyhead[l]{\papertitle~(Version: \pversion)}
\fancyhead[r]{\includegraphics[height=1em]{ngi-logo}}
\fancyfoot[r]{\thepage}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{1pt}
\renewcommand{\footrulewidth}{1pt}

\usepackage{titling}
\pretitle{\begin{center}\Large\bfseries}
\posttitle{\end{center}\vskip 0.5em}

\graphicspath{{./assets/}}

\title{\papertitle}

\author{Julian M. Kunkel
  \textit{University of Reading}
	\and
  Chris Hoffman
  \textit{National Center for Atmospheric Research}
}
\date{Version: \pversion; \today}

\begin{document}
\maketitle
\thispagestyle{fancy}

\section*{Abstract}
The efficient, convenient, and robust execution of data-driven workflows and enhanced data management are key for productivity in scientific computing and computer-aided Research, Development and Engineering (RD\&E).
As workflows become increasingly complex and blend beyond data centers while at the same time storage hierarchies becomes deeper, the community investigates means to re-organize storage access to fully utilize such environments.

In the NGI Forum, we believe workflows composed of data, compute and communication intensive tasks should drive the interfaces and hardware configurations to best support these programming models.
In this whitepaper, we describe how workflows for climate and weather could benefit from higher-level abstraction for data flows.
The document serves the purpose to motivate subsequent NGI activity, the implementation of described features will likely differ from the sketched solutions.

\section{Next Generation Workflows}
\label{sec:ngWorkflows}

This section illustrates how workflows could be run and optimized by smart runtime implementations.

\subsection{Climate/Weather in 2020+}

This workflow gives an idea how the climate and weather community could utilize emerging concepts in 2020+.
A generic overview of the workflows for climate and weather is given in \Cref{fig:climateWeather}.

Prediction of the weather in the near future (hours, known as nowcasting; days or weeks -- seasonal) is performed by executing workflows consisting of several processing steps. The execution of a workflow engine that orchestrates the individual execution steps is not explicitly included in the workflow description.

For weather prediction, these workflows are schedules on a regular basis, e.g., a prediction is performed every four hours.
The prediction is based on observational data and outputs data products that are processed further and distributed to customers.

\subsubsection{Execution of a weather workflow}

\begin{figure}[b]
  \includegraphics[width=\columnwidth]{climateweather-workflow}
  \caption{Future Climate and Weather Workflows}
  \label{fig:climateWeather}
\end{figure}

First, data sources providing up-to-date data (such as near-real-time satellite sources and weather stations, I1) are ingested into a local permanent repository (S1). These data may be supplemented by the increasing number of IoT and mobile sensor devices (I2) which data may be stored in a short-term storage (S2) with a limited life-time.
A weather service utilizing a HPC site ingests this data onto a local repository (S3). During this process, observational data is cleaned and reduced, e.g., by reducing resolution (subsampling).
An ensemble of forecasts (several parallel simulation applications) are started using a subset of the latest weather information -- typically each of the applications differs only in how the simulation is initialized.
They start with a warm up period using data from, e.g., 24 hours in the past. While a forecast runs it may use the data from I1/I2 to augment the simulated data with actual observations until the model time and the actual time matches and prediction of future weather starts.
While the ensemble of forecast runs, it computes various relevant output variables that may be stored in a transient data buffer (S4) and/or perform data transformation and analysis in-situ/in-transit. For example, a computed variables may be output, visualized and analyzed further by users even as the simulation continues. To identify patterns (events of interests for meteorologists), machine learning may be applied in the processing pipeline. Identified events may trigger further actions on the visualization, triggering additional data outputs, and controlling data reduction. To reduce the data volume for the ensemble, a data reduction selectively chooses data from the ensemble to preserve (on the repository S3, just before it is stored in the archive S5); this may be a typical run/minimum/maximum or runs that contain certain events as identified from the machine learning.
The generation of data products is either known beforehand, and calculated using a workflow during the ensemble run, or calculated post-mortem.
The former case is typical for weather applications. In the latter case, data products can be created offline from the stored output of the forecasts -- this allows users to request new or modified data products from the simulations via the web interface.
Machine learning can be applied in this workflow to identify relevant patterns in the data, supervised training involves the scientist selecting the phenomenon of interest presumably interactively using data visualization. The system should be able to quickly build a model and suggest other data with the similar phenomena to reduce the effort of the scientist to browse through all the data.

A selection of data products together are stored on a long-term (permanent) archive (S5).
Scientists over the world typically interact with web services to access data products on the long-term archive or to apply certain post-processing procedures to create new data products. This data is typically cached (S6) both at source before dissemination and by users in their workflow. When a new data product is requested it may actually trigger the recomputation of the data because it is cheaper than keeping all data products on a permanent storage.

Note that the box depicts “HPC-sites” that executes most parts of the workflow, however, not necessarily this must be a single data center. Scientists already work in a distributed fashion, downloading certain data products and performing subsequent data analysis locally, on other remote sites, data centers, or in the cloud. It is expected, thought, that the initial data production is kept on the site where the data is produced as the data volume is reduced significantly in this process.

\subsubsection{Execution of a climate workflow}

Generation of insight for a climate experiment differs from the weather case by several aspects: Firstly, the data (S2) from IOT devices (I2) is not used.
Secondly, in most cases scientists conduct new experiments, i.e., the result is a-prior unknown. Thus, a more exploratory and interactive use is expected.
For model intercomparison projects, once the variables and experiments to investigate are clear, a mass replication (execution) of the use case under different starting conditions takes place. These planned experiments come with less explorative nature, yet diagnosis of abnormal execution behavior may be investigated interactively.
Due to the vast amount of different experiments, the data management features of NGI are useful to simplify the data handling; for instance, by utilizing scientific metadata to specify the difference between different experiments differences between model output can be computed directly.

\subsection{Extreme Data Generation}

This section discusses a use case illustrating how the aforementioned concepts integrate into a smart system.
Consider the case in which a group of users conducts an experiment at large scale (a tightly coupled hero run on an HPC system) to gather new insight about small scale phenomenon.
This experiment generates raw data of a huge volume (e.g., several Petabytes on existing systems).
The user may know certain (diagnostic) post-processing that must be conducted on the data, but since the scientists work at the frontier of knowledge, it is usually unclear which features to look for and which data products to create.
Over a period of several month scientists will analyze a subset of data (e.g., two regions of 10 GiB each) potentially using machine learning to identify the further analysis steps that shall be conducted on the full dataset.
Once this is done, they register the analysis workflow and wait for all data products.

The user specifies the workflow defining the characteristics of the processing and the data and include that the final data product must be retained for 10 years.
The workflow also indicates that the data products are used in manual analysis and this may take several months.
Note that users do not deal with namespace issues but provide scientifically relevant metadata for the raw data and the data products.
At creation time of data products, these are semantically linked to the workflow and may be found when inspecting metadata for the created raw data.

\medskip

We envision the following support of the runtime for this use case:
The workflow description gives the scheduler various options; firstly, the final data product must be accessible for 10 years; it may decide that it is cost-efficient to store the data on a tape archive, a subset of the data region should be stored for in-depth analysis and likewise a low-resolution version of the should be created.
In another environment and depending on characteristics, the scheduler may have decided to not persist the huge data output but preserve a container for the application to allow recomputation -- but we will discuss the former case.

An I/O-enabled resource manager will dispatch the hero run exclusively once there is enough storage capacity available on the archive storage system such as tape and I/O bandwidth is available.
The system may schedule the post-processing, data reduction and subset creation, while  data transfer occurs in-transit on the client node or dedicated compute resources and picks a fast storage system for these products.
That means a data-driven streaming workflow is applied to minimize reading of data during the post-processing workflow.
The raw output is meanwhile directly stored on tape for archival purpose (e.g., as it must be kept for several years due to policies of the funding agency).
The sample data is a replicate of a subset of the raw data's domain and stored for inspection on a storage system supporting good random I/O workloads.
If a user inspects the metadata, one can see that fragments with a subset of data are stored on fast storage and the other parts on slow archival storage.
Once the user registers the final workflow, i.e., requests data processing on the full dataset, the system has to process the vast amount of data on tape.

Depending on the exact workflow, there are opportunities for the system:
Thanks to liquid computing, parts of the workflow can be executed directly on tape drives and the network thinning out the amount of data that needs to be transferred while reading.
Alternatively, we may retrieve the data from archive storage and create a copy on a faster storage system but -- since relevant parts of the workflow access pattern are known -- can reorganize the replicas of the fragment while data is fetched and, thus, optimize I/O for the subsequent accesses of the workflow.

\paragraph{In-Situ Analysis}

A user should be able any time to inspect and analyze data that is currently produced without changing the application code or even knowing a priori that this is an intent.
Let us extend the previous case of the hero run; now the user may decide to visually analyze a subset of data and a low-resolution output of the data while the data is generated.
We assume a node or visualization cluster has been allocated for this purpose.
An NGI enabled visualization tool would be able to query the metadata and search for the experiment currently conducted.
Then the region of interest is selected by the user.
In the best case, some derived data products could be reused but it is expected that the tool submits new tasks (code snippets for liquid computing) to the runtime to perform data analysis and reduction.
The system will create additional I/O streams and run the analysis codes in streaming mode as much as possible while it feeds the output to the application.
This would allow us to perform filter operations on the process creating the raw data.
If the volume of the produced data is too high, a workflow can be registered to create data products on the storage with good random I/O characteristics.
The visualization is then constantly updated according to the NGI semantics.

\section{Data Archival}
This scenario is a more conservative perspective utilizing NGI to replace data archival interfaces. 
This section is not meant to replace current archival systems, but instead of introduce capabilities that may not be present.
It also shows an incremental strategy to improve existing HPC environments.

\subsection{Introduction}
The archive strategy at many centers has been to have a large monolithic storage based system. 
These systems locked into vendor hardware and vendor software. 
There are typically minimal amount of additonal software that can be used with certain hardware. 
Features can be requested and implemented based on critical mass of user need or through non-recurring-engineering (NRE) funding. 
NRE is typically not available to many center besides the largest or the most well funded. 
HPC Centers may have a unique feature or capability that is wanted and while a request for enhancement is submitted, it may take a while to get critical mass, if at all.
Other features that may work with the underlying technology such as adding a new configuration or hardware component must be pre-approval from vendor before being integrated into the storage ecosystem.
This use case can work in a community driven model to add capabilities and features due to the FOSS nature.

Introduction of many physical tiers becomes troublesome for users to pre-migrate their data for usage or for archival purposes.
Physical tiers can be part of a namespace with many physical tiers into a single logical tier or an environment may have many logical tiers that are tied to their physical tier
This section will address ways to combine the many types of physical tiers into appropriate logical tiers, thus allows physical tiers have multiple uses.

The heterogeneous hierarchy of storage tiers have varying levels of performance and different economic models. 
For example, a burst buffer tier has been introduced for high IOPs and throughput to accelerate codes.
This tier is much smaller in capacity due to the cost of flash.
Various data retention boundaries may not align with an HPC storage system’s lifetime. 
This can be apparent when a storage system is years old and a new project spanning a decade.
Due to constantly changing use cases and generational technology shifts, a user may not be aware of proper use of a storage system.
For example, a user may store a data set in HPSS, which is commonly used for a long term archival solution.
This user then frequently recalls data from this solution.
A more optimized use case would be to use a near-line solution, such as Campaign Storage for warm active data sets.
In some cases, workflow may not be known until a project progresses.

Another unknown is future capabilities and layout of a storage system.
This storage system could introduce a new storage tier or combine multiple tiers into fewer logical tiers.
This also introduces significant effort in education, workflow optimization, and data stewardship.
This can cause extra effort of system administrators and users alike.
This future solution will address these challenges by giving a user more flexibility via workflow improvements.
We conclude that crucially in the HPC ecosystem, a middleware translation layer is needed which supports various backend storage solutions whether Commercial-off-the-shelf (COTS) solutions supplied or open source software (OSS).
This could be NGI.
A generic overview of this workflow is given in \Cref{fig:dataLifecycle}

\begin{figure}[b]
  \includegraphics[width=0.75\columnwidth]{datalifecycle-workflow}
  \caption{Future Data Lifecycle Workflows}
  \label{fig:dataLifecycle}
\end{figure}

\subsubsection{Execution of the workflow}

First data sources are read and written from/to an HPC supercomputer.
(S1) shows I/O between a compute and primary data store, that may be POSIX or another interface.
This storage has high-performance characteristics for bandwidth, latency and I/O operations.
This data will be selected by the user for archival or migration off this location via a data movement tool (this is depicted in S2).
Instead of using proprietary interfaces, NGI could be used.
The respective tools translate on the fly from the native interface of one data store to the NGI interface.
NGI is in charge to make the decision how to layout the data on one or multiple storage backends (S3).
These backends are fully pluggable and customizable solutions.
Under the hood a backend -- such as tape -- may need additional Data Movers or File Transfer Agents (FTA) to perform the work efficiently; we will not discuss this further in this paper.
(M1) is the smart data management powered by Machine Learning that also takes user-defined migration policies and rules into account.
(M1) works with the data movers to migrate data back and forth without manual user intervention.
A user is able to read, write and delete immutable data and then allow for various data retention policies.

\paragraph{Optimized Data Placement}
The user will have to define the relevant properties of the data (such as how long that data shall be stored) and workflow and the system administrator provides the definition of the extensible storage landscape.
NGI will utilize machine learning techniques based on empirical data models for optimized management and placement of the data itself.
The result is to use each storage tier in the intended use case with minimal effort of user.
These storage tier can be tape or spinning disk media.

For example, assume a large dataset of hundreds of TiBs was put on a spinning disk archive (Campaign Storage), but only a specific variable with 10 GiB of data is accessed.
A potential operation would be that through data policies migrate other variables, that are unused, to a tape based system.
This will allow for more efficient uses of economics of storage media based on the temperature of the variables.
Another strategy would be to store the first timestep(s) on disk while storing the remainder on tape.
This allows users to work already with the data while the remaining parts are retrieved in the background.

\paragraph{Increased Data Capability Momentum}
Data archives store a copy of data for long term storage and reference.
In many cases, the copy stored through the data archive interface is the only copy.
Users, teams and organizations through policy are required or recommended to make contigent plans of multiple copies of business critical data.
These procedures are audited and require many hours of compliance checking in addition to ensuring an additional copy.
Additonal copies also introduce failure of processes to other storage systems.
The additonal backups may not be cost effective being on multiple systems in full copy.
In this section we will discuss a system that addresses this data resilience issue through replicas (2x, 3x multiplier) and using erasure coded schema (1.2x, 1.4x mulitplier) across storge failure domains.

HPC Centers provide different storage systems within the storage ecosystem.
It can have a PFS, NFS Appliance, a Mid-tier Disk Store and a Tape Archival system.
The storage systems all have their own level of data resilience such as RAID5, RAID6, Distributed Parity schemas, or replicas that allow for multiple copies within the storage system.
The drawback with this setup is if a critical bug or catastrophic even happens and the data becomes unrecoverable.
In centers the loss of a complete storage system can be the loss of the only copy of data.
For data deem critical enough NGI will provide a mechanism to do replication or erasure coding across storage systems.
In the event of partial system failure, such as RAID failure with too many parity chunks being lost or a tape failing and a whole file lost one can do replication and more cost effective erasure coding within a storage system.

\section{Conclusions}
While these views are introduced capabilities are expected to change, grow or be removed completely this forum is expected to get new ideas introduced and feedback from community forum members.

\includegraphics[width=2cm]{ngi-logo}

\noindent\url{https://ngi.vi4io.org}

\end{document}

### 2020-01-18 paper/paper.tex
New paper

\documentclass[a4paper, twocolumn]{article}
\newcommand{\papertitle}{Whitepaper}

\usepackage[a4paper, margin=2cm]{geometry}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage[english]{babel}
\usepackage[colorlinks=true,urlcolor=red]{hyperref}
\usepackage{url}
\usepackage{cleveref}
\usepackage{todonotes}

\usepackage{multicol}
\setlength{\columnsep}{1cm}

\usepackage{fancyhdr}
\fancyhead{}
\fancyfoot{}
\fancyhead[l]{\papertitle}
\fancyhead[r]{\includegraphics[height=1em]{esiwacelogo_type_grey_left_2c_ext}}
\fancyfoot[r]{\thepage}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{1pt}
\renewcommand{\footrulewidth}{1pt}

\usepackage{titling}
\pretitle{\begin{center}\Large\bfseries}
\posttitle{\end{center}\vskip 0.5em}

\graphicspath{{./assets/}}

\title{\papertitle}

\author{Julian M. Kunkel
  \textit{University of Reading}
	\and
  Luciana R. Pedro
  \textit{University of Reading}
}
\date{\today}

\begin{document}
\maketitle
\thispagestyle{fancy}

\section*{Abstract}

\section{Introduction}

\section{Workflows}
\label{sec:background}

An example of a simple workflow can be seen in \Cref{fig:workflow}. In this workflow, there are the following components:

\begin{description}

\item[Input] The data inserted in the workflow. It is composed of files describing, for instance:

\begin{itemize}

\item The actual input data, for instance, the dates for the experiments, how many times the workflow has to run,

\item The dependencies among the tasks. In this case, Task 1 and Task 3 have no dependencies and Task 2 depends on the results of Task 1.

\item The I/O information, i.e., where the data will be storage in each step of the workflow.

\item The information about how long the final result has to be stored.

\item The steps in which manual decisions have to be made.

\end{itemize}

For simplicity, we will call these files as \texttt{input-workflow.conf}.

\item[Task 1] The task is specified by the user and the information we have about it are:

\begin{itemize}

\item The data input.

\item The data output.

\item How long the data has to be stored.

\item The dependencies regarding the other tasks.

\end{itemize}

In case of Task 1, we have:

\begin{itemize}

\item Task 1 has no prior dependencies other than the general input.

\item Task 1 is a dependency for running Task 2.

\item For simplification purposes, we will assume that Task 1 has the files \texttt{input1.conf}, \texttt{output1.conf}, and \texttt{storage1.conf} as descriptors.

\end{itemize}

\item[Task 2]

\begin{itemize}

\item Task 2 is a dependent of Task 1.

\item Task 2 has the files \texttt{input2.conf}, \texttt{output2.conf}, and \texttt{storage2.conf} as descriptors.

\end{itemize}

\item[Task 3]

\begin{itemize}

\item Task 3 is independent. It does not require any other task to be started and, after finished, no other task will use its results other than the general output.

\item Task 3 has the files \texttt{input3.conf}, \texttt{output3.conf}, and \texttt{storage3.conf} as descriptors.

\end{itemize}

\item[Output]

The output consists of all files generated by the workflow. Those files have to be stored for a specific amount of time previously described in one of the input files.

For simplicity, we will call these files as \texttt{output-workflow.conf}. In this specific example, it is composed by the files \texttt{output1.conf}, \texttt{output2.conf}, and \texttt{output3.conf}.

\end{description}

\begin{figure}[b]
  \centering
%  \includegraphics[width=0.75\columnwidth]{workflow}
  \caption{Example High-Level Workflow}
  \label{fig:workflow}
\end{figure}

For the purposes of this work, the files belonging to \texttt{input-workflow.conf} are the files inserted as input for the running Cylc. The Cylc tool receives, as predefined, information about the storage and dependencies of the files. These information are used to contructed and run the complete workflow, but it is not used to optimise intermediate steps. Some things are already predefined and they cannot change. For instante, if it is estabilished that Task 2 depends on Task 1, then we have to wait for Task 1 to be completed to start running Task 3. But several things can be optimised if we know the whole workflow in advance. For instance: Task 1 and Task 3 have no previous dependencies. How do we decide about when we should run those tasks? Task 1 first, because it is a dependency for Task 2? Task 3 first, because then we have a previous of a complete result, considering Task 3 has no dependencies at all? Can Task 1 and Task 3 run in parallel at the same time, and then the workflow only has to complete Task 2 to finish? On the contrary, Task 1 is really processing demanding, so it should run alone and then Task 2 and Task 3 can run concurrently without extra burden to the computing time? These are all concerns about processing time and the dependencies. But what about I/O? No decision is being made about an optimal way to store the data. Consider, for instance, that the files from \texttt{output1.conf} are divided into \texttt{output1-final.conf} and \texttt{output1-task2.conf}. Then, it is clear that \texttt{output1-task2.conf} will be required to run Task 2, and because of that, should be put in a fast storage to quickly being retrieved, of even kept in local storage. The \texttt{output1-final.conf} files, however, can go directly to tape, if the lifetime of the final output has to be stored for five years.

Those simple examples are here to demonstrate that, once we have the complete knowledge about how the workflow is expect to run, we can optimise steps that were before predefined by the user considering only the options for feasibility. The user input is one of the viable options to run the workflow, but probably it is not the best, assuming other options are available. The user might even have analysed other options, but s/he has done that in a naive way (usually trial and error). By automatising these tasks, specifically a smarter route for the tasks taking into consideration the architecture available and the way the files are stored during the intermediate steps of the workflow, we have a twofold gain:

\begin{itemize}

\item The user does not have to worry about the architecture that the workflow will use, removing from the specialist the decision-making process; Andy

\item The workflow will be optimised for the specific workflow without extra input from the user and the resources will, then, be better utilised.

\end{itemize}

In this paper, we approach two ways to have the extra information from the user to be able to make decisions about storage and the processing of the files/tasks.

\begin{itemize}

\item The user will insert this information as an extra file to the Cylc workflow management. In fact, the user is already doing that, but he is doing that in just one specific way, to make the workflow possible. Here, we are thinking about simplifying the information the user is already providing. So, instead of giving the complete workflow to Cylc, Here, we are expeting the user to provide really simple information, like:

\begin{description}

\item[Dependencies] Run Task 1 and Task 3 in any specific order and then run Task 2 after Task 1 is finished.

\item[Time/Processing]

\begin{itemize}

\item Task 1 usually requires ten hours running in a pc with standard configuration\footnote{Intel Core i5, 8 GB RAM, and 500 GB internal storage drive, for instance.}, requires 100 GB of storage that has to be persistent for no more than one week.

\item Task 2 usually requires two hours running in a pc with standard configuration, requires 150 GB of storage that has to be persistent for no more than two weeks week.

\item Task 3 usually requires five hours running in a pc with standard configuration and it does not require extra storage other than the storage for the final results.

\end{itemize}

\end{description}

\item The second option to acess this type of information from the user does not require any direct information at all. The idea here is to use the DDN IME Monitoring system and have an estimate for the parameters after one run of the complete workflow. This approach is benefical to the user, because s/he does not have to take any extra step that s/he is currently doing to run the workflow through Cylc, but it the price here is being paid with an extra run of the complete workflow, which can be demanding in both time and price (we are considering here the paid acess to a supercomputer, for example). In the end, it is up to the user and the previous knowledge s/he has about the intermediate stages of the workflow.

\end{itemize}

Now, comes the question. Assuming the information is available, who will be responsible to take the decisions for optimising the storage and the processing tasks? The answer here is simple: the ESDM Scheduler will do it. ESDM Scheduler can be used in different steps of the process. ESDM has already an interface with NetCDF. Assuming the user is able to provide extra information in advance, ESDM can interact in the process before the user provide the data to the Cylc tool, using the information to already contruct an optimised workflow with Cylc) and, once the workflow is constructed by Cylc, ESDM can interact again after the configuration of the scripts for both effective and efficient run of the parallel applications.

This paper intends to describe the information required for the user to run the tools employed in this extended workflow: Cylc (Section ), DDN IME (Section ), and ESDM Scheduler (Section ).

\subsection{Workflow Description}
\subsection{Task Definition}
\subsection{Heterogeneous Infrastructure}

\section{Challenges and Limitations}
\label{sec:challenges}
\paragraph{Low-level interfaces:}
\paragraph{Explicit workflows:}
\paragraph{Suboptimal Storage Tiering:}
\paragraph{Task and data equality:}
\paragraph{Manual optimization:}
\paragraph{Access pattern intransparency:}

\section{NGI Vision}
\label{sec:ngiVision}

\subsection{User Perspective}

\newcommand{\bnf}[1]{\textless #1\textgreater}

\paragraph{Data model:}
\paragraph{Namespace:}
\paragraph{Workflow specification:}
\paragraph{Behavioral contracts:}
\paragraph{Prescriptive analysis:}

\subsection{System Perspective}

\paragraph{Awareness:}
\paragraph{Resource management:}
\paragraph{Liquid computing:}
\paragraph{Smart scheduling:}
\paragraph{Extensible ecosystem:}

\section{Conclusions}
\label{sec:conclusions}

\end{document}

### 2020-01-20 paper/paper.tex
Updating

@@
-\documentclass[a4paper, twocolumn]{article}
+%\documentclass[a4paper, twocolumn]{article}
+\documentclass[a4paper]{article}
 \newcommand{\papertitle}{Whitepaper}
@@
 \usepackage{url}
+\usepackage{float}
 \usepackage{cleveref}
@@
 
-\item The user does not have to worry about the architecture that the workflow will use, removing from the specialist the decision-making process; Andy
+\item The user does not have to worry about the architecture that the workflow will use, removing from the specialist the decision-making process; and
 
@@
 
-\subsection{Workflow Description}
-\subsection{Task Definition}
-\subsection{Heterogeneous Infrastructure}
+\section{Nice Pictures (or not!)}
 
-\section{Challenges and Limitations}
-\label{sec:challenges}
-\paragraph{Low-level interfaces:}
-\paragraph{Explicit workflows:}
-\paragraph{Suboptimal Storage Tiering:}
-\paragraph{Task and data equality:}
-\paragraph{Manual optimization:}
-\paragraph{Access pattern intransparency:}
+Consider the Cylc suite configuration for a toy monthly cycling workflow: a warm-cycled atmospheric model (model) is followed by postprocessing (post), forecast verification (ver), and product generation (prod) tasks \ref{Cylc}.
+This suite presents the dependencies among the tasks, but it is missing the information about I/O and storage.
+The decision about where the data with be stored is usually made by the developers in stages prior to Cylc and then it is incorporated into the Cylc process.
+The decisions about storage have to be made taking into consideration the architecture in which the workflow will run.
+Applied scientists should not spent time understanding hardware characteristics, but using their time to develop their work, let the computer do the job the simplest way possible, and just collect and analyse the results.
+However, nowadays, to be able to run a job in a HPC environment, researchers have to have a profound knowledge about their workflow, which is expected, but also about decisions regarding storage, communication, and computing.
 
-\section{NGI Vision}
-\label{sec:ngiVision}
+We propose an approach to reduce the burden on researchers and, at the same time, optimise the decisions about jobs running in HPC systems.
+Once we have an automated decision about where the job will run and how the storage will be managed, the scientists can the run their workflows in any machine without further interactions and even without previous knowledge about the machine architecture.
 
-\subsection{User Perspective}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.9\columnwidth]{cylc}
+  \caption{Toy Example Workflow}
+  \label{fig:cylc}
+\end{figure}
 
-\newcommand{\bnf}[1]{\textless #1\textgreater}
+Consider one cycle of the workflow, as shown in Figure \ref{fig:cylc-1}.
+The tasks dependencies are represented by the arrows, but there is no information about the data that are generated.
+Tasks \texttt{ver} and \texttt{post} depend on Task \texttt{model}, but does Task \texttt{model} needs to be fully completed to satisfy the dependency?
+Of course once Task \texttt{model} is completed, then Tasks \texttt{ver} and \texttt{post} can proceed, and that is the information Cylc has to create and reproduce the workflow.
 
-\paragraph{Data model:}
-\paragraph{Namespace:}
-\paragraph{Workflow specification:}
-\paragraph{Behavioral contracts:}
-\paragraph{Prescriptive analysis:}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.4\columnwidth]{cylc-1}
+  \caption{First Stage Workflow}
+  \label{fig:cylc-1}
+\end{figure}
 
-\subsection{System Perspective}
+Assume, now, we have the scenario introduced in Figure \ref{fig:cylc-io-model-2}.
+In this case, we have extra information indicating why Tasks \texttt{ver} and \texttt{post} depend on Task \texttt{model}.
+In fact, the dependency is not to the complete task, but of some datasets produced by Task \texttt{model}.
+If this information is available a priori, then we can use it to start Task \texttt{ver} once we have the dataset \texttt{I/O 1} and start Task \texttt{post} once we have the dataset \texttt{I/O 3}, making the general process faster.
+Still considering this small example, we can assume that dataset \texttt{I/O 2} will not be used for other tasks.
+So this data can be stored in a "slow" storage or iot can even be deleted.
 
-\paragraph{Awareness:}
-\paragraph{Resource management:}
-\paragraph{Liquid computing:}
-\paragraph{Smart scheduling:}
-\paragraph{Extensible ecosystem:}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{cylc-io-model-2}
+  \caption{First Stage Workflow}
+  \label{fig:cylc-io-model-2}
+\end{figure}
 
-\section{Conclusions}
-\label{sec:conclusions}
+Aggregating the information about the datasets produced by every task, we can think about Tasks \texttt{ver}, \texttt{post}, and \texttt{model} as something as depicted in Table \ref{tab:cylc}.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{|c|c|c|}
+\hline
+  \includegraphics[width=0.3\columnwidth]{cylc-io-model} &
+  \includegraphics[width=0.3\columnwidth]{cylc-io-post} &
+  \includegraphics[width=0.3\columnwidth]{cylc-io-prod} \\
+  \hline
+\end{tabular}
+\caption{I/O Information}
+\label{tab:cylc}
+\end{table}
+
+Therefore, the cycle presented in Figure \ref{fig:cylc-1} becomes the extended cycle now introduced in Figure \ref{fig:cylc-io}.
+In this picture now we keep the tasks dependencies, we explicit the datasets dependencies regarding the tasks that generated them, but not the dependencies among the datasets and the next texts.
+The idea here is to embrance the concept that tasks dependencies are really datasets dependencies, and once we can insert the datasets explicitly in the workflow, it is easier to see that there are many optimisations that can be made.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.4\columnwidth]{cylc-io}
+  \caption{Example High-Level Workflow}
+  \label{fig:cylc-io}
+\end{figure}
+
+The trick part when we consider only the Tasks and not the datasets in the workflow is that it may lead to the false impression that the Tasks only depend on the immediate previous one.
+For that, we have an example in Figure \ref{fig:cylc-io-model-3} showing that Task \texttt{prod} depends directly of Task \texttt{post} (Dataset I/O 4) and indirectly of Task \texttt{model} (Dataset I/O 2).
+Once all this dependencies are represented in the workflow, then the storage for all datasets can be optimised.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{cylc-io-model-3}
+  \caption{First Stage Workflow}
+  \label{fig:cylc-io-model-3}
+\end{figure}
+
+The focus of this paper is to exemplify the I/O optimisations that can be made if the datasets dependencies are established.
+This information in fact is presented in the workflows prior to the Cylc application, so nothing else will be required from the user.
+But to have the information explicitly, Cylc need some changes to be able to:
+
+\begin{enumerate}
+
+\item deliver it together with the workflow, and
+
+\item adapt the resulting workflow to consider the I/O modifications.
+
+\end{enumerate}
+
+The information about the system architecture can be provided in advance.
+If that is not possible, we can use monitoring to identify both the architecture and the I/O patterns for the workflow.
+For that, all we need is to run one cycle of the workflow and detect the patterns associated with it.
+For this purpose, we intend to use Machine Learning to identify the patterns and adapt the workflow in case of unexpected situations (system is down, nodes are unresponsive, etc.).
+
+With this separation of concepts, we can abstract from the worflow what is essential and has to be provided by the user and what a machine should optimise to ensure that the resources available are being used smartly.
+Everytime parameters are manually tuned, that has a room from improvement by using automatised decision making.
+And Machine Learning is proven effective in replacing human-beings in several problems.
+
+About the I/O, the optimisations we can envision are related to the life cycle of datasets and the placement of such datasets into specific memory according to their utilisation.
+What happens in most current workflows is that usually they use only two types of file systems: \textbf{work} and \textbf{local}.
+When a task is set to computing, the corresponding dataset is moved to local, processed, and the resulting datasets are moved back to \textbf{work}.
+If the local filesystem achieves its capacity, then the datasets move back to \textbf{work} and keep running the task until is finished.
+If the I/O planning is made a priori, this situation can be avoided.
+
+\begin{description}
+
+\item{NFS} Relatively slow, persistent file system - Nightly incremental backups
+
+\item{Lustre} Large, fast distributed scratch file system - No backups
+
+\url{https://info.hpc.sussex.ac.uk/hpc-guide/storage.html}
+
+\end{description}
+
+In addition to the fyle systems described, many supercomputers have also available for the user \textbf{scratch} fyle systems (sometimes local and global).
+So, in a real scenario, if the architecture of the machine is known, we can better use its resources.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{move}
+  \caption{Some figure like this one indicating the movements among file systems. There are a lot of them in D4.2.}
+  \label{fig:move}
+\end{figure}
 

### 2020-01-20 paper/assets/convertSVGtoPDF.sh

@@ -1,0 +1,7 @@
+#!/bin/bash
+mkdir "$PWD"/pdf
+for file in $PWD/*.svg
+    do
+        filename=$(basename "$file")
+        inkscape "$file" -d 1200 -A "$PWD"/pdf/"${filename%.svg}.pdf"
+    done

### 2020-01-23 paper/article.tex
Paper - First version

--- 2020-01-18 18-whitepaper/article.tex
@@
 The efficient, convenient, and robust execution of data-driven workflows and enhanced data management are key for productivity in scientific computing and computer-aided Research Development and Engineering (RD\&E).
-Traditionally in HPC the concerns of storage and compute are separated and optimized independently from each other, and, moreover, independently from the needs of end-to-end user workflows that ultimately lead to insight.
-While there are good reasons for this strategy, there are workflows that run also efficiently with alternative organizations.
+Traditionally in HPC the concerns of storage and compute are separated and optimised independently from each other, and, moreover, independently from the needs of end-to-end user workflows that ultimately lead to insight.
+While there are good reasons for this strategy, there are workflows that also run efficiently with alternative organisations.
 For example, Big data tools integrate compute and storage capabilities into a holistic solution demonstrating the benefit of tight integrating for data-driven workflows.
-As workflows become increasingly complex and blend beyond data centers while at the same time storage hierarchies becomes deeper, the community investigates means to re-organize storage access to fully utilize such environments.
+As workflows become increasingly complex and blend beyond data centres while at the same time, storage hierarchies become deeper, the community investigates means to re-organise storage access to utilise such environments fully.
 
-Even within a single data center, utilizing homogeneous storage and compute infrastructure efficiently is complex for experts and needs a manual definition.
-The efficient management of data and compute capabilities in a heterogeneous environment, however, is an unresolved question as the execution of individual tasks from workflows may benefit from alternative hardware architectures and infrastructures.
+Even within a single data centre, utilising homogeneous storage and compute infrastructure efficiently is complex for experts and needs manual definition.
+The efficient management of data and compute capabilities in a heterogeneous environment.
+However, it is an unresolved question as the execution of individual tasks from workflows may benefit from alternative hardware architectures and infrastructures.
 
@@
 Within the NGI consortium, we fuse state-of-the-art concepts into a revolutionary interface by rethinking how data-driven I/O should be done.
-Basically, we believe workflows composed of data, compute and communication intensive tasks should drive the interfaces and hardware configurations to best support these programming models.
+Basically, we believe workflows composed of data, compute, and communication-intensive tasks should drive the interfaces and hardware configurations to best support these programming models.
 From the user perspective, NGI aims to provide an abstraction to data-driven processing and data management.
 The NGI interface ultimately will increase the opportunity of implementations for smarter scheduling of compute and storage in heterogeneous environments considering characteristics of workflows and data.
-From the system perspective, the vision of NGI aims to derive an execution plan that utilizes the available hardware and software infrastructure within and across data centers but also on small-scale systems.
+From the system perspective, the vision of NGI aims to derive an execution plan that utilises the available hardware and software infrastructure within and across data centres but also on small-scale systems.
 
@@
 Typical use-cases cover the numerical simulation of physical systems or the analysis of large-scale observational data.
-While the hardware of HPC systems is on the bleeding edge, the software landscape of the infrastructure is lacking behind modern concepts found in domains like cloud computing.
-There are various reasons for the lack of innovation, e.g., the market capitalization is limited; many HPC applications have a life time of sometimes several decades\footnote{For example, CTH at Sandia is more than 30 years old, but it has evolved over the years as hardware has changed.
-It is not uncommon that codes take 10 years to write and receive a first optimized setup.
+While the hardware of HPC systems is on the bleeding edge, the software landscape of the infrastructure lacks behind modern concepts found in domains like cloud computing.
+There are various reasons for the lack of innovation, e.g., the market capitalisation is limited; many HPC applications have a lifetime of sometimes several decades\footnote{For example, CTH at Sandia is more than 30 years old, but it has evolved over the years as hardware has changed.
+It is not uncommon that codes take ten years to write and receive a first optimised setup.
 Then they run for another 30 years with constant evolution based on platform changes and functionality expansions.}; the goal in HPC is to get science done and not to have modern and user-friendly systems.
-Therefore, software developed to target the HPC market is rarely used outside data centers; while there is no reason to use developed applications outside this scope, software to aid developers or for managing HPC infrastructure suffers the same cause.
+Therefore, software developed to target the HPC market is rarely used outside data centres; while there is no reason to use developed applications outside this scope, software to aid developers or for managing HPC infrastructure suffers the same cause.
 Nowadays, HPC adapts infrastructure software from the enterprise and cloud market as much as possible; at best, adapting (stretching) already existing software packages to the needs of HPC.
@@
 \begin{itemize}
-  \item Compute focus: HPC considers compute, storage, and network as separate resources and focuses on the compute aspects, e.g., I/O requires to retrieve/store data from a compute node.
-  While this provides a wonderful separation of concerns, it disregards the opportunity to process data closer to storage or between applications.
+  \item Compute focus: HPC considers computing, storage, and network as separate resources and focuses on the compute aspects, e.g., I/O requires to retrieve/store data from a compute node.
+  While this provides a beautiful separation of concerns, it disregards the opportunity to process data closer to storage or between applications.
   \item Complexity: a data-driven workflow is highly complex and may benefit from using a heterogeneous set of compute and storage technology at the same time.
-  Knowing the capabilities, interfaces, and performance characteristics of individual components is mandatory to make best use of them.
+  Knowing the capabilities, interfaces, and performance characteristics of individual components are mandatory to make the best use of them.
   \item Portability: moving from one infrastructure to another or even when modifying the infrastructure requires to rethink and manually adapt the workflow as the characteristics of the overall environment changes.
-  \item Cost-efficiency: it is time consuming to utilize low-level constructs often only increasing performance slightly; also, the cost of labor must be taken into account.
-  From the perspective of scientists and engineers, the time-to-solution from experimental design to insight matters and not the cost-savings for the public data center.
-  \item Conservative users: users are reluctant to change their workflows for the reasons that older workflows are proven to be correct and it is labor intensive to manually adapt workflows.
+  \item Cost-efficiency: it is time-consuming to utilise low-level constructs often only increasing performance slightly; also, the cost of labour must be taken into account.
+  From the perspective of scientists and engineers, the time-to-solution from experimental design to insight matters and not the cost-savings for the public data centre.
+  \item Conservative users: users are reluctant to change their workflows for the reasons that older workflows are proven to be correct, and it is labour intensive to adapt workflows manually.
 \end{itemize}
@@
 For these reasons, workflows are typically encoded manually and run suboptimally on HPC systems.
-Additionally, users of public data center services supply limited information about their experiments, typically the level of detail is like "I need X CPU hours and Y storage capacity" and they have not to proof their ability to use these systems efficiently.
-Still, the systems these workflows run on cost up to 100 million of dollars.
+Additionally, users of public data centre services supply limited information about their experiments, typically the level of detail is like "I need X CPU hours and Y storage capacity" and they have not to prove their ability to use these systems efficiently.
+Still, the systems these workflows run on cost up to 100 million dollars.
 Compare this to experiments conducted in physics like the Large Hadron Collider (LHC) at CERN or the Square Kilometer Array (SKA); execution of complex experiments may need years of planning involving the scientists.
-While it is important to keep some freedom to conduct ad-hoc experiments, we raise the question what could be achieved by systems and operators knowing workflows in more detail.
+While it is essential to keep some freedom to conduct ad-hoc experiments, we raise the question of what could be achieved by systems and operators knowing workflows in more detail.
 We argue that knowledge about workflow execution and overall experimental design helps to
-optimize execution of workflows and additionally increase the performance, throughput and  cost-efficiency of the environments, therewith, providing an incentive to users and data-centers that cannot be neglected.
+optimise execution of workflows and additionally increase the performance, throughput and cost-efficiency of the environments, in addition to that, providing an incentive to users and data-centres that cannot be neglected.
 
@@
 
-With the Next Generation Interfaces initiative, we aim to define a high-level interfaces that, firstly, meets the needs of users and data centers, and, secondly, that allows implementations to exploit the capabilities of current and future storage and compute landscapes.
+With the Next Generation Interfaces initiative, we aim to define a high-level interface that, firstly, meets the needs of users and data centres, and, secondly, that allows implementations to exploit the capabilities of current and future storage and compute landscapes.
 The remainder of this article is structured as follows: \\
@@
 
-To solve a computational problem, we can consider to describe the necessary steps from the start to the insight as a workflow.
+To solve a computational problem, we can consider describing the necessary steps from the start to the insight as a workflow.
 Such a workflow description must comprise of the individual data processing steps, the expected products, together with dependencies between the steps and to input.
 The nodes can be annotated with characteristics of the individual steps;
-data can be annotated with metadata describing its content, the expected data life cycle information, e.g., the value of data and how long should it be kept; tasks can be annotated with the information needed to start processing or with performance models that bound the  expectations for resources (time, memory, storage capacity).
+data can be annotated with metadata describing its content, the expected data life cycle information, e.g., the value of data and how long should it be kept; tasks can be annotated with the necessary information to start processing or with performance models that bound the expectations for resources (time, memory, storage capacity).
 
@@
 Most of the workflow can run automatically except the manual quality control of the products and the final data usage of Product\,3 at the end.
-The last step represents the typical uncertainty of data reuse, when this workflow is created it is unclear how Product\,3 will be used further, it is apparent that it will be.
+The last step represents the typical uncertainty of data reuse. When this workflow is created, it is unclear how Product\,3 will be used further, but apparently, it will be.
 Further workflows that use the data may be created similarly and use the product as input.
@@
 The high-level interaction of a task with its dependencies is generally specified in the workflow.
-A compute task is encoded in a programming language by a developer\footnote{We often treat developer and HPC user as synonyms as a fraction of HPC users are developers for parallel applications to some extent.}.
-A team of developers encodes the algorithm that solves the problem -- e.g., in many cases,  scientists that generate a numerical simulation implement the model programmaticially.
+A computing task is encoded in a programming language by a developer\footnote{We often treat developer and HPC user as synonyms as a fraction of HPC users are developers for parallel applications to some extent.}.
+A team of developers encodes the algorithm that solves the problem -- e.g., in many cases,  scientists that generate a numerical simulation implement the model programmatically.
 Often they use existing libraries that provide a piece of functionality or infrastructure.
 In the design phase of the program, a parallel programming paradigm and suitable data structures that allow efficient data processing are chosen.
-In the world of parallel computing, the developer must also consider strategies to parallelize the code and data processing; this often involves to use APIs and runtime systems that allow the efficient usage of supercomputers.
+In the world of parallel computing, the developer must also consider strategies to parallelise the code and data processing; this often involves to use APIs and runtime systems that allow the efficient usage of supercomputers.
 
-While most parts of the algorithm can be a tightly coupled simulation, as part of the algorithm, input data must be fetched and data products are generated which are processes that are executed outside of the main simulation.
-Typically, before the input data can be used some kind of preprocessing must be applied, similarly, in the output several known post-processing steps may be applied.
+While most parts of the algorithm can be a tightly coupled simulation, as part of the algorithm, input data must be fetched, and data products are generated which are processes that are executed outside of the first simulation.
+Typically, before the input data can be used some kind of preprocessing must be applied, similarly, in the output, several known post-processing steps may be applied.
 Also, the programmer is responsible for making the optimal decisions on how to map data structures from memory to I/O calls and place these files onto storage systems.
@@
 
-The variety of tasks executed by a single workflow may benefit from a heterogeneous storage and compute infrastructure and potentially even span multiple locations.
-\Cref{fig:heterogeneous} shows such an environment but focuses on the computation and storage.
-A variety of accelerators (GPU, TPU, FPGAs), active storage, in-memory, and in-network computing technologies provide processing capabilities across and beyond an HPC center optimized for certain workloads.
-Data centers typically provide more than one storage and compute class with different characteristics to optimize cost and efficiency.
-Likewise, storage is deployed data center wide, locally available to individual nodes, or subsets of nodes (e.g., racks); depending on the need, the characteristics range from predictable low-latency (in-memory storage, NVMe) to online storage (SSD, HDD), to cheap storage for long-term archival (tape).
+The variety of tasks executed by a single workflow may benefit from heterogeneous storage and compute infrastructure and potentially even span multiple locations.
+\Cref{fig:heterogeneous} shows such an environment but focuses on computation and storage.
+A variety of accelerators (GPU, TPU, FPGAs), active storage, in-memory, and in-network computing technologies provide processing capabilities across and beyond an HPC centre optimised for specific workloads.
+Data centres typically provide more than one storage and compute class with different characteristics to optimise cost and efficiency.
+Likewise, storage is deployed data centre wide, locally available to individual nodes, or subsets of nodes (e.g., racks); depending on the need, the characteristics range from predictable low-latency (in-memory storage, NVMe) to online storage (SSD, HDD), to cheap storage for long-term archival (tape).
 
 The processing across geographic locations is nowadays mandatory to tame the data volumes of observational or simulation data.
-Also, the lines between cloud computing and HPC are blurry, attempts are made to augment processing by flexibly migrate bursts of computation to cloud environments, ingest data from cloud sources or provided to customers.
-Consider a simple scenario where sensor data captured is pre-processed (cleaned) locally, then enriched with an additional data source in a small-scale HPC facility or by using fog computing, then transferred to a data center that performs the number crunching and generates data products.
+Also, the lines between cloud computing and HPC are blurry.
+Attempts are made to augment processing by flexibly migrate bursts of computation to cloud environments, ingest data from cloud sources or provided to customers.
+Consider a simple scenario where sensor data captured is pre-processed (cleaned) locally, then enriched with an additional data source in a small-scale HPC facility or by using fog computing, then transferred to a data centre that performs the number crunching and generates data products.
 These data products may be made available on cloud storage to customers for individual processing.
@@
 \label{sec:challenges}
-The current HPC environment faces several challenges and limitations in respect to the execution of data-driven workflows.
+The current HPC environment faces several challenges and limitations concerning the execution of data-driven workflows.
 
@@
 In a nutshell, a stream of bytes isn't a workable data model.
-HPC data management and data-flow processing utilizes low-level interfaces.
+HPC data management and data-flow processing utilise low-level interfaces.
 Storage relies still on traditional file system objects of directories and files -- provided by the POSIX interface or by object storage.
-Underlying storage systems are receiving or delivering streams of bytes, and it cannot make any intelligent decisions about how to optimize data layout for performance, or resiliency.
-It deals with data blocks and files and requires instructions on how to select and control available optimization strategies.
+Underlying storage systems are receiving or delivering streams of bytes, and it cannot make any intelligent decisions about how to optimise data layout for performance, or resiliency.
+It deals with data blocks and files and requires instructions on how to select and control available optimisation strategies.
 
-In practice, applications build up layers so that the application itself is usually working with the data mediated by a library which provides application specific semantics --- but at the lowest level such a library is limited by the underlying storage APIs (POSIX).
-This has performance implications due to synchronization in distributed environments but also severely limits the ability of the underlying storage systems to optimize I/O operations.
+In practice, applications build up layers so that the app itself is usually working with the data mediated by a library which provides application-specific semantics --- but at the lowest level, such a library is limited by the underlying storage APIs (POSIX).
+This has performance implications due to synchronisation in distributed environments but also severely limits the ability of the underlying storage systems to optimise I/O operations.
 
@@
 These loosely coupled workflows require that intermediate results are stored on persistent storage to be used by the next task.
-Data transfer, however, needs a dominant fraction of energy compared to computation favoring offloading of computation to storage.
+Data transfer, however, needs a dominant fraction of energy compared to computation favouring offloading of computation to storage.
 
 \paragraph{Suboptimal Storage Tiering:}
-Organizing the data placement on storage tiers is typically performed manually by the developer/user or via policies.
-This leads to suboptimal optimizations, is tedious and error prone.
+Organising the data placement on storage tiers is typically performed manually by the developer/user or via policies, leading to suboptimal optimisations, is tedious and error-prone.
 
-Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in form of files) on a particular storage system and typically moving data between storage by using scripts.
+Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of files) on a particular storage system and usually moving data between storage by using scripts.
 One limitation here is that decisions about how data structures are mapped and packaged into files are made once by the producing application, and cannot be changed without manual intervention by a downstream application.
@@
 
-Any hard-coded decision has significant implications for performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
+Any hard-coded decision has significant implications for the performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
 In any case, data migration needs resources on the source and the target storage system to copy data.
@@
 
-All tasks and data products are handled identically by an HPC system meaning there is no differentiation between individual jobs and data products.
-This is problematic as different use cases may not only differ in priority and deadline but as the speed they proceed depends on the critical path of the execution.
+All tasks and data products are handled identically by an HPC system, meaning there is no differentiation between individual jobs and data products.
+This lack of hierarchy is problematic as different use cases may not only differ in priority and deadline but as the speed, they depend on the critical path of the execution.
 
 Fair enough, batch systems understand a priority for jobs.
-However, once a set of nodes is allocated explicitly to a compute job, the job is run exclusively and may utilize shared resources such as storage systems and network as any other job.
-%-- quality of service features available in hardware are rarely used
+However, once a set of nodes is allocated explicitly to a compute job, the job is run exclusively.
+It may utilise shared resources such as storage systems and network as any other job.
+%-- the quality of service features available in hardware is rarely used.
 
-This is more sever for data products: a storage system provides a single level of performance and resilience and performance for all use cases.
-Theoretically, different fault-tolerance and performance policies can be realized with different storage systems and using management policies.
+This problem is even more severe for data products: a storage system provides a single level of performance and resilience and performance for all use cases.
+Theoretically, different fault-tolerance and performance policies can be realised with different storage systems and using management policies.
 However, these policies would be coarse-grained and would not account for the priority of the workflow and most importantly the value of data;
-the value of data depends on aspects like costs to reproduce the data (can it be re-produced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow.
-While value and priority should influence fault-tolerance strategies and imply quality of service for performance and availability, in practice this is not the case as the storage just sees blocks and files.
+the importance of data depends on aspects like costs to reproduce the data (can it be re-produced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow.
+While value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability, in practice, this is not the case as the storage just sees blocks and files.
 
-\paragraph{Manual optimization:}
-Utilizing the compute and storage characteristics of a heterogeneous environment as depicted in \Cref{fig:heterogeneous} is challenging at least.
-The optimization of a single parallel job already is quite challenging.
-To make optimal use of available features, programmers and users are forced to think about technological aspects and terms, e.g., setting low-level MPI hints for file striping, while they understand their workflow.
+\paragraph{Manual optimisation:}
+Utilising the compute and storage characteristics of a heterogeneous environment as depicted in \Cref{fig:heterogeneous} is challenging at least.
+The optimisation of a single parallel job already is quite challenging.
+To make optimal use of available features, programmers and users are forced to think about technical aspects and terms, e.g., setting low-level MPI hints for file striping, while they understand their workflow.
 Similarly, an efficient data placement is complex.
@@
 
-An explicit workflows combined with manual optimization is the root-cause of subsequent problems.
+A specific workflow combined with manual optimisation is the root-cause of subsequent problems.
 
-\paragraph{Access pattern intransparency:}
-As access patterns are primarily implicitly defined in application code and embedded in workflows, the understanding and debugability of behavior is limited.
-Firstly, there are many alternatives ways of programming data-intensive computation to achieve the identical data products.
-Secondly, many storage and compute interfaces are low-level.
-Thus, data centers have often limited insight into the access patterns as well.
-Moreover, some users are abusing the storage system in non-obvious ways, for example to exchange notifications.
-Together with the possibility to observe access patterns at low-level, these issues lead to the situation that understanding and improving of access patterns become difficult.
-Limited insight into behavior leads to suboptimal tuning.
+\paragraph{Access pattern in transparency:}
+As access patterns are primarily implicitly defined in application code and embedded in workflows, the understanding and debugability of behaviour are limited.
+Firstly, there are many alternatives ways of programming data-intensive computation to achieve identical data products.
+Secondly, many storages and compute interfaces are low-level.
+Thus, data centres have often limited insight into the access patterns as well.
+Moreover, some users are abusing the storage system in non-obvious ways, for example, to exchange notifications.
+Together with the possibility to observe access patterns at low-level, these issues lead to the situation that understanding and improvement of access patterns become difficult.
+Limited insight into behaviour leads to suboptimal tuning.
 
-Manual and hard-coded workflows cannot handle changes in the environment and are error prone.
+Manual and hard-coded workflows cannot handle changes in the environment and are error-prone.
 Therefore, users must be able to express their workflows in an abstract fashion that allows the system to generate (near-)optimal execution plans and monitor their execution.
@@
 %\paragraph{Data Model}
-NGI tackles the challenge of storage and data-flow computation in an holistic way accepting that data storage is only aiding the goal of generating insight.
-Besides providing a typical abstract data model, NGI treats domain metadata\footnote{Therewith, we mean non-technical information about the data, for instance, the name of an experiment.}, workflows, and information lifecycle management (ILM) as first-class citizens allowing the storage to understand these concepts.
-This user information is exploited by smart components of NGI which aim to utilize capabilities of  heterogeneous system landscapes.
+NGI tackles the challenge of storage and data-flow computation in a holistic way accepting that data storage is only aiding the goal of generating insight.
+Besides providing a typical abstract data model, NGI treats domain metadata\footnote{ in addition to that, we mean non-technical information about the data, for instance, the name of an experiment.}, workflows, and information lifecycle management (ILM) as first-class citizens allowing the storage to understand these concepts.
+This user information is exploited by smart components of NGI which aim to utilise capabilities of heterogeneous system landscapes.
 
 To describe the features and benefits of the approach further, we sketch the user and system perspective in the following, as this highlights how we believe we should work in the 21st century.
-The vision covers not only HPC but spans cloud, enterprise, and personalized IT.
+The vision covers not only HPC but spans cloud, enterprise, and personalised IT.
 
@@
 
-From the user's perspective, the interface and system matches the core requirements and expectations including reliability, availability, and usability.
+From the user's perspective, the interface and system match the core requirements and expectations, including reliability, availability, and usability.
 
 \paragraph{Data model:} the data model is based on state-of-the-art solutions like HDF5+X where X indicates an extension for ILM, workflows and processing.
-However, it abstracts from data localization and low-level concepts like files and directories and puts its focus on data organization and data structures.
+However, it abstracts from data localisation and low-level concepts like files and directories and puts its focus on data organisation and data structures.
 The integration into applications is shown in \Cref{fig:ngilayers}.
-Applications may utilize the NGI API directly, or utilize existing middleware\footnote{Which won't be able to utilize the full set of features if not extending its interface.}.
-From the perspective of a developer of middleware or an application like web publishing, NGI provides a common feature set that reduces the amount of code replication inside domain-specific solutions.
+Applications may utilise the NGI API directly, or utilise existing middleware\footnote{Which won't be able to use the full set of features if not extending its interface.}.
+From the perspective of a developer of middleware or an application like web publishing, NGI provides a standard feature set that reduces the amount of code replication inside domain-specific solutions.
 
 \paragraph{Namespace:} data is accessed using information from the domain metadata, i.e., similar to using methods of data indexing services.
-The system won't relying on a hierarchical namespace but a dynamic mapping to the POSIX namespace can be provided to enable compatibility for POSIX applications.
+The system won't be relying on a hierarchical namespace, but a dynamic mapping to the POSIX namespace can be provided to enable compatibility for POSIX applications.
 For example, a user may decide to see a hierarchy like an experiments \bnf{Date}/\bnf{Name}/\bnf{Timestep}/\bnf{Variable}.
 
-\paragraph{Workflow specification:} the user defines a workflow describing the intentions of data manipulation and the data lifecycle including manual (user) involvement in the data processing.
+\paragraph{Workflow specification:} the user defines a workflow describing the intentions of data manipulation and the data lifecycle, including manual (user) involvement in the data processing.
 This also includes the definition of constraints, for instance, deadlines for the generation of the final data products (insight).
-Workflows can be defined using arbitrary tools including Python.
-Note that a user does not provide a concrete mapping to system or infrastructure but expects that the system derives an optimized \textit{execution plan} that covers \textit{heterogeneous infrastructures}.
-It does not matter where data is processed and how and where data is stored but only that the desired result is computed with all constraints met.
-Workflows and data lineage of data products are recorded and can be inspected by the user upon demand allowing to reproduce and validate correct execution of experiments.
+Workflows can be defined using arbitrary tools, including Python.
+Note that a user does not provide an accurate mapping to system or infrastructure but expects that the system derives an optimised \textit{execution plan} that covers \textit{heterogeneous infrastructures}.
+It does not matter where data is processed and how and where information is stored but only that the desired result is computed with all constraints met.
+Workflows and data lineage of data products are recorded and can be inspected by the user upon demand, allowing to reproduce and validate the correct execution of experiments.
 
@@
 Besides the core characteristics, users expect the execution of their workflows under constraints (including cost and time constraints).
-The system is expected to always show the expected characteristics and shield the user from system-specifics and undesired events (like faults or data loss).
+The system is always expected to show the expected characteristics and shield the user from system-specifics and undesired events (like faults or data loss).
 The characteristics may improve over time, for instance, by expanding hardware capabilities, but the user won't have to change their workflow specification.
@@
 Sharing the workflow description with vendors allows them to conduct predictions for runtime and costs for their solutions.
-The ecosystem may also prescribe alternative systems to use in an open environment, e.g., this part of the workflow should be run on data center A and this part on cloud provider B.
+The ecosystem may also prescribe alternative systems to use in an open environment, e.g., this part of the workflow should be run on data centre A and this part on cloud provider B.
 
@@
 the system knows and understands user metadata and the overarching workflows (with the user intentions).
-It is self-aware, knowing the nominal characteristics of the hardware and software components, constantly comparing the system and workflow state with the expectations to perform mitigating actions to heal deviations.
-At best, each individual hardware component would ship with a behavioral model that can be integrated hierarchically up to a global system model.
-With behavioral models we not only mean performance models but reliability and cost-models as well.
+It is self-aware, knowing the nominal characteristics of the hardware and software components, continually comparing the system and workflow state with the expectations to perform mitigating actions to heal deviations.
+At best, each hardware component would ship with a behavioural model that can be integrated hierarchically up to a global system model.
+With behavioural models, we not only mean performance models but reliability and cost-models as well.
 
@@
 the system is in charge to manage and provision the available hardware resources autonomously.
-Ultimately, it aims to utilize the individual characteristics of heterogeneous compute and storage landscapes by utilizing intelligent scheduling.
-Data placement and localization is beyond mere tiering strategies, parts of data may be placed on certain storage technology or replicated in different representations across storage characteristics.
-Therewith, serving different access patterns and workflows at the same time.
-Naturally, the system will replicate (expand) upon the availability of storage space while reducing redundancy under shortage of space.
+Ultimately, it aims to utilise the individual characteristics of heterogeneous compute and storage landscapes by using intelligent scheduling.
+Data placement and localisation is beyond mere tiering strategies; parts of data may be placed on particular storage technology or replicated in different representations across storage characteristics.
+In addition to that, you are serving different access patterns and workflows at the same time.
+Naturally, the system will replicate (expand) upon the availability of storage space while reducing redundancy under the shortage of space.
 
@@
 the system is in charge to manage and provision the available hardware resources autonomously and to assign compute and storage resources to tasks.
-Therewith, it may use, e.g., GPU, CPU, and in-network computing capabilities to serve pieces of a single workflow.
-Moreover, as workflows stretch across data centers and over cloud and edge, an environment that schedules and dispatches work and storage efficiently is key.
+In addition to that, it may use, e.g., GPU, CPU, and in-network computing capabilities to serve pieces of a single workflow.
+Moreover, as workflows stretch across data centres and over cloud and edge, an environment that schedules and dispatches work and storage efficiently is vital.
 We call this concept \textbf{liquid computing} as the compute resources are dynamically provisioned and assigned; while data flows through the system, it is processed.
@@
 The programming of the processing will be based on a high-level functional language providing primitives for data-flow computing and compiled into a virtual machine or machine code when the code is to be dispatched.
-The capabilities of different hardware components may be limited realizing only a subset of the language which must be taken into account by the scheduler.
+The capabilities of different hardware components may be limited, realising only a subset of the language which must be taken into account by the scheduler.
 
 \paragraph{Smart scheduling:}
-supplied with user workflows and their characteristics, the system derives execution plans for the computation and make data placement decisions that aim to utilize the available hardware resources.
-As NGI is in charge for both, it may decide to trade computation vs.
-storage capacity or other characteristics (e.g., costs, energy consumption); for example, enabling automatic recomputation of intermediate states utilizing virtualization and container technologies.
+supplied with user workflows and their characteristics, the system derives execution plans for the computation and make data placement decisions that aim to utilise the available hardware resources.
+As NGI is in charge for both, it may decide to trade computation vs storage capacity or other characteristics (e.g., costs, energy consumption); for example, enabling automatic recomputation of intermediate states utilising virtualisation and container technologies.
 Similarly, the scheduler may decide to couple subsequent steps of a workflow directly without storing intermediate data products on persistent storage.
@@
 
-The automatic scheduling is key to efficient processing and will allow constant improvement without forcing user to adjust their workflows.
-In most cases, individual resources are not used exclusively by a single task but QoS methods ensure that constraints of workflows are met.
-Note that this significant change for HPC environments, as compute resources are typically dedicated for individual jobs.
+The automatic scheduling is key to efficient processing and will allow constant improvement without forcing the user to adjust their workflows.
+In most cases, individual resources are not used exclusively by a single task, but QoS methods ensure that constraints of workflows are met.
+Note that this significant change for HPC environments, as compute resources are typically dedicated to specific jobs.
 
 \paragraph{Extensible ecosystem:}
-the hardware and software landscape will be open allowing to integrate solutions from different hardware and software vendors.
-Systems connected may use different interfaces including traditional POSIX interfaces, NoSQL, or distributed hash tables.
-Imagine that a certain class of workflows performs inefficiently on a given infrastructure indicating a crucial processing or storage capability is missing.
+the hardware and software landscape will be open, allowing to integrate solutions from different hardware and software vendors.
+Systems connected may use different interfaces, including traditional POSIX interfaces, NoSQL, or distributed hash tables.
+Imagine that a specific class of workflows performs inefficiently on a given infrastructure indicating a crucial processing or storage capability is missing.
 Adding new capabilities or resources to the system will allow the scheduler to use them in subsequent user-workflows speeding up the processing; this should be as easy as adding an appliance to a business environment.
@@
 * But not all aspects together
-* Competing approaches; the standardization we describe here does not compete!
+* Competing approaches; the standardisation we describe here does not compete!
 

### 2020-01-23 paper/paper (copy).tex
Paper - First version

--- 2020-01-20 paper/paper.tex
@@
 Still considering this small example, we can assume that dataset \texttt{I/O 2} will not be used for other tasks.
-So this data can be stored in a "slow" storage or iot can even be deleted.
+So this data can be stored in a "slow" storage or it can even be deleted.
 
@@
 
+\section{Nice Pictures -- Version 2}
+
+Consider the Cylc suite configuration for a toy monthly cycling workflow in Figure \ref{fig:cylc}.
+A warm-cycled atmospheric model (\textbf{model}) is followed by postprocessing (\textbf{post}), forecast verification (\textbf{ver}), and product generation (\textbf{prod}) tasks.
+Also, consider, in addition, a task \textbf{check} that compares some verification metric against products from two cycles earlier.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.9\columnwidth]{cylc}
+  \caption{Toy Example Workflow}
+  \label{fig:cylc}
+\end{figure}
+
+The first cycle of the suite in Figure \ref{fig:cylc} is again represented in Figure \ref{fig:cycle1}.
+This suite presents the dependencies among the tasks, but it is missing the information about I/O and storage.
+The decision about where the data with be stored is usually made by the developers in stages prior to Cylc and then it is incorporated into the Cylc process.
+The decisions about storage have to be made taking into consideration the architecture in which the workflow will run.
+Applied scientists should not spent time understanding hardware characteristics, but using their time to develop their work, let the computer do the job the simplest way possible, and just collect and analyse the results.
+However, nowadays, to be able to run a job in a HPC environment, researchers have to have a profound knowledge about their workflow, which is expected, but also about decisions regarding storage, communication, and computing.
+In this paper, we propose an approach to reduce the burden on researchers and, at the same time, optimise the decisions about jobs running in HPC systems.
+Once we have an automated decision about where the job will run and how the storage will be managed, the scientists can the run their workflows in any machine without further interactions and even without previous knowledge about the machine architecture.
+
+Consider one cycle of the workflow, as shown in Figure \ref{fig:model7}.
+The tasks dependencies are represented by the arrows, but there is no information about the data that are generated.
+Tasks \texttt{ver} and \texttt{post} depend on Task \texttt{model}, but does Task \texttt{model} needs to be fully completed to satisfy the dependency?
+Of course once Task \texttt{model} is completed, then Tasks \texttt{ver} and \texttt{post} can proceed, and that is the information Cylc has to create and reproduce the workflow.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.4\columnwidth]{cycle1}
+  \caption{First stage workflow}
+  \label{fig:cycle1}
+\end{figure}
+
+The scenario prtesented in Figure \ref{fig:cycle-io} shows the same cycle of Figure \ref{fig:cycle-1} but now information about datasets is introduced.
+The way data moves through tasks is by creating datasets to pass on the information, including filename and storage destination.
+The datasets produced by every task are coded in the workflow scripts prior to Cylc evaluation and it is just replicated by this tool.
+The tasks can create a different number of datasets, and these datasets may be necessary for the immediate next task in the current cycle, or for Tasks in different cycles.
+Figure \ref{fig:cycle-io-dep} shows the dependencies among the datasets and the Tasks: datasets colours indicate with task will use the information and datasets in yellow represent information that will be used in a different cycle.
+Note that the information that is currently being used in the current cycle can also be used later.
+Also note that one dataset produced by Task \texttt{model} is later used in Task \texttt{prod gen}.
+This is not an immediate conclusion when we analyse just the arrows for the dependencies among Tasks.
+It is clear that Task \texttt{model} is a dependency for Task \texttt{post prod} which is a dependency for Task \texttt{prod gen}.
+The point here is that the knowledge that, for this example, Task \texttt{model} will generate, one dataset for Task \texttt{post proc}, one dataset for Task \texttt{prod gen}, two datasets for Task \texttt{verification}, and three dataset for a later cycle can be used to optimise how these datasets are stored during the workflow.
+This information today is static and manual, leaving opened many opportunities for automatisations and optimisations.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{cycle-io}
+  \caption{First stage workflow including datasets}
+  \label{fig:cycle-io}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{cycle-io-dep}
+  \caption{First stage workflow including datasets and dependencies}
+  \label{fig:cycle-io-dep}
+\end{figure}
+
+Figure \ref{fig:cycle-4} exemplifies another restriction of the way current workflows are being generated.
+Assume you have, for instance, four cycles of the model are running.
+Because the scripts for the workflow consider the individual tasks and the datasets generated by them are placed at the same place at storage every time.
+
+all datasets generated by Task \texttt{model} that will be used by Task \texttt{post prod}
+
+all datasets generated by Task \texttt{A} and used by Task \texttt{B} will have the same physical localion in the system, because there is one fixed script responsible for setting the configuration.
+Again, such configuration was done manually and with restricted information about the storage and system architecture.
+It would be interesting to explore the options for the datasets and have them placed in different storage systems.
+If the lifetime of such files is, for instance, two cycles, alternate the files into two schatch file systems is something would be simple job for a tool working on generating filenames and storage.
+However, manually that implies at having at least two scripts with information about the different storage and manual work to have a better utilisation of the available resources.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{cycle-4}
+  \caption{Four stages workflow example}
+  \label{fig:cycle-4}
+\end{figure}
+
+Figure \ref{fig:lifecycle} finalises the arguments in this paper with a vision of three possible life cycles for one specific dataset.
+The workflow scripts the way they are handle today have one option.
+The scientist responsible for these scripts tried to optimise the available resources \textbf{manuallly} and intuitively before deliver the final version.
+Wouldn't be better to have the machine automatising and making such decisions?
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{lifecycle}
+  \caption{Life cycle of one dataset: \textbf{C}reate, \textbf{M}igrate, \textbf{R}ead, and \textbf{D}elete}
+  \label{fig:lifecycle}
+\end{figure}
+
+In this case, we have extra information indicating why Tasks \texttt{ver} and \texttt{post} depend on Task \texttt{model}.
+In fact, the dependency is not to the complete task, but of some datasets produced by Task \texttt{model}.
+If this information is available a priori, then we can use it to start Task \texttt{ver} once we have the dataset \texttt{I/O 1} and start Task \texttt{post} once we have the dataset \texttt{I/O 3}, making the general process faster.
+Still considering this small example, we can assume that dataset \texttt{I/O 2} will not be used for other tasks.
+So this data can be stored in a persistent storage or it can even be deleted.
+
+Aggregating the information about the datasets produced by every task, we can think about Tasks \texttt{ver}, \texttt{post}, and \texttt{model} as something as depicted in Table \ref{tab:cylc}.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{|c|c|c|}
+\hline
+  \includegraphics[width=0.3\columnwidth]{cylc-io-model} &
+  \includegraphics[width=0.3\columnwidth]{cylc-io-post} &
+  \includegraphics[width=0.3\columnwidth]{cylc-io-prod} \\
+  \hline
+\end{tabular}
+\caption{I/O Information}
+\label{tab:cylc}
+\end{table}
+
+Therefore, the cycle presented in Figure \ref{fig:cylc-1} becomes the extended cycle now introduced in Figure \ref{fig:cylc-io}.
+In this picture now we keep the tasks dependencies, we explicit the datasets dependencies regarding the tasks that generated them, but not the dependencies among the datasets and the next texts.
+The idea here is to embrance the concept that tasks dependencies are really datasets dependencies, and once we can insert the datasets explicitly in the workflow, it is easier to see that there are many optimisations that can be made.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.4\columnwidth]{cylc-io}
+  \caption{Example High-Level Workflow}
+  \label{fig:cylc-io}
+\end{figure}
+
+The trick part when we consider only the Tasks and not the datasets in the workflow is that it may lead to the false impression that the Tasks only depend on the immediate previous one.
+For that, we have an example in Figure \ref{fig:cylc-io-model-3} showing that Task \texttt{prod} depends directly of Task \texttt{post} (Dataset I/O 4) and indirectly of Task \texttt{model} (Dataset I/O 2).
+Once all this dependencies are represented in the workflow, then the storage for all datasets can be optimised.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{cylc-io-model-3}
+  \caption{First Stage Workflow}
+  \label{fig:cylc-io-model-3}
+\end{figure}
+
+The focus of this paper is to exemplify the I/O optimisations that can be made if the datasets dependencies are established.
+This information in fact is presented in the workflows prior to the Cylc application, so nothing else will be required from the user.
+But to have the information explicitly, Cylc need some changes to be able to:
+
+\begin{enumerate}
+
+\item deliver it together with the workflow, and
+
+\item adapt the resulting workflow to consider the I/O modifications.
+
+\end{enumerate}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{model5}
+  \caption{5.}
+  \label{fig:model5}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{model6}
+  \caption{6.}
+  \label{fig:model6}
+\end{figure}
+
 \end{document}

### 2020-01-23 paper/paper.tex
Paper - First version

--- 2020-01-18 paper/paper.tex
@@
-\documentclass[a4paper, twocolumn]{article}
-\newcommand{\papertitle}{Whitepaper}
+%\documentclass[a4paper, twocolumn]{article}
+\documentclass[a4paper]{article}
+\newcommand{\papertitle}{Whitepaper -- Optimising Storage Inside Workflows}
 
@@
 \usepackage{url}
+\usepackage{float}
 \usepackage{cleveref}
 \usepackage{todonotes}
+\usepackage{comment}
 
@@
 
+{ \color{cyan}{
+
 \section*{Abstract}
+The efficient, convenient, and robust execution of data-driven workflows and enhanced data management are key for productivity in scientific computing.
+Traditionally, in HPC, the concerns of storage and computing are separated and optimised independently from each other, and, moreover, independently from the needs of end-to-end user workflows that ultimately lead to insight.
+As workflows become increasingly complex and blend beyond data centres while at the same time, storage hierarchies become deeper, the community investigates means to re-organise storage access to utilise such environments fully.
+
+In this white paper, we describe an initiative that aims to tackle storage and computing challenges mentioned above in a holistic approach by fusing state-of-the-art concepts into a revolutionary interface by rethinking how data-driven I/O should be done.
+The tools employed here to achieve this extended workflow are Cylc, DDN IME, and ESDM Scheduler.
+Basically, we believe workflows composed of data, computing, and communication-intensive tasks should drive the interfaces and hardware configurations to best support these programming models.
+From the user perspective, we aim to provide an abstraction to data-driven processing and data management.
+The proposed interface ultimately will increase the opportunity of implementations for smarter scheduling of computing and storage in heterogeneous environments considering characteristics of workflows and data.
+From the system perspective, the vision aims to derive an execution plan that utilises the available hardware and software infrastructure within and across data centres.
 
@@
 
-\section{Workflows}
-\label{sec:background}
+High-Performance Computing harnesses the fastest hardware components to enable
+the execution of tightly coupled applications from science and industry.
+Typical use-cases cover the numerical simulation of physical systems or the analysis of large-scale observational data.
+While the hardware of HPC systems is on the bleeding edge, the software landscape of the infrastructure lacks behind modern concepts found in domains like cloud computing.
+There are various reasons for the lack of innovation, e.g., the market capitalisation is limited; many HPC applications have a lifetime of sometimes several decades\footnote{For example, CTH at Sandia is more than 30 years old, but it has evolved over the years as hardware has changed.
+It is not uncommon that codes take ten years to write and receive a first optimised setup.
+After that, they run for another 30 years with constant evolution based on platform changes and functionality expansions.}; the goal in HPC is to get science done and not to have modern and user-friendly systems.
+Therefore, software developed to target the HPC market is rarely used outside data centres; while there is no reason to use developed applications outside this scope, software to aid developers or for managing HPC infrastructure suffers the same cause.
+Nowadays, HPC adapts infrastructure software from the enterprise and cloud market as much as possible; at best, adapting (stretching) already existing software packages to the needs of HPC.
 
-An example of a simple workflow can be seen in \Cref{fig:workflow}. In this workflow, there are the following components:
+From the perspective of big data and cloud solutions, the state-of-the-art in the HPC software ecosystem is perceived as conservative and outdated at least.
+For example, HPC still uses batch processing and dedicates nodes to the execution of individual applications which were the first concepts implemented at the dawn of computing.
+Similar to writing code in assembler, often it is argued that HPC trades user-friendliness and manageability to obtain the best possible performance.
+However, as the complexity of systems increases and various alternative storage and computing technologies provide unique characteristics, it is barely possible to define optimal data-driven workflows using low-level constructs for the following reasons:
 
@@
 
-\item[Input] The data inserted in the workflow. It is composed of files describing, for instance:
+  \item[Computing Focus] HPC considers computing, storage, and network as separate resources and focuses on the computing aspects, e.g., I/O requires to retrieve/store data from a computing node.
+  While there are good reasons for this strategy as they provide a beautiful separation of concerns,  there are workflows that run efficiently with alternative organisations because they take into consideration the opportunity to process data closer to storage or between applications.
+
+  \item[Complexity] A data-driven workflow is highly complex and may benefit from using a heterogeneous set of computing and storage technology at the same time.
+  While there are good reasons for this strategy, there are workflows that also run efficiently with alternative organisations.
+
+  Knowing the capabilities, interfaces, and performance characteristics of individual components are mandatory to make the best use of them.
+
+  \item[Portability] Moving from one infrastructure to another or even when modifying the infrastructure requires to rethink and manually adapt the workflow as the characteristics of the overall environment changes.
+
+  \item[Cost-efficiency] It is time-consuming to utilise low-level constructs and they often only increase the performance slightly. Also, the cost of labour must be taken into account.
+  From the perspective of scientists and engineers, the time-to-solution from experimental design to insight matters and the cost-savings for the public data centre can be disregarded.
+
+  \item[Conservative Users] Users are reluctant to change their workflows because they already know that older workflows are proven to be correct and, therefore, they want to avoid the intensive labour to adapt workflows manually.
+
+\end{description}
+
+For these reasons, workflows are typically encoded manually and run suboptimally on HPC systems.
+Additionally, users of public data centre services supply limited information about their experiments, typically the level of detail is like "I need X CPU hours and Y storage capacity" and they have not to prove their ability to use these systems efficiently.
+Still, the systems these workflows run on cost up to 100 million dollars.
+Comparing this amount to experiments conducted in physics like the Large Hadron Collider (LHC) at CERN or the Square Kilometer Array (SKA), it becomes clear that the execution of complex experiments may need years of planning involving the scientists.
+While it is essential to keep some freedom to conduct ad-hoc experiments, we raise the question of what could be achieved by systems and operators knowing workflows in more detail.
+We argue that knowledge about workflow execution and overall experimental design helps to
+optimise execution of workflows and additionally increase the performance, throughput and cost-efficiency of the environments, providing an incentive to users and data-centres that cannot be neglected.
+
+With the Next Generation Interfaces initiative, we aim to define a high-level interface that, firstly, meets the needs of users and data centres, and, secondly, that allows implementations to exploit the capabilities of current and future storage and computing landscapes.
+
+\section{Workflows}
+
+This section provides an introduction to workflows that solve computational problems, the mandatory steps to encode and setup tasks and describe data, and the relevance of heterogeneous infrastructure to run workflows.
+We focus on high-level considerations to illustrate mandatory concepts.
+
+\subsection{Workflow Description}
+
+To solve a computational problem, we can consider describing the necessary steps from the start to the insight as a workflow.
+Such a workflow description must comprise of the individual data processing steps, the dependencies among them, the input, and the expected products.
+An example execution of a workflow is illustrated in \Cref{fig:workflow} as a graph.
+Arrows indicate dependencies between nodes that represent tasks and data.
+In the example, Task\,1 needs two datasets to perform its work, it directly communicates with Task\,2 and produces Product\,1.
+Most of the workflow can run automatically except the manual quality control of the products and the final data usage of Product\,3 in the end.
+The last step represents the typical uncertainty of data reuse. When this workflow is analysed, it is unclear how Product\,3 will be used further, but, apparently, it will be.
+Further workflows that use the data may be created similarly and use products as input.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.4\columnwidth]{workflow}
+  \caption{Example of a high-level workflow}
+  \label{fig:workflow}
+\end{figure}
+
+Note that a realistic workflow will be much more complex and involves not a single parallel job execution, but many to generate a robust answer to the problem in question.
+Extra information can be, for instance:
 
@@
 
-\item The actual input data, for instance, the dates for the experiments, how many times the workflow has to run,
+\item The nodes can be annotated with characteristics of the individual steps;
 
-\item The dependencies among the tasks. In this case, Task 1 and Task 3 have no dependencies and Task 2 depends on the results of Task 1.
+\item Data can be annotated with metadata describing its content, the expected data life cycle information, e.g., the value of data and how long should it be kept;
 
-\item The I/O information, i.e., where the data will be storage in each step of the workflow.
+\item tasks can be annotated with the necessary information to start processing or with performance models that bound the expectations for resources (time, memory, storage capacity).
 
-\item The information about how long the final result has to be stored.
+\end{itemize}
+
+\subsection{Task Definition}
+
+The high-level interaction of a task with its dependencies is generally specified in the workflow.
+A computing task is encoded in a programming language by a developer\footnote{We often treat developer and HPC user as synonyms as a fraction of HPC users are developers for parallel applications to some extent.}.
+A team of developers encodes the algorithm that solves the problem. In many cases, scientists that generate a numerical simulation implement the model programmatically.
+Often, they use existing libraries that provide a piece of functionality or infrastructure.
+In the design phase of the program, a parallel programming paradigm and suitable data structures that allow efficient data processing are chosen.
+In the world of parallel computing, the developer must also consider strategies to parallelise the code and data processing, which often involve the use of APIs and runtime systems that allow the efficient usage of supercomputers.
+
+While most parts of the algorithm can be a tightly coupled simulation, as part of the algorithm, the input data must be fetched.
+
+Typically, before the input data can be used, some kind of pre-processing must be applied. Similarly, in the output, several known post-processing steps may be applied.
+Also, the programmer is responsible for making the optimal decisions on how to map data structures from memory to I/O calls and place these datasets onto storage systems.
+We consider that pre/post-processing are part of the data-driven scope of workflows.
+Indeed, we claim that the programmer should be able to define the input/output on an abstract fashion and connect processing steps as part of the general workflow.
+
+\subsection{Heterogeneous Infrastructure}
+
+The variety of tasks executed by a single workflow may benefit from heterogeneous storage and computing infrastructure and potentially even span multiple locations.
+\Cref{fig:heterogeneous} shows such an environment with a focus on computation and storage.
+A variety of accelerators (GPU, TPU, FPGAs), active storage, in-memory, and in-network computing technologies provide processing capabilities across and beyond a HPC centre optimised for specific workloads.
+Data centres typically provide more than one storage and computing class with different characteristics to optimise cost and efficiency.
+Likewise, storage is deployed data centre wide, locally available to individual nodes, or subsets of nodes (e.g., racks).
+Depending on the need, the characteristics range from predictable low-latency (in-memory storage, NVMe) to online storage (SSD, HDD), and cheap storage for long-term archival (tape).
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{system}
+  \caption{Example of an heterogeneous HPC landscape}
+  \label{fig:heterogeneous}
+\end{figure}
+
+The processing across geographic locations is nowadays mandatory to tame the data volumes of observational or simulation data.
+Also, the lines between cloud computing and HPC are blurry.
+Attempts are made to augment processing by flexibly migrate bursts of computation to cloud environments, ingest data from cloud sources or provided to customers.
+Consider a simple scenario where sensor data captured is pre-processed (cleaned) locally, then enriched with an additional data source in a small-scale HPC facility or by using fog computing, and then finally transferred to a data centre that performs the number crunching and generates data products.
+These data products can also be made available on cloud storage to customers for individual processing.
+
+{ \color{blue}{
+
+\subsection{User Perspective Workflow}
+
+From the user perspective, the workflow can be described by the following categories:
+
+\begin{description}
+
+\item[Input] The information inserted in the workflow. It is composed of datasets describing, for instance:
+
+\begin{itemize}
+
+\item The actual input data, for instance, the dates for the experiments, how many times the workflow has to run, etc.
+
+\item The dependencies among the tasks.
+
+\item The I/O information, i.e., where the datasets will be stored in each step of the workflow.
+
+\item The information about the lifetime of the final outcome.
 
@@
 
-For simplicity, we will call these files as \texttt{input-workflow.conf}.
+For simplicity, we will call these datasets as \texttt{input-workflow.conf}.
 
-\item[Task 1] The task is specified by the user and the information we have about it are:
+\item[Task 1] The task is specified by the user and the information we have about it is:
 
@@
 
-In case of Task 1, we have:
+In the case of Task 1, we have:
 
@@
 
-\item Task 1 has no prior dependencies other than the general input.
+\item Task\,1 has no prior dependencies other than the general input.
 
-\item Task 1 is a dependency for running Task 2.
+\item Task\,1 is a dependency for running Task\,2 and Task\,3.
 
-\item For simplification purposes, we will assume that Task 1 has the files \texttt{input1.conf}, \texttt{output1.conf}, and \texttt{storage1.conf} as descriptors.
+\item For simplification purposes, we will assume that Task\,1 has the datasets \texttt{input1.conf}, \texttt{output1.conf}, and \texttt{storage1.conf} as descriptors.
 
@@
 
-\item Task 2 is a dependent of Task 1.
+\item Task\,2 is dependent on the outcomes of Task\,1.
 
-\item Task 2 has the files \texttt{input2.conf}, \texttt{output2.conf}, and \texttt{storage2.conf} as descriptors.
+\item Task\,2 has the datasets \texttt{input2.conf}, \texttt{output2.conf}, and \texttt{storage2.conf} as descriptors.
 
@@
 
-\item Task 3 is independent. It does not require any other task to be started and, after finished, no other task will use its results other than the general output.
+\item Task 3 dependent on the outcomes of Task\,1 and Task\,2. It does not require any other task to be started and, after finished, no other task will use its results other than the overall output.
 
-\item Task 3 has the files \texttt{input3.conf}, \texttt{output3.conf}, and \texttt{storage3.conf} as descriptors.
+\item Task 3 has the datasets \texttt{input3.conf}, \texttt{output3.conf}, and \texttt{storage3.conf} as descriptors.
 
@@
 
-\item[Output]
+\item[Product]
 
-The output consists of all files generated by the workflow. Those files have to be stored for a specific amount of time previously described in one of the input files.
+A product is any of the possible outcomes of a task. Usually is composed of datasets that will be used by other tasks or generate the final output.
 
-For simplicity, we will call these files as \texttt{output-workflow.conf}.
-In this specific example, it is composed by the files \texttt{output1.conf}, \texttt{output2.conf}, and \texttt{output3.conf}.
+\item[Final Output]
+
+The final output consists of all datasets generated by the workflow. Those datasets have to be stored for a specific amount of time previously described in one of the input datasets.
+
+For simplicity, we will call these datasets as \texttt{output-workflow.conf}.
+In this specific example, it is composed by the datasets \texttt{output1.conf}, \texttt{output2.conf}, and \texttt{output3.conf}.
 
@@
 
-\begin{figure}[b]
-  \centering
-%  \includegraphics[width=0.75\columnwidth]{workflow}
-  \caption{Example High-Level Workflow}
-  \label{fig:workflow}
-\end{figure}
+For this work, the datasets belonging to \texttt{input-workflow.conf} are the datasets inserted as input for the running Cylc.
+The Cylc tool receives, as predefined, information about the storage and dependencies of the datasets.
+This information is used to construct and run the complete workflow, but it is not used to optimise intermediate steps.
+The configuration is already predefined, and it cannot change.
+For instance, if it is established that Task\,2 depends on Task\,1, then we have to wait for Task\,1 to be completed to start running Task\,2.
+But several options can be optimised if we know the whole workflow in advance.
+Thinking what about I/O, for instance, no decision is being made about an optimal way to store the data.
+Consider, as an example, that the datasets from \texttt{output1.conf} are divided into \texttt{output1-final.conf} and \texttt{output1-task2.conf}.
+Then, it is clear that \texttt{output1-task2.conf} will be required to run Task\,2, and, because of that, should be put in fast storage to quickly being retrieved, or even kept in local storage.
+The \texttt{output1-final.conf} dataset, however, can go directly to tape, if the lifetime of the final output has to be stored for a long time, but no access is required from the current workflow.
 
-For the purposes of this work, the files belonging to \texttt{input-workflow.conf} are the files inserted as input for the running Cylc.
-The Cylc tool receives, as predefined, information about the storage and dependencies of the files.
-These information are used to contructed and run the complete workflow, but it is not used to optimise intermediate steps.
-Some things are already predefined and they cannot change.
-For instante, if it is estabilished that Task 2 depends on Task 1, then we have to wait for Task 1 to be completed to start running Task 3.
-But several things can be optimised if we know the whole workflow in advance.
-For instance: Task 1 and Task 3 have no previous dependencies.
-How do we decide about when we should run those tasks?
-Task 1 first, because it is a dependency for Task 2?
-Task 3 first, because then we have a previous of a complete result, considering Task 3 has no dependencies at all?
-Can Task 1 and Task 3 run in parallel at the same time, and then the workflow only has to complete Task 2 to finish?
-On the contrary, Task 1 is really processing demanding, so it should run alone and then Task 2 and Task 3 can run concurrently without extra burden to the computing time?
-These are all concerns about processing time and the dependencies.
-But what about I/O?
-No decision is being made about an optimal way to store the data.
-Consider, for instance, that the files from \texttt{output1.conf} are divided into \texttt{output1-final.conf} and \texttt{output1-task2.conf}.
-Then, it is clear that \texttt{output1-task2.conf} will be required to run Task 2, and because of that, should be put in a fast storage to quickly being retrieved, of even kept in local storage.
-The \texttt{output1-final.conf} files, however, can go directly to tape, if the lifetime of the final output has to be stored for five years.
-
-Those simple examples are here to demonstrate that, once we have the complete knowledge about how the workflow is expect to run, we can optimise steps that were before predefined by the user considering only the options for feasibility.
+Those simple examples are here to demonstrate that, once we have the complete knowledge about how the workflow is expected to run, we can optimise steps that were before predefined by the user usually considering only options for feasibility.
 The user input is one of the viable options to run the workflow, but probably it is not the best, assuming other options are available.
-The user might even have analysed other options, but s/he has done that in a naive way (usually trial and error).
-By automatising these tasks, specifically a smarter route for the tasks taking into consideration the architecture available and the way the files are stored during the intermediate steps of the workflow, we have a twofold gain:
+The user might even have analysed other options, but s/he has naively done that (usually by trial and error).
+By automating these tasks, specifically a smarter route for the tasks taking into consideration the architecture available and the way the datasets are stored during the intermediate steps of the workflow, we have a twofold gain:
 
@@
 
-In this paper, we approach two ways to have the extra information from the user to be able to make decisions about storage and the processing of the files/tasks.
+In this paper, we approach two ways to have the extra information from the user to be able to make decisions about storage and the processing of the datasets/tasks.
 
@@
 
-\item The user will insert this information as an extra file to the Cylc workflow management.
+\item The user will insert this information as an additional file to the Cylc workflow management.
 In fact, the user is already doing that, but he is doing that in just one specific way, to make the workflow possible.
 Here, we are thinking about simplifying the information the user is already providing.
-So, instead of giving the complete workflow to Cylc, Here, we are expeting the user to provide really simple information, like:
-
-\begin{description}
-
-\item[Dependencies] Run Task 1 and Task 3 in any specific order and then run Task 2 after Task 1 is finished.
-
-\item[Time/Processing]
+So, instead of giving the complete workflow to Cylc, we are expecting the user to provide straightforward information, like:
 
@@
 
-\item Task 3 usually requires five hours running in a pc with standard configuration and it does not require extra storage other than the storage for the final results.
+\item Task 3 usually requires five hours running in a pc with standard configuration, and it does not require extra storage other than the storage for the final results.
 
@@
 
-\end{description}
-
-\item The second option to acess this type of information from the user does not require any direct information at all.
+\item The second option to access this type of information from the user does not require any direct information at all.
 The idea here is to use the DDN IME Monitoring system and have an estimate for the parameters after one run of the complete workflow.
-This approach is benefical to the user, because s/he does not have to take any extra step that s/he is currently doing to run the workflow through Cylc, but it the price here is being paid with an extra run of the complete workflow, which can be demanding in both time and price (we are considering here the paid acess to a supercomputer, for example).
+This approach is beneficial to the user because s/he does not have to take any extra step that s/he is currently doing to run the workflow through Cylc.
+However, the price here is being paid with an additional run of the complete workflow, which can be demanding at both time and expenses (we are considering here the paid access to a supercomputer, for example).
 In the end, it is up to the user and the previous knowledge s/he has about the intermediate stages of the workflow.
@@
 Now, comes the question.
-Assuming the information is available, who will be responsible to take the decisions for optimising the storage and the processing tasks?
+Assuming the information is available, who will be responsible for making the decisions for optimising the storage and the processing tasks?
 The answer here is simple: the ESDM Scheduler will do it.
-ESDM Scheduler can be used in different steps of the process.
+The ESDM Scheduler can be used in different steps of the process.
 ESDM has already an interface with NetCDF.
-Assuming the user is able to provide extra information in advance, ESDM can interact in the process before the user provide the data to the Cylc tool, using the information to already contruct an optimised workflow with Cylc) and, once the workflow is constructed by Cylc, ESDM can interact again after the configuration of the scripts for both effective and efficient run of the parallel applications.
+Assuming the user can provide extra information in advance, ESDM can interact in the process before the user give the data to the Cylc tool, using the information to construct an optimised workflow and, once Cylc builds the workflow, ESDM can interact again after the configuration of the scripts for both effective and efficient run of the parallel applications regarding the available storage and architecture.
 
-This paper intends to describe the information required for the user to run the tools employed in this extended workflow: Cylc (Section ), DDN IME (Section ), and ESDM Scheduler (Section ).
-
-\subsection{Workflow Description}
-\subsection{Task Definition}
-\subsection{Heterogeneous Infrastructure}
+}}
 
@@
 \label{sec:challenges}
-\paragraph{Low-level interfaces:}
-\paragraph{Explicit workflows:}
-\paragraph{Suboptimal Storage Tiering:}
-\paragraph{Task and data equality:}
-\paragraph{Manual optimization:}
-\paragraph{Access pattern intransparency:}
+The current HPC environment faces several challenges and limitations concerning the execution of data-driven workflows.
 
-\section{NGI Vision}
-\label{sec:ngiVision}
+\begin{description}
+
+\item[Low-level Interfaces]
+
+In a nutshell, a stream of bytes is not a workable data model.
+HPC data management and data-flow processing utilise low-level interfaces.
+Storage relies still on traditional file system objects of directories and files -- provided by the POSIX interface or by object storage.
+Underlying storage systems are receiving or delivering streams of bytes, and it cannot make any intelligent decisions about how to optimise data layout for performance, or resiliency.
+It deals with data blocks and files and requires instructions on how to select and control available optimisation strategies.
+
+In practice, applications build up layers so that the application itself is usually working with data mediated by a library which provides application-specific semantics.
+However, at the lowest level, such a library is limited by the underlying storage APIs (POSIX),
+which has performance implications due to the synchronisation in distributed environments by severely restricting the ability of the underlying storage systems to optimise I/O operations.
+Data-driven processing suffers similarly, from the high-level view of processes that are executed on a given computing node.
+
+\item[Explicit Workflows]
+
+Typically, as part of shell scripts, users explicitly encode data processing workflows as a sequence of steps consisting of input, computing, and output.
+The computing-intense parts (jobs) and their dependencies are managed by a cluster resource manager that dispatches them.
+These loosely coupled workflows require that intermediate results are stored on persistent storage to be used by the next task.
+Data transfer, however, needs a dominant fraction of energy compared to computation favouring offloading of computation to storage.
+
+\item[Suboptimal Storage Tiering]
+
+Organising the data placement on storage tiers is typically performed manually by the developer/user or via policies, leading to suboptimal optimisations.
+On top of that, it is a tedious and error-prone task.
+
+Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of files) on a particular storage system and, usually, moving data between storage by using scripts.
+One limitation here is that decisions about how data structures are mapped and packaged into files are made once by the producing application, and cannot be changed without manual intervention by a downstream application.
+
+A policy system (e.g. burst buffer) aims to simplify the data movement for the user, but typically migrates objects in the coarse granularity of files.
+However, the semantical information that can be used by a policy system to make decisions is limited, e.g., data location, file extension, age of the file, etc.
+
+Any hard-coded decision has significant implications for the performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
+In any case, data migration needs resources from the source and the target storage system to copy data.
+Thus, to avoid extra manual work and decisions, it is preferable to direct data to locations where it can be stored and directly used by subsequent workflows -- at best, coupling the data source producing data with a data consumer to prevent storing intermediate products on persistent storage.
+
+\item[Task and Data Equality]
+
+All tasks and data products are handled identically by a HPC system, meaning there is no differentiation between individual jobs and data products.
+This lack of hierarchy is problematic, as different use cases may not only differ in priority and deadline, but the computing speed usually depends on the critical path of the execution.
+
+Fair enough, batch systems understand the priority for jobs.
+However, once a set of nodes is allocated explicitly to a computing job, the job runs exclusively, and it may utilise shared resources such as storage systems and network as any other job.
+
+This problem is even more severe for data products: a storage system provides a single level of performance and resilience and performance for all use cases.
+Theoretically, different fault-tolerance and performance policies can be realised with different storage systems and using management policies.
+However, these policies would be coarse-grained and would not account for the priority of the workflow and, most importantly, the value of data.
+The importance of data depends on aspects like costs to reproduce the data (can it be reproduced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow.
+While value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability, in practice, this is not the case as the storage just sees blocks and files.
+
+\item[Manual Optimisation]
+
+Utilising the computing and storage characteristics of a heterogeneous environment as depicted in \Cref{fig:heterogeneous} is challenging.
+Even within a single data centre, utilising homogeneous storage and computing infrastructure efficiently is complex for experts and needs manual definition.
+The efficient management of data and computing capabilities in a heterogeneous environment and
+how the execution of individual tasks from workflows may benefit from alternative hardware architectures and infrastructures
+are still unresolved questions.
+
+The optimisation of a single parallel job already is quite challenging.
+To make optimal use of available features, programmers and users are forced to think about technical aspects and terms, e.g., setting low-level MPI hints for file striping, while they understand their workflow.
+Similarly, efficient data placement is complex.
+But a specific workflow combined with manual optimisation is the root-cause of subsequent problems, and it must be avoided and replaced with automated and optimised decisions.
+
+\item[Access Pattern in Transparency]
+
+As access patterns are primarily implicitly defined in application code and embedded in workflows, the understanding and debugability of behaviour are limited.
+Firstly, there are many alternative ways of programming data-intensive computation to achieve identical data products.
+Secondly, many storages and computing interfaces are low-level.
+Thus, data centres have often limited insight into the access patterns as well.
+Moreover, some users are abusing the storage system in non-obvious ways, for example, to exchange notifications.
+The understanding and improvement of access patterns are a limited and challenging insight into behaviour leads to suboptimal tuning.
+
+Manual and hard-coded workflows cannot handle changes in the environment and are error-prone.
+Therefore, users must be able to express their workflows in an abstract fashion that allows the system to generate (near-)optimal execution plans and monitor their execution.
+
+\end{description}
+
+\section{Vision}
+
+Our vision for the problems described in the first sections is a new approach for data management and data-driven computation.
+Our proposed model tackles the challenge of storage and data-flow computation in a holistic way accepting that data storage is only aiding the goal of generating insight.
+Besides providing a typical abstract data model, our insight treats domain metadata\footnote{ in addition to that, we mean non-technical information about the data, for instance, the name of an experiment.}, workflows, and information lifecycle management (ILM) as first-class citizens allowing the storage to understand these concepts.
+The user information will be exploited by smart components which aim to utilise capabilities of heterogeneous system landscapes.
+
+To describe the features and benefits of the approach further, we sketch the user and system perspective in the following, as this highlights how we believe we should work in the 21st century.
+The vision covers not only HPC but spans cloud, enterprise, and personalised IT.
 
@@
 
-\paragraph{Data model:}
-\paragraph{Namespace:}
-\paragraph{Workflow specification:}
-\paragraph{Behavioral contracts:}
-\paragraph{Prescriptive analysis:}
+From the user's perspective, the interface and system match the core requirements and expectations, including reliability, availability, and usability.
+
+\begin{description}
+
+\item[Data Model]
+
+The data model is based on state-of-the-art solutions like HDF5+X, where X indicates an extension for ILM, workflows and processing.
+However, it abstracts from data localisation and low-level concepts like files and directories and puts its focus on data organisation and data structures.
+The integration into applications is shown in \Cref{fig:apilayers}.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.5\columnwidth]{layers-api}
+  \caption{Layers with the proposed API}
+  \label{fig:apilayers}
+\end{figure}
+
+Applications may utilise the proposed API directly, or utilise existing middleware\footnote{Which will not be able to use the full set of features if not extending its interface.}.
+From the perspective of a developer of middleware or an application like web publishing, the tool provides a standard feature set that reduces the amount of code replication inside domain-specific solutions.
+
+\item[Namespace]
+
+Data is accessed using information from the domain metadata, i.e., similar to using methods of data indexing services.
+The system will not be relying on a hierarchical namespace, but a dynamic mapping to the POSIX namespace can be provided to enable compatibility for POSIX applications.
+For example, a user may decide to see a hierarchy like an experiment \bnf{Date}/\bnf{Name}/\bnf{Timestep}/\bnf{Task}/\bnf{Dataset}.
+A core observation here is that, although the user may provide suggested namespaces, it is up to the system to place the data into the storage in an optimised way and return the information to the user transparently.
+
+\item[Workflow Specification]
+
+The user defines a workflow describing the intentions of data manipulation and the data lifecycle, including manual (user) involvement in the data processing.
+The definition of constraints, for instance, deadlines for the generation of the final data products (insight), can also be included.
+Workflows can be defined using arbitrary tools, including Python.
+Note that a user will not provide an accurate mapping to system or infrastructure, but expects that the system derives an optimised execution plan that covers heterogeneous infrastructures.
+It does not matter where data is processed and how and where information is stored, but only that the desired result is computed with all constraints met.
+Workflows and data lineage of data products are recorded and can be inspected by the user upon demand, allowing reproduction and validation of the correct execution of experiments.
+
+\item[Behavioral Contracts]
+
+Besides the core characteristics, users expect the execution of their workflows under constraints (including cost and time constraints).
+The system is always expected to show the expected characteristics and shield the user from system-specifics and undesired events (like faults or data loss).
+The characteristics may improve over time, for instance, by expanding hardware capabilities, but the user will not have to change their workflow specification.
+Still, for expert users, interfaces for introspecting system state and decision making can always be provided.
+
+\item[Prescriptive Analysis]
+
+Similarly to an SQL Explain statement, a toolset for users and administrators that allows exploring scenarios provides an unprecedented level of introspection.
+For instance, a prediction for runtime and execution costs of a workflow on a given system is now easier to be achieved.
+Additionally, sharing the workflow description with vendors allows them to conduct predictions for runtime and costs for their solutions.
+The ecosystem may also prescribe alternative systems to use in an open environment, e.g., this part of the workflow should run on data centre A and this part on cloud provider B.
+
+\end{description}
 
@@
 
-\paragraph{Awareness:}
-\paragraph{Resource management:}
-\paragraph{Liquid computing:}
-\paragraph{Smart scheduling:}
-\paragraph{Extensible ecosystem:}
+An implementation of the proposed API exploits the user-supplied information and system characteristics intelligently to meet the user expectations and smartly use heterogeneous computing and storage landscapes.
+It is expected that machine learning techniques will be applied to serve the individual aspects and evolve the systems over time.
+
+\begin{description}
+
+\item[Awareness]
+
+The system knows and understands user metadata and the overarching workflows (with the user intentions).
+It is self-aware, knowing the nominal characteristics of the hardware and software components, continually comparing the system and workflow state with the expectations to perform mitigating actions to heal deviations.
+At best, each hardware component would ship with a behavioural model that can be integrated hierarchically up to a global system model.
+With behavioural models, we not only mean performance models but reliability and cost-models as well.
+
+\item[Resource Management]
+
+The system is in charge to manage and provision the available hardware resources autonomously.
+Ultimately, it aims to utilise the individual characteristics of heterogeneous computing and storage landscapes by using intelligent scheduling.
+Data placement and localisation is beyond mere tiering strategies; parts of data may be placed on particular storage technology or replicated in different representations across storage characteristics.
+In addition to that, you are serving different access patterns and workflows at the same time.
+Naturally, the system will replicate (expand) upon the availability of storage space while reducing redundancy under the shortage of space.
+
+\item[Liquid Computing]
+
+The system is in charge to manage and provision the available hardware resources autonomously and to assign computing and storage resources to tasks.
+In addition to that, it may use, e.g., GPU, CPU, and in-network computing capabilities to serve pieces of a single workflow.
+Moreover, as workflows stretch across data centres and over cloud and edge, an environment that schedules and dispatches work and storage efficiently is vital.
+We call this concept \textbf{liquid computing} as the computing resources are dynamically provisioned and assigned while data flows through the system are processed.
+Naturally, processing closer to the data sources is likely to improve system throughput.
+The programming for the processing will be based on a high-level functional language providing primitives for data-flow computing and compiled into a virtual machine or machine code when the code is to be dispatched.
+The capabilities of different hardware components may be limited, realising only a subset of the language which must be taken into account by the scheduler.
+
+\item[Smart Scheduling]
+
+Supplied with user workflows and their characteristics, the system derives execution plans for the computation and make data placement decisions that aim to utilise the available hardware resources efficiently.
+As the API is in charge of both tasks, it may decide to trade computation and storage capacity considering characteristics as costs and energy consumption, for example.
+Additionally, it can enable automatic recomputation of intermediate states utilising virtualisation and container technologies.
+Similarly, the scheduler may decide to couple subsequent steps of a workflow directly without storing intermediate data products on persistent storage.
+Data might be replicated but enables the system to rerun parts of the workflow in case of a data loss.
+
+Automatic scheduling is key to efficient processing and will allow constant improvement without forcing the user to adjust their workflows.
+In most cases, individual resources are not used exclusively by a single task, but QoS methods ensure that constraints of workflows are met.
+Note that this is a significant change for HPC environments, as computing resources are typically dedicated to specific jobs.
+
+\item[Extensible Ecosystem]
+
+The hardware and software landscape will be open, allowing to integrate solutions from different hardware and software vendors.
+Connected systems may use different interfaces, including traditional POSIX interfaces, NoSQL, or distributed hash tables.
+Imagine that a specific class of workflows performs inefficiently on a given infrastructure indicating that a crucial processing or storage capability is missing.
+Adding new capabilities or resources to the system will allow the scheduler to use them in subsequent user-workflows speeding up the processing, what should be now as easy as adding an appliance to a business environment.
+
+\end{description}
+
+{ \color{blue}{
+
+\section{Interaction with Cylc}
+
+The focus of this paper is to exemplify the I/O optimisations that can be made if the datasets dependencies are established.
+This information, in fact, is presented in the workflows before the Cylc application, so nothing else will be required from the user.
+But to have the information explicitly, Cylc needs some changes to be able to:
+
+\begin{enumerate}
+
+\item deliver it together with the workflow, and
+
+\item adapts the resulting workflow to consider the I/O modifications.
+
+\end{enumerate}
+
+Consider the Cylc suite configuration for a toy monthly cycling workflow in Figure \ref{fig:cylc}.
+A warm-cycled atmospheric model (\textbf{model}) is followed by postprocessing (\textbf{post}), forecast verification (\textbf{ver}), and product generation (\textbf{prod}) tasks.
+Also, consider, in addition, a task \textbf{check} that compares some verification metric against products from two cycles earlier.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.9\columnwidth]{cylc}
+  \caption{Toy example workflow}
+  \label{fig:cylc}
+\end{figure}
+
+This suite presents the dependencies among the tasks, but it is missing the information about I/O and storage.
+The developers usually make the configuration about where the data be stored in stages before Cylc and then it is incorporated into the Cylc process.
+The decisions about storage are created taking into consideration the architecture in which the workflow will run.
+Applied scientists should not spend time understanding hardware characteristics, but using their time to develop their work, let the computer do the job the simplest way possible, and just collect and analyse the results.
+However, nowadays, to be able to run a job in a HPC environment, researchers have to have profound knowledge about their workflow, which is expected, but also about decisions regarding storage, communication, and computing.
+In this paper, we propose an approach to reduce the burden on researchers and, at the same time, optimise the decisions about jobs running in HPC systems.
+Once we have an automated decision about where the job will run and how the storage will be managed, the scientists can then run their workflows in any machine without further interactions and even without previous knowledge about the machine architecture.
+
+Figure \ref{fig:cycle1} introduces one cycle of the suite in Figure \ref{fig:cylc} with a simplified notation.
+The arrows represent the tasks dependencies, but there is no information about the data that are generated.
+Tasks \textbf{verification} and \textbf{post\,proc} depend on Task \textbf{model}, but does Task \textbf{model} needs to be fully completed to satisfy the dependency?
+Of course, once Task \textbf{model} is completed, the dependent tasks can be executed, and that is the type of information Cylc has to create and reproduce the workflow.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.4\columnwidth]{cycle1}
+  \caption{First stage workflow}
+  \label{fig:cycle1}
+\end{figure}
+
+The scenario presented in Figure \ref{fig:cycle-io} shows the same cycle of Figure \ref{fig:cycle1}, but the extended information about datasets is introduced.
+In this picture, we have the tasks dependencies, and we explicit the datasets dependencies regarding the tasks that generated them.
+The way data moves through tasks is by creating datasets to pass on the information, including filename and storage destination.
+The datasets produced by every task are coded in the workflow scripts before Cylc evaluation, and it is just replicated by this tool.
+The tasks can create a different number of datasets, and these datasets may be necessary for the immediate next task in the current cycle, or tasks in different cycles.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{cycle-io}
+  \caption{First stage workflow including datasets}
+  \label{fig:cycle-io}
+\end{figure}
+
+Figure \ref{fig:cycle-io-dep} now introduces one example of dependencies among the datasets and tasks: datasets colours indicate which task will use their information and datasets in yellow represent information that will be used in a different cycle.
+The tricky part when we consider only tasks and not the datasets in the workflow is that it may lead to the false impression that the tasks only depend on the immediately previous one.
+Note, for instance, that one dataset produced by Task \textbf{model} is later used in Task \textbf{prod\,gen}.
+That is not an immediate conclusion when we analyse just the arrows for the dependencies.
+It is clear that Task \textbf{model} is a dependency for Task \textbf{post\,prod}, which is then a dependency for Task \textbf{prod\,gen}, but the crucial information here is that any dataset currently being used in the current cycle can also be used later.
+From the figure we can also see that Task \textbf{model} will generate one dataset for Task \textbf{post\,proc}, one dataset for Task \textbf{prod\,gen}, two datasets for Task \textbf{verification}, and three datasets for a later cycle.
+The access to this information before the process starts can be used to optimise how these datasets will be stored during the workflow.
+This information today is static and manual, leaving opportunities for automatisation and optimisations open.
+The idea here is to embrace the concept that tasks dependencies are really datasets dependencies.
+Once we can insert the datasets explicitly in the workflow, it is easier to see that many optimisations regarding storage can be made.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{cycle-io-dep}
+  \caption{First stage workflow including datasets and dependencies}
+  \label{fig:cycle-io-dep}
+\end{figure}
+
+Figure \ref{fig:cycle-4} exemplifies another restriction of the way current workflows are being generated.
+Assume you have, for instance, four cycles of the model you are running.
+Because the scripts for the workflow consider individual tasks, the datasets generated by them are placed at the same storage type every time.
+That happens because all datasets generated by Task \textbf{model} will be used by Task \textbf{post\,prod}, so they should stay "closer" when thinking about memory.
+However, although this decision is based on an optimised concept, it still restricts the way the system deals with the datasets and optimisations that are not made thinking only about the local structure.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.8\columnwidth]{cycle-4}
+  \caption{Four stages workflow example}
+  \label{fig:cycle-4}
+\end{figure}
+
+Datasets generated by Task \textbf{A} and used by Task \textbf{B} will have the same physical location in the system because there is one fixed script responsible for setting the configuration.
+Again, such configuration was done manually and with restricted information about the storage and system architecture.
+It would be interesting to explore the options for the datasets and have them placed in different storage systems.
+If the lifetime of such dataset is, for instance, two cycles, alternate the datasets into two scratch dataset systems is something that would be a simple job for a tool working on generating filenames and storage.
+However, manually, that implies at having at least two scripts with information about the different storage and extra work to have better utilisation of the available resources.
+
+The information about the system architecture can be provided in advance as a configuration file, and we can use monitoring to identify the I/O patterns for a specific workflow.
+For that, all we need is to run one cycle of the workflow and detect the patterns associated with it.
+For this purpose, we intend to use Machine Learning to identify the patterns and adapt the workflow in case of unexpected situations (the system is down, nodes are unresponsive, etc.).
+With this separation of concepts, we can abstract from the workflow what is essential and has to be provided by the user and what a machine should optimise to ensure that the resources available are being used smartly.
+Every time parameters are manually tuned, that has a room from improvement by using automatised decision making.
+And Machine Learning has been proven effective in replacing human-beings in several problems.
+
+Optimisations we can envision about the I/O are related to the life cycle of datasets and the placement of such datasets into specific memory according to their utilisation.
+What happens in most current workflows is that usually, they use only two types of file systems: \textbf{work} and \textbf{scratch}.
+When a task is set to computing, the corresponding dataset is moved to \textbf{scratch}, processed, and the resulting datasets are transferred back to \textbf{work}.
+If the \textbf{scratch} filesystem achieves its capacity, then the datasets move back to \textbf{work} and keep running the task until it is finished.
+If the I/O planning is made a priori, this situation can be avoided.
+In addition to the file systems described, many supercomputers also have available for the user \textbf{scratch2} fyle systems (sometimes local and global).
+Therefore, in a real scenario, if the architecture of the machine is known, we can better use its resources.
+
+Figure \ref{fig:lifecycle} finalises the arguments in this paper with a vision of three possible life cycles for one specific dataset.
+The workflow script defines the way the dataset is handled according to the storage, and the machine has only one of the possible scenarios.
+The scientist responsible for the script tried to optimise the available resources intuitively before deliver the final version.
+Would not be better to have the machine automating and making such decisions?
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{lifecycle}
+  \caption{Life cycle of one dataset: \textbf{C}reate, \textbf{M}igrate, \textbf{R}ead, and \textbf{D}elete}
+  \label{fig:lifecycle}
+\end{figure}
+
+}}
 
@@
 
+}}
+
+\begin{comment}
+
+and computer-aided Research Development and Engineering (RD\&E)
+
+and also on small-scale systems.
+
+For example, Big data tools integrate computing and storage capabilities into a holistic solution demonstrating the benefit of tight integrating for data-driven workflows.
+
+%-- the quality of service features available in hardware is rarely used.
+
+Cloud definition of desired state and contracts.
+
+Indeed many research prototypes address subproblems
+* But not all aspects together
+* Competing approaches; the standardisation we describe here does not compete!
+
+In this paper, we purposely do not mention too many of the ongoing RD\&E in storage and computing.
+
+To wrap up, NGI provides a high-level view on data and processing on heterogeneous platforms while shielding the user from execution details.
+
+Open ...
+
+%Kubernetes
+
+\begin{description}
+
+\item{NFS} Relatively slow, persistent file system - Nightly incremental backups
+
+\item{Lustre} Large, fast distributed scratch file system - No backups
+
+\url{https://info.hpc.sussex.ac.uk/hpc-guide/storage.html}
+
+\end{description}
+
+%, and data products are generated which are processes that are executed outside of the first simulation.
+
+\end{comment}
+
 \end{document}

### 2020-01-23 paper/paper.tex
Abstract

@@
 
-{ \color{cyan}{
+{ \color{black}{
 
@@
 Traditionally, in HPC, the concerns of storage and computing are separated and optimised independently from each other, and, moreover, independently from the needs of end-to-end user workflows that ultimately lead to insight.
-As workflows become increasingly complex and blend beyond data centres while at the same time, storage hierarchies become deeper, the community investigates means to re-organise storage access to utilise such environments fully.
+As climate and weather workflows become increasingly complex and blend beyond data centres while at the same time, storage hierarchies become deeper, the community investigates means to re-organise storage access to utilise such environments fully.
 
-In this white paper, we describe an initiative that aims to tackle storage and computing challenges mentioned above in a holistic approach by fusing state-of-the-art concepts into a revolutionary interface by rethinking how data-driven I/O should be done.
-The tools employed here to achieve this extended workflow are Cylc, DDN IME, and ESDM Scheduler.
-Basically, we believe workflows composed of data, computing, and communication-intensive tasks should drive the interfaces and hardware configurations to best support these programming models.
-From the user perspective, we aim to provide an abstraction to data-driven processing and data management.
-The proposed interface ultimately will increase the opportunity of implementations for smarter scheduling of computing and storage in heterogeneous environments considering characteristics of workflows and data.
-From the system perspective, the vision aims to derive an execution plan that utilises the available hardware and software infrastructure within and across data centres.
+The key contributions of this paper are:
+1) we sketch the visions of an integrated data-driven approach and discuss the challenges and implications of this strategy.
+2) we illustrate an architecture and roadmap that allows the seamless integration into existing workflows.
+The tools employed here to achieve this extended workflow are Cylc, XIOS, DDN IME, and ESDM.
+
+%Basically, we believe workflows composed of data, computing, and communication-intensive tasks should drive the interfaces and hardware configurations to best support the programming models.
+From the user perspective, we ultimately aim to provide an abstraction to data dependencies.
+The proposed changes increase the opportunity of implementations for smarter scheduling of computing and storage in heterogeneous storage environments considering characteristics of workflows and data.
+From the system perspective, the vision aims to derive an execution plan that utilises the available hardware and software infrastructure within a data centre.
 

### 2020-01-23 paper/paper.tex
Nai
Removed sections from NGI unneeded.

@@
 \documentclass[a4paper]{article}
-\newcommand{\papertitle}{Whitepaper -- Optimising Storage Inside Workflows}
+\newcommand{\papertitle}{Potential of IO-Aware Workflows in Climate and Weather}
 
@@
 
-\section{Workflows}
+\section{Workflows in Climate/Weather}
+
+This is how workflows in climate/weather look like with Cylc.
+Picture with XIOS software stack (julian' has that one.).
 
@@
 We focus on high-level considerations to illustrate mandatory concepts.
+
+\section{Vision for Workflows Considering IO}
+
+What changes now in terms of specification?
+
+\subsection{Potential Benefit}
 
@@
 
-{ \color{blue}{
+{ \color{black}{
 
@@
 
-For simplicity, we will call these datasets as \texttt{input-workflow.conf}.
+For simplicity, we will call these datasets as \texttt{input.ds}.
 
@@
 
-\item For simplification purposes, we will assume that Task\,1 has the datasets \texttt{input1.conf}, \texttt{output1.conf}, and \texttt{storage1.conf} as descriptors.
+\item For simplification purposes, we will assume that Task\,1 has the datasets \texttt{input1.ds}, \texttt{output1.ds}, and \texttt{storage1.ds} as descriptors.
 
@@
 
-\item Task\,2 has the datasets \texttt{input2.conf}, \texttt{output2.conf}, and \texttt{storage2.conf} as descriptors.
+\item Task\,2 has the datasets \texttt{input2.ds}, \texttt{output2.ds}, and \texttt{storage2.ds} as descriptors.
 
@@
 
-\item Task 3 has the datasets \texttt{input3.conf}, \texttt{output3.conf}, and \texttt{storage3.conf} as descriptors.
+\item Task 3 has the datasets \texttt{input3.ds}, \texttt{output3.ds}, and \texttt{storage3.ds} as descriptors.
 
@@
 
-For simplicity, we will call these datasets as \texttt{output-workflow.conf}.
-In this specific example, it is composed by the datasets \texttt{output1.conf}, \texttt{output2.conf}, and \texttt{output3.conf}.
+For simplicity, we will call these datasets as \texttt{output-workflow.ds}. In this specific example, it is composed by the datasets \texttt{output1.ds}, \texttt{output2.ds}, and \texttt{output3.ds}.
 
@@
 
-For this work, the datasets belonging to \texttt{input-workflow.conf} are the datasets inserted as input for the running Cylc.
+For this work, the datasets belonging to \texttt{input-workflow.ds} are the datasets inserted as input for the running Cylc.
 The Cylc tool receives, as predefined, information about the storage and dependencies of the datasets.
@@
 Thinking what about I/O, for instance, no decision is being made about an optimal way to store the data.
-Consider, as an example, that the datasets from \texttt{output1.conf} are divided into \texttt{output1-final.conf} and \texttt{output1-task2.conf}.
-Then, it is clear that \texttt{output1-task2.conf} will be required to run Task\,2, and, because of that, should be put in fast storage to quickly being retrieved, or even kept in local storage.
-The \texttt{output1-final.conf} dataset, however, can go directly to tape, if the lifetime of the final output has to be stored for a long time, but no access is required from the current workflow.
+Consider, as an example, that the datasets from \texttt{output1.ds} are divided into \texttt{output1-final.ds} and \texttt{output1-task2.ds}.
+Then, it is clear that \texttt{output1-task2.ds} will be required to run Task\,2, and, because of that, should be put in fast storage to quickly being retrieved, or even kept in local storage.
+The \texttt{output1-final.ds} dataset, however, can go directly to tape, if the lifetime of the final output has to be stored for a long time, but no access is required from the current workflow.
 
@@
 
-\section{Challenges and Limitations}
-\label{sec:challenges}
-The current HPC environment faces several challenges and limitations concerning the execution of data-driven workflows.
+{ \color{black}{
 
-\begin{description}
-
-\item[Low-level Interfaces]
-
-In a nutshell, a stream of bytes is not a workable data model.
-HPC data management and data-flow processing utilise low-level interfaces.
-Storage relies still on traditional file system objects of directories and files -- provided by the POSIX interface or by object storage.
-Underlying storage systems are receiving or delivering streams of bytes, and it cannot make any intelligent decisions about how to optimise data layout for performance, or resiliency.
-It deals with data blocks and files and requires instructions on how to select and control available optimisation strategies.
-
-In practice, applications build up layers so that the application itself is usually working with data mediated by a library which provides application-specific semantics.
-However, at the lowest level, such a library is limited by the underlying storage APIs (POSIX),
-which has performance implications due to the synchronisation in distributed environments by severely restricting the ability of the underlying storage systems to optimise I/O operations.
-Data-driven processing suffers similarly, from the high-level view of processes that are executed on a given computing node.
-
-\item[Explicit Workflows]
-
-Typically, as part of shell scripts, users explicitly encode data processing workflows as a sequence of steps consisting of input, computing, and output.
-The computing-intense parts (jobs) and their dependencies are managed by a cluster resource manager that dispatches them.
-These loosely coupled workflows require that intermediate results are stored on persistent storage to be used by the next task.
-Data transfer, however, needs a dominant fraction of energy compared to computation favouring offloading of computation to storage.
-
-\item[Suboptimal Storage Tiering]
-
-Organising the data placement on storage tiers is typically performed manually by the developer/user or via policies, leading to suboptimal optimisations.
-On top of that, it is a tedious and error-prone task.
-
-Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of files) on a particular storage system and, usually, moving data between storage by using scripts.
-One limitation here is that decisions about how data structures are mapped and packaged into files are made once by the producing application, and cannot be changed without manual intervention by a downstream application.
-
-A policy system (e.g. burst buffer) aims to simplify the data movement for the user, but typically migrates objects in the coarse granularity of files.
-However, the semantical information that can be used by a policy system to make decisions is limited, e.g., data location, file extension, age of the file, etc.
-
-Any hard-coded decision has significant implications for the performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
-In any case, data migration needs resources from the source and the target storage system to copy data.
-Thus, to avoid extra manual work and decisions, it is preferable to direct data to locations where it can be stored and directly used by subsequent workflows -- at best, coupling the data source producing data with a data consumer to prevent storing intermediate products on persistent storage.
-
-\item[Task and Data Equality]
-
-All tasks and data products are handled identically by a HPC system, meaning there is no differentiation between individual jobs and data products.
-This lack of hierarchy is problematic, as different use cases may not only differ in priority and deadline, but the computing speed usually depends on the critical path of the execution.
-
-Fair enough, batch systems understand the priority for jobs.
-However, once a set of nodes is allocated explicitly to a computing job, the job runs exclusively, and it may utilise shared resources such as storage systems and network as any other job.
-
-This problem is even more severe for data products: a storage system provides a single level of performance and resilience and performance for all use cases.
-Theoretically, different fault-tolerance and performance policies can be realised with different storage systems and using management policies.
-However, these policies would be coarse-grained and would not account for the priority of the workflow and, most importantly, the value of data.
-The importance of data depends on aspects like costs to reproduce the data (can it be reproduced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow.
-While value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability, in practice, this is not the case as the storage just sees blocks and files.
-
-\item[Manual Optimisation]
-
-Utilising the computing and storage characteristics of a heterogeneous environment as depicted in \Cref{fig:heterogeneous} is challenging.
-Even within a single data centre, utilising homogeneous storage and computing infrastructure efficiently is complex for experts and needs manual definition.
-The efficient management of data and computing capabilities in a heterogeneous environment and
-how the execution of individual tasks from workflows may benefit from alternative hardware architectures and infrastructures
-are still unresolved questions.
-
-The optimisation of a single parallel job already is quite challenging.
-To make optimal use of available features, programmers and users are forced to think about technical aspects and terms, e.g., setting low-level MPI hints for file striping, while they understand their workflow.
-Similarly, efficient data placement is complex.
-But a specific workflow combined with manual optimisation is the root-cause of subsequent problems, and it must be avoided and replaced with automated and optimised decisions.
-
-\item[Access Pattern in Transparency]
-
-As access patterns are primarily implicitly defined in application code and embedded in workflows, the understanding and debugability of behaviour are limited.
-Firstly, there are many alternative ways of programming data-intensive computation to achieve identical data products.
-Secondly, many storages and computing interfaces are low-level.
-Thus, data centres have often limited insight into the access patterns as well.
-Moreover, some users are abusing the storage system in non-obvious ways, for example, to exchange notifications.
-The understanding and improvement of access patterns are a limited and challenging insight into behaviour leads to suboptimal tuning.
-
-Manual and hard-coded workflows cannot handle changes in the environment and are error-prone.
-Therefore, users must be able to express their workflows in an abstract fashion that allows the system to generate (near-)optimal execution plans and monitor their execution.
-
-\end{description}
-
-\section{Vision}
-
-Our vision for the problems described in the first sections is a new approach for data management and data-driven computation.
-Our proposed model tackles the challenge of storage and data-flow computation in a holistic way accepting that data storage is only aiding the goal of generating insight.
-Besides providing a typical abstract data model, our insight treats domain metadata\footnote{ in addition to that, we mean non-technical information about the data, for instance, the name of an experiment.}, workflows, and information lifecycle management (ILM) as first-class citizens allowing the storage to understand these concepts.
-The user information will be exploited by smart components which aim to utilise capabilities of heterogeneous system landscapes.
-
-To describe the features and benefits of the approach further, we sketch the user and system perspective in the following, as this highlights how we believe we should work in the 21st century.
-The vision covers not only HPC but spans cloud, enterprise, and personalised IT.
-
-\subsection{User Perspective}
-
-\newcommand{\bnf}[1]{\textless #1\textgreater}
-
-From the user's perspective, the interface and system match the core requirements and expectations, including reliability, availability, and usability.
-
-\begin{description}
-
-\item[Data Model]
-
-The data model is based on state-of-the-art solutions like HDF5+X, where X indicates an extension for ILM, workflows and processing.
-However, it abstracts from data localisation and low-level concepts like files and directories and puts its focus on data organisation and data structures.
-The integration into applications is shown in \Cref{fig:apilayers}.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.5\columnwidth]{layers-api}
-  \caption{Layers with the proposed API}
-  \label{fig:apilayers}
-\end{figure}
-
-Applications may utilise the proposed API directly, or utilise existing middleware\footnote{Which will not be able to use the full set of features if not extending its interface.}.
-From the perspective of a developer of middleware or an application like web publishing, the tool provides a standard feature set that reduces the amount of code replication inside domain-specific solutions.
-
-\item[Namespace]
-
-Data is accessed using information from the domain metadata, i.e., similar to using methods of data indexing services.
-The system will not be relying on a hierarchical namespace, but a dynamic mapping to the POSIX namespace can be provided to enable compatibility for POSIX applications.
-For example, a user may decide to see a hierarchy like an experiment \bnf{Date}/\bnf{Name}/\bnf{Timestep}/\bnf{Task}/\bnf{Dataset}.
-A core observation here is that, although the user may provide suggested namespaces, it is up to the system to place the data into the storage in an optimised way and return the information to the user transparently.
-
-\item[Workflow Specification]
-
-The user defines a workflow describing the intentions of data manipulation and the data lifecycle, including manual (user) involvement in the data processing.
-The definition of constraints, for instance, deadlines for the generation of the final data products (insight), can also be included.
-Workflows can be defined using arbitrary tools, including Python.
-Note that a user will not provide an accurate mapping to system or infrastructure, but expects that the system derives an optimised execution plan that covers heterogeneous infrastructures.
-It does not matter where data is processed and how and where information is stored, but only that the desired result is computed with all constraints met.
-Workflows and data lineage of data products are recorded and can be inspected by the user upon demand, allowing reproduction and validation of the correct execution of experiments.
-
-\item[Behavioral Contracts]
-
-Besides the core characteristics, users expect the execution of their workflows under constraints (including cost and time constraints).
-The system is always expected to show the expected characteristics and shield the user from system-specifics and undesired events (like faults or data loss).
-The characteristics may improve over time, for instance, by expanding hardware capabilities, but the user will not have to change their workflow specification.
-Still, for expert users, interfaces for introspecting system state and decision making can always be provided.
-
-\item[Prescriptive Analysis]
-
-Similarly to an SQL Explain statement, a toolset for users and administrators that allows exploring scenarios provides an unprecedented level of introspection.
-For instance, a prediction for runtime and execution costs of a workflow on a given system is now easier to be achieved.
-Additionally, sharing the workflow description with vendors allows them to conduct predictions for runtime and costs for their solutions.
-The ecosystem may also prescribe alternative systems to use in an open environment, e.g., this part of the workflow should run on data centre A and this part on cloud provider B.
-
-\end{description}
-
-\subsection{System Perspective}
-
-An implementation of the proposed API exploits the user-supplied information and system characteristics intelligently to meet the user expectations and smartly use heterogeneous computing and storage landscapes.
-It is expected that machine learning techniques will be applied to serve the individual aspects and evolve the systems over time.
-
-\begin{description}
-
-\item[Awareness]
-
-The system knows and understands user metadata and the overarching workflows (with the user intentions).
-It is self-aware, knowing the nominal characteristics of the hardware and software components, continually comparing the system and workflow state with the expectations to perform mitigating actions to heal deviations.
-At best, each hardware component would ship with a behavioural model that can be integrated hierarchically up to a global system model.
-With behavioural models, we not only mean performance models but reliability and cost-models as well.
-
-\item[Resource Management]
-
-The system is in charge to manage and provision the available hardware resources autonomously.
-Ultimately, it aims to utilise the individual characteristics of heterogeneous computing and storage landscapes by using intelligent scheduling.
-Data placement and localisation is beyond mere tiering strategies; parts of data may be placed on particular storage technology or replicated in different representations across storage characteristics.
-In addition to that, you are serving different access patterns and workflows at the same time.
-Naturally, the system will replicate (expand) upon the availability of storage space while reducing redundancy under the shortage of space.
-
-\item[Liquid Computing]
-
-The system is in charge to manage and provision the available hardware resources autonomously and to assign computing and storage resources to tasks.
-In addition to that, it may use, e.g., GPU, CPU, and in-network computing capabilities to serve pieces of a single workflow.
-Moreover, as workflows stretch across data centres and over cloud and edge, an environment that schedules and dispatches work and storage efficiently is vital.
-We call this concept \textbf{liquid computing} as the computing resources are dynamically provisioned and assigned while data flows through the system are processed.
-Naturally, processing closer to the data sources is likely to improve system throughput.
-The programming for the processing will be based on a high-level functional language providing primitives for data-flow computing and compiled into a virtual machine or machine code when the code is to be dispatched.
-The capabilities of different hardware components may be limited, realising only a subset of the language which must be taken into account by the scheduler.
-
-\item[Smart Scheduling]
-
-Supplied with user workflows and their characteristics, the system derives execution plans for the computation and make data placement decisions that aim to utilise the available hardware resources efficiently.
-As the API is in charge of both tasks, it may decide to trade computation and storage capacity considering characteristics as costs and energy consumption, for example.
-Additionally, it can enable automatic recomputation of intermediate states utilising virtualisation and container technologies.
-Similarly, the scheduler may decide to couple subsequent steps of a workflow directly without storing intermediate data products on persistent storage.
-Data might be replicated but enables the system to rerun parts of the workflow in case of a data loss.
-
-Automatic scheduling is key to efficient processing and will allow constant improvement without forcing the user to adjust their workflows.
-In most cases, individual resources are not used exclusively by a single task, but QoS methods ensure that constraints of workflows are met.
-Note that this is a significant change for HPC environments, as computing resources are typically dedicated to specific jobs.
-
-\item[Extensible Ecosystem]
-
-The hardware and software landscape will be open, allowing to integrate solutions from different hardware and software vendors.
-Connected systems may use different interfaces, including traditional POSIX interfaces, NoSQL, or distributed hash tables.
-Imagine that a specific class of workflows performs inefficiently on a given infrastructure indicating that a crucial processing or storage capability is missing.
-Adding new capabilities or resources to the system will allow the scheduler to use them in subsequent user-workflows speeding up the processing, what should be now as easy as adding an appliance to a business environment.
-
-\end{description}
-
-{ \color{blue}{
-
-\section{Interaction with Cylc}
+\section{Extending Existing Workflows}
 
@@
 
+\section{Challenges and Limitations}
+\label{sec:challenges}
+The current HPC environment faces several challenges and limitations concerning the execution of data-driven workflows.
+
+\begin{description}
+\item[Explicit Workflows]
+
+Typically, as part of shell scripts, users explicitly encode data processing workflows as a sequence of steps consisting of input, computing, and output.
+The computing-intense parts (jobs) and their dependencies are managed by a cluster resource manager that dispatches them.
+These loosely coupled workflows require that intermediate results are stored on persistent storage to be used by the next task.
+Data transfer, however, needs a dominant fraction of energy compared to computation favouring offloading of computation to storage.
+
+\item[Suboptimal Storage Tiering]
+
+Organising the data placement on storage tiers is typically performed manually by the developer/user or via policies, leading to suboptimal optimisations.
+On top of that, it is a tedious and error-prone task.
+
+Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of files) on a particular storage system and, usually, moving data between storage by using scripts.
+One limitation here is that decisions about how data structures are mapped and packaged into files are made once by the producing application, and cannot be changed without manual intervention by a downstream application.
+
+A policy system (e.g. burst buffer) aims to simplify the data movement for the user, but typically migrates objects in the coarse granularity of files.
+However, the semantical information that can be used by a policy system to make decisions is limited, e.g., data location, file extension, age of the file, etc.
+
+Any hard-coded decision has significant implications for the performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
+In any case, data migration needs resources from the source and the target storage system to copy data.
+Thus, to avoid extra manual work and decisions, it is preferable to direct data to locations where it can be stored and directly used by subsequent workflows -- at best, coupling the data source producing data with a data consumer to prevent storing intermediate products on persistent storage.
+
+\item[Task and Data Equality]
+
+All tasks and data products are handled identically by a HPC system, meaning there is no differentiation between individual jobs and data products.
+This lack of hierarchy is problematic, as different use cases may not only differ in priority and deadline, but the computing speed usually depends on the critical path of the execution.
+
+This problem is even more severe for data products: a storage system provides a single level of performance and resilience and performance for stored files.
+Theoretically, different fault-tolerance and performance policies can be realised with different storage systems and using management policies.
+However, these policies would be coarse-grained and would not account for the priority of the workflow and, most importantly, the value of data.
+The importance of data depends on aspects like costs to reproduce the data (can it be reproduced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow.
+While value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability, in practice, this is not the case as the storage just sees blocks and files.
+
+\item[Manual Optimisation]
+
+Utilising the computing and storage characteristics of a heterogeneous environment as depicted in \Cref{fig:heterogeneous} is challenging.
+Even within a single data centre, utilising homogeneous storage and computing infrastructure efficiently is complex for experts and needs manual definition.
+The efficient management of data and computing capabilities in a heterogeneous environment and
+how the execution of individual tasks from workflows may benefit from alternative hardware architectures and infrastructures
+are still unresolved questions.
+
+The optimisation of a single parallel job already is quite challenging.
+To make optimal use of available features, programmers and users are forced to think about technical aspects and terms, e.g., setting low-level MPI hints for file striping, while they understand their workflow.
+Similarly, efficient data placement is complex.
+But a specific workflow combined with manual optimisation is the root-cause of subsequent problems, and it must be avoided and replaced with automated and optimised decisions.
+
+Manual and hard-coded workflows cannot handle changes in the environment and are error-prone.
+Therefore, users must be able to express their workflows in an abstract fashion that allows the system to generate (near-)optimal execution plans and monitor their execution.
+
+\end{description}
+
+\section{Vision}
+
+Our vision for the problems described in the first sections is a new approach for data management and data-driven computation.
+
+\begin{description}
+\item[Workflow Specification]
+
+The user defines a workflow describing the intentions of data manipulation and the data lifecycle, including manual (user) involvement in the data processing.
+The definition of constraints, for instance, deadlines for the generation of the final data products (insight), can also be included.
+Workflows can be defined using arbitrary tools, including Python.
+Note that a user will not provide an accurate mapping to system or infrastructure, but expects that the system derives an optimised execution plan that covers heterogeneous infrastructures.
+It does not matter where data is processed and how and where information is stored, but only that the desired result is computed with all constraints met.
+Workflows and data lineage of data products are recorded and can be inspected by the user upon demand, allowing reproduction and validation of the correct execution of experiments.
+
+\item[Smart Scheduling]
+
+Supplied with user workflows and their characteristics, the system derives execution plans for the computation and make data placement decisions that aim to utilise the available hardware resources efficiently.
+As the API is in charge of both tasks, it may decide to trade computation and storage capacity considering characteristics as costs and energy consumption, for example.
+Additionally, it can enable automatic recomputation of intermediate states utilising virtualisation and container technologies.
+Similarly, the scheduler may decide to couple subsequent steps of a workflow directly without storing intermediate data products on persistent storage.
+Data might be replicated but enables the system to rerun parts of the workflow in case of a data loss.
+
+Automatic scheduling is key to efficient processing and will allow constant improvement without forcing the user to adjust their workflows.
+In most cases, individual resources are not used exclusively by a single task, but QoS methods ensure that constraints of workflows are met.
+Note that this is a significant change for HPC environments, as computing resources are typically dedicated to specific jobs.
+
+\end{description}
+
 \end{document}

### 2020-01-23 paper/paper.tex
Intro

@@
 \usepackage{comment}
+
+\newcommand{\jk}[1]{\todo[inline]{JK: #1}}
 
@@
 
-High-Performance Computing harnesses the fastest hardware components to enable
-the execution of tightly coupled applications from science and industry.
+High-Performance Computing harnesses the fastest hardware components to enable the execution of tightly coupled applications from science and industry.
 Typical use-cases cover the numerical simulation of physical systems or the analysis of large-scale observational data.
-While the hardware of HPC systems is on the bleeding edge, the software landscape of the infrastructure lacks behind modern concepts found in domains like cloud computing.
-There are various reasons for the lack of innovation, e.g., the market capitalisation is limited; many HPC applications have a lifetime of sometimes several decades\footnote{For example, CTH at Sandia is more than 30 years old, but it has evolved over the years as hardware has changed.
-It is not uncommon that codes take ten years to write and receive a first optimised setup.
-After that, they run for another 30 years with constant evolution based on platform changes and functionality expansions.}; the goal in HPC is to get science done and not to have modern and user-friendly systems.
-Therefore, software developed to target the HPC market is rarely used outside data centres; while there is no reason to use developed applications outside this scope, software to aid developers or for managing HPC infrastructure suffers the same cause.
-Nowadays, HPC adapts infrastructure software from the enterprise and cloud market as much as possible; at best, adapting (stretching) already existing software packages to the needs of HPC.
+In the domain of climate and weather, there is a huge demand for the orchestration of ensembles of simulation models and the generation of data products.
+A weather forecast service such as ECMWF generates .... \jk{Cylc team, please provide some facts}
+Likewise ensembles in climate can ...
 
-From the perspective of big data and cloud solutions, the state-of-the-art in the HPC software ecosystem is perceived as conservative and outdated at least.
-For example, HPC still uses batch processing and dedicates nodes to the execution of individual applications which were the first concepts implemented at the dawn of computing.
-Similar to writing code in assembler, often it is argued that HPC trades user-friendliness and manageability to obtain the best possible performance.
-However, as the complexity of systems increases and various alternative storage and computing technologies provide unique characteristics, it is barely possible to define optimal data-driven workflows using low-level constructs for the following reasons:
-
-\begin{description}
-
-  \item[Computing Focus] HPC considers computing, storage, and network as separate resources and focuses on the computing aspects, e.g., I/O requires to retrieve/store data from a computing node.
-  While there are good reasons for this strategy as they provide a beautiful separation of concerns,  there are workflows that run efficiently with alternative organisations because they take into consideration the opportunity to process data closer to storage or between applications.
-
-  \item[Complexity] A data-driven workflow is highly complex and may benefit from using a heterogeneous set of computing and storage technology at the same time.
-  While there are good reasons for this strategy, there are workflows that also run efficiently with alternative organisations.
+Based on their needs, the community has developed a software ecosystem that supports them to execute their large-scale workflows.
+While a big leap forward, many processes still require experts.
+For example, porting a workflow from one system to another requires to adjust runtime parameters of applications and to decide how data is managed.
+Since performance is of key importance to large-scale workflows, careful attention must be paid to exploit the system characteristics of the target supercomputer.
+To obtain best performance, a specific supercomputer typically requires substantial changes to the workflow that tailor the workflow to the specific machine.
+For instance, a data-driven workflow may benefit from using a heterogeneous set of computing and storage technology at the same time.
 
 Knowing the capabilities, interfaces, and performance characteristics of individual components are mandatory to make the best use of them.
+As the complexity of systems increases and various alternative storage and computing technologies provide unique characteristics, it becomes increasingly difficult even for experts to manually optimize IO in workflows.
 
-  \item[Portability] Moving from one infrastructure to another or even when modifying the infrastructure requires to rethink and manually adapt the workflow as the characteristics of the overall environment changes.
+In many cases, modifications are not performed because: 1) they are labor intense: any change to the workflow requires careful validation which may not pay off for small scale runs; 2) users are not aware of the potential of the complex system.
 
-  \item[Cost-efficiency] It is time-consuming to utilise low-level constructs and they often only increase the performance slightly. Also, the cost of labour must be taken into account.
-  From the perspective of scientists and engineers, the time-to-solution from experimental design to insight matters and the cost-savings for the public data centre can be disregarded.
+In this paper, we show that knowledge about workflow execution and overall experimental design helps to
+optimise execution of climate and weather workflows and may increase the performance, throughput and cost-efficiency of the environments, providing an incentive to users and data-centres that cannot be neglected.
 
-  \item[Conservative Users] Users are reluctant to change their workflows because they already know that older workflows are proven to be correct and, therefore, they want to avoid the intensive labour to adapt workflows manually.
-
-\end{description}
-
-For these reasons, workflows are typically encoded manually and run suboptimally on HPC systems.
-Additionally, users of public data centre services supply limited information about their experiments, typically the level of detail is like "I need X CPU hours and Y storage capacity" and they have not to prove their ability to use these systems efficiently.
-Still, the systems these workflows run on cost up to 100 million dollars.
-Comparing this amount to experiments conducted in physics like the Large Hadron Collider (LHC) at CERN or the Square Kilometer Array (SKA), it becomes clear that the execution of complex experiments may need years of planning involving the scientists.
-While it is essential to keep some freedom to conduct ad-hoc experiments, we raise the question of what could be achieved by systems and operators knowing workflows in more detail.
-We argue that knowledge about workflow execution and overall experimental design helps to
-optimise execution of workflows and additionally increase the performance, throughput and cost-efficiency of the environments, providing an incentive to users and data-centres that cannot be neglected.
-
-With the Next Generation Interfaces initiative, we aim to define a high-level interface that, firstly, meets the needs of users and data centres, and, secondly, that allows implementations to exploit the capabilities of current and future storage and computing landscapes.
+The paper is organized as follows:
+\jk{TODO}
 
@@
 This is how workflows in climate/weather look like with Cylc.
-Picture with XIOS software stack (julian' has that one.).
+
+\begin{figure}[bt]
+  \includegraphics[width=0.3\textwidth]{layers-xios}
+  \caption{IO Path for an MPI-Parallel Application}
+  \label{fig:layers}
+\end{figure}
 
@@
 We focus on high-level considerations to illustrate mandatory concepts.
+
+Consider the Cylc suite configuration for a toy monthly cycling workflow in Figure \ref{fig:cylc}.
+A warm-cycled atmospheric model (\textbf{model}) is followed by postprocessing (\textbf{post}), forecast verification (\textbf{ver}), and product generation (\textbf{prod}) tasks.
+Also, consider, in addition, a task \textbf{check} that compares some verification metric against products from two cycles earlier.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.9\columnwidth]{cylc}
+  \caption{Example workflow\cite{TODO}}
+  \label{fig:cylc}
+\end{figure}
+
+This suite presents the dependencies among the tasks, but it is missing the information about I/O and storage.
+The developers usually make the configuration about where the data be stored in stages before Cylc and then it is incorporated into the Cylc process.
+The decisions about storage are created taking into consideration the architecture in which the workflow will run.
+Applied scientists should not spend time understanding hardware characteristics, but using their time to develop their work, let the computer do the job the simplest way possible, and just collect and analyse the results.
+However, nowadays, to be able to run a job in a HPC environment, researchers have to have profound knowledge about their workflow, which is expected, but also about decisions regarding storage, communication, and computing.
+In this paper, we propose an approach to reduce the burden on researchers and, at the same time, optimise the decisions about jobs running in HPC systems.
+Once we have an automated decision about where the job will run and how the storage will be managed, the scientists can then run their workflows in any machine without further interactions and even without previous knowledge about the machine architecture.
 
@@
 
-Consider the Cylc suite configuration for a toy monthly cycling workflow in Figure \ref{fig:cylc}.
-A warm-cycled atmospheric model (\textbf{model}) is followed by postprocessing (\textbf{post}), forecast verification (\textbf{ver}), and product generation (\textbf{prod}) tasks.
-Also, consider, in addition, a task \textbf{check} that compares some verification metric against products from two cycles earlier.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.9\columnwidth]{cylc}
-  \caption{Toy example workflow}
-  \label{fig:cylc}
-\end{figure}
-
-This suite presents the dependencies among the tasks, but it is missing the information about I/O and storage.
-The developers usually make the configuration about where the data be stored in stages before Cylc and then it is incorporated into the Cylc process.
-The decisions about storage are created taking into consideration the architecture in which the workflow will run.
-Applied scientists should not spend time understanding hardware characteristics, but using their time to develop their work, let the computer do the job the simplest way possible, and just collect and analyse the results.
-However, nowadays, to be able to run a job in a HPC environment, researchers have to have profound knowledge about their workflow, which is expected, but also about decisions regarding storage, communication, and computing.
-In this paper, we propose an approach to reduce the burden on researchers and, at the same time, optimise the decisions about jobs running in HPC systems.
-Once we have an automated decision about where the job will run and how the storage will be managed, the scientists can then run their workflows in any machine without further interactions and even without previous knowledge about the machine architecture.
-
 Figure \ref{fig:cycle1} introduces one cycle of the suite in Figure \ref{fig:cylc} with a simplified notation.

### 2020-01-24 paper/paper.tex
Nai
Intermediate

@@
 The paper is organized as follows:
+In \Cref{sec:workflows}, we provide an overview for the specification of workflows in climate and weather, and the involved software stack.
+
 \jk{TODO}
@@
 \section{Workflows in Climate/Weather}
+\label{sec:workflows}
 
-This is how workflows in climate/weather look like with Cylc.
+In this section, we describe the environment workflows are executed in and the typical software stack.
 
-\begin{figure}[bt]
-  \includegraphics[width=0.3\textwidth]{layers-xios}
+\subsection{Data Center Infrastructure}
+
+To satisfy the needs for different workflows, data centers are providing a infrastructure consisting of compute and storage devices with different characteristics -- that make them more efficient for specific tasks.
+Take for example the supercomputer Mistral at DKRZ that consists of about 3,000 nodes.
+It offers two types of compute nodes equipped with different CPUs and a range of GPU nodes.
+Each node provides a SSD for local storage.
+DKRZ has additionally two shared Lustre file systems with different performance characteristics.
+
+Future centers are likely to have even more heterogeneity: A variety of accelerators (GPU, TPU, FPGAs), active storage, in-memory, and in-network computing technologies provide storage and processing capabilities.
+\Cref{fig:heterogeneous} shows such an environment with a focus on computation and storage.
+Some of these technologies might be local to specific compute nodes or globally available.
+Depending on the need, the characteristics of storage range from predictable low-latency (in-memory storage, NVMe) to online storage (SSD, HDD), and cheap storage for long-term archival (tape).
+\jk{Mention BBs, e.g., DDN IME}
+Naturally, the variety of tasks executed by a single workflow may benefit from the different storage and computing infrastructure.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{system}
+  \caption{Example of an heterogeneous HPC landscape}
+  \label{fig:heterogeneous}
+\end{figure}
+
+\subsection{Software Stack}
+
+The relevant software stack involved in executing workflows is depicted in \Cref{XXX}
+\jk{insert picture about software stack here}
+
+\subsection{IO Stack}
+
+A typical IO stack for a single parallel application like a climate model is shown in \Cref{fig:layers}.
+
+\begin{minipage}{0.2\textwidth}
+\begin{figure}[H]
+  \includegraphics[width=\textwidth]{layers-xios}
   \caption{IO Path for an MPI-Parallel Application}
@@
 \end{figure}
+\end{minipage}
+\qquad
+\begin{minipage}{0.7\textwidth}
+In our example, we assume the application is parallelized for performance reasons and uses MPI to coordinate the application.
+It may use XIOS to gather data from individual fields which then uses NetCDF4 to store the data as a file.
+XIOS provides a domain-level semantics and ...\jk{TODO}
+Under the hood, NetCDF4 used the HDF5 API and file format.
+Internally, HDF5 uses MPI again and its data types to specify the type of the data stored.
+Finally, data is stored on a parallel file system like Lustre, which -- on the server-side -- stores data in a local file system that is stored on block devices and storage media such as SSDs and HDDs.
 
-This section provides an introduction to workflows that solve computational problems, the mandatory steps to encode and setup tasks and describe data, and the relevance of heterogeneous infrastructure to run workflows.
-We focus on high-level considerations to illustrate mandatory concepts.
+Different applications involved in a workflow may use a different IO stack to store their outputs.
+Naturally, the application that has to input previously generated data must use a compatible API to be able to read the specific data format.
+In this figure, for example, XIOS is beneficial for parallel I/O and a process might directly rely on the NetCDF API to read in previous data.
+\end{minipage}
+
+\subsection{Cylc}
+
+Cylc is in charge to execute and monitor workflows of which each step is submitted to the batch scheduler of a data center.
+ %, and the relevance of heterogeneous infrastructure to run workflows.
+%We focus on high-level considerations to illustrate mandatory concepts.
 
 Consider the Cylc suite configuration for a toy monthly cycling workflow in Figure \ref{fig:cylc}.
-A warm-cycled atmospheric model (\textbf{model}) is followed by postprocessing (\textbf{post}), forecast verification (\textbf{ver}), and product generation (\textbf{prod}) tasks.
+In this workflow, a warm-cycled atmospheric model (\textbf{model}) simulates the physics from a current state to predict the future after a timestep, for example, after six hours.
+This process is repeated in the model to simulate month and years into the future.
+Once the simulation of a month is computed, the data of this month can be analyzed.
+A user has specified a detailed workflow in a Cylc suite file (shown on the left).
+In this workflow, the \textbf{model} is followed by postprocessing (\textbf{post}), forecast verification (\textbf{ver}), and product generation (\textbf{prod}) tasks.
 Also, consider, in addition, a task \textbf{check} that compares some verification metric against products from two cycles earlier.
@@
 This suite presents the dependencies among the tasks, but it is missing the information about I/O and storage.
-The developers usually make the configuration about where the data be stored in stages before Cylc and then it is incorporated into the Cylc process.
-The decisions about storage are created taking into consideration the architecture in which the workflow will run.
+The developers make the configuration about parallelism and where the data be stored in a script file that is executed for the specific task.
+The decisions about storage must take into consideration the architecture in which the workflow will run.
+
+\section{State of the Art}
+
+Other tools for workflow handling, how is this information transported?
+dataspaces
+ADIOS
+
+\section{Potential for IO-Aware Workflows}
+
 Applied scientists should not spend time understanding hardware characteristics, but using their time to develop their work, let the computer do the job the simplest way possible, and just collect and analyse the results.
-However, nowadays, to be able to run a job in a HPC environment, researchers have to have profound knowledge about their workflow, which is expected, but also about decisions regarding storage, communication, and computing.
+However, nowadays, to be able to run a job in a HPC environment efficiently, researchers have to have profound knowledge about their workflow, which is expected, but also about decisions regarding storage, communication, and computing.
+
 In this paper, we propose an approach to reduce the burden on researchers and, at the same time, optimise the decisions about jobs running in HPC systems.
@@
 What changes now in terms of specification?
-
-\subsection{Potential Benefit}
 
@@
 \item tasks can be annotated with the necessary information to start processing or with performance models that bound the expectations for resources (time, memory, storage capacity).
-
 \end{itemize}
 
-\subsection{Task Definition}
-
-The high-level interaction of a task with its dependencies is generally specified in the workflow.
-A computing task is encoded in a programming language by a developer\footnote{We often treat developer and HPC user as synonyms as a fraction of HPC users are developers for parallel applications to some extent.}.
-A team of developers encodes the algorithm that solves the problem. In many cases, scientists that generate a numerical simulation implement the model programmatically.
-Often, they use existing libraries that provide a piece of functionality or infrastructure.
-In the design phase of the program, a parallel programming paradigm and suitable data structures that allow efficient data processing are chosen.
-In the world of parallel computing, the developer must also consider strategies to parallelise the code and data processing, which often involve the use of APIs and runtime systems that allow the efficient usage of supercomputers.
-
-While most parts of the algorithm can be a tightly coupled simulation, as part of the algorithm, the input data must be fetched.
-
-Typically, before the input data can be used, some kind of pre-processing must be applied. Similarly, in the output, several known post-processing steps may be applied.
-Also, the programmer is responsible for making the optimal decisions on how to map data structures from memory to I/O calls and place these datasets onto storage systems.
-We consider that pre/post-processing are part of the data-driven scope of workflows.
-Indeed, we claim that the programmer should be able to define the input/output on an abstract fashion and connect processing steps as part of the general workflow.
-
-\subsection{Heterogeneous Infrastructure}
-
-The variety of tasks executed by a single workflow may benefit from heterogeneous storage and computing infrastructure and potentially even span multiple locations.
-\Cref{fig:heterogeneous} shows such an environment with a focus on computation and storage.
-A variety of accelerators (GPU, TPU, FPGAs), active storage, in-memory, and in-network computing technologies provide processing capabilities across and beyond a HPC centre optimised for specific workloads.
-Data centres typically provide more than one storage and computing class with different characteristics to optimise cost and efficiency.
-Likewise, storage is deployed data centre wide, locally available to individual nodes, or subsets of nodes (e.g., racks).
-Depending on the need, the characteristics range from predictable low-latency (in-memory storage, NVMe) to online storage (SSD, HDD), and cheap storage for long-term archival (tape).
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.6\columnwidth]{system}
-  \caption{Example of an heterogeneous HPC landscape}
-  \label{fig:heterogeneous}
-\end{figure}
-
-The processing across geographic locations is nowadays mandatory to tame the data volumes of observational or simulation data.
-Also, the lines between cloud computing and HPC are blurry.
-Attempts are made to augment processing by flexibly migrate bursts of computation to cloud environments, ingest data from cloud sources or provided to customers.
-Consider a simple scenario where sensor data captured is pre-processed (cleaned) locally, then enriched with an additional data source in a small-scale HPC facility or by using fog computing, and then finally transferred to a data centre that performs the number crunching and generates data products.
-These data products can also be made available on cloud storage to customers for individual processing.
+\subsection{Potential Benefit}
 

### 2020-01-24 paper/paper.tex
Further

@@
 DKRZ has additionally two shared Lustre file systems with different performance characteristics.
+The user can access both file systems as a work and scratch file system.
+While data is kept on the work file system indefinitely its available space is limited by a quota.
+The scratch file system allows to store more data but data is automatically deleted after a period of time.
 
@@
 \jk{insert picture about software stack here}
+To solve a computational problem, we can consider describing the necessary steps from the start to the insight as a workflow.
+Cylc executes a workflow description comprising of the individual data processing steps and the dependencies among them.
 
@@
 
+\subsection{Data Management}
+
+The scripts for the tasks in Cylc define how a dataset is stored on the available storage systems.
+\Cref{fig:lifecycle} shows three possible life cycles for a specific dataset and the access pattern to the dataset.
+Consider that the machine has three file systems: a fast scratch file system on which data may reside only for a week, a slower work file system, and each node has a local file system.
+The dataset could be stored on the local storage first to avoid congestion on the work file system, then be migrated to work file system where subsequent tasks of the workflow may read it multiple times.
+At the end, this dataset might be an intermediate product that can then be deleted.
+Alternatively, it could be stored on the scratch file system immediately and accessed there.
+However, that would require that the last read access is within the 7-day period before files on scratch are automatically removed.
+The last scenario shows the case where data is created on work by a task but then copied to a local node to allow read accesses of subsequent tasks.
+This might be beneficial for small random accesses.
+However, it would require that subsequent tasks are placed on the same machine where data is now copied to.
+
+The scientist responsible for the workflow must optimise the available resources intuitively.
+These are just examples and scientists perform such kind of optimizations already manually.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.6\columnwidth]{lifecycle}
+  \caption{Life cycle of one dataset: \textbf{C}reate, \textbf{M}igrate, \textbf{R}ead, and \textbf{D}elete}
+  \label{fig:lifecycle}
+\end{figure}
+
 \section{State of the Art}
@@
 
-\section{Potential for IO-Aware Workflows}
+\section{Vision for IO-Aware Workflows}
 
@@
 
-In this paper, we propose an approach to reduce the burden on researchers and, at the same time, optimise the decisions about jobs running in HPC systems.
+We propose an approach to reduce the burden on researchers and, at the same time, optimise the decisions about jobs running in HPC systems.
 Once we have an automated decision about where the job will run and how the storage will be managed, the scientists can then run their workflows in any machine without further interactions and even without previous knowledge about the machine architecture.
-
-\section{Vision for Workflows Considering IO}
-
-What changes now in terms of specification?
 
@@
 
-To solve a computational problem, we can consider describing the necessary steps from the start to the insight as a workflow.
-Such a workflow description must comprise of the individual data processing steps, the dependencies among them, the input, and the expected products.
-An example execution of a workflow is illustrated in \Cref{fig:workflow} as a graph.
+So far, Cylc allowed to specify tasks, and the dependencies among them.
+We aim to enrich this information with characteristics for input, and the products.
+An example workflow containing data products is illustrated in \Cref{fig:workflow}.
 Arrows indicate dependencies between nodes that represent tasks and data.
@@
 Most of the workflow can run automatically except the manual quality control of the products and the final data usage of Product\,3 in the end.
-The last step represents the typical uncertainty of data reuse. When this workflow is analysed, it is unclear how Product\,3 will be used further, but, apparently, it will be.
-Further workflows that use the data may be created similarly and use products as input.
+The last step represents the typical uncertainty of data reuse, i.e., it is unclear how Product\,3 will be used further.
+In this approach, each task is annotated with the required inputs and the generated products.
+Each data product is annotated with metadata such as data life cycle information, e.g., the value of data and how long should it be kept.
 
@@
 
-Note that a realistic workflow will be much more complex and involves not a single parallel job execution, but many to generate a robust answer to the problem in question.
-Extra information can be, for instance:
+\subsection{Potential Benefit}
 
-\begin{itemize}
-
-\item The nodes can be annotated with characteristics of the individual steps;
-
-\item Data can be annotated with metadata describing its content, the expected data life cycle information, e.g., the value of data and how long should it be kept;
-
-\item tasks can be annotated with the necessary information to start processing or with performance models that bound the expectations for resources (time, memory, storage capacity).
-\end{itemize}
-
-\subsection{Potential Benefit}
+\section{Design}
+What changes now in terms of specification?
 
@@
 
-Figure \ref{fig:lifecycle} finalises the arguments in this paper with a vision of three possible life cycles for one specific dataset.
-The workflow script defines the way the dataset is handled according to the storage, and the machine has only one of the possible scenarios.
-The scientist responsible for the script tried to optimise the available resources intuitively before deliver the final version.
-Would not be better to have the machine automating and making such decisions?
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.6\columnwidth]{lifecycle}
-  \caption{Life cycle of one dataset: \textbf{C}reate, \textbf{M}igrate, \textbf{R}ead, and \textbf{D}elete}
-  \label{fig:lifecycle}
-\end{figure}
-
 }}

### 2020-01-26 paper/script.sh
Script

@@ -1,0 +1,24 @@
+#!/bin/bash
+
+# export CYLC_TASK_CYCLE_POINT=2012-08
+OF=$(cylc cycle-point --offset-month=1 --template=/home/k202020/experiment1/cycle-temperature-CCYY-MM.nc)
+
+IF1=$(cylc cycle-point --offset-month=-1 --template=/home/k202020/experiment1/athmosphere-CCYY-MM.nc)
+IF2=$(cylc cycle-point --offset-month=-2 --template=/home/k202020/experiment1/athmosphere-CCYY-MM.nc)
+IF3=$(cylc cycle-point --offset-month=-3 --template=/home/k202020/experiment1/athmosphere-CCYY-MM.nc)
+IF4=$(cylc cycle-point --offset-month=-4 --template=/home/k202020/experiment1/athmosphere-CCYY-MM.nc)
+
+
+# Model needs input data wind speeds for athmosphere (start data) and topology
+# What do I need 5 files athmosphere Cycle now up to -5
+# I generate one output file current cycle-temperate.nc
+
+echo """
+SIMULATED_TIME=1 month
+CURRENT_TIME=$CYLC_TASK_CYCLE_POINT
+DRIVER_ATHMOSPHERE=[$IF1, $IF2, $IF3, $IF4]
+TOPOLOGY_FILE=/work/pool/echam/topology15degree.nc
+OUTPUT_FILE=$OF
+""" > model.conf
+
+mpiexec -np 10 ./model model.conf

### 2020-01-26 paper/paper.tex
Lucy + me

@@
 
+Now:
+\begin{enumerate}
+  \item User runs Cylc to start a suite
+  \item Cylc parses the suite file, generate dependencies and generate a schedule for the execution
+  \item Once a task can be executed because dependencies are ready: Cylc generates on the fly a Slurm script with the required metadata for SLURM and invoking the user-defined script for that task and then runs Slurm to queue the script
+  \item Slurm queues the task and once the task is ready to be dispatched it is started on the supercomputer
+  \item The shell-script executes the users-provided instructions, some of those are to generate file names using Cylc and the variables that are generated by Cylc.
+These filenames use templates provide by the user that define the storage location to be used.
+  \item User-specified applications run taking the filenames as instructed.
+  The application may use NetCDF or XIOS to perform the IO.
+
+\end{enumerate}
+
+Our Plan:
+
+\begin{enumerate}
+  \item User runs Cylc to start a suite.
+  \item Cylc parses the suite file, generate dependencies and generate a schedule for the execution.
+
+  Now, Cylc will call into EWIOScheduler to potentially reprioritize the order of execution of tasks.
+  The EWIOScheduler will read the suite file and the suite-io.rc file to make suggestions but won't enforce anything.
+
+  \item Once a task can be executed because dependencies are ready: Cylc generates on the fly a Slurm script with the required metadata for SLURM and invoking the user-defined script for that task.
+
+  Potentially, Cylc will change the metadata of the script telling Slurm to place a job on the same nodes as a previous job to reusue local data.
+  Potentially, Cylc may also add as metadata information to migrate data.
+
+  Then runs Slurm to queue the task.
+
+  \item Slurm queues the task and once the task is ready to be dispatched it is started on the supercomputer
+
+  If a migration has to be performed before or after a job is run, it would do so.
+
+  \item The shell-script executes the users-provided instructions, some of those are to generate file names using Cylc and the variables that are generated by Cylc.
+  These filenames use templates provide by the user that define the storage location to be used.
+
+  The way filenames are generated is now slightly different, using a new command that then uses ESWIOSchedule.
+  The guy reads the suite file(s) and generates information for ESDM where to prioritize to place or access data. The filename hence encoded additional information.
+  Potentially, we generate an esdm.conf file to reflect the specific hardware we want to use.
+
+  \item User-specified applications run taking the filenames as instructed.
+    The application may use NetCDF, XIOS
+
+    OR ESDM
+
+    to perform the IO.
+
+  ESDM loads the system configuration knowing the potential targets and to schedule application-local IO.
+  Application uses ESDM with the script filename, ESDM extracts the long-term schedule information and considers it during "application-local" IO scheduling.
+
+  \item A data management service may run and purge or migrate data according to user-specifications of the information life cycle.
+
+\end{enumerate}
+
 }}

### 2020-01-27 paper/paper.tex
Some changes worth of committing

@@
 \documentclass[a4paper]{article}
-\newcommand{\papertitle}{Potential of IO-Aware Workflows in Climate and Weather}
+\newcommand{\papertitle}{Potential of I/O-Aware Workflows in Climate and Weather}
 
@@
 \usepackage{comment}
+\usepackage{ulem}
 
@@
 
-{ \color{black}{
-
 \section*{Abstract}
 The efficient, convenient, and robust execution of data-driven workflows and enhanced data management are key for productivity in scientific computing.
-Traditionally, in HPC, the concerns of storage and computing are separated and optimised independently from each other, and, moreover, independently from the needs of end-to-end user workflows that ultimately lead to insight.
-As climate and weather workflows become increasingly complex and blend beyond data centres while at the same time, storage hierarchies become deeper, the community investigates means to re-organise storage access to utilise such environments fully.
+\sout{{\color{blue}{
+Traditionally, in HPC, the concerns of storage and computing are separated and optimised independently from each other, and, moreover, independently from the needs of end-to-end user workflows that ultimately lead to insight.}}}
+{\color{cyan}{
+Traditionally, in HPC, the concerns of storage and computing are separated and optimised independently from each other and from the needs of end-to-end user.}}
+As climate and weather workflows become increasingly complex and blend beyond data centres while, at the same time, storage hierarchies become deeper, the community investigates means to re-organise storage access to utilise such environments fully.
 
@@
 1) we sketch the visions of an integrated data-driven approach and discuss the challenges and implications of this strategy.
-2) we illustrate an architecture and roadmap that allows the seamless integration into existing workflows.
+2) we illustrate architecture and roadmap that allows the seamless integration into existing workflows.
 The tools employed here to achieve this extended workflow are Cylc, XIOS, DDN IME, and ESDM.
@@
 
-High-Performance Computing harnesses the fastest hardware components to enable the execution of tightly coupled applications from science and industry.
-Typical use-cases cover the numerical simulation of physical systems or the analysis of large-scale observational data.
-In the domain of climate and weather, there is a huge demand for the orchestration of ensembles of simulation models and the generation of data products.
+High-Performance Computing (HPC) harnesses the fastest hardware components to enable the execution of tightly coupled applications from science and industry.
+Typical use-cases cover the numerical simulation of physical systems and the analysis of large-scale observational data.
+In the domain of climate and weather, there is a considerable demand for the orchestration of ensembles of simulation models and the generation of data products.
 A weather forecast service such as ECMWF generates .... \jk{Cylc team, please provide some facts}
-Likewise ensembles in climate can ...
+Likewise, ensembles in climate can ...
 
-Based on their needs, the community has developed a software ecosystem that supports them to execute their large-scale workflows.
-While a big leap forward, many processes still require experts.
+Based on their needs, the {\color{cyan}{HPC}} community has developed a software ecosystem that supports them to execute their large-scale workflows.
+While the current advances correspond to a big leap forward, many processes still require experts.
 For example, porting a workflow from one system to another requires to adjust runtime parameters of applications and to decide how data is managed.
-Since performance is of key importance to large-scale workflows, careful attention must be paid to exploit the system characteristics of the target supercomputer.
-To obtain best performance, a specific supercomputer typically requires substantial changes to the workflow that tailor the workflow to the specific machine.
+Since performance is of crucial importance to large-scale workflows, careful attention must be paid to exploit the system characteristics of the target supercomputer.
+To obtain the best performance, a specific supercomputer typically requires substantial changes to the workflow to tailor the workflow to the particular machine.
 For instance, a data-driven workflow may benefit from using a heterogeneous set of computing and storage technology at the same time.
@@
 Knowing the capabilities, interfaces, and performance characteristics of individual components are mandatory to make the best use of them.
-As the complexity of systems increases and various alternative storage and computing technologies provide unique characteristics, it becomes increasingly difficult even for experts to manually optimize IO in workflows.
-
-In many cases, modifications are not performed because: 1) they are labor intense: any change to the workflow requires careful validation which may not pay off for small scale runs; 2) users are not aware of the potential of the complex system.
+As the complexity of systems increases and alternative storage and computing technologies provide unique characteristics, it becomes increasingly difficult even for experts to manually optimise I/O in workflows.
+{\color{cyan}{First time I/O is mentioned.
+I would change to something like...
+it becomes increasingly difficult even for experts to manually optimise resources (other stuff) and, in particular, I/O in such workflows.}}
+In many cases, modifications are not performed because: 1) they are labour intense: any change to the workflow requires careful validation which may not pay off for small scale runs; 2) users are not aware of the potential of the complex system.
 
 In this paper, we show that knowledge about workflow execution and overall experimental design helps to
-optimise execution of climate and weather workflows and may increase the performance, throughput and cost-efficiency of the environments, providing an incentive to users and data-centres that cannot be neglected.
+optimise the execution of climate and weather workflows and may increase the performance, throughput and cost-efficiency of the environments, providing an incentive to users and data-centres that cannot be neglected.
 
-The paper is organized as follows:
-In \Cref{sec:workflows}, we provide an overview for the specification of workflows in climate and weather, and the involved software stack.
+The paper is organised as follows:
+In \Cref{sec:workflows}, we provide an overview of the specification of workflows in climate and weather and the involved software stack.
 
@@
 
-In this section, we describe the environment workflows are executed in and the typical software stack.
+In this section, we describe how the environment workflows are executed in the typical software stack.
 
@@
 
-To satisfy the needs for different workflows, data centers are providing a infrastructure consisting of compute and storage devices with different characteristics -- that make them more efficient for specific tasks.
-Take for example the supercomputer Mistral at DKRZ that consists of about 3,000 nodes.
+To satisfy the needs for different workflows, data centres are providing a infrastructure consisting of computing and storage devices with different characteristics which makes them more efficient for specific tasks.
+Take, for example, the supercomputer Mistral at DKRZ that consists of about {\color{cyan}{about?}} 3,000 nodes.
 It offers two types of compute nodes equipped with different CPUs and a range of GPU nodes.
-Each node provides a SSD for local storage.
-DKRZ has additionally two shared Lustre file systems with different performance characteristics.
-The user can access both file systems as a work and scratch file system.
-While data is kept on the work file system indefinitely its available space is limited by a quota.
-The scratch file system allows to store more data but data is automatically deleted after a period of time.
+Each node provides an SSD for local storage and DKRZ has additionally two shared Lustre file systems with different performance characteristics.
+The user can access both file systems \sout{{\color{blue}{as a work and scratch file system}}}. {\color{cyan}{as work and scratch file systems.}}
+While data is kept on the work file system indefinitely, its available space is limited by a quota.
+The scratch file system allows to store more data, but data is automatically deleted after some time.
 
-Future centers are likely to have even more heterogeneity: A variety of accelerators (GPU, TPU, FPGAs), active storage, in-memory, and in-network computing technologies provide storage and processing capabilities.
+Future centres are expected to have even more heterogeneity.
+A variety of accelerators (GPU, TPU, FPGAs), active storage, in-memory, and in-network computing technologies provide storage and processing capabilities.
 \Cref{fig:heterogeneous} shows such an environment with a focus on computation and storage.
 Some of these technologies might be local to specific compute nodes or globally available.
-Depending on the need, the characteristics of storage range from predictable low-latency (in-memory storage, NVMe) to online storage (SSD, HDD), and cheap storage for long-term archival (tape).
+Depending on the need, the storage characteristics range from predictable low-latency (in-memory storage, NVMe) to online storage (SSD, HDD), and cheap storage for long-term archival (tape).
 \jk{Mention BBs, e.g., DDN IME}
@@
 \jk{insert picture about software stack here}
-To solve a computational problem, we can consider describing the necessary steps from the start to the insight as a workflow.
+To solve a computational problem, we can consider describing the necessary steps from the start to the insight {\color{cyan}{insight?
+I really don't undestand some of your uses of this word.}} as a workflow.
 Cylc executes a workflow description comprising of the individual data processing steps and the dependencies among them.
 
-\subsection{IO Stack}
+\subsection{I/O Stack}
 
-A typical IO stack for a single parallel application like a climate model is shown in \Cref{fig:layers}.
+A typical I/O stack for a single parallel application like a climate model is shown in \Cref{fig:layers}.
 
@@
   \includegraphics[width=\textwidth]{layers-xios}
-  \caption{IO Path for an MPI-Parallel Application}
+  \caption{I/O Path for an MPI-Parallel Application}
   \label{fig:layers}
@@
 \begin{minipage}{0.7\textwidth}
-In our example, we assume the application is parallelized for performance reasons and uses MPI to coordinate the application.
+In our example, we assume the application is parallelised for performance reasons and uses MPI to \sout{{\color{blue}{coordinate}}} {\color{cyan}{run}} the application.
 It may use XIOS to gather data from individual fields which then uses NetCDF4 to store the data as a file.
@@
 Under the hood, NetCDF4 used the HDF5 API and file format.
-Internally, HDF5 uses MPI again and its data types to specify the type of the data stored.
-Finally, data is stored on a parallel file system like Lustre, which -- on the server-side -- stores data in a local file system that is stored on block devices and storage media such as SSDs and HDDs.
+Internally, HDF5 uses MPI again and its data types to specify the nature of the data stored.
+\sout{{\color{blue}{Finally, data is stored on a parallel file system like Lustre, which, on the server-side, stores data in a local file system that is stored on block devices and storage media such as SSDs and HDDs.}}}
+{\color{cyan}{Finally, data is stored on a parallel file system like Lustre, which, on the server-side, stores data in local file system as block devices on storage media such as SSDs and HDDs.}}
 
-Different applications involved in a workflow may use a different IO stack to store their outputs.
-Naturally, the application that has to input previously generated data must use a compatible API to be able to read the specific data format.
-In this figure, for example, XIOS is beneficial for parallel I/O and a process might directly rely on the NetCDF API to read in previous data.
+Different applications involved in a workflow may use a different I/O stack to store their outputs.
+Naturally, the application that inputs previously generated data must use a compatible API to be able to read the specific data format.
+In this figure, for example, XIOS is beneficial for parallel I/O and a process might directly rely on the NetCDF API to read previous data.
 \end{minipage}
@@
 
-Cylc is in charge to execute and monitor workflows of which each step is submitted to the batch scheduler of a data center.
+Cylc is in charge to execute and monitor workflows in which each step is submitted to the batch scheduler of a data centre.
  %, and the relevance of heterogeneous infrastructure to run workflows.
@@
 This process is repeated in the model to simulate month and years into the future.
-Once the simulation of a month is computed, the data of this month can be analyzed.
+Once the simulation of a month is computed, the data for this month can be analysed.
+{\color{cyan}{This is wuite strange for me.
+It means that only after a month is completed one will use the information to start previsions?
+I would expect it to be more like a window-month than a fixed month, no?}}
 A user has specified a detailed workflow in a Cylc suite file (shown on the left).
@@
   \includegraphics[width=0.9\columnwidth]{cylc}
-  \caption{Example workflow\cite{TODO}}
+  \caption{Example workflow\cite{8675433}}
   \label{fig:cylc}
@@
 This suite presents the dependencies among the tasks, but it is missing the information about I/O and storage.
-The developers make the configuration about parallelism and where the data be stored in a script file that is executed for the specific task.
-The decisions about storage must take into consideration the architecture in which the workflow will run.
+Decisions about storage must take into consideration the architecture in which the workflow will run.
+The developers make the configuration about parallelism and where the data is stored in a script file that is executed for each specific task.
 
@@
 
-The scripts for the tasks in Cylc define how a dataset is stored on the available storage systems.
-\Cref{fig:lifecycle} shows three possible life cycles for a specific dataset and the access pattern to the dataset.
-Consider that the machine has three file systems: a fast scratch file system on which data may reside only for a week, a slower work file system, and each node has a local file system.
-The dataset could be stored on the local storage first to avoid congestion on the work file system, then be migrated to work file system where subsequent tasks of the workflow may read it multiple times.
-At the end, this dataset might be an intermediate product that can then be deleted.
-Alternatively, it could be stored on the scratch file system immediately and accessed there.
-However, that would require that the last read access is within the 7-day period before files on scratch are automatically removed.
-The last scenario shows the case where data is created on work by a task but then copied to a local node to allow read accesses of subsequent tasks.
-This might be beneficial for small random accesses.
-However, it would require that subsequent tasks are placed on the same machine where data is now copied to.
-
+The scripts for the tasks in Cylc define how the data is stored on the available storage systems.
+\Cref{fig:lifecycle} shows three possible life cycles and the access pattern for a specific dataset.
+Consider that the machine has three file systems: a fast \textbf{scratch} file system on which data may reside only for a week, a slower \textbf{work} file system, and each node has a \textbf{local} file system.
+The dataset could be stored on the \textbf{local} storage first to avoid congestion on the \textbf{work} file system, then be migrated to \textbf{work} file system where subsequent tasks of the workflow may read it multiple times.
+In the end, this dataset might be an intermediate product that can then be deleted.
+Alternatively, it could be stored on the \textbf{scratch} file system immediately and accessed there.
+However, that would require that the last read access happens before files on \textbf{scratch} are automatically removed.
+The last scenario shows the case where data is created on \textbf{work} by a task, but then it is copied to a \textbf{local} node to allow read accesses of subsequent tasks which might be beneficial for small random accesses.
+However, it would require that the subsequent tasks are placed on the same machine where data is now stored.
 The scientist responsible for the workflow must optimise the available resources intuitively.
-These are just examples and scientists perform such kind of optimizations already manually.
+These are just examples and currently the scientists perform such kind of optimisations manually.
 
@@
 Other tools for workflow handling, how is this information transported?
+
 dataspaces
+
 ADIOS
 
-\section{Vision for IO-Aware Workflows}
+\section{Vision for I/O-Aware Workflows}
 
-Applied scientists should not spend time understanding hardware characteristics, but using their time to develop their work, let the computer do the job the simplest way possible, and just collect and analyse the results.
+Applied scientists should not spend time understanding hardware characteristics, but using their time to develop their work, let the computer do the job the simplest way as possible, and just collect and analyse the results.
 However, nowadays, to be able to run a job in a HPC environment efficiently, researchers have to have profound knowledge about their workflow, which is expected, but also about decisions regarding storage, communication, and computing.
@@
 We propose an approach to reduce the burden on researchers and, at the same time, optimise the decisions about jobs running in HPC systems.
-Once we have an automated decision about where the job will run and how the storage will be managed, the scientists can then run their workflows in any machine without further interactions and even without previous knowledge about the machine architecture.
+Once we have an automated decision about where the job will run and how the storage will be managed, scientists can then run their workflows in any machine without further interactions and even without previous knowledge about the machine architecture.
 
@@
 
-So far, Cylc allowed to specify tasks, and the dependencies among them.
-We aim to enrich this information with characteristics for input, and the products.
+So far, Cylc allowed specifying tasks and the dependencies among them.
+We aim to enrich this information with characteristics for input, and the products. {\color{cyan}{What?!}}
 An example workflow containing data products is illustrated in \Cref{fig:workflow}.
-Arrows indicate dependencies between nodes that represent tasks and data.
+\sout{{\color{blue}{Arrows indicate dependencies between nodes that represent tasks and data.}}}
+{\color{cyan}{Nodes represent tasks and data and arrows indicate dependencies among tasks.}}
 In the example, Task\,1 needs two datasets to perform its work, it directly communicates with Task\,2 and produces Product\,1.
-Most of the workflow can run automatically except the manual quality control of the products and the final data usage of Product\,3 in the end.
-The last step represents the typical uncertainty of data reuse, i.e., it is unclear how Product\,3 will be used further.
+Most of the workflow can run automatically, except for the manual quality control of the products and the final data usage of Product\,3.
+This last step represents the typical uncertainty of data reuse, i.e., it is unclear how Product\,3 will be used further.
 In this approach, each task is annotated with the required inputs and the generated products.
@@
 What changes now in terms of specification?
-
-{ \color{black}{
 
@@
 ESDM has already an interface with NetCDF.
-Assuming the user can provide extra information in advance, ESDM can interact in the process before the user give the data to the Cylc tool, using the information to construct an optimised workflow and, once Cylc builds the workflow, ESDM can interact again after the configuration of the scripts for both effective and efficient run of the parallel applications regarding the available storage and architecture.
-
-}}
-
-{ \color{black}{
+Assuming the user can provide extra information in advance, ESDM can interact in the process before the user give the data to the Cylc tool, using the information to construct an optimised workflow.
+Once Cylc builds the workflow, ESDM can interact again after the configuration of the scripts for both effective and efficient run of the parallel applications regarding the available storage and architecture.
 
@@
 That is not an immediate conclusion when we analyse just the arrows for the dependencies.
-It is clear that Task \textbf{model} is a dependency for Task \textbf{post\,prod}, which is then a dependency for Task \textbf{prod\,gen}, but the crucial information here is that any dataset currently being used in the current cycle can also be used later.
+It is clear that Task \textbf{model} is a dependency for Task \textbf{post\,prod}, which is then a dependency for Task \textbf{prod\,gen}.
+Still, the crucial information here is that any dataset currently being used in the current cycle can also be used later.
 From the figure we can also see that Task \textbf{model} will generate one dataset for Task \textbf{post\,proc}, one dataset for Task \textbf{prod\,gen}, two datasets for Task \textbf{verification}, and three datasets for a later cycle.
@@
 
-}}
-
 \section{Conclusions}
@@
 \begin{enumerate}
-  \item User runs Cylc to start a suite
+  \item user runs Cylc to start a suite
   \item Cylc parses the suite file, generate dependencies and generate a schedule for the execution
   \item Once a task can be executed because dependencies are ready: Cylc generates on the fly a Slurm script with the required metadata for SLURM and invoking the user-defined script for that task and then runs Slurm to queue the script
-  \item Slurm queues the task and once the task is ready to be dispatched it is started on the supercomputer
+  \item Slurm queues the task, and once the task is ready to be dispatched it is started on the supercomputer
   \item The shell-script executes the users-provided instructions, some of those are to generate file names using Cylc and the variables that are generated by Cylc.
-These filenames use templates provide by the user that define the storage location to be used.
+These filenames use templates provided by the user that defines the storage location to be used.
   \item User-specified applications run taking the filenames as instructed.
-  The application may use NetCDF or XIOS to perform the IO.
+  The application may use NetCDF or XIOS to perform the I/O.
 
@@
 \begin{enumerate}
-  \item User runs Cylc to start a suite.
+  \item user runs Cylc to start a suite.
   \item Cylc parses the suite file, generate dependencies and generate a schedule for the execution.
 
-  Now, Cylc will call into EWIOScheduler to potentially reprioritize the order of execution of tasks.
+  Now, Cylc will call into EWIOScheduler to potentially reprioritise the order of execution of tasks.
   The EWIOScheduler will read the suite file and the suite-io.rc file to make suggestions but won't enforce anything.
@@
 
-  Potentially, Cylc will change the metadata of the script telling Slurm to place a job on the same nodes as a previous job to reusue local data.
+  Potentially, Cylc will change the metadata of the script telling Slurm to place a job on the same nodes as a previous job to reuse local data.
   Potentially, Cylc may also add as metadata information to migrate data.
@@
 
-  \item Slurm queues the task and once the task is ready to be dispatched it is started on the supercomputer
+  \item Slurm queues the task, and once the task is ready to be dispatched it is started on the supercomputer
 
-  If a migration has to be performed before or after a job is run, it would do so.
+  If migration has to be performed before or after a job is run, it would do so.
 
   \item The shell-script executes the users-provided instructions, some of those are to generate file names using Cylc and the variables that are generated by Cylc.
-  These filenames use templates provide by the user that define the storage location to be used.
+  These filenames use templates provided by the user that defines the storage location to be used.
 
-  The way filenames are generated is now slightly different, using a new command that then uses ESWIOSchedule.
-  The guy reads the suite file(s) and generates information for ESDM where to prioritize to place or access data. The filename hence encoded additional information.
+  The way filenames are generated now slightly different, using a new command that then uses ESWIOSchedule.
+  The guy reads the suite file(s) and generates information for ESDM where to prioritise to place or access data. The filename hence encoded additional information.
   Potentially, we generate an esdm.conf file to reflect the specific hardware we want to use.
@@
 
-    to perform the IO.
+    to perform the I/O.
 
-  ESDM loads the system configuration knowing the potential targets and to schedule application-local IO.
-  Application uses ESDM with the script filename, ESDM extracts the long-term schedule information and considers it during "application-local" IO scheduling.
+Knowing the potential targets, ESDM loads the system configuration to schedule application-local I/O.
+  The application uses ESDM with the script filename, ESDM extracts the long-term schedule information and considers it during "application-local" I/O scheduling.
 
@@
 \end{enumerate}
-
-}}
 
@@
 
-%, and data products are generated which are processes that are executed outside of the first simulation.
+%, and data products are generated, which are processes that are executed outside of the first simulation.
 
@@
 This problem is even more severe for data products: a storage system provides a single level of performance and resilience and performance for stored files.
-Theoretically, different fault-tolerance and performance policies can be realised with different storage systems and using management policies.
+Theoretically, different fault-tolerance and performance policies can be realised with varying storage systems and using management policies.
 However, these policies would be coarse-grained and would not account for the priority of the workflow and, most importantly, the value of data.
@@
 
+\bibliographystyle{alpha}
+\bibliography{paper}
+
 \end{document}

### 2020-01-27 paper/paper.tex
Further changings

@@
 I would expect it to be more like a window-month than a fixed month, no?}}
-A user has specified a detailed workflow in a Cylc suite file (shown on the left).
+A user has specified a detailed workflow in a Cylc suite file (\texttt{suite.rc}).
 In this workflow, the \textbf{model} is followed by postprocessing (\textbf{post}), forecast verification (\textbf{ver}), and product generation (\textbf{prod}) tasks.
@@
   \centering
-  \includegraphics[width=0.9\columnwidth]{cylc}
+  \includegraphics[width=0.9\columnwidth]{cylc1}
   \caption{Example workflow\cite{8675433}}
@@
 Consider that the machine has three file systems: a fast \textbf{scratch} file system on which data may reside only for a week, a slower \textbf{work} file system, and each node has a \textbf{local} file system.
-The dataset could be stored on the \textbf{local} storage first to avoid congestion on the \textbf{work} file system, then be migrated to \textbf{work} file system where subsequent tasks of the workflow may read it multiple times.
+As shown in \Cref{fig:lifecycle}(a), the dataset could be stored on the \textbf{local} storage first to avoid congestion on the \textbf{work} file system, then be migrated to \textbf{work} file system where subsequent tasks of the workflow may read it multiple times.
 In the end, this dataset might be an intermediate product that can then be deleted.
-Alternatively, it could be stored on the \textbf{scratch} file system immediately and accessed there.
+Alternatively (\Cref{fig:lifecycle}(b)), it could be stored on the \textbf{scratch} file system immediately and accessed there.
 However, that would require that the last read access happens before files on \textbf{scratch} are automatically removed.
-The last scenario shows the case where data is created on \textbf{work} by a task, but then it is copied to a \textbf{local} node to allow read accesses of subsequent tasks which might be beneficial for small random accesses.
+The last scenario (\Cref{fig:lifecycle}(c)) presents the case where data is created on \textbf{work} by a task, but then it is copied to a \textbf{local} node to allow read accesses of subsequent tasks which might be beneficial for small random accesses.
 However, it would require that the subsequent tasks are placed on the same machine where data is now stored.

### 2020-01-27 paper/paper.tex
Further modifications

@@
 
-\section{Conclusions}
-\label{sec:conclusions}
+\section{Stages}
 
-Now:
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=1.4]{stages}
+  \caption{Stages}
+  \label{fig:work2}
+\end{figure}
+
+Now.
+
 \begin{enumerate}
-  \item user runs Cylc to start a suite
-  \item Cylc parses the suite file, generate dependencies and generate a schedule for the execution
-  \item Once a task can be executed because dependencies are ready: Cylc generates on the fly a Slurm script with the required metadata for SLURM and invoking the user-defined script for that task and then runs Slurm to queue the script
-  \item Slurm queues the task, and once the task is ready to be dispatched it is started on the supercomputer
-  \item The shell-script executes the users-provided instructions, some of those are to generate file names using Cylc and the variables that are generated by Cylc.
-These filenames use templates provided by the user that defines the storage location to be used.
-  \item User-specified applications run taking the filenames as instructed.
-  The application may use NetCDF or XIOS to perform the I/O.
+
+  \item \textbf{Suite}\ \ The user runs Cylc to start a suite. The file \texttt{suite.rc} is created.
+
+  \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
+
+  \item \textbf{Slurm}\ \ Once a task can be executed (dependencies are fulfiled), Cylc generates on the fly a Slurm script with the required metadata for Slurm, invokes the user-defined script for that task, and runs Slurm to queue the process.
+
+  \item \textbf{Task}\ \ Once the task is ready to be dispatched by Slurm, it is started on the supercomputer.
+
+  \item \textbf{Script}\ \ The shell-script executes the users-provided instructions allowing Cylc to generate its internal variables.
+To create the filenames, Cylc reads templates provided by the user that defines the storage location to be used.
+
+  \item \textbf{Application}\ \ The user-specified application runs taking the filenames as instructed. The application may use NetCDF or XIOS to perform the I/O.
 
@@
 
-Our Plan:
+Our Plan.
 
 \begin{enumerate}
-  \item user runs Cylc to start a suite.
-  \item Cylc parses the suite file, generate dependencies and generate a schedule for the execution.
+
+  \item \textbf{Suite}\ \ The user runs Cylc to start a suite. The file \texttt{suite.rc} is created.
+
+  \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
 
@@
 
-  \item Once a task can be executed because dependencies are ready: Cylc generates on the fly a Slurm script with the required metadata for SLURM and invoking the user-defined script for that task.
+  \item \textbf{Slurm}\ \ Once a task can be executed (dependencies are fulfiled), Cylc generates on the fly a Slurm script with the required metadata for Slurm, invokes the user-defined script for that task, and runs Slurm to queue the process.
 
@@
 
-  \item Slurm queues the task, and once the task is ready to be dispatched it is started on the supercomputer
+  \item \textbf{Task}\ \ Once the task is ready to be dispatched by Slurm, it is started on the supercomputer.
 
@@
 
-  \item The shell-script executes the users-provided instructions, some of those are to generate file names using Cylc and the variables that are generated by Cylc.
-  These filenames use templates provided by the user that defines the storage location to be used.
+  \item \textbf{Script}\ \ The shell-script executes the users-provided instructions allowing Cylc to generate its internal variables.
+To create the filenames, Cylc reads templates provided by the user that defines the storage location to be used.
 
@@
 
-  \item User-specified applications run taking the filenames as instructed.
-    The application may use NetCDF, XIOS
+  \item \textbf{Application}\ \ The user-specified application runs taking the filenames as instructed. The application may use NetCDF or XIOS to perform the I/O.
 
-    OR ESDM
-
-    to perform the I/O.
+  The application may use NetCDF, XIOS, or ESDM to perform the I/O.
 
@@
 \end{enumerate}
+
+\section{Conclusions}
+\label{sec:conclusions}
 
@@
 
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.6]{workflow1-v2}
+  \caption{Workflow 1.}
+  \label{fig:work1}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.6]{workflow2-v2}
+  \caption{Workflow 2.}
+  \label{fig:work2}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.6]{workflow-all}
+  \caption{Workflow 2.}
+  \label{fig:work2}
+\end{figure}
+
+\begin{comment}
+
+\begin{table}[H]
+  \centering
+  \begin{tabular}{cc}
+  \includegraphics[scale=0.6]{workflow1-v2} &
+  \includegraphics[scale=0.6]{workflow2-v2}
+  \end{tabular}
+  \caption{Workflows 1 and 2.}
+  \label{fig:work1}
+\end{table}
+
+\end{comment}
+
 \bibliographystyle{alpha}

### 2020-01-27 paper/paper.tex
More changing in the text.

@@
 \usepackage{ulem}
+\usepackage{array}
 
@@
 
-\subsection{Potential Benefit}
-
-\section{Design}
-What changes now in terms of specification?
-
 \subsection{User Perspective Workflow}
 
-From the user perspective, the workflow can be described by the following categories:
+From the user perspective, a general workflow can be described by three categories described in Table \ref{tab:workflow}.
 
-\begin{description}
+\begin{table}[H]
+\centering
+\begin{tabular}{|m{2cm}|m{5.5cm}|m{3.5cm}|m{3.5cm}|}
+\hline
+& \parbox{5.5cm}{\centering \textbf{Input}} & \parbox{3.5cm}{\centering \textbf{Tasks}} & \parbox{3.5cm}{\centering \textbf{Products}} \\ \hline \hline
 
-\item[Input] The information inserted in the workflow. It is composed of datasets describing, for instance:
+Description & Information inserted in the workflow. & The tasks specified by the user. & Possible outcomes of the tasks. \\ \hline
 
-\begin{itemize}
+File & \texttt{input.ds} & \texttt{tasks.ds} & \texttt{output.ds} \\ \hline
 
-\item The actual input data, for instance, the dates for the experiments, how many times the workflow has to run, etc.
+Composed of & Datasets/Scripts & Datasets/Scripts & Datasets \\ \hline
 
-\item The dependencies among the tasks.
+Information &
 
-\item The I/O information, i.e., where the datasets will be stored in each step of the workflow.
+\begin{tabular}{m{5cm}}
+\textbullet\ Dates for the experiments, how many times the workflow has to run, etc.\\
+\textbullet\ Dependencies among the tasks.\\
+\textbullet\ Information about the lifetime of the final outcome.\\
+\textbullet\ Steps in which manual decisions have to be made.\\
+\textbullet\ {{\color{cyan}{I/O information. (Where the datasets will be stored in each step of the workflow.)}}}\\
+\end{tabular}
 
-\item The information about the lifetime of the final outcome.
+&
 
-\item The steps in which manual decisions have to be made.
+\begin{tabular}{m{3cm}}
+\textbullet\ The data input.\\ \\
+\textbullet\ The data output.\\ \\
+\textbullet\ The data lifetime.\\ \\
+\textbullet\ The dependencies regarding the other tasks.\\ \\
+\textbullet\ {{\color{cyan}{I/O information.}}}
+\end{tabular}
 
-\end{itemize}
+&
 
-For simplicity, we will call these datasets as \texttt{input.ds}.
+\begin{tabular}{m{3cm}}
+\textbullet\ Datasets that will be used by other tasks.\\ \\
+\textbullet\ Datasets that will generate the final output.\\
+\end{tabular}
 
-\item[Task 1] The task is specified by the user and the information we have about it is:
+\\ \hline
+\end{tabular}
+\caption{Components of the workflow in the user perspective}
+\label{tab:workflow}
+\end{table}
 
-\begin{itemize}
-
-\item The data input.
-
-\item The data output.
-
-\item How long the data has to be stored.
-
-\item The dependencies regarding the other tasks.
-
-\end{itemize}
-
-In the case of Task 1, we have:
-
-\begin{itemize}
-
-\item Task\,1 has no prior dependencies other than the general input.
-
-\item Task\,1 is a dependency for running Task\,2 and Task\,3.
-
-\item For simplification purposes, we will assume that Task\,1 has the datasets \texttt{input1.ds}, \texttt{output1.ds}, and \texttt{storage1.ds} as descriptors.
-
-\end{itemize}
-
-\item[Task 2]
-
-\begin{itemize}
-
-\item Task\,2 is dependent on the outcomes of Task\,1.
-
-\item Task\,2 has the datasets \texttt{input2.ds}, \texttt{output2.ds}, and \texttt{storage2.ds} as descriptors.
-
-\end{itemize}
-
-\item[Task 3]
-
-\begin{itemize}
-
-\item Task 3 dependent on the outcomes of Task\,1 and Task\,2. It does not require any other task to be started and, after finished, no other task will use its results other than the overall output.
-
-\item Task 3 has the datasets \texttt{input3.ds}, \texttt{output3.ds}, and \texttt{storage3.ds} as descriptors.
-
-\end{itemize}
-
-\item[Product]
-
-A product is any of the possible outcomes of a task. Usually is composed of datasets that will be used by other tasks or generate the final output.
-
-\item[Final Output]
-
-The final output consists of all datasets generated by the workflow. Those datasets have to be stored for a specific amount of time previously described in one of the input datasets.
-
-For simplicity, we will call these datasets as \texttt{output-workflow.ds}. In this specific example, it is composed by the datasets \texttt{output1.ds}, \texttt{output2.ds}, and \texttt{output3.ds}.
-
-\end{description}
-
-For this work, the datasets belonging to \texttt{input-workflow.ds} are the datasets inserted as input for the running Cylc.
-The Cylc tool receives, as predefined, information about the storage and dependencies of the datasets.
-This information is used to construct and run the complete workflow, but it is not used to optimise intermediate steps.
+For this work, Cylc is the essential tool regarding not only the execution of the workflow, but the exchange of information with the EWIOScheduler about I/O and storage.
+Cylc receives, as predefined, information about the storage and dependencies of the datasets.
+The current information is used to construct and run the complete workflow, but it is not employed in optimising intermediate steps.
 The configuration is already predefined, and it cannot change.
 For instance, if it is established that Task\,2 depends on Task\,1, then we have to wait for Task\,1 to be completed to start running Task\,2.
-But several options can be optimised if we know the whole workflow in advance.
-Thinking what about I/O, for instance, no decision is being made about an optimal way to store the data.
-Consider, as an example, that the datasets from \texttt{output1.ds} are divided into \texttt{output1-final.ds} and \texttt{output1-task2.ds}.
-Then, it is clear that \texttt{output1-task2.ds} will be required to run Task\,2, and, because of that, should be put in fast storage to quickly being retrieved, or even kept in local storage.
-The \texttt{output1-final.ds} dataset, however, can go directly to tape, if the lifetime of the final output has to be stored for a long time, but no access is required from the current workflow.
+Thinking what about I/O, no decision is being made about an optimal way to store the data.
+Consider, as an example, that the datasets from \texttt{product1.ds} (the final product of task 1) are divided into \texttt{product1-final.ds} and \texttt{product1-task2.ds}.
+Then, it is clear that \texttt{product1-task2.ds} will be required to run Task\,2, and, because of that, should be put in fast storage to quickly being retrieved, or even kept in local storage.
+The \texttt{product1-final.ds} dataset, however, can go directly to tape, if the lifetime of the final output has to be stored for a long time, but no access is required from the current workflow.
+The bottom line here is: several options can be optimised if we know the whole workflow in advance.
+
+\subsection{Potential Benefit}
 
@@
 
+\section{Design}
+What changes now in terms of specification?
+
 In this paper, we approach two ways to have the extra information from the user to be able to make decisions about storage and the processing of the datasets/tasks.
@@
 
-\item Task 1 usually requires ten hours running in a pc with standard configuration\footnote{Intel Core i5, 8 GB RAM, and 500 GB internal storage drive, for instance.}, requires 100 GB of storage that has to be persistent for no more than one week.
+\item Task\,1 usually requires ten hours running in a pc with standard configuration\footnote{Intel Core i5, 8 GB RAM, and 500 GB internal storage drive, for instance.}, requires 100 GB of storage that has to be persistent for no more than one week.
 
-\item Task 2 usually requires two hours running in a pc with standard configuration, requires 150 GB of storage that has to be persistent for no more than two weeks week.
+\item Task\,2 usually requires two hours running in a pc with standard configuration, requires 150 GB of storage that has to be persistent for no more than two weeks week.
 
-\item Task 3 usually requires five hours running in a pc with standard configuration, and it does not require extra storage other than the storage for the final results.
+\item Task\,3 usually requires five hours running in a pc with standard configuration, and it does not require extra storage other than the storage for the final results.
 
@@
 
-Datasets generated by Task \textbf{A} and used by Task \textbf{B} will have the same physical location in the system because there is one fixed script responsible for setting the configuration.
+Datasets generated by Task\,\textbf{A} and used by Task\,\textbf{B} will have the same physical location in the system because there is one fixed script responsible for setting the configuration.
 Again, such configuration was done manually and with restricted information about the storage and system architecture.
@@
 
+Figure \ref{fig:stages} illustrates the stages of a workflow being processed by Cylc.
+Each stage has a description about what happens with the current methodologies and the increments we propose in this vision paper.
+
 \begin{figure}[H]
@@
   \caption{Stages}
-  \label{fig:work2}
+  \label{fig:stages}
 \end{figure}
-
-Now.
 
@@
 
+  \textbf{Suite + I/O Vision}\ \ Together with the file \texttt{suite.rc}, the user will be responsible to insert the file \texttt{suite-io.rc}, as described in Figure/Section.
+
   \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
 
-  \item \textbf{Slurm}\ \ Once a task can be executed (dependencies are fulfiled), Cylc generates on the fly a Slurm script with the required metadata for Slurm, invokes the user-defined script for that task, and runs Slurm to queue the process.
+  \textbf{Cylc + I/O Vision}\ \ Cylc will also call EWIOScheduler to potentially reprioritise the execution order of the tasks.
+  EWIOScheduler will read both suite files to make suggestions, without enforcing any changings at this moment {\color{cyan}{why?}}.
+
+  \item \textbf{Slurm}\ \ Once a task can be executed (dependencies are fulfilled), Cylc generates on the fly a Slurm script with the required metadata for Slurm, invokes the user-defined script for that task, and runs Slurm to queue the process.
+
+  \textbf{Slurm + I/O Vision}\ \ Assuming the suggestions made by EWIOScheduler are acepted, Cylc will change the metadata of the Slurm script.
+Optimisation in this stage can be, for instance, defining places for a job on the same nodes as a previous job to reuse local data {\color{cyan}{NextgenIO?}} and adding to the metadata information about migrating the data.
 
   \item \textbf{Task}\ \ Once the task is ready to be dispatched by Slurm, it is started on the supercomputer.
+
+  \textbf{Task + I/O Vision}\ \ The optimisation defined in the previous stage are implemented.
+For instance, if migrations have to be performed before or after a job is run, \sout{{\color{cyan}{it}}} {\color{cyan}{Slurm}} will administer them.
 
@@
 
-  \item \textbf{Application}\ \ The user-specified application runs taking the filenames as instructed. The application may use NetCDF or XIOS to perform the I/O.
-
-\end{enumerate}
-
-Our Plan.
-
-\begin{enumerate}
-
-  \item \textbf{Suite}\ \ The user runs Cylc to start a suite. The file \texttt{suite.rc} is created.
-
-  \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
-
-  Now, Cylc will call into EWIOScheduler to potentially reprioritise the order of execution of tasks.
-  The EWIOScheduler will read the suite file and the suite-io.rc file to make suggestions but won't enforce anything.
-
-  \item \textbf{Slurm}\ \ Once a task can be executed (dependencies are fulfiled), Cylc generates on the fly a Slurm script with the required metadata for Slurm, invokes the user-defined script for that task, and runs Slurm to queue the process.
-
-  Potentially, Cylc will change the metadata of the script telling Slurm to place a job on the same nodes as a previous job to reuse local data.
-  Potentially, Cylc may also add as metadata information to migrate data.
-
-  Then runs Slurm to queue the task.
-
-  \item \textbf{Task}\ \ Once the task is ready to be dispatched by Slurm, it is started on the supercomputer.
-
-  If migration has to be performed before or after a job is run, it would do so.
-
-  \item \textbf{Script}\ \ The shell-script executes the users-provided instructions allowing Cylc to generate its internal variables.
-To create the filenames, Cylc reads templates provided by the user that defines the storage location to be used.
-
-  The way filenames are generated now slightly different, using a new command that then uses ESWIOSchedule.
-  The guy reads the suite file(s) and generates information for ESDM where to prioritise to place or access data. The filename hence encoded additional information.
-  Potentially, we generate an esdm.conf file to reflect the specific hardware we want to use.
+  \textbf{Script + I/O Vision}\ \ The way the filenames are generated are now slightly different.
+We will use a new command that invokes ESWIOSchedule again {\color{cyan}{Slurm is calling ESWIOSchedule?}} to read the suite files and generate information for ESDM about where to prioritise {\color{cyan}{prioritise ==> Is this the right word here?}} data placement according to data accessment.
+The filename hence encodes additional information.
+  Potentially, we also can generate an \texttt{esdm.conf} file to reflect the specific hardware we have available to run the workflow and employ ESDM to perform the I/O.
 
@@
 
-  The application may use NetCDF, XIOS, or ESDM to perform the I/O.
-
+  \textbf{Application + I/O Vision}\ \ If the \texttt{esdm.conf} file is generated in the last stage, now the application can also use ESDM to perform the I/O.
 Knowing the potential targets, ESDM loads the system configuration to schedule application-local I/O.
-  The application uses ESDM with the script filename, ESDM extracts the long-term schedule information and considers it during "application-local" I/O scheduling.
-
-  \item A data management service may run and purge or migrate data according to user-specifications of the information life cycle.
+The application uses ESDM with the script filename, ESDM extracts the long-term schedule information and considers it during I/O scheduling to exploit the data locality.
+A data management service may run and purge or migrate data according to \sout{{\color{blue}{user-specifications}}} {\color{cyan}{user or ESDM specifications}} of the information life cycle.
 
@@
   \includegraphics[scale=0.6]{workflow-all}
-  \caption{Workflow 2.}
+  \caption{Workflow All.}
   \label{fig:work2}
@@
 
+\begin{comment}
+
+\begin{description}
+
+\item[Input] The information inserted in the workflow.
+
+\item[Tasks] The tasks specified by the user.
+
+\item[Products] Any of the possible outcomes of the tasks.
+
+For simplicity, we will call these datasets as \texttt{product-workflow.ds}.
+In this specific example, it is composed by the datasets \texttt{product1.ds}, \texttt{product2.ds}, and \texttt{product3.ds}.
+
+\end{description}
+
+\end{comment}
+
+\begin{comment}
+
+Now.
+
+\begin{enumerate}
+
+  \item \textbf{Suite}\ \ The user runs Cylc to start a suite. The file \texttt{suite.rc} is created.
+
+  \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
+
+  \item \textbf{Slurm}\ \ Once a task can be executed (dependencies are fulfiled), Cylc generates on the fly a Slurm script with the required metadata for Slurm, invokes the user-defined script for that task, and runs Slurm to queue the process.
+
+  \item \textbf{Task}\ \ Once the task is ready to be dispatched by Slurm, it is started on the supercomputer.
+
+  \item \textbf{Script}\ \ The shell-script executes the users-provided instructions allowing Cylc to generate its internal variables.
+To create the filenames, Cylc reads templates provided by the user that defines the storage location to be used.
+
+  \item \textbf{Application}\ \ The user-specified application runs taking the filenames as instructed. The application may use NetCDF or XIOS to perform the I/O.
+
+\end{enumerate}
+
+Our Plan.
+
+\end{comment}
+
 \bibliographystyle{alpha}

### 2020-01-28 paper/paper.tex
Tiny increment.

@@
 \usepackage{array}
+\usepackage{subcaption}
 
 \newcommand{\jk}[1]{\todo[inline]{JK: #1}}
+\newcommand{\lr}[1]{\todo[inline]{LR: #1}}
 
@@
 The efficient, convenient, and robust execution of data-driven workflows and enhanced data management are key for productivity in scientific computing.
-\sout{{\color{blue}{
-Traditionally, in HPC, the concerns of storage and computing are separated and optimised independently from each other, and, moreover, independently from the needs of end-to-end user workflows that ultimately lead to insight.}}}
-{\color{cyan}{
-Traditionally, in HPC, the concerns of storage and computing are separated and optimised independently from each other and from the needs of end-to-end user.}}
+Traditionally, in HPC, the concerns of storage and computing are separated and optimised independently from each other and from the needs of end-to-end user.
 As climate and weather workflows become increasingly complex and blend beyond data centres while, at the same time, storage hierarchies become deeper, the community investigates means to re-organise storage access to utilise such environments fully.
@@
 The key contributions of this paper are:
-1) we sketch the visions of an integrated data-driven approach and discuss the challenges and implications of this strategy.
+1) we sketch the visions of an integrated data-driven approach and discuss the challenges and implications of this strategy;
 2) we illustrate architecture and roadmap that allows the seamless integration into existing workflows.
@@
 
-Based on their needs, the {\color{cyan}{HPC}} community has developed a software ecosystem that supports them to execute their large-scale workflows.
+Based on their needs, the HPC community has developed a software ecosystem that supports them to execute their large-scale workflows.
 While the current advances correspond to a big leap forward, many processes still require experts.
@@
 Knowing the capabilities, interfaces, and performance characteristics of individual components are mandatory to make the best use of them.
-As the complexity of systems increases and alternative storage and computing technologies provide unique characteristics, it becomes increasingly difficult even for experts to manually optimise I/O in workflows.
-{\color{cyan}{First time I/O is mentioned.
-I would change to something like...
-it becomes increasingly difficult even for experts to manually optimise resources (other stuff) and, in particular, I/O in such workflows.}}
+As the complexity of systems increases and alternative storage and computing technologies provide unique characteristics, it becomes increasingly difficult even for experts to manually optimise resource usage in workflows.
 In many cases, modifications are not performed because: 1) they are labour intense: any change to the workflow requires careful validation which may not pay off for small scale runs; 2) users are not aware of the potential of the complex system.
 
-In this paper, we show that knowledge about workflow execution and overall experimental design helps to
+In this paper, we show that knowledge about Input/Output of workflow tasks and overall experimental design helps to
 optimise the execution of climate and weather workflows and may increase the performance, throughput and cost-efficiency of the environments, providing an incentive to users and data-centres that cannot be neglected.
@@
 To satisfy the needs for different workflows, data centres are providing a infrastructure consisting of computing and storage devices with different characteristics which makes them more efficient for specific tasks.
-Take, for example, the supercomputer Mistral at DKRZ that consists of about {\color{cyan}{about?}} 3,000 nodes.
+Take, for example, the supercomputer Mistral at DKRZ that consists of 3,321 nodes.
 It offers two types of compute nodes equipped with different CPUs and a range of GPU nodes.
 Each node provides an SSD for local storage and DKRZ has additionally two shared Lustre file systems with different performance characteristics.
-The user can access both file systems \sout{{\color{blue}{as a work and scratch file system}}}. {\color{cyan}{as work and scratch file systems.}}
-While data is kept on the work file system indefinitely, its available space is limited by a quota.
+Individual users and projects are mapped to one file system explicitly and users can access it as work or scratch file system: While data is kept on the work file system indefinitely, its available space is limited by a quota.
 The scratch file system allows to store more data, but data is automatically deleted after some time.
@@
 \jk{insert picture about software stack here}
-To solve a computational problem, we can consider describing the necessary steps from the start to the insight {\color{cyan}{insight?
-I really don't undestand some of your uses of this word.}} as a workflow.
+To solve a computational problem, we can consider describing the necessary steps from the start to the scientific insight generated by the workflow.
 Cylc executes a workflow description comprising of the individual data processing steps and the dependencies among them.
@@
 \begin{minipage}{0.7\textwidth}
-In our example, we assume the application is parallelised for performance reasons and uses MPI to \sout{{\color{blue}{coordinate}}} {\color{cyan}{run}} the application.
+In our example, we assume the application is parallelised for performance reasons and uses MPI to run the application.
 It may use XIOS to gather data from individual fields which then uses NetCDF4 to store the data as a file.
@@
 Internally, HDF5 uses MPI again and its data types to specify the nature of the data stored.
-\sout{{\color{blue}{Finally, data is stored on a parallel file system like Lustre, which, on the server-side, stores data in a local file system that is stored on block devices and storage media such as SSDs and HDDs.}}}
-{\color{cyan}{Finally, data is stored on a parallel file system like Lustre, which, on the server-side, stores data in local file system as block devices on storage media such as SSDs and HDDs.}}
+Finally, data is stored on a parallel file system like Lustre, which, on the server-side, stores data in local file system as block devices on storage media such as SSDs and HDDs.
 
@@
 This process is repeated in the model to simulate month and years into the future.
-Once the simulation of a month is computed, the data for this month can be analysed.
-{\color{cyan}{This is wuite strange for me.
-It means that only after a month is completed one will use the information to start previsions?
-I would expect it to be more like a window-month than a fixed month, no?}}
+Once the simulation of any month is computed, the data for this month becomes available and can be analysed.
 A user has specified a detailed workflow in a Cylc suite file (\texttt{suite.rc}).
@@
   \centering
-  \includegraphics[width=0.9\columnwidth]{cylc1}
+  \includegraphics[width=0.9\columnwidth]{cylc1.png}
   \caption{Example workflow\cite{8675433}}
@@
 
-\begin{figure}[H]
+\begin{figure}[tb]
   \centering
-  \includegraphics[width=0.6\columnwidth]{lifecycle}
-  \caption{Life cycle of one dataset: \textbf{C}reate, \textbf{M}igrate, \textbf{R}ead, and \textbf{D}elete}
+  \begin{subfigure}{.49\textwidth}
+  \includegraphics[width=\columnwidth]{lifecycle-1}
+  \caption{Local and work}
+  \end{subfigure}
+  \begin{subfigure}{.49\textwidth}
+  \includegraphics[width=\columnwidth]{lifecycle-2}
+  \caption{Scratch only}
+  \end{subfigure}
+
+  \caption{Alternative life cycles for mapping a dataset to storage and the operations: \textbf{C}reate, \textbf{M}igrate, \textbf{R}ead, and \textbf{D}elete}
   \label{fig:lifecycle}
@@
 
-dataspaces
-
-ADIOS
+dataspaces + ADIOS + in-situ
 

### 2020-01-28 paper/paper.tex
Minor

@@
   \centering
-  \begin{subfigure}{.49\textwidth}
-  \includegraphics[width=\columnwidth]{lifecycle-1}
+  \begin{subfigure}{.45\textwidth}
+  \includegraphics[width=0.9\columnwidth]{lifecycle-1}
   \caption{Local and work}
   \end{subfigure}
-  \begin{subfigure}{.49\textwidth}
-  \includegraphics[width=\columnwidth]{lifecycle-2}
+  \begin{subfigure}{.45\textwidth}
+  \includegraphics[width=0.9\columnwidth]{lifecycle-2}
   \caption{Scratch only}
@@
 \hline
-& \parbox{5.5cm}{\centering \textbf{Input}} & \parbox{3.5cm}{\centering \textbf{Tasks}} & \parbox{3.5cm}{\centering \textbf{Products}} \\ \hline \hline
+& \parbox{5.5cm}{\centering \textbf{Input}} & \parbox{3.5cm}{\centering \textbf{Computing}} & \parbox{3.5cm}{\centering \textbf{Products}} \\ \hline \hline
 
-Description & Information inserted in the workflow. & The tasks specified by the user. & Possible outcomes of the tasks. \\ \hline
+Description & Information inserted in the workflow. & Tasks specified by the user. & Possible outcomes of the tasks. \\ \hline
 
-File & \texttt{input.ds} & \texttt{tasks.ds} & \texttt{output.ds} \\ \hline
+File & \texttt{input.ds} & \texttt{computing.ds} & \texttt{product.ds} \\ \hline
 
@@
 
-  \item \textbf{Script}\ \ The shell-script executes the users-provided instructions allowing Cylc to generate its internal variables.
+  \item \textbf{Script}\ \ The shell script executes the users-provided instructions allowing Cylc to generate its internal variables.
 To create the filenames, Cylc reads templates provided by the user that defines the storage location to be used.
@@
 
-Typically, as part of shell scripts, users explicitly encode data processing workflows as a sequence of steps consisting of input, computing, and output.
-The computing-intense parts (jobs) and their dependencies are managed by a cluster resource manager that dispatches them.
+Typically, as part of shell scripts, users explicitly encode data processing workflows as a sequence of steps consisting of input, computing, and \sout{{\color{blue}{output}}} {\color{cyan}{products}}.
+The computing-intense parts ({\color{blue}{jobs}} {\color{cyan}{tasks?}}) and their dependencies are administrated by a cluster resource manager that dispatches them.
 These loosely coupled workflows require that intermediate results are stored on persistent storage to be used by the next task.
@@
 
-Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of files) on a particular storage system and, usually, moving data between storage by using scripts.
-One limitation here is that decisions about how data structures are mapped and packaged into files are made once by the producing application, and cannot be changed without manual intervention by a downstream application.
+Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of \sout{{\color{blue}{files}}} {\color{cyan}{datasets}}) on a particular storage system and, usually, moving data between storage by using scripts.
+One limitation here is that decisions about how data structures are mapped and packaged into \sout{{\color{blue}{files}}} {\color{cyan}{datasets}}) are made once by the producing application, and cannot be changed without manual intervention by a downstream application.
 
@@
 Any hard-coded decision has significant implications for the performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
-In any case, data migration needs resources from the source and the target storage system to copy data.
-Thus, to avoid extra manual work and decisions, it is preferable to direct data to locations where it can be stored and directly used by subsequent workflows -- at best, coupling the data source producing data with a data consumer to prevent storing intermediate products on persistent storage.
+In any case, data migration needs resources from the source {\color{cyan}{source? the machine?}} and the target storage system to copy data.
+Thus, to avoid extra manual work and decisions, it is preferable to direct data to locations where it can be stored and directly used by subsequent workflows -- at best, coupling \sout{{\color{blue}{the data source producing data with a data consumer}}} {\color{cyan}{the data source with the data consumer}}) to prevent storing intermediate products on persistent storage.
 
@@
 
-This problem is even more severe for data products: a storage system provides a single level of performance and resilience and performance for stored files.
+This problem is even more severe for data products: a storage system provides a single level of performance and resilience \sout{{\color{blue}{and performance}}} for stored files.
+\sout{{\color{blue}{files}}} {\color{cyan}{datasets}})
 Theoretically, different fault-tolerance and performance policies can be realised with varying storage systems and using management policies.
@@
 While value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability, in practice, this is not the case as the storage just sees blocks and files.
+{\color{cyan}{I don't think we will be dealing with that in the next 10 years!}}
 

### 2020-01-28 paper/paper.tex
Software stack

--- 2020-01-28 paper/paper.tex
@@
 
-\subsection{Software Stack}
-
-The relevant software stack involved in executing workflows is depicted in \Cref{XXX}
-\jk{insert picture about software stack here}
-To solve a computational problem, we can consider describing the necessary steps from the start to the scientific insight generated by the workflow.
-Cylc executes a workflow description comprising of the individual data processing steps and the dependencies among them.
-
 \subsection{I/O Stack}
@@
 \end{minipage}
+
+\subsection{Software Stack}
+
+The relevant software stack involved in executing a workflow is depicted in \ref{fig:stages}.
+As a preliminary, the user had specified a Cylc suite representing the workflow and provided scripts files for the individual tasks to execute; now the user enacts Cylc to start the workflow.
+In the following, each stage of the execution is described further.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=1.4]{stages}
+  \caption{Software Stack and Stages of Execution}
+  \label{fig:stages}
+\end{figure}
+
+\begin{enumerate}
+  \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
+  Cylc monitors the progress of the workflow.
+  Once a task can be executed (dependencies are fulfilled), Cylc generates a Slurm script on the fly with the required metadata for Slurm that will invoke the user-defined script for that task, and runs Slurm to queue the job.
+
+  \item \textbf{Slurm}\ \ Schedules the execution of the job considering the scheduling policy of the data center.
+  Once the job is ready to be dispatched, i.e., resources are available and the job priority is the highest, it is started on the supercomputer.
+
+  \item \textbf{Job}\ \ The user-provided script is executed on a single node assigned to the job.
+  The script has set environment variables containing information about the Cylc execution (e.g., task in the workflow) and Slurm job (e.g., number of nodes).
+
+  \item \textbf{Script}\ \ The shell-script executes the sequence of commands including applications that are parallel programs.
+  Typically, to create a filename, the script calls Cylc commands with a template provided by the user and feeds this information to the application.
+  This ultimately defines the storage location to be used.
+  A script typically consist of multiple fine-grained commands and may even call multiple parallel applications.
+
+  \item \textbf{Application}\ \ The user-specified parallel application runs taking the filenames as defined in a configuration file or as arguments.
+  The application may use NetCDF or XIOS to perform the I/O.
+\end{enumerate}
 
@@
 
+\jk{TODO}
+
 \section{Vision for I/O-Aware Workflows}
@@
 So far, Cylc allowed specifying tasks and the dependencies among them.
-We aim to enrich this information with characteristics for input, and the products. {\color{cyan}{What?!}}
-An example workflow containing data products is illustrated in \Cref{fig:workflow}.
-\sout{{\color{blue}{Arrows indicate dependencies between nodes that represent tasks and data.}}}
-{\color{cyan}{Nodes represent tasks and data and arrows indicate dependencies among tasks.}}
+We aim to enrich this information with characteristics for input, and output, i.e., the datasets used.
+An example workflow containing input data sets and (intermediate) products is illustrated in \Cref{fig:workflow}.
+Nodes represent tasks or data; arrows indicate dependencies among between them.
 In the example, Task\,1 needs two datasets to perform its work, it directly communicates with Task\,2 and produces Product\,1.
@@
 
-File & \texttt{input.ds} & \texttt{tasks.ds} & \texttt{output.ds} \\ \hline
+\jk{?} File & \texttt{input.ds} & \texttt{tasks.ds} & \texttt{output.ds} \\ \hline
 

### 2020-01-28 paper/paper.tex

--- 2020-01-28 paper/paper.tex
@@
 
-File & \texttt{input.ds} & \texttt{computing.ds} & \texttt{product.ds} \\ \hline
+\jk{?} File & \texttt{input.ds} & \texttt{tasks.ds} & \texttt{output.ds} \\ \hline
 

### 2020-01-28 paper/paper.tex
Pain

@@
 
-\subsection{I/O Stack}
+\subsection{I/O Stack of a Parallel Application}
 
@@
 
-\begin{figure}[H]
+\begin{figure}[bt]
   \centering
@@
   \caption{Scratch only}
+  \end{subfigure}
+
+  \begin{subfigure}{.45\textwidth}
+  \includegraphics[width=0.9\columnwidth]{lifecycle-3}
+  \caption{Local and work}
   \end{subfigure}
@@
 
-\section{Stages}
+\subsection{Modified Software Stack}
 
-Figure \ref{fig:stages} illustrates the stages of a workflow being processed by Cylc.
-Each stage has a description about what happens with the current methodologies and the increments we propose in this vision paper.
+The relevant software stack involved in executing a workflow with our IO-aware scheduler is depicted in \Cref{fig:stages-io}.
+In the following, we describe the modifications to the current execution (described in \Cref{fig:stages}) we propose in this vision paper.
+Firstly, the end-user now has to provide an additional file that covers the IO information for each task; slight changes have to be made to the scripts that are executed, these will be discussed below.
 
-\begin{figure}[H]
+\begin{figure}[tb]
   \centering
-  \includegraphics[scale=1.4]{stages}
-  \caption{Stages}
-  \label{fig:stages}
+  \includegraphics[scale=1.4]{stages-io}
+  \caption{Software Stack and Stages of Execution with our IO-Aware Scheduler}
+  \label{fig:stages-io}
 \end{figure}
@@
 \begin{enumerate}
+  \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
+  Now, Cylc calls EIOS to identify potential to optimize the schedule.
+  EIOS may have decided that subsequent jobs shall be placed on the same node or to reorder the execution of some jobs.
+  Also it may have provided information for conducting data migration.
 
-  \item \textbf{Suite}\ \ The user runs Cylc to start a suite. The file \texttt{suite.rc} is created.
+  \item \textbf{EIOS}\ \
+  Reads information the workflow: the original suite file and the IO suit file.
+  Acts depending on the stage of execution, i.e., we could be preparing execution in Cylc or we are already inside a script.
+  It loads the \texttt{esdm.conf} file that contains the information about the available storage technology in the data center and their characteristics.
+  Provides the information how to prioritize IO targets.
 
-  \textbf{Suite + I/O Vision}\ \ Together with the file \texttt{suite.rc}, the user will be responsible to insert the file \texttt{suite-io.rc}, as described in Figure/Section.
+  A data management service may run as a service and purge or migrate data according to user life cycle specification and EIOS information.
 
-  \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
+  \item \textbf{Slurm}\ \ Schedules the execution of the job considering the scheduling policy of the data center.
+  Cylc may now have added a constraint to place a job on the same nodes as a previous job which is handled by the modified Slurm.
+  Also, if migrations have to be performed before or after a job is run, Slurm will administer them.
 
-  \textbf{Cylc + I/O Vision}\ \ Cylc will also call EWIOScheduler to potentially reprioritise the execution order of the tasks.
-  EWIOScheduler will read both suite files to make suggestions, without enforcing any changings at this moment {\color{cyan}{why?}}.
+  \item \textbf{Job}\ \ The user-provided script is executed on a single node assigned to the job.
 
-  \item \textbf{Slurm}\ \ Once a task can be executed (dependencies are fulfilled), Cylc generates on the fly a Slurm script with the required metadata for Slurm, invokes the user-defined script for that task, and runs Slurm to queue the process.
+  The script has set environment variables containing information about the Cylc execution (e.g., task in the workflow) and Slurm job (e.g., number of nodes).
 
-  \textbf{Slurm + I/O Vision}\ \ Assuming the suggestions made by EWIOScheduler are acepted, Cylc will change the metadata of the Slurm script.
-Optimisation in this stage can be, for instance, defining places for a job on the same nodes as a previous job to reuse local data {\color{cyan}{NextgenIO?}} and adding to the metadata information about migrating the data.
+  \item \textbf{Script}\ \ The shell-script executes the sequence of commands.
 
-  \item \textbf{Task}\ \ Once the task is ready to be dispatched by Slurm, it is started on the supercomputer.
+  Typically, to create a filename, the script calls Cylc commands with a template provided by the user and feeds this information to the application.
+  This ultimately defines the storage location to be used.
+  A script typically consist of multiple fine-grained commands and may even call multiple parallel applications.
 
-  \textbf{Task + I/O Vision}\ \ The optimisation defined in the previous stage are implemented.
-For instance, if migrations have to be performed before or after a job is run, \sout{{\color{cyan}{it}}} {\color{cyan}{Slurm}} will administer them.
+  The way the filenames are generated is now different; we will now use a replacement command that invokes a tool of EIOS to generate a pseudo filename.
+  This filename will encode additional information for ESDM about how to prioritise data placement according to data access.
+  Potentially, we also can generate an \texttt{esdm.conf} file to reflect the specific hardware we have available to run the specific application.
 
-  \item \textbf{Script}\ \ The shell script executes the users-provided instructions allowing Cylc to generate its internal variables.
-To create the filenames, Cylc reads templates provided by the user that defines the storage location to be used.
-
-  \textbf{Script + I/O Vision}\ \ The way the filenames are generated are now slightly different.
-We will use a new command that invokes ESWIOSchedule again {\color{cyan}{Slurm is calling ESWIOSchedule?}} to read the suite files and generate information for ESDM about where to prioritise {\color{cyan}{prioritise ==> Is this the right word here?}} data placement according to data accessment.
-The filename hence encodes additional information.
-  Potentially, we also can generate an \texttt{esdm.conf} file to reflect the specific hardware we have available to run the workflow and employ ESDM to perform the I/O.
-
-  \item \textbf{Application}\ \ The user-specified application runs taking the filenames as instructed. The application may use NetCDF or XIOS to perform the I/O.
-
-  \textbf{Application + I/O Vision}\ \ If the \texttt{esdm.conf} file is generated in the last stage, now the application can also use ESDM to perform the I/O.
-Knowing the potential targets, ESDM loads the system configuration to schedule application-local I/O.
-The application uses ESDM with the script filename, ESDM extracts the long-term schedule information and considers it during I/O scheduling to exploit the data locality.
-A data management service may run and purge or migrate data according to \sout{{\color{blue}{user-specifications}}} {\color{cyan}{user or ESDM specifications}} of the information life cycle.
+  \item \textbf{Application}\ \ The user-specified parallel application runs.
+  The application may either use NetCDF with ESDM support or ESDM directly to access data sets.
+  Hence, in \Cref{fig:layers}, the HDF5 layer is replaced with ESDM.
+  ESDM loads the \texttt{esdm.conf} file that contains the information about the available storage backends to use and their configuration.
+  Knowing the potential targets, ESDM schedules application-local I/O.
+  ESDM extracts the long-term schedule information from the generated filenames and considers it during I/O scheduling to exploit the data locality between tasks.
 

### 2020-01-28 paper/paper.tex
Luciana + me. Section 4

@@
 \subsection{Data Management}
+\label{sec:datamanagement}
 
 The scripts for the tasks in Cylc define how the data is stored on the available storage systems.
+Consider that the machine has three file systems: a fast \textbf{scratch} file system on which data may reside only for a week, a slower \textbf{work} file system, and each node has a \textbf{local} file system.
+
+What happens in many current workflows is that they use only \textbf{work} and \textbf{scratch}.
+When a task is set to computing, the corresponding dataset is moved to \textbf{scratch}, processed, and the resulting datasets are transferred back to \textbf{work}.
+If the \textbf{scratch} filesystem achieves its capacity, then the datasets move back to \textbf{work} and keep running the task until it is finished.
+If the I/O planning is made a priori, this situation can be avoided.
+In addition to the file systems described, many supercomputers also have available for the user \textbf{scratch2} file systems (sometimes local and global).
+Therefore, in a real scenario, if the architecture of the machine is known, we can better utilize its resources.
+
 \Cref{fig:lifecycle} shows three possible life cycles and the access pattern for a specific dataset.
-Consider that the machine has three file systems: a fast \textbf{scratch} file system on which data may reside only for a week, a slower \textbf{work} file system, and each node has a \textbf{local} file system.
 As shown in \Cref{fig:lifecycle}(a), the dataset could be stored on the \textbf{local} storage first to avoid congestion on the \textbf{work} file system, then be migrated to \textbf{work} file system where subsequent tasks of the workflow may read it multiple times.
@@
 
-\subsection{Workflow Description}
+Our vision requires two additional information: firstly, the user must augment the workflow description with information about IO requirements and explicitly annotate dependencies to data sets; secondly, the data center (or expert user) must provide a description of the storage architecture and characteristics for each storage system.
+
+There are many approaches to implement the technology for this vision, it is apparent that changes need to be made to several software components to realize this vision.
+In the next section, we will discuss the specific design for our transitional roadmap for climate and weather.
+
+\subsection{Extended Workflow Description}
 
@@
 
+Indeed, by monitoring I/O accesses we can automatically draft such IO descriptions on behalf of the user to simplify the specification.
+All we need is to run one cycle of the specific workflow, considering that a workflow cycle is executed many times in climate and weather, this is a cheap approach.
+From the recorded IO accesses, we analyze the IO patterns and dependencies associated with it to generate the configuration file.
+
 \begin{figure}[H]
@@
   \includegraphics[width=0.4\columnwidth]{workflow}
-  \caption{Example of a high-level workflow}
+  \caption{Example of a high-level workflow with tasks and data dependencies}
   \label{fig:workflow}
@@
 
-\subsection{User Perspective Workflow}
+\subsection{System Information}
+
+The system information shall comprise of all available storage systems, the system topology, and the characteristics of each of the involved components.
+This may include simplified or complex models for the component to approximate performance for specific I/O patterns allowing to predict performance behavior.
+It is expected that the data center (or expert users) can create such a configuration file, for instance, by using vendor provided information or by executing benchmarks.
+
+\subsection{Smarter IO Scheduling}
+
+With this separation of concerns, we can abstract from the workflow what is essential and what a machine should optimise to ensure that the resources available are being used smartly.
+In this paper, we focus on utilizing a heterogeneous storage environment instead of exhaustively discussing the potential that is unleashed with the right level of abstraction in place by researchers that exploit the available opportunities.
+Every time parameters are manually tuned right now is room from improvement by using automatised decision making.
+And Machine Learning has been proven effective in replacing human-beings and adapt the workflow dynamically in case of unexpected situations (the system is down, nodes are unresponsive, etc.).
+
+Optimisations we discuss in this paper are related to the life cycle of datasets and the placement of such datasets into specific storage according to system performance characteristics and the workflow specification.
+Firstly, the IO scheduler could make the placement decision for the individual data sets and manage their relocation during the life cycle as discussed in \Cref{sec:datamanagement}.
+
+Consider the theoretical case to have just two tasks A and B that need to be repeated many times:
+Datasets generated by Task\,\textbf{A} are used by Task\,\textbf{B} and hence both tasks must have access to the same data in the system.
+Because there is one script responsible for setting the configuration, the decision in which directory to store data is somewhat fixed.
+Such configuration was done manually and with restricted information about the storage and system architecture.
+It would be interesting to explore storage options for the datasets and, e.g., have different cycles stored placed in different storage systems.
+If the lifetime of such dataset is, for instance, two cycles, alternate the datasets into two scratch file systems is something that would be a simple job an I/O-aware scheduler.
+However, manually, that implies at having at least two scripts\footnote{Complicated scripts would have allowed to change the storage type depending on the cycle but it is big burden to the user.} for each task with information about the different storage.
+
+Figure \ref{fig:cycle-4} exemplifies a restriction of our example climate/weather workflow the way it was being generated.
+Assume you have, for instance, four cycles of the model running with all the required post-processing steps.
+Typically, the user has generated the scripts for each task by generating a filename using the same prefix that is then placed at the same storage type every time.
+That happens because all datasets generated by Task \textbf{model} will be used by Task \textbf{post\,prod}, so they should stay on the same node.
+
+The user-specific storage locations are a viable option to run a workflow.
+The user might even have analysed various scenarios such as depicted in
+\Cref{sec:datamanagement}.
+However, by automating these tasks, we provide a smarter route taking into consideration the architecture available and the way the datasets are stored during the intermediate steps of the workflow.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.8\columnwidth]{cycle-4}
+  \caption{Four stages workflow example}
+  \label{fig:cycle-4}
+\end{figure}
+
+\subsection{Benefit}
+
+Once we have the complete knowledge about how the workflow is expected to run, we can optimise steps that were before predefined by the user usually considering only options for feasibility.
+
+\begin{itemize}
+\item Abstraction: The user does not have to worry about the architecture that the workflow will run, removing the specialist from the decision-making process.
+
+\item Optimization: The workflow will be optimised for the specific available system characteristics considering potentially even runtime characteristics.
+This will increase performance and system utilization.
+
+\item Performance-portability: The user has to specify the IO requirements only once for the tasks of the workflow.
+The workflow can now be run on any system without user intervention.
+Even more, if the system characteristics change, e.g..
+it gets upgraded and an additional storage tier becomes available or if storage degrades, the I/O-aware workflow could automatically adapt and make use of this new environment.
+\end{itemize}
+
+\section{Specification of the Workflow from User Perspective}
 
@@
 The bottom line here is: several options can be optimised if we know the whole workflow in advance.
-
-\subsection{Potential Benefit}
-
-Those simple examples are here to demonstrate that, once we have the complete knowledge about how the workflow is expected to run, we can optimise steps that were before predefined by the user usually considering only options for feasibility.
-The user input is one of the viable options to run the workflow, but probably it is not the best, assuming other options are available.
-The user might even have analysed other options, but s/he has naively done that (usually by trial and error).
-By automating these tasks, specifically a smarter route for the tasks taking into consideration the architecture available and the way the datasets are stored during the intermediate steps of the workflow, we have a twofold gain:
-
-\begin{itemize}
-
-\item The user does not have to worry about the architecture that the workflow will use, removing from the specialist the decision-making process; and
-
-\item The workflow will be optimised for the specific workflow without extra input from the user and the resources will, then, be better utilised.
-
-\end{itemize}
 
@@
 \end{figure}
-
-Figure \ref{fig:cycle-4} exemplifies another restriction of the way current workflows are being generated.
-Assume you have, for instance, four cycles of the model you are running.
-Because the scripts for the workflow consider individual tasks, the datasets generated by them are placed at the same storage type every time.
-That happens because all datasets generated by Task \textbf{model} will be used by Task \textbf{post\,prod}, so they should stay "closer" when thinking about memory.
-However, although this decision is based on an optimised concept, it still restricts the way the system deals with the datasets and optimisations that are not made thinking only about the local structure.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.8\columnwidth]{cycle-4}
-  \caption{Four stages workflow example}
-  \label{fig:cycle-4}
-\end{figure}
-
-Datasets generated by Task\,\textbf{A} and used by Task\,\textbf{B} will have the same physical location in the system because there is one fixed script responsible for setting the configuration.
-Again, such configuration was done manually and with restricted information about the storage and system architecture.
-It would be interesting to explore the options for the datasets and have them placed in different storage systems.
-If the lifetime of such dataset is, for instance, two cycles, alternate the datasets into two scratch dataset systems is something that would be a simple job for a tool working on generating filenames and storage.
-However, manually, that implies at having at least two scripts with information about the different storage and extra work to have better utilisation of the available resources.
-
-The information about the system architecture can be provided in advance as a configuration file, and we can use monitoring to identify the I/O patterns for a specific workflow.
-For that, all we need is to run one cycle of the workflow and detect the patterns associated with it.
-For this purpose, we intend to use Machine Learning to identify the patterns and adapt the workflow in case of unexpected situations (the system is down, nodes are unresponsive, etc.).
-With this separation of concepts, we can abstract from the workflow what is essential and has to be provided by the user and what a machine should optimise to ensure that the resources available are being used smartly.
-Every time parameters are manually tuned, that has a room from improvement by using automatised decision making.
-And Machine Learning has been proven effective in replacing human-beings in several problems.
-
-Optimisations we can envision about the I/O are related to the life cycle of datasets and the placement of such datasets into specific memory according to their utilisation.
-What happens in most current workflows is that usually, they use only two types of file systems: \textbf{work} and \textbf{scratch}.
-When a task is set to computing, the corresponding dataset is moved to \textbf{scratch}, processed, and the resulting datasets are transferred back to \textbf{work}.
-If the \textbf{scratch} filesystem achieves its capacity, then the datasets move back to \textbf{work} and keep running the task until it is finished.
-If the I/O planning is made a priori, this situation can be avoided.
-In addition to the file systems described, many supercomputers also have available for the user \textbf{scratch2} fyle systems (sometimes local and global).
-Therefore, in a real scenario, if the architecture of the machine is known, we can better use its resources.
 

### 2020-01-28 paper/paper.tex
Cleanup of old stuff

@@
 
+Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of files on a particular storage system and, usually, moving data between storage by using scripts.
+One limitation here is that decisions about how data structures are mapped and packaged into files are made once by the producing application, and cannot be changed without manual intervention by a downstream application.
+
+A policy system (e.g., deployed on a burst buffer) aims to simplify the data movement for the user, but typically migrates objects in the coarse granularity of files.
+However, the semantical information that can be used by a policy system to make decisions is limited, e.g., data location, file extension, age of the file, etc.
+
 \jk{TODO}
@@
 However, nowadays, to be able to run a job in a HPC environment efficiently, researchers have to have profound knowledge about their workflow, which is expected, but also about decisions regarding storage, communication, and computing.
+
+Data system do not respect the importance of data depends at all; aspects like costs to reproduce the data (can it be reproduced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow are not considered.
+Value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability.
 
@@
 That happens because all datasets generated by Task \textbf{model} will be used by Task \textbf{post\,prod}, so they should stay on the same node.
-
 The user-specific storage locations are a viable option to run a workflow.
@@
 \Cref{sec:datamanagement}.
-However, by automating these tasks, we provide a smarter route taking into consideration the architecture available and the way the datasets are stored during the intermediate steps of the workflow.
+
+Additionally, by knowing how to rerun the workflow behind the data creation, a smarter storage system could trade data security with potential recomputation opportunities to optimize the cost-efficiency of the system.
+Intermediate states could be rerun by utilising virtualisation and container technologies.
+Similarly, the scheduler may decide to couple subsequent steps of a workflow directly without storing intermediate data products on persistent storage.
+Finally, data might be replicated but enables the system to rerun parts of the workflow in case of a data loss.
+
+Ultimately, by abstracting and automatizing these tasks, we provide a smarter route taking into consideration the architecture available and the way the datasets are stored during the intermediate steps of the workflow.
 
@@
 
-\item Optimization: The workflow will be optimised for the specific available system characteristics considering potentially even runtime characteristics.
+\item Optimization: The workflow will be optimised for the specific available system characteristics considering potentially even runtime characteristics and storage resilience.
 This will increase performance and system utilization.
+By knowing the workflow behind the data creation, the system could trade data security with potential recomputation opportunities to optimize the cost-efficiency.
 
@@
 it gets upgraded and an additional storage tier becomes available or if storage degrades, the I/O-aware workflow could automatically adapt and make use of this new environment.
+
 \end{itemize}
+
+\section{Design}
+
+This section describes our approach to incrementally extent the current software stack for climate and weather workflows that realizes parts our vision.
+
+In order to automatically make scheduling decisions the software stack needs to:
+\begin{enumerate}
+\item deliver information about information life cycle together with the workflow, and
+\item adapt the resulting workflow, individual script and application execution to consider the automatic IO schedule.
+\end{enumerate}
+
+\subsection{Transiting to Data Dependencies}
+
+Figure \ref{fig:cycle1} introduces one cycle of the suite in Figure \ref{fig:cylc} with a simplified notation.
+The arrows represent the tasks dependencies, but there is no information about the data that are generated.
+Tasks \textbf{verification} and \textbf{post\,proc} depend on Task \textbf{model}, but does Task \textbf{model} needs to be fully completed to satisfy the dependency?
+Of course, once Task \textbf{model} is completed, the dependent tasks can be executed, and that is the type of information Cylc has to create and reproduce the workflow.
+
+\begin{figure}[bt]
+  \begin{subfigure}{0.45\textwidth}
+    \centering
+    \includegraphics[width=\columnwidth]{cycle1}
+    \caption{Original}
+    \label{fig:cycle1}
+  \end{subfigure}
+  \qquad
+  \begin{subfigure}{0.45\textwidth}
+    \centering
+    \includegraphics[width=\columnwidth]{cycle-io-dep}
+    \caption{Including datasets and dependencies}
+    \label{fig:cycle-io-dep}
+  \end{subfigure}
+
+  \caption{Simplified Cylc Workflow for the first stage}
+\end{figure}
+
+The scenario presented in \Cref{fig:cycle-io-dep} shows the same cycle of Figure \ref{fig:cycle1}, but  extended information about datasets is added.
+In this picture, we have the tasks dependencies, and we explicitly list the datasets dependencies regarding the tasks that generated them.
+The way data moves through tasks is by creating datasets to pass on the information, including filename and storage destination.
+The datasets produced by every task are coded in the workflow scripts before Cylc evaluation, and it is just replicated by this tool.
+The tasks can create a different number of datasets, and these datasets may be necessary for the immediate next task in the current cycle, or tasks in different cycles.
+
+Datasets colours indicate which task will use their information and datasets in yellow represent information that will be used in a different cycle.
+The tricky part when we consider only tasks and not the datasets in the workflow is that it may lead to the false impression that the tasks only depend on the immediately previous one.
+Note, for instance, that one dataset produced by Task \textbf{model} is later used in Task \textbf{prod\,gen}.
+That is not an immediate conclusion when we analyse just the arrows for the dependencies.
+It is clear that Task \textbf{model} is a dependency for Task \textbf{post\,prod}, which is then a dependency for Task \textbf{prod\,gen}.
+Still, the crucial information here is that any dataset currently being used in the current cycle can also be used later.
+From the figure we can also see that Task \textbf{model} will generate one dataset for Task \textbf{post\,proc}, one dataset for Task \textbf{prod\,gen}, two datasets for Task \textbf{verification}, and three datasets for a later cycle.
+The access to this information before the process starts can be used to optimise how these datasets will be stored during the workflow.
+This information today is static and manual, leaving opportunities for automatisation and optimisations open.
+The idea here is to embrace the concept that tasks dependencies are really datasets dependencies.
+Once we can insert the datasets explicitly in the workflow, it is easier to see that many optimisations regarding storage can be made.
+
+\subsection{Extended Workflow Description}
+
+The user to be able to make decisions about storage and the processing of the datasets/tasks.
+user will insert this information as an additional file to the Cylc workflow management.
+In fact, the user is already doing that, but he is doing that in just one specific way, to make the workflow possible.
+Here, we are thinking about simplifying the information the user is already providing.
+So, instead of giving the complete workflow to Cylc, we are expecting the user to provide straightforward information, like:
+
+\begin{itemize}
+
+\item Task\,1 usually requires ten hours running in a pc with standard configuration\footnote{Intel Core i5, 8 GB RAM, and 500 GB internal storage drive, for instance.}, requires 100 GB of storage that has to be persistent for no more than one week.
+
+\item Task\,2 usually requires two hours running in a pc with standard configuration, requires 150 GB of storage that has to be persistent for no more than two weeks week.
+
+\item Task\,3 usually requires five hours running in a pc with standard configuration, and it does not require extra storage other than the storage for the final results.
+
+\end{itemize}
+
+\subsection{System Information}
+
+\subsection{Smarter Scheduling}
+
+Now, comes the question.
+Assuming the information is available, who will be responsible for making the decisions for optimising the storage and the processing tasks?
+The answer here is simple: the ESDM Scheduler will do it.
+The ESDM Scheduler can be used in different steps of the process.
+ESDM has already an interface with NetCDF.
+Assuming the user can provide extra information in advance, ESDM can interact in the process before the user give the data to the Cylc tool, using the information to construct an optimised workflow.
+Once Cylc builds the workflow, ESDM can interact again after the configuration of the scripts for both effective and efficient run of the parallel applications regarding the available storage and architecture.
+
+\subsection{Modified Software Stack}
+
+The relevant software stack involved in executing a workflow with our IO-aware scheduler is depicted in \Cref{fig:stages-io}.
+In the following, we describe the modifications to the current execution (described in \Cref{fig:stages}) we propose in this vision paper.
+Firstly, the end-user now has to provide an additional file that covers the IO information for each task; slight changes have to be made to the scripts that are executed, these will be discussed below.
+
+\begin{figure}[tb]
+  \centering
+  \includegraphics[scale=1.4]{stages-io}
+  \caption{Software Stack and Stages of Execution with our IO-Aware Scheduler}
+  \label{fig:stages-io}
+\end{figure}
+
+\begin{enumerate}
+  \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
+  Now, Cylc calls EIOS to identify potential to optimize the schedule.
+  EIOS may have decided that subsequent jobs shall be placed on the same node or to reorder the execution of some jobs.
+  Also it may have provided information for conducting data migration.
+
+  \item \textbf{EIOS}\ \
+  Reads information the workflow: the original suite file and the IO suit file.
+  Acts depending on the stage of execution, i.e., we could be preparing execution in Cylc or we are already inside a script.
+  It loads the \texttt{esdm.conf} file that contains the information about the available storage technology in the data center and their characteristics.
+  Provides the information how to prioritize IO targets.
+
+  A data management service may run as a service and purge or migrate data according to user life cycle specification and EIOS information.
+
+  \item \textbf{Slurm}\ \ Schedules the execution of the job considering the scheduling policy of the data center.
+  Cylc may now have added a constraint to place a job on the same nodes as a previous job which is handled by the modified Slurm.
+  Also, if migrations have to be performed before or after a job is run, Slurm will administer them.
+
+  \item \textbf{Job}\ \ The user-provided script is executed on a single node assigned to the job.
+
+  The script has set environment variables containing information about the Cylc execution (e.g., task in the workflow) and Slurm job (e.g., number of nodes).
+
+  \item \textbf{Script}\ \ The shell-script executes the sequence of commands.
+
+  Typically, to create a filename, the script calls Cylc commands with a template provided by the user and feeds this information to the application.
+  This ultimately defines the storage location to be used.
+  A script typically consist of multiple fine-grained commands and may even call multiple parallel applications.
+
+  The way the filenames are generated is now different; we will now use a replacement command that invokes a tool of EIOS to generate a pseudo filename.
+  This filename will encode additional information for ESDM about how to prioritise data placement according to data access.
+  Potentially, we also can generate an \texttt{esdm.conf} file to reflect the specific hardware we have available to run the specific application.
+
+  \item \textbf{Application}\ \ The user-specified parallel application runs.
+  The application may either use NetCDF with ESDM support or ESDM directly to access data sets.
+  Hence, in \Cref{fig:layers}, the HDF5 layer is replaced with ESDM.
+  ESDM loads the \texttt{esdm.conf} file that contains the information about the available storage backends to use and their configuration.
+  Knowing the potential targets, ESDM schedules application-local I/O.
+  ESDM extracts the long-term schedule information from the generated filenames and considers it during I/O scheduling to exploit the data locality between tasks.
+\end{enumerate}
+
+\section{Conclusions}
+\label{sec:conclusions}
+
+Organising the data placement on storage tiers is typically performed manually by the developer/user or via policies, leading to suboptimal optimisations.
+%On top of that, it is a tedious, non-portable, and error-prone task.
+%Any hard-coded decision has significant implications for the performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
+Manual and hard-coded workflows cannot handle changes in the environment and are error-prone.
+Therefore, users must be able to express their workflows in an abstract fashion that allows the system to generate (near-)optimal execution plans and monitor their execution.
+
+\begin{comment}
+
+and computer-aided Research Development and Engineering (RD\&E)
+
+and also on small-scale systems.
+
+For example, Big data tools integrate computing and storage capabilities into a holistic solution demonstrating the benefit of tight integrating for data-driven workflows.
+
+%-- the quality of service features available in hardware is rarely used.
+
+Cloud definition of desired state and contracts.
+
+Indeed many research prototypes address subproblems
+* But not all aspects together
+* Competing approaches; the standardisation we describe here does not compete!
+
+In this paper, we purposely do not mention too many of the ongoing RD\&E in storage and computing.
+
+To wrap up, NGI provides a high-level view on data and processing on heterogeneous platforms while shielding the user from execution details.
+
+Open ...
+
+%Kubernetes
+
+\begin{description}
+
+\item{NFS} Relatively slow, persistent file system - Nightly incremental backups
+
+\item{Lustre} Large, fast distributed scratch file system - No backups
+
+\url{https://info.hpc.sussex.ac.uk/hpc-guide/storage.html}
+
+\end{description}
+
+%, and data products are generated, which are processes that are executed outside of the first simulation.
+
+\end{comment}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.6]{workflow1-v2}
+  \caption{Workflow 1.}
+  \label{fig:work1}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.6]{workflow2-v2}
+  \caption{Workflow 2.}
+  \label{fig:work2}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.6]{workflow-all}
+  \caption{Workflow All.}
+  \label{fig:work2}
+\end{figure}
+
+\begin{comment}
+
+\begin{table}[H]
+  \centering
+  \begin{tabular}{cc}
+  \includegraphics[scale=0.6]{workflow1-v2} &
+  \includegraphics[scale=0.6]{workflow2-v2}
+  \end{tabular}
+  \caption{Workflows 1 and 2.}
+  \label{fig:work1}
+\end{table}
+
+\end{comment}
 
@@
 
-\section{Design}
-What changes now in terms of specification?
-
-In this paper, we approach two ways to have the extra information from the user to be able to make decisions about storage and the processing of the datasets/tasks.
-
-\begin{itemize}
-
-\item The user will insert this information as an additional file to the Cylc workflow management.
-In fact, the user is already doing that, but he is doing that in just one specific way, to make the workflow possible.
-Here, we are thinking about simplifying the information the user is already providing.
-So, instead of giving the complete workflow to Cylc, we are expecting the user to provide straightforward information, like:
-
-\begin{itemize}
-
-\item Task\,1 usually requires ten hours running in a pc with standard configuration\footnote{Intel Core i5, 8 GB RAM, and 500 GB internal storage drive, for instance.}, requires 100 GB of storage that has to be persistent for no more than one week.
-
-\item Task\,2 usually requires two hours running in a pc with standard configuration, requires 150 GB of storage that has to be persistent for no more than two weeks week.
-
-\item Task\,3 usually requires five hours running in a pc with standard configuration, and it does not require extra storage other than the storage for the final results.
-
-\end{itemize}
-
-\item The second option to access this type of information from the user does not require any direct information at all.
-The idea here is to use the DDN IME Monitoring system and have an estimate for the parameters after one run of the complete workflow.
-This approach is beneficial to the user because s/he does not have to take any extra step that s/he is currently doing to run the workflow through Cylc.
-However, the price here is being paid with an additional run of the complete workflow, which can be demanding at both time and expenses (we are considering here the paid access to a supercomputer, for example).
-In the end, it is up to the user and the previous knowledge s/he has about the intermediate stages of the workflow.
-
-\end{itemize}
-
-Now, comes the question.
-Assuming the information is available, who will be responsible for making the decisions for optimising the storage and the processing tasks?
-The answer here is simple: the ESDM Scheduler will do it.
-The ESDM Scheduler can be used in different steps of the process.
-ESDM has already an interface with NetCDF.
-Assuming the user can provide extra information in advance, ESDM can interact in the process before the user give the data to the Cylc tool, using the information to construct an optimised workflow.
-Once Cylc builds the workflow, ESDM can interact again after the configuration of the scripts for both effective and efficient run of the parallel applications regarding the available storage and architecture.
-
-\section{Extending Existing Workflows}
-
-The focus of this paper is to exemplify the I/O optimisations that can be made if the datasets dependencies are established.
-This information, in fact, is presented in the workflows before the Cylc application, so nothing else will be required from the user.
-But to have the information explicitly, Cylc needs some changes to be able to:
-
-\begin{enumerate}
-
-\item deliver it together with the workflow, and
-
-\item adapts the resulting workflow to consider the I/O modifications.
-
-\end{enumerate}
-
-Figure \ref{fig:cycle1} introduces one cycle of the suite in Figure \ref{fig:cylc} with a simplified notation.
-The arrows represent the tasks dependencies, but there is no information about the data that are generated.
-Tasks \textbf{verification} and \textbf{post\,proc} depend on Task \textbf{model}, but does Task \textbf{model} needs to be fully completed to satisfy the dependency?
-Of course, once Task \textbf{model} is completed, the dependent tasks can be executed, and that is the type of information Cylc has to create and reproduce the workflow.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.4\columnwidth]{cycle1}
-  \caption{First stage workflow}
-  \label{fig:cycle1}
-\end{figure}
-
-The scenario presented in Figure \ref{fig:cycle-io} shows the same cycle of Figure \ref{fig:cycle1}, but the extended information about datasets is introduced.
-In this picture, we have the tasks dependencies, and we explicit the datasets dependencies regarding the tasks that generated them.
-The way data moves through tasks is by creating datasets to pass on the information, including filename and storage destination.
-The datasets produced by every task are coded in the workflow scripts before Cylc evaluation, and it is just replicated by this tool.
-The tasks can create a different number of datasets, and these datasets may be necessary for the immediate next task in the current cycle, or tasks in different cycles.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.6\columnwidth]{cycle-io}
-  \caption{First stage workflow including datasets}
-  \label{fig:cycle-io}
-\end{figure}
-
-Figure \ref{fig:cycle-io-dep} now introduces one example of dependencies among the datasets and tasks: datasets colours indicate which task will use their information and datasets in yellow represent information that will be used in a different cycle.
-The tricky part when we consider only tasks and not the datasets in the workflow is that it may lead to the false impression that the tasks only depend on the immediately previous one.
-Note, for instance, that one dataset produced by Task \textbf{model} is later used in Task \textbf{prod\,gen}.
-That is not an immediate conclusion when we analyse just the arrows for the dependencies.
-It is clear that Task \textbf{model} is a dependency for Task \textbf{post\,prod}, which is then a dependency for Task \textbf{prod\,gen}.
-Still, the crucial information here is that any dataset currently being used in the current cycle can also be used later.
-From the figure we can also see that Task \textbf{model} will generate one dataset for Task \textbf{post\,proc}, one dataset for Task \textbf{prod\,gen}, two datasets for Task \textbf{verification}, and three datasets for a later cycle.
-The access to this information before the process starts can be used to optimise how these datasets will be stored during the workflow.
-This information today is static and manual, leaving opportunities for automatisation and optimisations open.
-The idea here is to embrace the concept that tasks dependencies are really datasets dependencies.
-Once we can insert the datasets explicitly in the workflow, it is easier to see that many optimisations regarding storage can be made.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.6\columnwidth]{cycle-io-dep}
-  \caption{First stage workflow including datasets and dependencies}
-  \label{fig:cycle-io-dep}
-\end{figure}
-
-\subsection{Modified Software Stack}
-
-The relevant software stack involved in executing a workflow with our IO-aware scheduler is depicted in \Cref{fig:stages-io}.
-In the following, we describe the modifications to the current execution (described in \Cref{fig:stages}) we propose in this vision paper.
-Firstly, the end-user now has to provide an additional file that covers the IO information for each task; slight changes have to be made to the scripts that are executed, these will be discussed below.
-
-\begin{figure}[tb]
-  \centering
-  \includegraphics[scale=1.4]{stages-io}
-  \caption{Software Stack and Stages of Execution with our IO-Aware Scheduler}
-  \label{fig:stages-io}
-\end{figure}
-
-\begin{enumerate}
-  \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
-  Now, Cylc calls EIOS to identify potential to optimize the schedule.
-  EIOS may have decided that subsequent jobs shall be placed on the same node or to reorder the execution of some jobs.
-  Also it may have provided information for conducting data migration.
-
-  \item \textbf{EIOS}\ \
-  Reads information the workflow: the original suite file and the IO suit file.
-  Acts depending on the stage of execution, i.e., we could be preparing execution in Cylc or we are already inside a script.
-  It loads the \texttt{esdm.conf} file that contains the information about the available storage technology in the data center and their characteristics.
-  Provides the information how to prioritize IO targets.
-
-  A data management service may run as a service and purge or migrate data according to user life cycle specification and EIOS information.
-
-  \item \textbf{Slurm}\ \ Schedules the execution of the job considering the scheduling policy of the data center.
-  Cylc may now have added a constraint to place a job on the same nodes as a previous job which is handled by the modified Slurm.
-  Also, if migrations have to be performed before or after a job is run, Slurm will administer them.
-
-  \item \textbf{Job}\ \ The user-provided script is executed on a single node assigned to the job.
-
-  The script has set environment variables containing information about the Cylc execution (e.g., task in the workflow) and Slurm job (e.g., number of nodes).
-
-  \item \textbf{Script}\ \ The shell-script executes the sequence of commands.
-
-  Typically, to create a filename, the script calls Cylc commands with a template provided by the user and feeds this information to the application.
-  This ultimately defines the storage location to be used.
-  A script typically consist of multiple fine-grained commands and may even call multiple parallel applications.
-
-  The way the filenames are generated is now different; we will now use a replacement command that invokes a tool of EIOS to generate a pseudo filename.
-  This filename will encode additional information for ESDM about how to prioritise data placement according to data access.
-  Potentially, we also can generate an \texttt{esdm.conf} file to reflect the specific hardware we have available to run the specific application.
-
-  \item \textbf{Application}\ \ The user-specified parallel application runs.
-  The application may either use NetCDF with ESDM support or ESDM directly to access data sets.
-  Hence, in \Cref{fig:layers}, the HDF5 layer is replaced with ESDM.
-  ESDM loads the \texttt{esdm.conf} file that contains the information about the available storage backends to use and their configuration.
-  Knowing the potential targets, ESDM schedules application-local I/O.
-  ESDM extracts the long-term schedule information from the generated filenames and considers it during I/O scheduling to exploit the data locality between tasks.
-
-\end{enumerate}
-
-\section{Conclusions}
-\label{sec:conclusions}
-
-\begin{comment}
-
-and computer-aided Research Development and Engineering (RD\&E)
-
-and also on small-scale systems.
-
-For example, Big data tools integrate computing and storage capabilities into a holistic solution demonstrating the benefit of tight integrating for data-driven workflows.
-
-%-- the quality of service features available in hardware is rarely used.
-
-Cloud definition of desired state and contracts.
-
-Indeed many research prototypes address subproblems
-* But not all aspects together
-* Competing approaches; the standardisation we describe here does not compete!
-
-In this paper, we purposely do not mention too many of the ongoing RD\&E in storage and computing.
-
-To wrap up, NGI provides a high-level view on data and processing on heterogeneous platforms while shielding the user from execution details.
-
-Open ...
-
-%Kubernetes
-
-\begin{description}
-
-\item{NFS} Relatively slow, persistent file system - Nightly incremental backups
-
-\item{Lustre} Large, fast distributed scratch file system - No backups
-
-\url{https://info.hpc.sussex.ac.uk/hpc-guide/storage.html}
-
-\end{description}
-
-%, and data products are generated, which are processes that are executed outside of the first simulation.
-
-\end{comment}
-
-\section{Challenges and Limitations}
-\label{sec:challenges}
-The current HPC environment faces several challenges and limitations concerning the execution of data-driven workflows.
-
-\begin{description}
-\item[Explicit Workflows]
-
-Typically, as part of shell scripts, users explicitly encode data processing workflows as a sequence of steps consisting of input, computing, and \sout{{\color{blue}{output}}} {\color{cyan}{products}}.
-The computing-intense parts ({\color{blue}{jobs}} {\color{cyan}{tasks?}}) and their dependencies are administrated by a cluster resource manager that dispatches them.
-These loosely coupled workflows require that intermediate results are stored on persistent storage to be used by the next task.
-Data transfer, however, needs a dominant fraction of energy compared to computation favouring offloading of computation to storage.
-
-\item[Suboptimal Storage Tiering]
-
-Organising the data placement on storage tiers is typically performed manually by the developer/user or via policies, leading to suboptimal optimisations.
-On top of that, it is a tedious and error-prone task.
-
-Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of \sout{{\color{blue}{files}}} {\color{cyan}{datasets}}) on a particular storage system and, usually, moving data between storage by using scripts.
-One limitation here is that decisions about how data structures are mapped and packaged into \sout{{\color{blue}{files}}} {\color{cyan}{datasets}}) are made once by the producing application, and cannot be changed without manual intervention by a downstream application.
-
-A policy system (e.g. burst buffer) aims to simplify the data movement for the user, but typically migrates objects in the coarse granularity of files.
-However, the semantical information that can be used by a policy system to make decisions is limited, e.g., data location, file extension, age of the file, etc.
-
-Any hard-coded decision has significant implications for the performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
-In any case, data migration needs resources from the source {\color{cyan}{source? the machine?}} and the target storage system to copy data.
-Thus, to avoid extra manual work and decisions, it is preferable to direct data to locations where it can be stored and directly used by subsequent workflows -- at best, coupling \sout{{\color{blue}{the data source producing data with a data consumer}}} {\color{cyan}{the data source with the data consumer}}) to prevent storing intermediate products on persistent storage.
-
-\item[Task and Data Equality]
-
-All tasks and data products are handled identically by a HPC system, meaning there is no differentiation between individual jobs and data products.
-This lack of hierarchy is problematic, as different use cases may not only differ in priority and deadline, but the computing speed usually depends on the critical path of the execution.
-
-This problem is even more severe for data products: a storage system provides a single level of performance and resilience \sout{{\color{blue}{and performance}}} for stored files.
-\sout{{\color{blue}{files}}} {\color{cyan}{datasets}})
-Theoretically, different fault-tolerance and performance policies can be realised with varying storage systems and using management policies.
-However, these policies would be coarse-grained and would not account for the priority of the workflow and, most importantly, the value of data.
-The importance of data depends on aspects like costs to reproduce the data (can it be reproduced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow.
-While value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability, in practice, this is not the case as the storage just sees blocks and files.
-{\color{cyan}{I don't think we will be dealing with that in the next 10 years!}}
-
-\item[Manual Optimisation]
-
-Utilising the computing and storage characteristics of a heterogeneous environment as depicted in \Cref{fig:heterogeneous} is challenging.
-Even within a single data centre, utilising homogeneous storage and computing infrastructure efficiently is complex for experts and needs manual definition.
-The efficient management of data and computing capabilities in a heterogeneous environment and
-how the execution of individual tasks from workflows may benefit from alternative hardware architectures and infrastructures
-are still unresolved questions.
-
-The optimisation of a single parallel job already is quite challenging.
-To make optimal use of available features, programmers and users are forced to think about technical aspects and terms, e.g., setting low-level MPI hints for file striping, while they understand their workflow.
-Similarly, efficient data placement is complex.
-But a specific workflow combined with manual optimisation is the root-cause of subsequent problems, and it must be avoided and replaced with automated and optimised decisions.
-
-Manual and hard-coded workflows cannot handle changes in the environment and are error-prone.
-Therefore, users must be able to express their workflows in an abstract fashion that allows the system to generate (near-)optimal execution plans and monitor their execution.
-
-\end{description}
-
-\section{Vision}
-
-Our vision for the problems described in the first sections is a new approach for data management and data-driven computation.
-
-\begin{description}
-\item[Workflow Specification]
-
-The user defines a workflow describing the intentions of data manipulation and the data lifecycle, including manual (user) involvement in the data processing.
-The definition of constraints, for instance, deadlines for the generation of the final data products (insight), can also be included.
-Workflows can be defined using arbitrary tools, including Python.
-Note that a user will not provide an accurate mapping to system or infrastructure, but expects that the system derives an optimised execution plan that covers heterogeneous infrastructures.
-It does not matter where data is processed and how and where information is stored, but only that the desired result is computed with all constraints met.
-Workflows and data lineage of data products are recorded and can be inspected by the user upon demand, allowing reproduction and validation of the correct execution of experiments.
-
-\item[Smart Scheduling]
-
-Supplied with user workflows and their characteristics, the system derives execution plans for the computation and make data placement decisions that aim to utilise the available hardware resources efficiently.
-As the API is in charge of both tasks, it may decide to trade computation and storage capacity considering characteristics as costs and energy consumption, for example.
-Additionally, it can enable automatic recomputation of intermediate states utilising virtualisation and container technologies.
-Similarly, the scheduler may decide to couple subsequent steps of a workflow directly without storing intermediate data products on persistent storage.
-Data might be replicated but enables the system to rerun parts of the workflow in case of a data loss.
-
-Automatic scheduling is key to efficient processing and will allow constant improvement without forcing the user to adjust their workflows.
-In most cases, individual resources are not used exclusively by a single task, but QoS methods ensure that constraints of workflows are met.
-Note that this is a significant change for HPC environments, as computing resources are typically dedicated to specific jobs.
-
-\end{description}
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[scale=0.6]{workflow1-v2}
-  \caption{Workflow 1.}
-  \label{fig:work1}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[scale=0.6]{workflow2-v2}
-  \caption{Workflow 2.}
-  \label{fig:work2}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[scale=0.6]{workflow-all}
-  \caption{Workflow All.}
-  \label{fig:work2}
-\end{figure}
-
-\begin{comment}
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{cc}
-  \includegraphics[scale=0.6]{workflow1-v2} &
-  \includegraphics[scale=0.6]{workflow2-v2}
-  \end{tabular}
-  \caption{Workflows 1 and 2.}
-  \label{fig:work1}
-\end{table}
-
-\end{comment}
-
-\begin{comment}
-
-\begin{description}
-
-\item[Input] The information inserted in the workflow.
-
-\item[Tasks] The tasks specified by the user.
-
-\item[Products] Any of the possible outcomes of the tasks.
-
-For simplicity, we will call these datasets as \texttt{product-workflow.ds}.
-In this specific example, it is composed by the datasets \texttt{product1.ds}, \texttt{product2.ds}, and \texttt{product3.ds}.
-
-\end{description}
-
-\end{comment}
-
-\begin{comment}
-
-Now.
-
-\begin{enumerate}
-
-  \item \textbf{Suite}\ \ The user runs Cylc to start a suite. The file \texttt{suite.rc} is created.
-
-  \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
-
-  \item \textbf{Slurm}\ \ Once a task can be executed (dependencies are fulfiled), Cylc generates on the fly a Slurm script with the required metadata for Slurm, invokes the user-defined script for that task, and runs Slurm to queue the process.
-
-  \item \textbf{Task}\ \ Once the task is ready to be dispatched by Slurm, it is started on the supercomputer.
-
-  \item \textbf{Script}\ \ The shell script executes the users-provided instructions allowing Cylc to generate its internal variables.
-To create the filenames, Cylc reads templates provided by the user that defines the storage location to be used.
-
-  \item \textbf{Application}\ \ The user-specified application runs taking the filenames as instructed. The application may use NetCDF or XIOS to perform the I/O.
-
-\end{enumerate}
-
-Our Plan.
-
-\end{comment}
-
 \bibliographystyle{alpha}

### 2020-01-28 paper/paper.tex
Cleaning

@@
 
-\author{Julian M. Kunkel
-  \textit{University of Reading}
-    \and
-  Luciana R. Pedro
-  \textit{University of Reading}
+\author{
+%  Julian M. Kunkel
+%  \textit{University of Reading}
+%    \and
+%  Luciana R. Pedro
+%  \textit{University of Reading}
 }
@@
 The efficient, convenient, and robust execution of data-driven workflows and enhanced data management are key for productivity in scientific computing.
-Traditionally, in HPC, the concerns of storage and computing are separated and optimised independently from each other and from the needs of end-to-end user.
+Traditionally, in HPC, the concerns of storage and computing are separated and optimised independently from each other and the needs of the end-to-end user.
 As climate and weather workflows become increasingly complex and blend beyond data centres while, at the same time, storage hierarchies become deeper, the community investigates means to re-organise storage access to utilise such environments fully.
@@
 It offers two types of compute nodes equipped with different CPUs and a range of GPU nodes.
-Each node provides an SSD for local storage and DKRZ has additionally two shared Lustre file systems with different performance characteristics.
-Individual users and projects are mapped to one file system explicitly and users can access it as work or scratch file system: While data is kept on the work file system indefinitely, its available space is limited by a quota.
+Each node provides an SSD for local storage, and DKRZ has additionally two shared Lustre file systems with different performance characteristics.
+Individual users and projects are mapped to one file system explicitly, and users can access it as work or scratch file system: While data is kept on the work file system indefinitely, its available space is limited by a quota.
 The scratch file system allows to store more data, but data is automatically deleted after some time.
@@
 Internally, HDF5 uses MPI again and its data types to specify the nature of the data stored.
-Finally, data is stored on a parallel file system like Lustre, which, on the server-side, stores data in local file system as block devices on storage media such as SSDs and HDDs.
+Finally, data is stored on a parallel file system like Lustre, which, on the server-side, stores data in the local file system as block devices on storage media such as SSDs and HDDs.
 
@@
 
-The relevant software stack involved in executing a workflow is depicted in \ref{fig:stages}.
+The appropriate software stack involved in executing a workflow is depicted in \Cref{fig:stages}.
 As a preliminary, the user had specified a Cylc suite representing the workflow and provided scripts files for the individual tasks to execute; now the user enacts Cylc to start the workflow.
@@
 
-\begin{figure}[bt]
+\begin{figure}[H]
   \centering
@@
 \begin{enumerate}
-  \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
+  \item \textbf{Cylc} Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
   Cylc monitors the progress of the workflow.
-  Once a task can be executed (dependencies are fulfilled), Cylc generates a Slurm script on the fly with the required metadata for Slurm that will invoke the user-defined script for that task, and runs Slurm to queue the job.
+  Once a task can be executed (dependencies are fulfilled), Cylc generates a Slurm script on the fly with the required metadata for Slurm that will invoke the user-defined script for that task and runs Slurm to queue the job.
 
-  \item \textbf{Slurm}\ \ Schedules the execution of the job considering the scheduling policy of the data center.
-  Once the job is ready to be dispatched, i.e., resources are available and the job priority is the highest, it is started on the supercomputer.
+  \item \textbf{Slurm} Schedules the execution of the job considering the scheduling policy of the data centre.
+  Once the job is ready to be dispatched, i.e., resources are available, and the job priority is the highest, it is started on the supercomputer.
 
-  \item \textbf{Job}\ \ The user-provided script is executed on a single node assigned to the job.
-  The script has set environment variables containing information about the Cylc execution (e.g., task in the workflow) and Slurm job (e.g., number of nodes).
+  \item \textbf{Job} The user-provided script is executed on a single node assigned to the job.
+  The script has set environment variables containing information about the Cylc execution (e.g., a task in the workflow) and the Slurm job (e.g., number of nodes).
 
-  \item \textbf{Script}\ \ The shell-script executes the sequence of commands including applications that are parallel programs.
-  Typically, to create a filename, the script calls Cylc commands with a template provided by the user and feeds this information to the application.
-  This ultimately defines the storage location to be used.
-  A script typically consist of multiple fine-grained commands and may even call multiple parallel applications.
+  \item \textbf{Script} The shell-script executes the sequence of commands, including applications that are parallel programs.
+  Typically, to create a filename, the script calls Cylc commands with a template provided by the user and feeds this information to the application which, ultimately, defines the storage location to be used.
+  A script typically consists of multiple fine-grained commands and may even call multiple parallel applications.
 
-  \item \textbf{Application}\ \ The user-specified parallel application runs taking the filenames as defined in a configuration file or as arguments.
+  \item \textbf{Application} The user-specified parallel application runs taking the filenames as defined in a configuration file or as arguments.
   The application may use NetCDF or XIOS to perform the I/O.
@@
 
-Consider the Cylc suite configuration for a toy monthly cycling workflow in Figure \ref{fig:cylc}.
+Consider the Cylc suite configuration for a toy monthly cycling workflow in \Cref{fig:cylc}.
 In this workflow, a warm-cycled atmospheric model (\textbf{model}) simulates the physics from a current state to predict the future after a timestep, for example, after six hours.
@@
   \centering
-  \includegraphics[width=0.9\columnwidth]{cylc1.png}
+  \includegraphics[width=0.9\columnwidth]{cylc1}
   \caption{Example workflow\cite{8675433}}
@@
 
-What happens in many current workflows is that they use only \textbf{work} and \textbf{scratch}.
+What happens in many current workflows is that they use the only \textbf{work} and \textbf{scratch}.
 When a task is set to computing, the corresponding dataset is moved to \textbf{scratch}, processed, and the resulting datasets are transferred back to \textbf{work}.
@@
 In addition to the file systems described, many supercomputers also have available for the user \textbf{scratch2} file systems (sometimes local and global).
-Therefore, in a real scenario, if the architecture of the machine is known, we can better utilize its resources.
+Therefore, in a real scenario, if the architecture of the machine is known, we can better utilise its resources.
 
@@
 However, that would require that the last read access happens before files on \textbf{scratch} are automatically removed.
-The last scenario (\Cref{fig:lifecycle}(c)) presents the case where data is created on \textbf{work} by a task, but then it is copied to a \textbf{local} node to allow read accesses of subsequent tasks which might be beneficial for small random accesses.
+The last scenario (\Cref{fig:lifecycle}(c)) presents the case where data is created on \textbf{work} by a task.
+Still, then it is copied to a \textbf{local} node to allow read accesses of subsequent tasks which might be beneficial for small random accesses.
 However, it would require that the subsequent tasks are placed on the same machine where data is now stored.
 The scientist responsible for the workflow must optimise the available resources intuitively.
-These are just examples and currently the scientists perform such kind of optimisations manually.
+These are just examples, and currently, the scientists perform such kind of optimisations manually.
 
-\begin{figure}[tb]
+\begin{figure}[H]
   \centering
@@
 
-Other tools for workflow handling, how is this information transported?
-
-dataspaces + ADIOS + in-situ
-
 Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of files on a particular storage system and, usually, moving data between storage by using scripts.
@@
 
+Other tools for workflow handling, how is this information transported?
+
+dataspaces + ADIOS + in-situ
+
 \jk{TODO}
+
+\cite{Vladimirov2014FileIO}
 
@@
 
-Applied scientists should not spend time understanding hardware characteristics, but using their time to develop their work, let the computer do the job the simplest way as possible, and just collect and analyse the results.
+Applied scientists should not spend time understanding hardware characteristics, but using their time to develop their work, let the computer do the job the most straightforward way as possible, and just collect and analyse the results.
 However, nowadays, to be able to run a job in a HPC environment efficiently, researchers have to have profound knowledge about their workflow, which is expected, but also about decisions regarding storage, communication, and computing.
 
-Data system do not respect the importance of data depends at all; aspects like costs to reproduce the data (can it be reproduced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow are not considered.
+Data system do not respect the importance of data depends at all.
+Aspects like costs to reproduce the data (can it be reproduced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow are not considered.
 Value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability.
@@
 
-Our vision requires two additional information: firstly, the user must augment the workflow description with information about IO requirements and explicitly annotate dependencies to data sets; secondly, the data center (or expert user) must provide a description of the storage architecture and characteristics for each storage system.
+Our vision requires two additional information: firstly, the user must augment the workflow description with information about I/O requirements and explicitly annotate dependencies to data sets; secondly, the data centre (or expert user) must provide a description of the storage architecture and characteristics for each storage system.
 
-There are many approaches to implement the technology for this vision, it is apparent that changes need to be made to several software components to realize this vision.
+There are many approaches to implement the technology for this vision. It is apparent that changes need to be made to several software components to realise this vision.
 In the next section, we will discuss the specific design for our transitional roadmap for climate and weather.
@@
 
-Indeed, by monitoring I/O accesses we can automatically draft such IO descriptions on behalf of the user to simplify the specification.
-All we need is to run one cycle of the specific workflow, considering that a workflow cycle is executed many times in climate and weather, this is a cheap approach.
-From the recorded IO accesses, we analyze the IO patterns and dependencies associated with it to generate the configuration file.
+Indeed, by monitoring I/O accesses, we can automatically draft such I/O descriptions on behalf of the user to simplify the specification.
+All we need is to run one cycle of the specific workflow, considering that a workflow cycle is executed many times in climate and weather; this is a cheap approach.
+From the recorded I/O accesses, we analyse the I/O patterns and dependencies associated with it to generate the configuration file.
 
@@
 
-The system information shall comprise of all available storage systems, the system topology, and the characteristics of each of the involved components.
-This may include simplified or complex models for the component to approximate performance for specific I/O patterns allowing to predict performance behavior.
-It is expected that the data center (or expert users) can create such a configuration file, for instance, by using vendor provided information or by executing benchmarks.
+The system information shall comprise of all available storage systems, the system topology, and the characteristics of each of its required components.
+This may include simplified or complex models of the components to approximate performance for specific I/O patterns allowing to predict performance behaviour.
+It is expected that the data centre (or expert users) can create such a configuration file, for instance, by using vendor-provided information or by executing benchmarks.
 
-\subsection{Smarter IO Scheduling}
+\subsection{Smarter I/O Scheduling}
 
 With this separation of concerns, we can abstract from the workflow what is essential and what a machine should optimise to ensure that the resources available are being used smartly.
-In this paper, we focus on utilizing a heterogeneous storage environment instead of exhaustively discussing the potential that is unleashed with the right level of abstraction in place by researchers that exploit the available opportunities.
-Every time parameters are manually tuned right now is room from improvement by using automatised decision making.
+In this paper, we focus on utilising a heterogeneous storage environment instead of exhaustively discussing the potential that is unleashed with the right level of abstraction in place by researchers that exploit the available opportunities.
+Every time parameters are manually tuned right now there is a room from improvement by using automatised decision making.
 And Machine Learning has been proven effective in replacing human-beings and adapt the workflow dynamically in case of unexpected situations (the system is down, nodes are unresponsive, etc.).
@@
 Optimisations we discuss in this paper are related to the life cycle of datasets and the placement of such datasets into specific storage according to system performance characteristics and the workflow specification.
-Firstly, the IO scheduler could make the placement decision for the individual data sets and manage their relocation during the life cycle as discussed in \Cref{sec:datamanagement}.
+Firstly, the I/O scheduler could make the placement decision for the individual data sets and manage their relocation during the life cycle, as discussed in \Cref{sec:datamanagement}.
 
-Consider the theoretical case to have just two tasks A and B that need to be repeated many times:
+Consider the hypothetical case to have just two tasks A and B that need to be repeated many times:
 Datasets generated by Task\,\textbf{A} are used by Task\,\textbf{B} and hence both tasks must have access to the same data in the system.
@@
 If the lifetime of such dataset is, for instance, two cycles, alternate the datasets into two scratch file systems is something that would be a simple job an I/O-aware scheduler.
-However, manually, that implies at having at least two scripts\footnote{Complicated scripts would have allowed to change the storage type depending on the cycle but it is big burden to the user.} for each task with information about the different storage.
+However, manually, that implies at having at least two scripts\footnote{Complicated scripts would have allowed changing the storage type depending on the cycle.
+Still, it is a significant burden to the user.} for each task with information about the different storage.
 
-Figure \ref{fig:cycle-4} exemplifies a restriction of our example climate/weather workflow the way it was being generated.
-Assume you have, for instance, four cycles of the model running with all the required post-processing steps.
-Typically, the user has generated the scripts for each task by generating a filename using the same prefix that is then placed at the same storage type every time.
+\Cref{fig:cycle4} exemplifies a restriction of our example climate/weather workflow the way it was being generated.
+Assume you have, for instance, four cycles of the model running with all the necessary post-processing steps.
+Typically, the user has written the scripts for each task by defining a filename using the same prefix that is then placed at the same storage type every time.
 That happens because all datasets generated by Task \textbf{model} will be used by Task \textbf{post\,prod}, so they should stay on the same node.
@@
 
-Additionally, by knowing how to rerun the workflow behind the data creation, a smarter storage system could trade data security with potential recomputation opportunities to optimize the cost-efficiency of the system.
+Additionally, by knowing how to rerun the workflow behind the data creation, a smarter storage system could trade data security with potential recomputation opportunities to optimise the cost-efficiency of the system.
 Intermediate states could be rerun by utilising virtualisation and container technologies.
@@
   \centering
-  \includegraphics[width=0.8\columnwidth]{cycle-4}
+  \includegraphics[width=0.8\columnwidth]{cycle4}
   \caption{Four stages workflow example}
-  \label{fig:cycle-4}
+  \label{fig:cycle4}
 \end{figure}
@@
 
-Once we have the complete knowledge about how the workflow is expected to run, we can optimise steps that were before predefined by the user usually considering only options for feasibility.
+Once we have the complete knowledge about how the workflow is expected to run, we can optimise steps that were before predefined by the user, usually considering only options for feasibility.
 
@@
 
-\item Optimization: The workflow will be optimised for the specific available system characteristics considering potentially even runtime characteristics and storage resilience.
-This will increase performance and system utilization.
-By knowing the workflow behind the data creation, the system could trade data security with potential recomputation opportunities to optimize the cost-efficiency.
+\item Optimisation: The workflow will be optimised for the specific available system characteristics considering potentially runtime characteristics and storage resilience which will increase performance and system utilisation.
+By knowing the workflow behind the data creation, the system could trade data security with potential recomputation opportunities to optimise the cost-efficiency.
 
-\item Performance-portability: The user has to specify the IO requirements only once for the tasks of the workflow.
+\item Performance-portability: The user has to specify the I/O requirements only once for the tasks of the workflow.
 The workflow can now be run on any system without user intervention.
-Even more, if the system characteristics change, e.g..
-it gets upgraded and an additional storage tier becomes available or if storage degrades, the I/O-aware workflow could automatically adapt and make use of this new environment.
+Even more, if the system characteristics change, e.g., it gets upgraded, and an additional storage tier becomes available, or if storage degrades, the I/O-aware workflow could automatically adapt and make use of this new environment.
 
@@
 
-This section describes our approach to incrementally extent the current software stack for climate and weather workflows that realizes parts our vision.
+This section describes our approach to incrementally extend the current software stack for climate and weather workflows that realises parts our vision.
 
-In order to automatically make scheduling decisions the software stack needs to:
+To automatically make scheduling decisions, the software stack needs to:
 \begin{enumerate}
-\item deliver information about information life cycle together with the workflow, and
-\item adapt the resulting workflow, individual script and application execution to consider the automatic IO schedule.
+\item delivers information about information life cycle together with the workflow, and
+\item adapts the resulting workflow, individual script and application execution to consider the automatic I/O schedule.
 \end{enumerate}
@@
 
-Figure \ref{fig:cycle1} introduces one cycle of the suite in Figure \ref{fig:cylc} with a simplified notation.
+\Cref{fig:cycle1} introduces one cycle of the suite in \Cref{fig:cylc} with a simplified notation.
 The arrows represent the tasks dependencies, but there is no information about the data that are generated.
@@
 
-\begin{figure}[bt]
+\begin{figure}[H]
   \begin{subfigure}{0.45\textwidth}
@@
 
-The scenario presented in \Cref{fig:cycle-io-dep} shows the same cycle of Figure \ref{fig:cycle1}, but  extended information about datasets is added.
+The scenario presented in \Cref{fig:cycle-io-dep} shows the same cycle of \Cref{fig:cycle1}, but  extended information about datasets is added.
 In this picture, we have the tasks dependencies, and we explicitly list the datasets dependencies regarding the tasks that generated them.
@@
 
-The user to be able to make decisions about storage and the processing of the datasets/tasks.
-user will insert this information as an additional file to the Cylc workflow management.
+The user can make decisions about storage, and the processing of the datasets/tasks and s/he will insert this information as an additional file to the Cylc workflow management.
 In fact, the user is already doing that, but he is doing that in just one specific way, to make the workflow possible.
@@
 ESDM has already an interface with NetCDF.
-Assuming the user can provide extra information in advance, ESDM can interact in the process before the user give the data to the Cylc tool, using the information to construct an optimised workflow.
+Assuming the user can provide extra information in advance, ESDM can interact in the process before the user gives the data to the Cylc tool, using the information to construct an optimised workflow.
 Once Cylc builds the workflow, ESDM can interact again after the configuration of the scripts for both effective and efficient run of the parallel applications regarding the available storage and architecture.
@@
 
-The relevant software stack involved in executing a workflow with our IO-aware scheduler is depicted in \Cref{fig:stages-io}.
+The appropriate software stack involved in executing a workflow with our I/O-aware scheduler is depicted in \Cref{fig:stages-io}.
 In the following, we describe the modifications to the current execution (described in \Cref{fig:stages}) we propose in this vision paper.
-Firstly, the end-user now has to provide an additional file that covers the IO information for each task; slight changes have to be made to the scripts that are executed, these will be discussed below.
+Firstly, the end-user now has to provide an additional file that covers the I/O information for each task; slight changes have to be made to the scripts that are executed; these will be discussed below.
 
-\begin{figure}[tb]
+\begin{figure}[H]
   \centering
   \includegraphics[scale=1.4]{stages-io}
-  \caption{Software Stack and Stages of Execution with our IO-Aware Scheduler}
+  \caption{Software Stack and Stages of Execution with our I/O-Aware Scheduler}
   \label{fig:stages-io}
@@
   \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
-  Now, Cylc calls EIOS to identify potential to optimize the schedule.
+  Now, Cylc calls EIOS to identify potential to optimise the schedule.
   EIOS may have decided that subsequent jobs shall be placed on the same node or to reorder the execution of some jobs.
-  Also it may have provided information for conducting data migration.
+  Also, it may have provided information for conducting data migration.
 
   \item \textbf{EIOS}\ \
-  Reads information the workflow: the original suite file and the IO suit file.
-  Acts depending on the stage of execution, i.e., we could be preparing execution in Cylc or we are already inside a script.
-  It loads the \texttt{esdm.conf} file that contains the information about the available storage technology in the data center and their characteristics.
-  Provides the information how to prioritize IO targets.
+  Reads information the workflow: the original suite file and the I/O suit file.
+  Acts depending on the stage of execution, i.e., we could be preparing execution in Cylc, or we are already inside a script.
+  It loads the \texttt{esdm.conf} file that contains the information about the available storage technology in the data centre and their characteristics.
+  Provides the information on how to prioritise I/O targets.
 
@@
 
-  \item \textbf{Slurm}\ \ Schedules the execution of the job considering the scheduling policy of the data center.
+  \item \textbf{Slurm}\ \ Schedules the execution of the job considering the scheduling policy of the data centre.
   Cylc may now have added a constraint to place a job on the same nodes as a previous job which is handled by the modified Slurm.
@@
 
-  The script has set environment variables containing information about the Cylc execution (e.g., task in the workflow) and Slurm job (e.g., number of nodes).
+  The script has set environment variables containing information about the Cylc execution (e.g., a task in the workflow) and the Slurm job (e.g., number of nodes).
 
@@
   This ultimately defines the storage location to be used.
-  A script typically consist of multiple fine-grained commands and may even call multiple parallel applications.
+  A script typically consists of multiple fine-grained commands and may even call multiple parallel applications.
 
@@
 Therefore, users must be able to express their workflows in an abstract fashion that allows the system to generate (near-)optimal execution plans and monitor their execution.
+
+\section{Extra Figures}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.6]{workflow1-v2}
+  \caption{Workflow 1.}
+  \label{fig:work1}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.6]{workflow2-v2}
+  \caption{Workflow 2.}
+  \label{fig:work2}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.6]{workflow-all}
+  \caption{Workflow All.}
+  \label{fig:work}
+\end{figure}
+
+\begin{comment}
+
+\begin{table}[H]
+  \centering
+  \begin{tabular}{cc}
+  \includegraphics[scale=0.6]{workflow1-v2} &
+  \includegraphics[scale=0.6]{workflow2-v2}
+  \end{tabular}
+  \caption{Workflows 1 and 2.}
+  \label{fig:work1}
+\end{table}
+
+\end{comment}
+
+\section{Specification of the Workflow from User Perspective}
+
+From the user perspective, a general workflow can be described by three categories described in \Cref{tab:workflow}.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{|m{2cm}|m{5.5cm}|m{3.5cm}|m{3.5cm}|}
+\hline
+& \parbox{5.5cm}{\centering \textbf{Input}} & \parbox{3.5cm}{\centering \textbf{Computing}} & \parbox{3.5cm}{\centering \textbf{Products}} \\ \hline \hline
+
+Description & information inserted in the workflow. & Tasks specified by the user. & Possible outcomes of the tasks. \\ \hline
+
+\jk{?} File & \texttt{input.ds} & \texttt{tasks.ds} & \texttt{output.ds} \\ \hline
+
+Composed of & Datasets/Scripts & Datasets/Scripts & Datasets \\ \hline
+
+Information &
+
+\begin{tabular}{m{5cm}}
+\textbullet\ Dates for the experiments, how many times the workflow has to run, etc.\\
+\textbullet\ Dependencies among the tasks.\\
+\textbullet\ Information about the lifetime of the final outcome.\\
+\textbullet\ Steps in which manual decisions have to be made.\\
+\textbullet\ {{\color{cyan}{I/O information. (Where the datasets will be stored in each step of the workflow.)}}}\\
+\end{tabular}
+
+&
+
+\begin{tabular}{m{3cm}}
+\textbullet\ The data input.\\ \\
+\textbullet\ The data output.\\ \\
+\textbullet\ The data lifetime.\\ \\
+\textbullet\ The dependencies regarding the other tasks.\\ \\
+\textbullet\ {{\color{cyan}{I/O information.}}}
+\end{tabular}
+
+&
+
+\begin{tabular}{m{3cm}}
+\textbullet\ Datasets that will be used by other tasks.\\ \\
+\textbullet\ Datasets that will generate the final output.\\
+\end{tabular}
+
+\\ \hline
+\end{tabular}
+\caption{Components of the workflow in the user perspective}
+\label{tab:workflow}
+\end{table}
+
+For this work, Cylc is an essential tool regarding not only the execution of the workflow but the exchange of information with the EIOS about I/O and storage.
+Cylc receives, as predefined, information about the storage and dependencies of the datasets.
+The current information is used to construct and run the complete workflow, but it is not employed in optimising intermediate steps.
+The configuration is already predefined, and it cannot change.
+For instance, if it is established that Task\,2 depends on Task\,1, then we have to wait for Task\,1 to be completed to start running Task\,2.
+Thinking what about I/O, no decision is being made about an optimal way to store the data.
+Consider, as an example, that the datasets from \texttt{product1.ds} (the final product of task 1) are divided into \texttt{product1-final.ds} and \texttt{product1-task2.ds}.
+Then, it is clear that \texttt{product1-task2.ds} will be required to run Task\,2, and, because of that, should be put in fast storage to quickly being retrieved, or even kept in local storage.
+The \texttt{product1-final.ds} dataset, however, can go directly to tape, if the lifetime of the final output has to be stored for a long time, but no access is required from the current workflow.
+The bottom line here is: several options can be optimised if we know the whole workflow in advance.
 
@@
 
-\begin{figure}[H]
-  \centering
-  \includegraphics[scale=0.6]{workflow1-v2}
-  \caption{Workflow 1.}
-  \label{fig:work1}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[scale=0.6]{workflow2-v2}
-  \caption{Workflow 2.}
-  \label{fig:work2}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[scale=0.6]{workflow-all}
-  \caption{Workflow All.}
-  \label{fig:work2}
-\end{figure}
-
-\begin{comment}
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{cc}
-  \includegraphics[scale=0.6]{workflow1-v2} &
-  \includegraphics[scale=0.6]{workflow2-v2}
-  \end{tabular}
-  \caption{Workflows 1 and 2.}
-  \label{fig:work1}
-\end{table}
-
-\end{comment}
-
-\section{Specification of the Workflow from User Perspective}
-
-From the user perspective, a general workflow can be described by three categories described in Table \ref{tab:workflow}.
-
-\begin{table}[H]
-\centering
-\begin{tabular}{|m{2cm}|m{5.5cm}|m{3.5cm}|m{3.5cm}|}
-\hline
-& \parbox{5.5cm}{\centering \textbf{Input}} & \parbox{3.5cm}{\centering \textbf{Computing}} & \parbox{3.5cm}{\centering \textbf{Products}} \\ \hline \hline
-
-Description & Information inserted in the workflow. & Tasks specified by the user. & Possible outcomes of the tasks. \\ \hline
-
-\jk{?} File & \texttt{input.ds} & \texttt{tasks.ds} & \texttt{output.ds} \\ \hline
-
-Composed of & Datasets/Scripts & Datasets/Scripts & Datasets \\ \hline
-
-Information &
-
-\begin{tabular}{m{5cm}}
-\textbullet\ Dates for the experiments, how many times the workflow has to run, etc.\\
-\textbullet\ Dependencies among the tasks.\\
-\textbullet\ Information about the lifetime of the final outcome.\\
-\textbullet\ Steps in which manual decisions have to be made.\\
-\textbullet\ {{\color{cyan}{I/O information. (Where the datasets will be stored in each step of the workflow.)}}}\\
-\end{tabular}
-
-&
-
-\begin{tabular}{m{3cm}}
-\textbullet\ The data input.\\ \\
-\textbullet\ The data output.\\ \\
-\textbullet\ The data lifetime.\\ \\
-\textbullet\ The dependencies regarding the other tasks.\\ \\
-\textbullet\ {{\color{cyan}{I/O information.}}}
-\end{tabular}
-
-&
-
-\begin{tabular}{m{3cm}}
-\textbullet\ Datasets that will be used by other tasks.\\ \\
-\textbullet\ Datasets that will generate the final output.\\
-\end{tabular}
-
-\\ \hline
-\end{tabular}
-\caption{Components of the workflow in the user perspective}
-\label{tab:workflow}
-\end{table}
-
-For this work, Cylc is the essential tool regarding not only the execution of the workflow, but the exchange of information with the EWIOScheduler about I/O and storage.
-Cylc receives, as predefined, information about the storage and dependencies of the datasets.
-The current information is used to construct and run the complete workflow, but it is not employed in optimising intermediate steps.
-The configuration is already predefined, and it cannot change.
-For instance, if it is established that Task\,2 depends on Task\,1, then we have to wait for Task\,1 to be completed to start running Task\,2.
-Thinking what about I/O, no decision is being made about an optimal way to store the data.
-Consider, as an example, that the datasets from \texttt{product1.ds} (the final product of task 1) are divided into \texttt{product1-final.ds} and \texttt{product1-task2.ds}.
-Then, it is clear that \texttt{product1-task2.ds} will be required to run Task\,2, and, because of that, should be put in fast storage to quickly being retrieved, or even kept in local storage.
-The \texttt{product1-final.ds} dataset, however, can go directly to tape, if the lifetime of the final output has to be stored for a long time, but no access is required from the current workflow.
-The bottom line here is: several options can be optimised if we know the whole workflow in advance.
-
 \bibliographystyle{alpha}

### 2020-01-28 superfri-template/example-without-bibtex.tex
Cleaning

%For correct displaying bibliography please use \bibliographystyle{splncs03_unsrt}
%Journal's leadership do not recommend to change .cls and .bst files.

\documentclass{superfri}
\usepackage[hidelinks]{hyperref}
% ------------

\begin{document}

%\classify{MSC?}
\author{Alice~M.~Scientist\footnote{\label{susu}South Ural State University, Chelyabinsk, Russian Federation}$^{,2}$\orcidID{0000-1111-2222-3333} \and  Bob~A.~Scientist\footnoteref{susu} \and Derek~R.~Author\footnote{\label{msu}Lomonosov Moscow State University, Moscow, Russian Federation}\orcidID{1111-2222-3333-4444}
}

\title{SuperFri Article Template\thanks{The paper is recommended for publication by the Program Committee of the International Scientific Conference ``Parallel computational technologies (PCT) 2018''.}}

\maketitle{}

\begin{abstract}%
This template is devoted to helping you format your article the proper way in
\LaTeX.

\keywords{superfri, template, article formatting}
\end{abstract}

% -----------------------------------------------------------------------
\section*{Introduction}
\label{sec:intro}
The template demonstrates~\cite{DBLP:conf/ics/2015} how to use the \emph{superfri} class. It could be
used as a basis~\cite{DBLP:conf/iccS/AltintasNLKDS16} for your article and meets all the formatting requirements of
the SuperFri journal. Consult the author
guidelines for details on that matter.

\section{The Class}
\label{sec:class}

\subsection{Installation}
\label{sec:install}
The class consists of a single file \verb=superfri.cls= which should reside
along with your manuscript file.

\subsection{Features}
\label{sec:features}
The class is based on the standard~\cite{DBLP:books/sp/byeon2010/RiedelSMWL10} article class and supports all of its
features, as well as these:
\begin{itemize}
\item title and author formatting;
\item abstract and keywords;
\item theorems, definitions, and proofs.
\end{itemize}

Begin your paper~\cite{DBLP:series/cogtech/354089408} with the title and authors' names. After filling the
necessary info in with \verb=\title= and \verb=\author= commands put it all
together~\cite{SIV2008} using the \verb=\maketitle= command.

Define the structure of your article using up to three levels of nesting with
the \verb=\section=, \verb=\subsection=, and \verb=\subsubsection= commands.
You can use the environments \verb=theorem=, \verb=proof=, \verb=lemma=, etc.

For further directions read the comments in \verb=superfri.cls= as a manual~\cite{DBLP:journals/pcs/SokolinskyS16}.

\subsubsection{Subsubsection title}

The equations look like
\begin{equation}
a^2 + b^2 = c^2 + \int_1^2 x+y \, \mathrm{d}y,
\end{equation}
where $2$ is a number.

An example of a figure~\cite{DBLP:journals/corr/Zhang16b} is shown in~\figref{pic/pic}, while a table
is in~\tabref{tab-mytable}.

Figure~\ref{fig:fig} demonstrates formatting of a complex figure.

\fig{width=5cm}{pic/pic}{An example of a simple figure}

\begin{figure}[!ht]
	\begin{minipage}{.5\linewidth}
		\centering
		\includegraphics[width=0.65\linewidth]{pic/pic}
		\subcaption{This is subcaption for A subfigure}\label{subfig:a}
	\end{minipage}
	\begin{minipage}{.5\linewidth}
		\centering
		\includegraphics[width=0.65\linewidth]{pic/pic}
		\subcaption{This is subcaption for B subfigure}\label{subfig:b}
	\end{minipage}
	\vspace{3pt}
	\caption{An example of a complex figure}
	\label{fig:fig}
\end{figure}

\tab{tab-mytable}{An example of a table}{
	\begin{tabular}{rcl}
	\hline
	one & two & three \\
	\hline
	a & b & c \\
	$\alpha$ & $\beta$ & $\gamma$ \\
	$x$ & $y$ & $z$ \\
	\hline
	\end{tabular}
}

\section*{Acknowledgements}
Put the acknowledgements after the last section, like this.

\openaccess

%To make hyperlink to network resourse without BiBTeX use \bibitem{id} ...., \link{url_to_resourse}.
%Usage examples are located below.

\bibliographystyle{superfri}
\begin{thebibliography}{9}

	\bibitem{DBLP:conf/iccS/AltintasNLKDS16}
	Altintas, I., Normal, M., Lees, M., Krzhizhanovskaya, V.V., Dongarra, J.J., Sloot, P.M.A.: Data through the computational lens, preface for ICCS 2016. In: Connolly, M. (ed.) International Conference on Computational Science 2016, ICCS 2016, 6-8 June 2016, San
	Diego, California, USA. Procedia Computer Science, vol. 80, pp. 1--7. Elsevier (2016),  DOI:~\href{http://dx.doi.org/10.1016/j.procs.2016.05.426}{10.1016/j.procs.2016.05.426}

	\bibitem{DBLP:conf/ics/2015}
	Bhuyan, L.N., Chong, F., Sarkar, V. (eds.): Proceedings of the 29th ACM on International Conference on Supercomputing, ICS'15, Newport Beach/Irvine, CA, USA, June 08 - 11, 2015. ACM (2015), \url{http://dl.acm.org/citation.cfm?id=2751205}

	\bibitem{DBLP:series/cogtech/354089408}
	Crocker, M.W., Siekmann, J.H. (eds.): Resource-Adaptive Cognitive Processes. Cognitive Technologies, Springer (2011), DOI:~\href{http://dx.doi.org/10.1007/978-3-540-89408-7}{10.1007/978-3-540-89408-7}

	\bibitem{SIV2008}
	Harkins, D.: Synthetic Initialization Vector (SIV) Authenticated Encryption Using the Advanced Encryption Standard (AES). \url{http://tools.ietf.org/html/rfc5297} (2008), accessed: 2010-09-30

	\bibitem{DBLP:books/sp/byeon2010/RiedelSMWL10}
	Riedel, M., Streit, A., Mallmann, D., Wolf, F., Lippert, T.: Experiences and requirements for interoperability between HTC and hpc-driven e-science infrastructure. In: Byeon, O., Kwon, J.H., Dunning, T.H., Cho, K., Savoy-Navarro, A. (eds.) Future Application and Middleware Technology on e-Science, pp. 113--123. Springer (2010), DOI:~\href{http://dx.doi.org/10.1007/978-1-4419-1719-5\_12}{10.1007/978-1-4419-1719-5\_12}

	\bibitem{DBLP:journals/pcs/SokolinskyS16}
	Sokolinsky, L.B., Shamakina, A.V.: Methods of resource management in problem-oriented computing environment. Programming and Computer Software 42(1), 17--26 (2016),  DOI:~\href{http://dx.doi.org/10.1134/S0361768816010084}{10.1134/S0361768816010084}

	\bibitem{DBLP:journals/corr/Zhang16b}
	Zhang, S.: Distributed stochastic optimization for deep learning (thesis). CoRR abs/1605.02216 (2016), \url{http://arxiv.org/abs/1605.02216}

\end{thebibliography}

%\received{September 25, 2013}

\end{document}

### 2020-01-28 superfri-template/example.tex
Cleaning

--- 2020-01-28 superfri-template/example-without-bibtex.tex
@@
-%For correct displaying bibliography please use \bibliographystyle{splncs03_unsrt}
-%Journal's leadership do not recommend to change .cls and .bst files.
+%% An example of the article based on superfri.cls class file of
+%% ``Supercomputing Frontiers and Innovations. An International Journal''
+%% http://superfri.org/.
 
 \documentclass{superfri}
+\usepackage{graphicx}
 \usepackage[hidelinks]{hyperref}
+
 % ------------
+\bibliographystyle{superfri}
 
@@
 
-%\classify{MSC?}
 \author{Alice~M.~Scientist\footnote{\label{susu}South Ural State University, Chelyabinsk, Russian Federation}$^{,2}$\orcidID{0000-1111-2222-3333} \and  Bob~A.~Scientist\footnoteref{susu} \and Derek~R.~Author\footnote{\label{msu}Lomonosov Moscow State University, Moscow, Russian Federation}\orcidID{1111-2222-3333-4444}
@@
 
-\title{SuperFri Article Template\thanks{The paper is recommended for publication by the Program Committee of the International Scientific Conference ``Parallel computational technologies (PCT) 2018''.}}
+\title{SuperFrI Article Template\thanks{The paper is recommended for publication by the Program Committee of the International Scientific Conference ``Parallel computational technologies (PCT) 2018''.}}
 
@@
 
-An example of a figure~\cite{DBLP:journals/corr/Zhang16b} is shown in~\figref{pic/pic}, while a table
+An example of a simple figure~\cite{DBLP:journals/corr/Zhang16b} is shown in~\figref{pic/pic}, while a table
 is in~\tabref{tab-mytable}.
 
+\fig{width=5cm}{pic/pic}{An example of a simple figure}
+
 Figure~\ref{fig:fig} demonstrates formatting of a complex figure.
-
-\fig{width=5cm}{pic/pic}{An example of a simple figure}
 
@@
 
-%To make hyperlink to network resourse without BiBTeX use \bibitem{id} ...., \link{url_to_resourse}.
-%Usage examples are located below.
-
-\bibliographystyle{superfri}
-\begin{thebibliography}{9}
-
-	\bibitem{DBLP:conf/iccS/AltintasNLKDS16}
-	Altintas, I., Normal, M., Lees, M., Krzhizhanovskaya, V.V., Dongarra, J.J., Sloot, P.M.A.: Data through the computational lens, preface for ICCS 2016.
-In: Connolly, M.
-(ed.) International Conference on Computational Science 2016, ICCS 2016, 6-8 June 2016, San
-	Diego, California, USA. Procedia Computer Science, vol. 80, pp. 1--7. Elsevier (2016),  DOI:~\href{http://dx.doi.org/10.1016/j.procs.2016.05.426}{10.1016/j.procs.2016.05.426}
-
-	\bibitem{DBLP:conf/ics/2015}
-	Bhuyan, L.N., Chong, F., Sarkar, V.
-(eds.): Proceedings of the 29th ACM on International Conference on Supercomputing, ICS'15, Newport Beach/Irvine, CA, USA, June 08 - 11, 2015.
-ACM (2015), \url{http://dl.acm.org/citation.cfm?id=2751205}
-
-	\bibitem{DBLP:series/cogtech/354089408}
-	Crocker, M.W., Siekmann, J.H.
-(eds.): Resource-Adaptive Cognitive Processes.
-Cognitive Technologies, Springer (2011), DOI:~\href{http://dx.doi.org/10.1007/978-3-540-89408-7}{10.1007/978-3-540-89408-7}
-
-	\bibitem{SIV2008}
-	Harkins, D.: Synthetic Initialization Vector (SIV) Authenticated Encryption Using the Advanced Encryption Standard (AES). \url{http://tools.ietf.org/html/rfc5297} (2008), accessed: 2010-09-30
-
-	\bibitem{DBLP:books/sp/byeon2010/RiedelSMWL10}
-	Riedel, M., Streit, A., Mallmann, D., Wolf, F., Lippert, T.: Experiences and requirements for interoperability between HTC and hpc-driven e-science infrastructure.
-In: Byeon, O., Kwon, J.H., Dunning, T.H., Cho, K., Savoy-Navarro, A.
-(eds.) Future Application and Middleware Technology on e-Science, pp.
-113--123.
-Springer (2010), DOI:~\href{http://dx.doi.org/10.1007/978-1-4419-1719-5\_12}{10.1007/978-1-4419-1719-5\_12}
-
-	\bibitem{DBLP:journals/pcs/SokolinskyS16}
-	Sokolinsky, L.B., Shamakina, A.V.: Methods of resource management in problem-oriented computing environment.
-Programming and Computer Software 42(1), 17--26 (2016),  DOI:~\href{http://dx.doi.org/10.1134/S0361768816010084}{10.1134/S0361768816010084}
-
-	\bibitem{DBLP:journals/corr/Zhang16b}
-	Zhang, S.: Distributed stochastic optimization for deep learning (thesis). CoRR abs/1605.02216 (2016), \url{http://arxiv.org/abs/1605.02216}
-
-\end{thebibliography}
-
-%\received{September 25, 2013}
-
+\bibliography{example}
 \end{document}

### 2020-01-29 paper/paper.tex
Changes in chapter 5.

@@
 
-The paper is organised as follows:
-In \Cref{sec:workflows}, we provide an overview of the specification of workflows in climate and weather and the involved software stack.
+%The paper is organised as follows:
+%In \Cref{sec:workflows}, we provide an overview of the specification of workflows in climate and weather and the involved software stack.
 
@@
   \includegraphics[width=\textwidth]{layers-xios}
-  \caption{I/O Path for an MPI-Parallel Application}
+  \caption{I/O path for an MPI-parallel application}
   \label{fig:layers}
@@
   \includegraphics[scale=1.4]{stages}
-  \caption{Software Stack and Stages of Execution}
+  \caption{Software stack and stages of execution}
   \label{fig:stages}
@@
 
-  \item \textbf{Slurm} Schedules the execution of the job considering the scheduling policy of the data centre.
+  \item \textbf{Slurm} Slurm schedules the execution of the job considering the scheduling policy of the data centre.
   Once the job is ready to be dispatched, i.e., resources are available, and the job priority is the highest, it is started on the supercomputer.
@@
 
-Consider the Cylc suite configuration for a toy monthly cycling workflow in \Cref{fig:cylc}.
+Consider the Cylc suite configuration for a toy monthly cycling workflow in \Cref{fig:cylc} \cite{8675433}.
 In this workflow, a warm-cycled atmospheric model (\textbf{model}) simulates the physics from a current state to predict the future after a timestep, for example, after six hours.
@@
   \includegraphics[width=0.9\columnwidth]{cylc1}
-  \caption{Example workflow\cite{8675433}}
+  \caption{Example workflow}
   \label{fig:cylc}
@@
 \Cref{fig:lifecycle} shows three possible life cycles and the access pattern for a specific dataset.
-As shown in \Cref{fig:lifecycle}(a), the dataset could be stored on the \textbf{local} storage first to avoid congestion on the \textbf{work} file system, then be migrated to \textbf{work} file system where subsequent tasks of the workflow may read it multiple times.
+As shown in \Cref{fig:lifecycle1}, the dataset could be stored on the \textbf{local} storage first to avoid congestion on the \textbf{work} file system, then be migrated to \textbf{work} file system where subsequent tasks of the workflow may read it multiple times.
 In the end, this dataset might be an intermediate product that can then be deleted.
-Alternatively (\Cref{fig:lifecycle}(b)), it could be stored on the \textbf{scratch} file system immediately and accessed there.
+Alternatively (\Cref{fig:lifecycle2}, it could be stored on the \textbf{scratch} file system immediately and accessed there.
 However, that would require that the last read access happens before files on \textbf{scratch} are automatically removed.
-The last scenario (\Cref{fig:lifecycle}(c)) presents the case where data is created on \textbf{work} by a task.
+The last scenario (\Cref{fig:lifecycle3}) presents the case where data is created on \textbf{work} by a task.
 Still, then it is copied to a \textbf{local} node to allow read accesses of subsequent tasks which might be beneficial for small random accesses.
@@
   \centering
-  \begin{subfigure}{.45\textwidth}
+  \begin{subfigure}{.3\textwidth}
   \includegraphics[width=0.9\columnwidth]{lifecycle-1}
-  \caption{Local and work}
+  \caption{\label{fig:lifecycle1} Local and work file systems}
   \end{subfigure}
-  \begin{subfigure}{.45\textwidth}
+  \begin{subfigure}{.3\textwidth}
   \includegraphics[width=0.9\columnwidth]{lifecycle-2}
-  \caption{Scratch only}
+  \caption{\label{fig:lifecycle2} Scratch file system only}
   \end{subfigure}
-
-  \begin{subfigure}{.45\textwidth}
+  \begin{subfigure}{.3\textwidth}
   \includegraphics[width=0.9\columnwidth]{lifecycle-3}
-  \caption{Local and work}
+  \caption{\label{fig:lifecycle3} Local and work file systems}
   \end{subfigure}
-
   \caption{Alternative life cycles for mapping a dataset to storage and the operations: \textbf{C}reate, \textbf{M}igrate, \textbf{R}ead, and \textbf{D}elete}
@@
 Each data product is annotated with metadata such as data life cycle information, e.g., the value of data and how long should it be kept.
+
+\lr{This next paragraph should be migrate to 4.3 and again mentioned in 5.3.}
 
@@
 
-This section describes our approach to incrementally extend the current software stack for climate and weather workflows that realises parts our vision.
+This section describes our approach to incrementally extend the current software stack for climate and weather workflows which realises parts of our vision.
+To automatically make scheduling decisions, the software stack needs to:
 
-To automatically make scheduling decisions, the software stack needs to:
 \begin{enumerate}
-\item delivers information about information life cycle together with the workflow, and
-\item adapts the resulting workflow, individual script and application execution to consider the automatic I/O schedule.
+
+\item Delivers information about storage life cycle together with the workflow, and
+
+\item Adapts the resulting workflow, individual scripts and application executions to consider the automatic I/O schedule.
+
 \end{enumerate}
 
-\subsection{Transiting to Data Dependencies}
+\subsection{Agregatting Data Dependencies}
 
@@
 The arrows represent the tasks dependencies, but there is no information about the data that are generated.
-Tasks \textbf{verification} and \textbf{post\,proc} depend on Task \textbf{model}, but does Task \textbf{model} needs to be fully completed to satisfy the dependency?
-Of course, once Task \textbf{model} is completed, the dependent tasks can be executed, and that is the type of information Cylc has to create and reproduce the workflow.
+Tasks \textbf{verification} and \textbf{post\,proc} depend on Task \textbf{model}, but does Task \textbf{model} needs to be fully completed to satisfy the dependencies?
+Of course, once Task \textbf{model} is completed, the dependent tasks can be executed, and that is the type of information Cylc has to create to reproduce the workflow.
+
+\begin{comment}
+
+\begin{figure}[H]
+\centering
+\begin{tabular}{ccc}
+\includegraphics[scale=0.6]{cycle1} & \qquad &
+\includegraphics[scale=0.6]{cycle-io-dep} \\
+(a) Original Cycle & & (b) Cycle including datasets and dependencies
+\end{tabular}
+\caption{ \label{fig:cycle} Simplified Cylc workflow for the first cycle}
+\end{figure}
+
+\end{comment}
 
@@
     \includegraphics[width=\columnwidth]{cycle1}
-    \caption{Original}
+    \caption{Original Cycle}
     \label{fig:cycle1}
@@
     \includegraphics[width=\columnwidth]{cycle-io-dep}
-    \caption{Including datasets and dependencies}
+    \caption{Cycle including datasets and dependencies}
     \label{fig:cycle-io-dep}
   \end{subfigure}
-
-  \caption{Simplified Cylc Workflow for the first stage}
+  \caption{Simplified Cylc workflow for the first cycle}
 \end{figure}
 
-The scenario presented in \Cref{fig:cycle-io-dep} shows the same cycle of \Cref{fig:cycle1}, but  extended information about datasets is added.
+The scenario presented in \Cref{fig:cycle1} shows the same cycle of \Cref{fig:cycle-io-dep}, but  extended information about datasets is added.
 In this picture, we have the tasks dependencies, and we explicitly list the datasets dependencies regarding the tasks that generated them.
@@
 Datasets colours indicate which task will use their information and datasets in yellow represent information that will be used in a different cycle.
-The tricky part when we consider only tasks and not the datasets in the workflow is that it may lead to the false impression that the tasks only depend on the immediately previous one.
+The tricky part when we consider only tasks and not the datasets in the workflow is that it may lead to the false impression that one task depends only on the immediately previous one.
 Note, for instance, that one dataset produced by Task \textbf{model} is later used in Task \textbf{prod\,gen}.
@@
 Still, the crucial information here is that any dataset currently being used in the current cycle can also be used later.
-From the figure we can also see that Task \textbf{model} will generate one dataset for Task \textbf{post\,proc}, one dataset for Task \textbf{prod\,gen}, two datasets for Task \textbf{verification}, and three datasets for a later cycle.
+From the diagram we can also notice that Task \textbf{model} will generate one dataset for Task \textbf{post\,proc}, one dataset for Task \textbf{prod\,gen}, two datasets for Task \textbf{verification}, and three datasets for later cycles.
 The access to this information before the process starts can be used to optimise how these datasets will be stored during the workflow.
 This information today is static and manual, leaving opportunities for automatisation and optimisations open.
-The idea here is to embrace the concept that tasks dependencies are really datasets dependencies.
+The idea here is to embrace the concept that tasks dependencies are really imposed by datasets dependencies.
 Once we can insert the datasets explicitly in the workflow, it is easier to see that many optimisations regarding storage can be made.
+
+\lr{The end of the paragraph is all about motivation...}
 
@@
 
-The user can make decisions about storage, and the processing of the datasets/tasks and s/he will insert this information as an additional file to the Cylc workflow management.
-In fact, the user is already doing that, but he is doing that in just one specific way, to make the workflow possible.
+The user can make decisions about storage, and the processing of the datasets/tasks. S/he will insert this information as an additional file to the Cylc workflow management.
+In fact, the user is already doing that, but he is doing just to make the workflow feasible.
 Here, we are thinking about simplifying the information the user is already providing.
 So, instead of giving the complete workflow to Cylc, we are expecting the user to provide straightforward information, like:
+
+\lr{Now this part contradicts the previous part, because it talks about only one dataset per task.
+I wish we could keep something that high-level, but it might not be possible for the text and, worse, it might not be feasible for us.}
+
+\lr{Checking 4.2, we shoukd have something concrete here. High-level and concrete would be amazing.}
 
@@
 
+\lr{Does Cylc have the information about the system? If so, we can just use it. If not, how to get it for our purpose? And how it's done now?}
+
 \subsection{Smarter Scheduling}
+
+\lr{It's clear that I was mixing ESDM and EIOS in a very confusing way. :( The main information that should be here is, imho, in the next section. Consider just drop it.}
 
@@
 
-The appropriate software stack involved in executing a workflow with our I/O-aware scheduler is depicted in \Cref{fig:stages-io}.
-In the following, we describe the modifications to the current execution (described in \Cref{fig:stages}) we propose in this vision paper.
-Firstly, the end-user now has to provide an additional file that covers the I/O information for each task; slight changes have to be made to the scripts that are executed; these will be discussed below.
+The proposed I/O-aware scheduler, called here EIOS (ESDM I/O Scheduler), is now introduced.
+\lr{Note that this is the first mention to ESDM in the whole text! I think the E was for ESiWACE, but I think it has to be ESDM.}
+The appropriate software stack involved in executing a workflow with EIOS is depicted in \Cref{fig:stages-io}.
+In the following, we describe the modifications we propose in this vision paper to the current software stack (\Cref{fig:stages}).
+The end-user has now to provide an additional file that covers the I/O information for each task and slight changes (discussed below) have to be made to the scripts that currently are executed.
 
@@
   \includegraphics[scale=1.4]{stages-io}
-  \caption{Software Stack and Stages of Execution with our I/O-Aware Scheduler}
+  \caption{Software stack and stages of execution with our I/O-aware scheduler -- EIOS}
   \label{fig:stages-io}
@@
 \begin{enumerate}
-  \item \textbf{Cylc}\ \ Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
-  Now, Cylc calls EIOS to identify potential to optimise the schedule.
-  EIOS may have decided that subsequent jobs shall be placed on the same node or to reorder the execution of some jobs.
-  Also, it may have provided information for conducting data migration.
 
-  \item \textbf{EIOS}\ \
-  Reads information the workflow: the original suite file and the I/O suit file.
-  Acts depending on the stage of execution, i.e., we could be preparing execution in Cylc, or we are already inside a script.
-  It loads the \texttt{esdm.conf} file that contains the information about the available storage technology in the data centre and their characteristics.
-  Provides the information on how to prioritise I/O targets.
+  \item \textbf{Cylc}
+  %Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
+  Now, Cylc calls EIOS to identify potential optimisations in the schedule.
+  EIOS may decide that subsequent jobs shall be placed on the same node or to reorder the execution of some jobs.
+  Also, it may provide information for conducting data migration.
 
-  A data management service may run as a service and purge or migrate data according to user life cycle specification and EIOS information.
+  \item \textbf{EIOS} The proposed system reads the information about the workflow (original and I/O suit files) and acts depending on the stage of the execution.
+Two options are possible: we could be preparing the execution in Cylc, or we are already inside a script.
+  \lr{It's not clear to me the consequences of having to options in calling EIOS. It seems the next text is just about one of them.}
+  EIOS loads the \texttt{esdm.conf} file that contains information about the available technology in the data centre and it provides the information on how to prioritise I/O targets.
+  A data management plan may run as a service and purge or migrate data according to the user life cycle specification and the novelty information from EIOS.
 
-  \item \textbf{Slurm}\ \ Schedules the execution of the job considering the scheduling policy of the data centre.
-  Cylc may now have added a constraint to place a job on the same nodes as a previous job which is handled by the modified Slurm.
-  Also, if migrations have to be performed before or after a job is run, Slurm will administer them.
+  \item \textbf{Slurm}
+  %Schedules the execution of the job considering the scheduling policy of the data centre.
+  Cylc may now have added a constraint to place a job on the same node as a previous job and that is now handled by a modified Slurm.
+  Also, if migrations have to be performed before or after a job, Slurm will administer them.
 
-  \item \textbf{Job}\ \ The user-provided script is executed on a single node assigned to the job.
+  \item \textbf{Job}
+  %The user-provided script is executed on a single node assigned to the job.
+  % The script has set environment variables containing information about the Cylc execution (e.g., a task in the workflow) and the Slurm job (e.g., number of nodes).
 
-  The script has set environment variables containing information about the Cylc execution (e.g., a task in the workflow) and the Slurm job (e.g., number of nodes).
+\lr{Any novelty here? It's better say it's not than say nothing.}
 
-  \item \textbf{Script}\ \ The shell-script executes the sequence of commands.
+  \item \textbf{Script}
+  %The shell-script executes the sequence of commands.
+  % Typically, to create a filename, the script calls Cylc commands with a template provided by the user and feeds this information to the application.
+  % This ultimately defines the storage location to be used.
+  % A script typically consists of multiple fine-grained commands and may even call multiple parallel applications.
+  The way the filenames are generated is now different. A replacement command invokes EIOS to generate a pseudo filename.
+  This filename will encode additional information for ESDM about how to prioritise data placement according to data access.
+  Potentially, we can also generate an \texttt{esdm.conf} file to reflect the specific hardware we have available to run the specific application.
 
-  Typically, to create a filename, the script calls Cylc commands with a template provided by the user and feeds this information to the application.
-  This ultimately defines the storage location to be used.
-  A script typically consists of multiple fine-grained commands and may even call multiple parallel applications.
+\lr{You already talked about esdm.conf in 2, almost the same information. And 6 will use it, so it has to be available.}
 
-  The way the filenames are generated is now different; we will now use a replacement command that invokes a tool of EIOS to generate a pseudo filename.
-  This filename will encode additional information for ESDM about how to prioritise data placement according to data access.
-  Potentially, we also can generate an \texttt{esdm.conf} file to reflect the specific hardware we have available to run the specific application.
+  \item \textbf{Application}
+  %The user-specified parallel application runs.
+  The application may either use NetCDF with ESDM support or ESDM directly to access datasets.
+  Hence, in \Cref{fig:layers}, the HDF5 layer is replaced with ESDM.
+  ESDM loads the \texttt{esdm.conf} file that contains the information about the available storage backends and their configuration.
+  % Knowing the potential targets, ESDM schedules application-local I/O.
+  ESDM extracts the long-term schedule information from the generated filenames and considers it during the I/O scheduling to exploit the data locality between tasks.
 
-  \item \textbf{Application}\ \ The user-specified parallel application runs.
-  The application may either use NetCDF with ESDM support or ESDM directly to access data sets.
-  Hence, in \Cref{fig:layers}, the HDF5 layer is replaced with ESDM.
-  ESDM loads the \texttt{esdm.conf} file that contains the information about the available storage backends to use and their configuration.
-  Knowing the potential targets, ESDM schedules application-local I/O.
-  ESDM extracts the long-term schedule information from the generated filenames and considers it during I/O scheduling to exploit the data locality between tasks.
+  \lr{Now I'm not sure I'm comfortable with ESDM being enforced here.
+Is the replacement mandatory to our model works?
+I don't think so.
+It's nice, sure, but it seems tie-in sale, which is not good.
+Also note that I commented a lost sentence here, that can be used somewhere else.}
+
 \end{enumerate}

### 2020-01-29 paper/paper.tex
Comments cleanup except 5.

@@
 \newcommand{\jk}[1]{\todo[inline]{JK: #1}}
-\newcommand{\lr}[1]{\todo[inline]{LR: #1}}
+\newcommand{\lr}[1]{\textcolor{blue}{LR: #1}}
 
@@
 In this figure, for example, XIOS is beneficial for parallel I/O and a process might directly rely on the NetCDF API to read previous data.
+
+Within ESiWACE, we are developing ESDM, ... \jk{TODO}
 \end{minipage}
@@
 
-\lr{This next paragraph should be migrate to 4.3 and again mentioned in 5.3.}
+%It was correctly placed here
 
-Indeed, by monitoring I/O accesses, we can automatically draft such I/O descriptions on behalf of the user to simplify the specification.
+By monitoring I/O accesses, we can automatically draft such I/O descriptions on behalf of the user to simplify the specification.
 All we need is to run one cycle of the specific workflow, considering that a workflow cycle is executed many times in climate and weather; this is a cheap approach.
@@
 From the diagram we can also notice that Task \textbf{model} will generate one dataset for Task \textbf{post\,proc}, one dataset for Task \textbf{prod\,gen}, two datasets for Task \textbf{verification}, and three datasets for later cycles.
-The access to this information before the process starts can be used to optimise how these datasets will be stored during the workflow.
+
+The access to this information before the process starts can be used to optimise where compute jobs are executed and how these datasets will be stored during the workflow.
 This information today is static and manual, leaving opportunities for automatisation and optimisations open.
@@
 Once we can insert the datasets explicitly in the workflow, it is easier to see that many optimisations regarding storage can be made.
-
-\lr{The end of the paragraph is all about motivation...}
 
@@
 The proposed I/O-aware scheduler, called here EIOS (ESDM I/O Scheduler), is now introduced.
-\lr{Note that this is the first mention to ESDM in the whole text! I think the E was for ESiWACE, but I think it has to be ESDM.}
 The appropriate software stack involved in executing a workflow with EIOS is depicted in \Cref{fig:stages-io}.
@@
 
-  \item \textbf{EIOS} The proposed system reads the information about the workflow (original and I/O suit files) and acts depending on the stage of the execution.
-Two options are possible: we could be preparing the execution in Cylc, or we are already inside a script.
-  \lr{It's not clear to me the consequences of having to options in calling EIOS. It seems the next text is just about one of them.}
-  EIOS loads the \texttt{esdm.conf} file that contains information about the available technology in the data centre and it provides the information on how to prioritise I/O targets.
-  A data management plan may run as a service and purge or migrate data according to the user life cycle specification and the novelty information from EIOS.
+  \item \textbf{EIOS}
+  The proposed system reads the information about the workflow (original and I/O suit files) and acts depending on the stage of the execution.
+  EIOS consists of several subcomponents:
+    \begin{itemize}
+      \item The high-level scheduler that interfaces with Cylc.
+      \item A tool that is run inside a script of a specific job/task -- as a replacement to the \texttt{cylc cycle-point} call.
+      It generates pseudo file names that are used by the applications with ESDM support.
+      \item A data management service (not shown on the figure) that purges or migrate data according to the user life cycle specification and the novelty information from EIOS.
+    \end{itemize}
+  EIOS components use the \texttt{esdm.conf} file that contains information about the available technology in the data centre and it provides the information on how to prioritise I/O targets.
 
@@
   \item \textbf{Job}
+  A job might be scheduled on the same nodes than a previous job to utilize local storage.
   %The user-provided script is executed on a single node assigned to the job.
   % The script has set environment variables containing information about the Cylc execution (e.g., a task in the workflow) and the Slurm job (e.g., number of nodes).
-
-\lr{Any novelty here? It's better say it's not than say nothing.}
 
@@
   This filename will encode additional information for ESDM about how to prioritise data placement according to data access.
-  Potentially, we can also generate an \texttt{esdm.conf} file to reflect the specific hardware we have available to run the specific application.
-
-\lr{You already talked about esdm.conf in 2, almost the same information. And 6 will use it, so it has to be available.}
+  Potentially, EIOS may generate an \texttt{esdm.conf} file for the specific application to adjust the mapping of data to available storage targets.
 
@@
   ESDM extracts the long-term schedule information from the generated filenames and considers it during the I/O scheduling to exploit the data locality between tasks.
-
-  \lr{Now I'm not sure I'm comfortable with ESDM being enforced here.
-Is the replacement mandatory to our model works?
-I don't think so.
-It's nice, sure, but it seems tie-in sale, which is not good.
-Also note that I commented a lost sentence here, that can be used somewhere else.}
 

### 2020-01-29 paper/paper.tex
Nai

@@
 
-The user can make decisions about storage, and the processing of the datasets/tasks. S/he will insert this information as an additional file to the Cylc workflow management.
-In fact, the user is already doing that, but he is doing just to make the workflow feasible.
-Here, we are thinking about simplifying the information the user is already providing.
-So, instead of giving the complete workflow to Cylc, we are expecting the user to provide straightforward information, like:
+The user provides information about datasets required for input and the generated output in a separate file similarly to Cylc's suite file.
+This information contains dataset information for each Cylc task:
+\begin{itemize}
+  \item name: the name of the field generated/required, this may include a pattern such as supported by Cylc
+  \item size: an estimate of the file size
+  \item lifetime: how long the data must be retained on storage (if at all)
+  \item type: may specify the type of the data, i.e., checkpoint, diagnostics, temporary
+\end{itemize}
+% Expressions will support generic variables such as the number of nodes a job runs...
 
-\lr{Now this part contradicts the previous part, because it talks about only one dataset per task.
-I wish we could keep something that high-level, but it might not be possible for the text and, worse, it might not be feasible for us.}
-
-\lr{Checking 4.2, we shoukd have something concrete here. High-level and concrete would be amazing.}
-
-\begin{itemize}
-
-\item Task\,1 usually requires ten hours running in a pc with standard configuration\footnote{Intel Core i5, 8 GB RAM, and 500 GB internal storage drive, for instance.}, requires 100 GB of storage that has to be persistent for no more than one week.
-
-\item Task\,2 usually requires two hours running in a pc with standard configuration, requires 150 GB of storage that has to be persistent for no more than two weeks week.
-
-\item Task\,3 usually requires five hours running in a pc with standard configuration, and it does not require extra storage other than the storage for the final results.
-
-\end{itemize}
+% \begin{itemize}
+% \item Task\,1 usually requires ten hours running in a pc with standard configuration\footnote{Intel Core i5, 8 GB RAM, and 500 GB internal storage drive, for instance.}, requires 100 GB of storage that has to be persistent for no more than one week.
+% \item Task\,2 usually requires two hours running in a pc with standard configuration, requires 150 GB of storage that has to be persistent for no more than two weeks week.
+% \item Task\,3 usually requires five hours running in a pc with standard configuration, and it does not require extra storage other than the storage for the final results.
+% \end{itemize}
 
@@
 
-\lr{Does Cylc have the information about the system? If so, we can just use it. If not, how to get it for our purpose? And how it's done now?}
+The information about the system is provided using a configuration file for ESDM.
 
@@
 
-\lr{It's clear that I was mixing ESDM and EIOS in a very confusing way. :( The main information that should be here is, imho, in the next section. Consider just drop it.}
-
-Now, comes the question.
-Assuming the information is available, who will be responsible for making the decisions for optimising the storage and the processing tasks?
-The answer here is simple: the ESDM Scheduler will do it.
+The ESDM Scheduler will make the schedule.
 The ESDM Scheduler can be used in different steps of the process.

### 2020-02-04 paper/paper.tex
Updating

@@
-%\documentclass[a4paper, twocolumn]{article}
-\documentclass[a4paper]{article}
-\newcommand{\papertitle}{Potential of I/O-Aware Workflows in Climate and Weather}
+%% This is an article based on superfri.cls class file of
+%% ``Supercomputing Frontiers and Innovations. An International Journal''
+%% http://superfri.org/.
 
-\usepackage[a4paper, margin=2cm]{geometry}
+\documentclass{superfri}
+\usepackage{graphicx}
+\usepackage[hidelinks]{hyperref}
 
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage[english]{babel}
-\usepackage[colorlinks=true,urlcolor=red]{hyperref}
-\usepackage{url}
-\usepackage{float}
+\usepackage{todonotes}
 \usepackage{cleveref}
-\usepackage{todonotes}
+\usepackage{subcaption}
 \usepackage{comment}
-\usepackage{ulem}
-\usepackage{array}
-\usepackage{subcaption}
 
@@
 \newcommand{\lr}[1]{\textcolor{blue}{LR: #1}}
+\pagestyle{plain}
 
-\usepackage{multicol}
-\setlength{\columnsep}{1cm}
-
-\usepackage{fancyhdr}
-\fancyhead{}
-\fancyfoot{}
-\fancyhead[l]{\papertitle}
-\fancyhead[r]{\includegraphics[height=1em]{esiwacelogo_type_grey_left_2c_ext}}
-\fancyfoot[r]{\thepage}
-\pagestyle{fancy}
-\renewcommand{\headrulewidth}{1pt}
-\renewcommand{\footrulewidth}{1pt}
-
-\usepackage{titling}
-\pretitle{\begin{center}\Large\bfseries}
-\posttitle{\end{center}\vskip 0.5em}
-
-\graphicspath{{./assets/}}
-
-\title{\papertitle}
-
-\author{
-%  Julian M. Kunkel
-%  \textit{University of Reading}
-%    \and
-%  Luciana R. Pedro
-%  \textit{University of Reading}
-}
-\date{\today}
+% ------------
+\bibliographystyle{superfri}
 
 \begin{document}
-\maketitle
-\thispagestyle{fancy}
 
-\section*{Abstract}
+\author{}
+%\author{Julian M. Kunkel\footnote{\label{uread}University of Reading, Reading, United Kingdom}\orcidID{0000-0002-6915-1179} \and Luciana R. Pedro\footnoteref{uread}\orcidID{0000-0001-8365-6264}
+%}
+
+\title{Potential of I/O-Aware Workflows in Climate and Weather}
+
+\maketitle{}
+
+\begin{abstract}%
 The efficient, convenient, and robust execution of data-driven workflows and enhanced data management are key for productivity in scientific computing.
@@
 The key contributions of this paper are:
-1) we sketch the visions of an integrated data-driven approach and discuss the challenges and implications of this strategy;
-2) we illustrate architecture and roadmap that allows the seamless integration into existing workflows.
+1) we sketch the visions of an integrated data-driven approach and discuss the challenges and implications of this strategy, and 2) we illustrate architecture and roadmap that allows the seamless integration into current climate and weather workflows.
 The tools employed here to achieve this extended workflow are Cylc, XIOS, DDN IME, and ESDM.
 
-%Basically, we believe workflows composed of data, computing, and communication-intensive tasks should drive the interfaces and hardware configurations to best support the programming models.
-From the user perspective, we ultimately aim to provide an abstraction to data dependencies.
-The proposed changes increase the opportunity of implementations for smarter scheduling of computing and storage in heterogeneous storage environments considering characteristics of workflows and data.
-From the system perspective, the vision aims to derive an execution plan that utilises the available hardware and software infrastructure within a data centre.
+We believe workflows composed of data, computing, and communication-intensive tasks should drive the interfaces and hardware configurations to best support the programming models.
+The changes proposed in this work increase the opportunity of implementations for smarter scheduling of computing and storage in heterogeneous storage environments considering characteristics of workflows and data.
 
-\section{Introduction}
+\keywords{workflow, climate, weather, heterogeneous storage, data-driven}
+\end{abstract}
+
+% -----------------------------------------------------------------------
+\section*{Introduction}
+\label{sec:intro}
 
@@
 While the current advances correspond to a big leap forward, many processes still require experts.
-For example, porting a workflow from one system to another requires to adjust runtime parameters of applications and to decide how data is managed.
+For example, porting a workflow from one system to another still requires adjusting runtime parameters of applications and deciding on how data is managed.
 Since performance is of crucial importance to large-scale workflows, careful attention must be paid to exploit the system characteristics of the target supercomputer.
@@
 As the complexity of systems increases and alternative storage and computing technologies provide unique characteristics, it becomes increasingly difficult even for experts to manually optimise resource usage in workflows.
-In many cases, modifications are not performed because: 1) they are labour intense: any change to the workflow requires careful validation which may not pay off for small scale runs; 2) users are not aware of the potential of the complex system.
+In many cases, modifications are not performed because: 1) they are labour intense: any change to the workflow requires careful validation which may not pay off for small scale runs, and 2) users are not aware of the potential of the complex system.
 
-In this paper, we show that knowledge about Input/Output of workflow tasks and overall experimental design helps to
-optimise the execution of climate and weather workflows and may increase the performance, throughput and cost-efficiency of the environments, providing an incentive to users and data-centres that cannot be neglected.
-
-%The paper is organised as follows:
-%In \Cref{sec:workflows}, we provide an overview of the specification of workflows in climate and weather and the involved software stack.
-
-\jk{TODO}
+In this paper, we illustrate how knowing the Input/Output (I/O) characteristics of workflow tasks and overall experimental design help to optimise the execution of climate and weather workflows.
+This enlightenment may increase the performance, throughput and cost-efficiency of the environments, providing an incentive to users and data-centres that cannot be neglected any longer.
+Our approach intends to reduce the burden on researchers and, at the same time, optimise the decisions about jobs running in HPC systems.
+Once we have an automated decision about where the job will run and how the storage will be managed, climate and weather scientists can then run their workflows in any machine efficiently without prior knowledge about its architecture.
 
@@
 A variety of accelerators (GPU, TPU, FPGAs), active storage, in-memory, and in-network computing technologies provide storage and processing capabilities.
-\Cref{fig:heterogeneous} shows such an environment with a focus on computation and storage.
+\figref{pic/system} shows such an environment with a focus on computation and storage.
 Some of these technologies might be local to specific compute nodes or globally available.
@@
 
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.6\columnwidth]{system}
-  \caption{Example of an heterogeneous HPC landscape}
-  \label{fig:heterogeneous}
-\end{figure}
+\fig{width=0.6\columnwidth}{pic/system}{Example of an heterogeneous HPC landscape}
 
@@
 
-A typical I/O stack for a single parallel application like a climate model is shown in \Cref{fig:layers}.
+A typical I/O stack for a single parallel application like a climate model is shown in \figref{pic/layers-xios}.
 
-\begin{minipage}{0.2\textwidth}
-\begin{figure}[H]
-  \includegraphics[width=\textwidth]{layers-xios}
-  \caption{I/O path for an MPI-parallel application}
-  \label{fig:layers}
-\end{figure}
-\end{minipage}
-\qquad
-\begin{minipage}{0.7\textwidth}
+\fig{width=0.2\textwidth}{pic/layers-xios}{I/O path for an MPI-parallel application}
+
 In our example, we assume the application is parallelised for performance reasons and uses MPI to run the application.
@@
 Within ESiWACE, we are developing ESDM, ... \jk{TODO}
-\end{minipage}
 
@@
 
-The appropriate software stack involved in executing a workflow is depicted in \Cref{fig:stages}.
-As a preliminary, the user had specified a Cylc suite representing the workflow and provided scripts files for the individual tasks to execute; now the user enacts Cylc to start the workflow.
+The appropriate software stack involved in executing a general workflow is depicted in \figref{pic/stages}.
+As a preliminary, the user has to specify a configuration representing the workflow and provide scripts for the individual tasks to be executed.
+After that, the user enacts a workflow engine to start the workflow.
 In the following, each stage of the execution is described further.
 
-\begin{figure}[H]
-  \centering
-  \includegraphics[scale=1.4]{stages}
-  \caption{Software stack and stages of execution}
-  \label{fig:stages}
-\end{figure}
+\fig{width=0.8\textwidth}{pic/stages}{Software stack and stages of execution}
 
 \begin{enumerate}
-  \item \textbf{Cylc} Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
-  Cylc monitors the progress of the workflow.
-  Once a task can be executed (dependencies are fulfilled), Cylc generates a Slurm script on the fly with the required metadata for Slurm that will invoke the user-defined script for that task and runs Slurm to queue the job.
 
-  \item \textbf{Slurm} Slurm schedules the execution of the job considering the scheduling policy of the data centre.
+  \item \textbf{Scientist} specifies a configuration representing the workflow and provides scripts for each task.
+
+  \item \textbf{Workflow Engine} parses the workflow configuration file, generate tasks dependencies, defines a schedule for the execution, and monitors the progress of the workflow.
+  Once a task can be executed (dependencies are fulfilled), the workflow engine generates a Slurm script on the fly with the required metadata that will load the user-defined script for each task.
+Then, Slurm runs to queue the job.
+
+  \item \textbf{Slurm} \cite{Jette02slurm:simple} schedules the execution of the job, considering the scheduling policy of the data centre.
   Once the job is ready to be dispatched, i.e., resources are available, and the job priority is the highest, it is started on the supercomputer.
 
-  \item \textbf{Job} The user-provided script is executed on a single node assigned to the job.
-  The script has set environment variables containing information about the Cylc execution (e.g., a task in the workflow) and the Slurm job (e.g., number of nodes).
+  \item \textbf{Job} runs the user-provided script on one single node assigned by Slurm.
+  The script has set environment variables containing information about the execution (e.g., a task in the workflow) and the Slurm job (e.g., number of nodes).
 
-  \item \textbf{Script} The shell-script executes the sequence of commands, including applications that are parallel programs.
-  Typically, to create a filename, the script calls Cylc commands with a template provided by the user and feeds this information to the application which, ultimately, defines the storage location to be used.
+  \item \textbf{Script} typically calls commands to create a filename with a template provided by the user and feeds this information to the application which, ultimately, defines the storage location.
   A script typically consists of multiple fine-grained commands and may even call multiple parallel applications.
 
-  \item \textbf{Application} The user-specified parallel application runs taking the filenames as defined in a configuration file or as arguments.
-  The application may use NetCDF or XIOS to perform the I/O.
+  \item \textbf{Application} runs taking the generated filenames that the script has set.
+  The user-specified parallel application may use XIOS \cite{xios} or NetCDF \cite{Jenter92netcdf:a} to perform the I/O.
+
 \end{enumerate}
-
-\subsection{Cylc}
-
-Cylc is in charge to execute and monitor workflows in which each step is submitted to the batch scheduler of a data centre.
- %, and the relevance of heterogeneous infrastructure to run workflows.
-%We focus on high-level considerations to illustrate mandatory concepts.
-
-Consider the Cylc suite configuration for a toy monthly cycling workflow in \Cref{fig:cylc} \cite{8675433}.
-In this workflow, a warm-cycled atmospheric model (\textbf{model}) simulates the physics from a current state to predict the future after a timestep, for example, after six hours.
-This process is repeated in the model to simulate month and years into the future.
-Once the simulation of any month is computed, the data for this month becomes available and can be analysed.
-A user has specified a detailed workflow in a Cylc suite file (\texttt{suite.rc}).
-In this workflow, the \textbf{model} is followed by postprocessing (\textbf{post}), forecast verification (\textbf{ver}), and product generation (\textbf{prod}) tasks.
-Also, consider, in addition, a task \textbf{check} that compares some verification metric against products from two cycles earlier.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.9\columnwidth]{cylc1}
-  \caption{Example workflow}
-  \label{fig:cylc}
-\end{figure}
-
-This suite presents the dependencies among the tasks, but it is missing the information about I/O and storage.
-Decisions about storage must take into consideration the architecture in which the workflow will run.
-The developers make the configuration about parallelism and where the data is stored in a script file that is executed for each specific task.
 
@@
 
-The scripts for the tasks in Cylc define how the data is stored on the available storage systems.
+Usually, the scripts for tasks in climate and weather workflows define how the data is stored on the available storage systems.
+What happens in many current workflows is that they ignore the benefits of using multiple file systems concurrently or exploiting the data locality between tasks by co-locating the tasks.
+On top of that, in the current state-of-the-art, scientists have to optimise the available storage resources intuitively and insert the information about this decision-making process manually.
+Once we have the complete knowledge about how the workflow is expected to run, we can optimise steps that were before predefined by the user, usually considering only options for feasibility.
+Machine Learning has been proven effective in replacing human-beings, and it may adapt the workflow dynamically in case of unexpected situations (the system is down, nodes are unresponsive, etc.).
+
 Consider that the machine has three file systems: a fast \textbf{scratch} file system on which data may reside only for a week, a slower \textbf{work} file system, and each node has a \textbf{local} file system.
-
 What happens in many current workflows is that they use the only \textbf{work} and \textbf{scratch}.
@@
 
-\Cref{fig:lifecycle} shows three possible life cycles and the access pattern for a specific dataset.
-As shown in \Cref{fig:lifecycle1}, the dataset could be stored on the \textbf{local} storage first to avoid congestion on the \textbf{work} file system, then be migrated to \textbf{work} file system where subsequent tasks of the workflow may read it multiple times.
+\figref{fig:lifecycle} shows three possible life cycles and the access pattern for a specific dataset.
+As shown in \figref{fig:lifecycle1}, the dataset could be stored on the \textbf{local} storage first to avoid congestion on the \textbf{work} file system, then be migrated to \textbf{work} file system where subsequent tasks of the workflow may read it multiple times.
 In the end, this dataset might be an intermediate product that can then be deleted.
-Alternatively (\Cref{fig:lifecycle2}, it could be stored on the \textbf{scratch} file system immediately and accessed there.
+Alternatively (\figref{fig:lifecycle2}, it could be stored on the \textbf{scratch} file system immediately and accessed there.
 However, that would require that the last read access happens before files on \textbf{scratch} are automatically removed.
-The last scenario (\Cref{fig:lifecycle3}) presents the case where data is created on \textbf{work} by a task.
+The last scenario (\figref{fig:lifecycle3}) presents the case where data is created on \textbf{work} by a task.
 Still, then it is copied to a \textbf{local} node to allow read accesses of subsequent tasks which might be beneficial for small random accesses.
@@
 
-\begin{figure}[H]
+\begin{figure}[b]
+    \begin{minipage}{.33\linewidth}
         \centering
-  \begin{subfigure}{.3\textwidth}
-  \includegraphics[width=0.9\columnwidth]{lifecycle-1}
-  \caption{\label{fig:lifecycle1} Local and work file systems}
-  \end{subfigure}
-  \begin{subfigure}{.3\textwidth}
-  \includegraphics[width=0.9\columnwidth]{lifecycle-2}
-  \caption{\label{fig:lifecycle2} Scratch file system only}
-  \end{subfigure}
-  \begin{subfigure}{.3\textwidth}
-  \includegraphics[width=0.9\columnwidth]{lifecycle-3}
-  \caption{\label{fig:lifecycle3} Local and work file systems}
-  \end{subfigure}
+        \includegraphics[width=0.9\columnwidth]{pic/lifecycle-1}
+        \subcaption{Local and work file systems}\label{fig:lifecycle1}
+    \end{minipage}
+    \begin{minipage}{.33\linewidth}
+        \centering
+        \includegraphics[width=0.9\columnwidth]{pic/lifecycle-2}
+        \subcaption{Scratch file system only}\label{fig:lifecycle2}
+    \end{minipage}
+    \begin{minipage}{.33\linewidth}
+        \centering
+        \includegraphics[width=0.9\columnwidth]{pic/lifecycle-3}
+        \subcaption{Local and work file systems}\label{fig:lifecycle3}
+    \end{minipage}
+    \vspace{5pt}
     \caption{Alternative life cycles for mapping a dataset to storage and the operations: \textbf{C}reate, \textbf{M}igrate, \textbf{R}ead, and \textbf{D}elete}
@@
 \section{Vision for I/O-Aware Workflows}
+\label{sec:vision}
+
+There are several approaches to implement the technology for this vision, and changes need to be made to software components to realise it.
+In the next sections, we will discuss a specific design for our transitional roadmap considering climate and weather workflows and the tools scientists from this field already use in their routine research.
 
@@
 
-Our vision requires two additional information: firstly, the user must augment the workflow description with information about I/O requirements and explicitly annotate dependencies to data sets; secondly, the data centre (or expert user) must provide a description of the storage architecture and characteristics for each storage system.
-
-There are many approaches to implement the technology for this vision. It is apparent that changes need to be made to several software components to realise this vision.
-In the next section, we will discuss the specific design for our transitional roadmap for climate and weather.
+Our vision for I/O-aware workflows requires two additional information.
+Firstly, the user must augment the workflow description with information about I/O requirements and explicitly annotate dependencies to data sets.
+Secondly, the data centre (or expert user) must provide details about the storage architecture.
 
@@
 
-So far, Cylc allowed specifying tasks and the dependencies among them.
-We aim to enrich this information with characteristics for input, and output, i.e., the datasets used.
-An example workflow containing input data sets and (intermediate) products is illustrated in \Cref{fig:workflow}.
-Nodes represent tasks or data; arrows indicate dependencies among between them.
+In general, climate and weather workflows allow specifying tasks and the dependencies among them.
+We aim to enable enriching this information with characteristics for input, and output, i.e., the datasets.
+An example workflow containing input datasets and (intermediate) products is illustrated in \figref{pic/workflow}.
+Nodes represent tasks or data, and arrows indicate the dependencies.
 In the example, Task\,1 needs two datasets to perform its work, it directly communicates with Task\,2 and produces Product\,1.
@@
 This last step represents the typical uncertainty of data reuse, i.e., it is unclear how Product\,3 will be used further.
-In this approach, each task is annotated with the required inputs and the generated products.
+In the approach proposed in this work, each task is annotated with the required input datasets and the generated products with their characteristics.
 Each data product is annotated with metadata such as data life cycle information, e.g., the value of data and how long should it be kept.
-
-%It was correctly placed here
 
@@
 
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.4\columnwidth]{workflow}
-  \caption{Example of a high-level workflow with tasks and data dependencies}
-  \label{fig:workflow}
-\end{figure}
+\fig{width=0.4\columnwidth}{pic/workflow}{Example of a high-level workflow with tasks and data dependencies}
 
@@
 
-The system information shall comprise of all available storage systems, the system topology, and the characteristics of each of its required components.
+While many optimisations are possible once the abstraction is in place, here we focus on utilising a heterogeneous storage environment.
+The optimisations we discuss are related to the life cycle of datasets and the placement of such datasets into specific storage according to system performance characteristics and the workflow specification.
+
+The system information shall comprise of all available storage systems, the system topology, and details of each of its required components.
 This may include simplified or complex models of the components to approximate performance for specific I/O patterns allowing to predict performance behaviour.
-It is expected that the data centre (or expert users) can create such a configuration file, for instance, by using vendor-provided information or by executing benchmarks.
+It is expected that the data centre (or expert users) can create such a configuration file, e.g., by using vendor-provided information or by executing benchmarks.
+With this information, the I/O scheduler can make the placement and migration decisions for individual datasets and manage their relocation during their life cycle, as discussed in \figref{sec:datamanagement}.\lr{Check ref.}
+This separation of concerns allows us to abstract from the workflow what is essential and what a machine should optimise to ensure that the resources available are being used smartly.
 
@@
 
-With this separation of concerns, we can abstract from the workflow what is essential and what a machine should optimise to ensure that the resources available are being used smartly.
 In this paper, we focus on utilising a heterogeneous storage environment instead of exhaustively discussing the potential that is unleashed with the right level of abstraction in place by researchers that exploit the available opportunities.
-Every time parameters are manually tuned right now there is a room from improvement by using automatised decision making.
-And Machine Learning has been proven effective in replacing human-beings and adapt the workflow dynamically in case of unexpected situations (the system is down, nodes are unresponsive, etc.).
-
-Optimisations we discuss in this paper are related to the life cycle of datasets and the placement of such datasets into specific storage according to system performance characteristics and the workflow specification.
-Firstly, the I/O scheduler could make the placement decision for the individual data sets and manage their relocation during the life cycle, as discussed in \Cref{sec:datamanagement}.
 
@@
 
-\Cref{fig:cycle4} exemplifies a restriction of our example climate/weather workflow the way it was being generated.
-Assume you have, for instance, four cycles of the model running with all the necessary post-processing steps.
+\figref{pic/cycle4} exemplifies a restriction of our example climate/weather workflow the way it was being generated.
+Assume you have, for instance, five cycles of the model running with all the necessary post-processing steps.
 Typically, the user has written the scripts for each task by defining a filename using the same prefix that is then placed at the same storage type every time.
@@
 The user might even have analysed various scenarios such as depicted in
-\Cref{sec:datamanagement}.
+Section \ref{sec:datamanagement}.
 
@@
 
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.8\columnwidth]{cycle4}
-  \caption{Four stages workflow example}
-  \label{fig:cycle4}
-\end{figure}
+\fig{width=0.8\columnwidth}{pic/cycle4}{Four stages workflow example}
 
@@
 
-Once we have the complete knowledge about how the workflow is expected to run, we can optimise steps that were before predefined by the user, usually considering only options for feasibility.
+The benefits of the proposed vision are:
 
 \begin{itemize}
-\item Abstraction: The user does not have to worry about the architecture that the workflow will run, removing the specialist from the decision-making process.
 
-\item Optimisation: The workflow will be optimised for the specific available system characteristics considering potentially runtime characteristics and storage resilience which will increase performance and system utilisation.
-By knowing the workflow behind the data creation, the system could trade data security with potential recomputation opportunities to optimise the cost-efficiency.
+\item Abstraction: The user does not have to worry about the architecture that the workflow will run, removing the specialist from the decision-making in the storage process.
 
-\item Performance-portability: The user has to specify the I/O requirements only once for the tasks of the workflow.
-The workflow can now be run on any system without user intervention.
-Even more, if the system characteristics change, e.g., it gets upgraded, and an additional storage tier becomes available, or if storage degrades, the I/O-aware workflow could automatically adapt and make use of this new environment.
+\item Optimisation: The workflow will be optimised specifically for the available system infrastructure considering potentially runtime characteristics and storage resilience which will increase performance and system utilisation.
+By knowing the workflow behind the data creation, the system could even trade data security with potential recomputing opportunities to optimise the cost-efficiency.
+
+\lr{Introduce the picture as an example?}
+
+\item Performance-portability: The user has to specify the I/O requirements only once for the tasks of a specific workflow, and the workflow can now run on any system without user intervention.
+Even more, if the system characteristics change, e.g., it gets upgraded, an additional storage tier becomes available, or if storage degrades, the I/O-aware workflow could automatically adapt and make use of this new environment.
 
@@
 
-\item Adapts the resulting workflow, individual scripts and application executions to consider the automatic I/O schedule.
+\item Adapts the resulting workflow, individual scripts and application executions to consider the automatic I/O scheduler.
 
@@
 
+\subsection{Cylc}
+
+Cylc is in charge to execute and monitor workflows in which each step is submitted to the batch scheduler of a data centre.
+ %, and the relevance of heterogeneous infrastructure to run workflows.
+%We focus on high-level considerations to illustrate mandatory concepts.
+
+Consider the Cylc workflow for a toy monthly cycling workflow in \figref{pic/cylc1} \cite{8675433}.
+In this workflow, a warm-cycled atmospheric model (\textbf{model}) simulates the physics from a current state to predict the future after a timestep, for example, after six hours.
+This process is repeated in the model to simulate month and years into the future.
+Once the simulation of any month is computed, the data for this month becomes available and can be analysed.
+A user has specified a detailed workflow in a Cylc configuration file (\texttt{conf.rc}).
+In this workflow, the \textbf{model} is followed by postprocessing (\textbf{post}), forecast verification (\textbf{ver}), and product generation (\textbf{prod}) tasks.
+Also, consider, in addition, a task \textbf{check} that compares some verification metric against products from two cycles earlier.
+
+\fig{width=0.9\columnwidth}{pic/cylc1}{Example workflow}
+
+This workflow presents the dependencies among the tasks, but it is missing the information about I/O and storage.
+Decisions about storage must take into consideration the architecture in which the workflow will run.
+The developers make the configuration about parallelism and where the data is stored in a script file that is executed for each specific task.
+
 \subsection{Agregatting Data Dependencies}
 
-\Cref{fig:cycle1} introduces one cycle of the suite in \Cref{fig:cylc} with a simplified notation.
+\figref{fig:cycle1} introduces one cycle of the workflow in \figref{pic/cylc1} with a simplified notation.
 The arrows represent the tasks dependencies, but there is no information about the data that are generated.
@@
 
-\begin{comment}
-
-\begin{figure}[H]
+\begin{figure}[b]
+    \begin{minipage}{.45\linewidth}
         \centering
-\begin{tabular}{ccc}
-\includegraphics[scale=0.6]{cycle1} & \qquad &
-\includegraphics[scale=0.6]{cycle-io-dep} \\
-(a) Original Cycle & & (b) Cycle including datasets and dependencies
-\end{tabular}
-\caption{ \label{fig:cycle} Simplified Cylc workflow for the first cycle}
+        \includegraphics[width=0.9\linewidth]{pic/cycle1}
+        \subcaption{Including tasks and dependencies}\label{fig:cycle1}
+    \end{minipage}
+    \begin{minipage}{.45\linewidth}
+        \centering
+        \includegraphics[width=0.9\linewidth]{pic/cycle-io-dep}
+        \subcaption{Including tasks, datasets and dependencies}\label{fig:cycle-io-dep}
+    \end{minipage}
+    \vspace{5pt}
+    \caption{Example of a high-level workflow}
+    \label{fig:cycle}
 \end{figure}
 
-\end{comment}
-
-\begin{figure}[H]
-  \begin{subfigure}{0.45\textwidth}
-    \centering
-    \includegraphics[width=\columnwidth]{cycle1}
-    \caption{Original Cycle}
-    \label{fig:cycle1}
-  \end{subfigure}
-  \qquad
-  \begin{subfigure}{0.45\textwidth}
-    \centering
-    \includegraphics[width=\columnwidth]{cycle-io-dep}
-    \caption{Cycle including datasets and dependencies}
-    \label{fig:cycle-io-dep}
-  \end{subfigure}
-  \caption{Simplified Cylc workflow for the first cycle}
-\end{figure}
-
-The scenario presented in \Cref{fig:cycle1} shows the same cycle of \Cref{fig:cycle-io-dep}, but  extended information about datasets is added.
+The scenario presented in \figref{fig:cycle1} shows the same cycle of \figref{fig:cycle-io-dep}, but  extended information about datasets is added.
 In this picture, we have the tasks dependencies, and we explicitly list the datasets dependencies regarding the tasks that generated them.
@@
 
-The user provides information about datasets required for input and the generated output in a separate file similarly to Cylc's suite file.
+The user provides information about datasets required for input and the generated output in a separate file similarly to Cylc's workflow configuration file.
 This information contains dataset information for each Cylc task:
-\begin{itemize}
-  \item name: the name of the field generated/required, this may include a pattern such as supported by Cylc
-  \item size: an estimate of the file size
-  \item lifetime: how long the data must be retained on storage (if at all)
-  \item type: may specify the type of the data, i.e., checkpoint, diagnostics, temporary
-\end{itemize}
+
+\begin{description}
+
+  \item[Name] The name of the field generated/required, which may include a pattern such as supported by Cylc.
+  \item[Size] An estimate of the file size.
+  \item[Lifetime] How long the data must be retained on storage (if at all).
+  \item[Type] It may specify the type of data, i.e., checkpoint, diagnostics, temporary.
+
+\end{description}
+
 % Expressions will support generic variables such as the number of nodes a job runs...
@@
 
-The ESDM Scheduler will make the schedule.
-The ESDM Scheduler can be used in different steps of the process.
 ESDM has already an interface with NetCDF.
+The ESDM Scheduler will make the schedule, and it can be used in different steps of the process.
 Assuming the user can provide extra information in advance, ESDM can interact in the process before the user gives the data to the Cylc tool, using the information to construct an optimised workflow.
@@
 The proposed I/O-aware scheduler, called here EIOS (ESDM I/O Scheduler), is now introduced.
-The appropriate software stack involved in executing a workflow with EIOS is depicted in \Cref{fig:stages-io}.
-In the following, we describe the modifications we propose in this vision paper to the current software stack (\Cref{fig:stages}).
-The end-user has now to provide an additional file that covers the I/O information for each task and slight changes (discussed below) have to be made to the scripts that currently are executed.
+The appropriate software stack involved in executing a workflow with EIOS is depicted in \figref{pic/stages-io}.
+The suggested alterations can be seen in boxes pointed by red arrows, and the remaining components are the current state-of-the-art for general workflows (\figref{pic/stages}).
 
-\begin{figure}[H]
-  \centering
-  \includegraphics[scale=1.4]{stages-io}
-  \caption{Software stack and stages of execution with our I/O-aware scheduler -- EIOS}
-  \label{fig:stages-io}
-\end{figure}
+The end-user has now to provide an additional file that covers the I/O information for each task, and slight changes (discussed below) have to be made to the scripts currently being executed.
+Here, we use ESDM \cite{esdm} as the middleware that aims to exploit heterogeneous system environments and Cylc \cite{8675433} as the workflow engine.
+In the following, we describe the modifications we propose in this vision paper in each stage of the software stack.
+
+\fig{scale=1.4}{pic/stages-io}{Software stack and stages of execution with our I/O-aware scheduler -- EIOS}
 
@@
 
+  \item \textbf{Scientist} In addition, now the user also has to provide data dependencies and characteristics.
+
   \item \textbf{Cylc}
-  %Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
-  Now, Cylc calls EIOS to identify potential optimisations in the schedule.
+  Cylc invokes EIOS to identify potential optimisations in the schedule before generating the Slurm script.
   EIOS may decide that subsequent jobs shall be placed on the same node or to reorder the execution of some jobs.
@@
 
-  \item \textbf{EIOS}
-  The proposed system reads the information about the workflow (original and I/O suit files) and acts depending on the stage of the execution.
+  \item \textbf{\color{red}{EIOS}} reads the information about the workflow (original and I/O workflow configuration files) and acts depending on the stage of the execution.
   EIOS consists of several subcomponents:
+
     \begin{itemize}
+
       \item The high-level scheduler that interfaces with Cylc.
-      \item A tool that is run inside a script of a specific job/task -- as a replacement to the \texttt{cylc cycle-point} call.
-      It generates pseudo file names that are used by the applications with ESDM support.
-      \item A data management service (not shown on the figure) that purges or migrate data according to the user life cycle specification and the novelty information from EIOS.
+
+      \item A tool that interacts with Cylc and calls job scripts to generate pseudo-file names used by applications with ESDM support.
+
+      \item A data management service (not shown on the figure) that purges or migrate data according to information from EIOS, e.g., the user life cycle specification.
+
     \end{itemize}
-  EIOS components use the \texttt{esdm.conf} file that contains information about the available technology in the data centre and it provides the information on how to prioritise I/O targets.
+
+  EIOS components use ESDM through an \texttt{esdm.conf} file that contains information about the available technology in the data centre and its I/O characteristics which contains information on how to prioritise I/O targets.
 
   \item \textbf{Slurm}
-  %Schedules the execution of the job considering the scheduling policy of the data centre.
-  Cylc may now have added a constraint to place a job on the same node as a previous job and that is now handled by a modified Slurm.
-  Also, if migrations have to be performed before or after a job, Slurm will administer them.
+  Cylc may now have added a constraint to place a job on the same node as a previous job, and that is now handled by a modified Slurm.
+  Also, if migrations have to be performed, Slurm will administer them according to the specification in the job script.
 
   \item \textbf{Job}
-  A job might be scheduled on the same nodes than a previous job to utilize local storage.
-  %The user-provided script is executed on a single node assigned to the job.
-  % The script has set environment variables containing information about the Cylc execution (e.g., a task in the workflow) and the Slurm job (e.g., number of nodes).
+  A job might run on the same node than a previous job to utilise local storage.
 
   \item \textbf{Script}
-  %The shell-script executes the sequence of commands.
-  % Typically, to create a filename, the script calls Cylc commands with a template provided by the user and feeds this information to the application.
-  % This ultimately defines the storage location to be used.
-  % A script typically consists of multiple fine-grained commands and may even call multiple parallel applications.
-  The way the filenames are generated is now different. A replacement command invokes EIOS to generate a pseudo filename.
+  Filenames are now generated by a replacement command that calls EIOS to create a pseudo filename.
   This filename will encode additional information for ESDM about how to prioritise data placement according to data access.
@@
   \item \textbf{Application}
-  %The user-specified parallel application runs.
-  The application may either use NetCDF with ESDM support or ESDM directly to access datasets.
-  Hence, in \Cref{fig:layers}, the HDF5 layer is replaced with ESDM.
+  The application may either use XIOS, NetCDF with ESDM support or ESDM directly to access datasets.
+  Hence, in \figref{pic/layers-xios}, the HDF5 layer is replaced with ESDM.
   ESDM loads the \texttt{esdm.conf} file that contains the information about the available storage backends and their configuration.
-  % Knowing the potential targets, ESDM schedules application-local I/O.
-  ESDM extracts the long-term schedule information from the generated filenames and considers it during the I/O scheduling to exploit the data locality between tasks.
+  ESDM extracts the long-term schedule information from the generated pseudo filenames and employs it during the I/O scheduling to optimise the storage considering data locality between tasks.
 
@@
 
+The changes proposed in this work increase the opportunity for smarter scheduling of computing and storage in heterogeneous storage environments considering characteristics of workflows and data.
 Organising the data placement on storage tiers is typically performed manually by the developer/user or via policies, leading to suboptimal optimisations.
-%On top of that, it is a tedious, non-portable, and error-prone task.
-%Any hard-coded decision has significant implications for the performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
-Manual and hard-coded workflows cannot handle changes in the environment and are error-prone.
+Manual and hard-coded workflows cannot handle changes in the environment. On top of that, it is a tedious, non-portable, and error-prone task.
+Any hard-coded decision has significant implications for the performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
+By increasing the abstraction level for scientists, not only tedious manual optimisation could be automatised, but also novel strategies for further optimisations can be harnessed.
 Therefore, users must be able to express their workflows in an abstract fashion that allows the system to generate (near-)optimal execution plans and monitor their execution.
 
-\section{Extra Figures}
+\section*{Acknowledgements}
 
-\begin{figure}[H]
-  \centering
-  \includegraphics[scale=0.6]{workflow1-v2}
-  \caption{Workflow 1.}
-  \label{fig:work1}
-\end{figure}
+\small
+This project is funded by the European Union's Horizon 2020 research and innovation programme under grant agreement No. 823988.
 
-\begin{figure}[H]
-  \centering
-  \includegraphics[scale=0.6]{workflow2-v2}
-  \caption{Workflow 2.}
-  \label{fig:work2}
-\end{figure}
+\openaccess
 
-\begin{figure}[H]
-  \centering
-  \includegraphics[scale=0.6]{workflow-all}
-  \caption{Workflow All.}
-  \label{fig:work}
-\end{figure}
-
-\begin{comment}
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{cc}
-  \includegraphics[scale=0.6]{workflow1-v2} &
-  \includegraphics[scale=0.6]{workflow2-v2}
-  \end{tabular}
-  \caption{Workflows 1 and 2.}
-  \label{fig:work1}
-\end{table}
-
-\end{comment}
-
-\section{Specification of the Workflow from User Perspective}
-
-From the user perspective, a general workflow can be described by three categories described in \Cref{tab:workflow}.
-
-\begin{table}[H]
-\centering
-\begin{tabular}{|m{2cm}|m{5.5cm}|m{3.5cm}|m{3.5cm}|}
-\hline
-& \parbox{5.5cm}{\centering \textbf{Input}} & \parbox{3.5cm}{\centering \textbf{Computing}} & \parbox{3.5cm}{\centering \textbf{Products}} \\ \hline \hline
-
-Description & information inserted in the workflow. & Tasks specified by the user. & Possible outcomes of the tasks. \\ \hline
-
-\jk{?} File & \texttt{input.ds} & \texttt{tasks.ds} & \texttt{output.ds} \\ \hline
-
-Composed of & Datasets/Scripts & Datasets/Scripts & Datasets \\ \hline
-
-Information &
-
-\begin{tabular}{m{5cm}}
-\textbullet\ Dates for the experiments, how many times the workflow has to run, etc.\\
-\textbullet\ Dependencies among the tasks.\\
-\textbullet\ Information about the lifetime of the final outcome.\\
-\textbullet\ Steps in which manual decisions have to be made.\\
-\textbullet\ {{\color{cyan}{I/O information. (Where the datasets will be stored in each step of the workflow.)}}}\\
-\end{tabular}
-
-&
-
-\begin{tabular}{m{3cm}}
-\textbullet\ The data input.\\ \\
-\textbullet\ The data output.\\ \\
-\textbullet\ The data lifetime.\\ \\
-\textbullet\ The dependencies regarding the other tasks.\\ \\
-\textbullet\ {{\color{cyan}{I/O information.}}}
-\end{tabular}
-
-&
-
-\begin{tabular}{m{3cm}}
-\textbullet\ Datasets that will be used by other tasks.\\ \\
-\textbullet\ Datasets that will generate the final output.\\
-\end{tabular}
-
-\\ \hline
-\end{tabular}
-\caption{Components of the workflow in the user perspective}
-\label{tab:workflow}
-\end{table}
-
-For this work, Cylc is an essential tool regarding not only the execution of the workflow but the exchange of information with the EIOS about I/O and storage.
-Cylc receives, as predefined, information about the storage and dependencies of the datasets.
-The current information is used to construct and run the complete workflow, but it is not employed in optimising intermediate steps.
-The configuration is already predefined, and it cannot change.
-For instance, if it is established that Task\,2 depends on Task\,1, then we have to wait for Task\,1 to be completed to start running Task\,2.
-Thinking what about I/O, no decision is being made about an optimal way to store the data.
-Consider, as an example, that the datasets from \texttt{product1.ds} (the final product of task 1) are divided into \texttt{product1-final.ds} and \texttt{product1-task2.ds}.
-Then, it is clear that \texttt{product1-task2.ds} will be required to run Task\,2, and, because of that, should be put in fast storage to quickly being retrieved, or even kept in local storage.
-The \texttt{product1-final.ds} dataset, however, can go directly to tape, if the lifetime of the final output has to be stored for a long time, but no access is required from the current workflow.
-The bottom line here is: several options can be optimised if we know the whole workflow in advance.
-
-\begin{comment}
-
-and computer-aided Research Development and Engineering (RD\&E)
-
-and also on small-scale systems.
-
-For example, Big data tools integrate computing and storage capabilities into a holistic solution demonstrating the benefit of tight integrating for data-driven workflows.
-
-%-- the quality of service features available in hardware is rarely used.
-
-Cloud definition of desired state and contracts.
-
-Indeed many research prototypes address subproblems
-* But not all aspects together
-* Competing approaches; the standardisation we describe here does not compete!
-
-In this paper, we purposely do not mention too many of the ongoing RD\&E in storage and computing.
-
-To wrap up, NGI provides a high-level view on data and processing on heterogeneous platforms while shielding the user from execution details.
-
-Open ...
-
-%Kubernetes
-
-\begin{description}
-
-\item{NFS} Relatively slow, persistent file system - Nightly incremental backups
-
-\item{Lustre} Large, fast distributed scratch file system - No backups
-
-\url{https://info.hpc.sussex.ac.uk/hpc-guide/storage.html}
-
-\end{description}
-
-%, and data products are generated, which are processes that are executed outside of the first simulation.
-
-\end{comment}
-
-\bibliographystyle{alpha}
 \bibliography{paper}
-
 \end{document}

### 2020-02-06 paper/paper.tex
First version

@@
 \usepackage{comment}
+\usepackage{float}
 
@@
 We believe workflows composed of data, computing, and communication-intensive tasks should drive the interfaces and hardware configurations to best support the programming models.
-The changes proposed in this work increase the opportunity of implementations for smarter scheduling of computing and storage in heterogeneous storage environments considering characteristics of workflows and data.
+The changes proposed here increase the opportunity of implementations for smarter scheduling of computing and storage in heterogeneous storage environments.
 
@@
 % -----------------------------------------------------------------------
-\section*{Introduction}
+\section{Introduction}  // superfri template requires section*
 \label{sec:intro}
@@
 Since performance is of crucial importance to large-scale workflows, careful attention must be paid to exploit the system characteristics of the target supercomputer.
-To obtain the best performance, a specific supercomputer typically requires substantial changes to the workflow to tailor the workflow to the particular machine.
 For instance, a data-driven workflow may benefit from using a heterogeneous set of computing and storage technology at the same time.
+To obtain the best performance, a specific supercomputer typically requires substantial changes to the workflow to tailor it to the particular machine.
 
@@
 As the complexity of systems increases and alternative storage and computing technologies provide unique characteristics, it becomes increasingly difficult even for experts to manually optimise resource usage in workflows.
-In many cases, modifications are not performed because: 1) they are labour intense: any change to the workflow requires careful validation which may not pay off for small scale runs, and 2) users are not aware of the potential of the complex system.
+In many cases, modifications are not performed because: 1) They are labour intense: any change to the workflow requires careful validation which may not pay off for small scale runs, and 2) Users are not aware of the potential of the complex system.
 
 In this paper, we illustrate how knowing the Input/Output (I/O) characteristics of workflow tasks and overall experimental design help to optimise the execution of climate and weather workflows.
-This enlightenment may increase the performance, throughput and cost-efficiency of the environments, providing an incentive to users and data-centres that cannot be neglected any longer.
-Our approach intends to reduce the burden on researchers and, at the same time, optimise the decisions about jobs running in HPC systems.
+The access of this information may increase the performance, throughput and cost-efficiency of the environments, providing an incentive to users and data-centres that cannot be neglected any longer.
+Our approach intends to reduce the burden on researchers and, at the same time, optimise the decisions about jobs running on HPC systems.
 Once we have an automated decision about where the job will run and how the storage will be managed, climate and weather scientists can then run their workflows in any machine efficiently without prior knowledge about its architecture.
@@
 
-In this section, we describe how the environment workflows are executed in the typical software stack.
+In this section, we describe how the environment workflows are executed in a typical software stack.
 
@@
 Each node provides an SSD for local storage, and DKRZ has additionally two shared Lustre file systems with different performance characteristics.
-Individual users and projects are mapped to one file system explicitly, and users can access it as work or scratch file system: While data is kept on the work file system indefinitely, its available space is limited by a quota.
-The scratch file system allows to store more data, but data is automatically deleted after some time.
+Individual users and projects are mapped to one file system explicitly, and users can access it as work or scratch file system.
+While data is kept on the work file system indefinitely, its available space is limited by a quota.
+The scratch file system allows storing more data, but data is automatically deleted after some time.
 
@@
 
-A typical I/O stack for a single parallel application like a climate model is shown in \figref{pic/layers-xios}.
-
 \fig{width=0.2\textwidth}{pic/layers-xios}{I/O path for an MPI-parallel application}
 
-In our example, we assume the application is parallelised for performance reasons and uses MPI to run the application.
+A typical I/O stack for a single parallel application like a climate model is shown in \figref{pic/layers-xios}.
+In our example, we assume the application is parallelised for performance reasons using MPI.
 It may use XIOS to gather data from individual fields which then uses NetCDF4 to store the data as a file.
 XIOS provides a domain-level semantics and ...\jk{TODO}
-Under the hood, NetCDF4 used the HDF5 API and file format.
+
+Under the hood, NetCDF4 uses the HDF5 API and file format.
 Internally, HDF5 uses MPI again and its data types to specify the nature of the data stored.
-Finally, data is stored on a parallel file system like Lustre, which, on the server-side, stores data in the local file system as block devices on storage media such as SSDs and HDDs.
+Finally, data is stored on a parallel file system like Lustre which, on the server-side, stores data in the local file system as block devices on storage media such as SSDs and HDDs.
 
@@
 Naturally, the application that inputs previously generated data must use a compatible API to be able to read the specific data format.
-In this figure, for example, XIOS is beneficial for parallel I/O and a process might directly rely on the NetCDF API to read previous data.
+In this figure, for example, XIOS is beneficial\lr{beneficial?} for parallel I/O and a process might directly rely on the NetCDF API to read previous data.
 
@@
 After that, the user enacts a workflow engine to start the workflow.
-In the following, each stage of the execution is described further.
+In the following, each stage of the execution is further described.
 
@@
   Once a task can be executed (dependencies are fulfilled), the workflow engine generates a Slurm script on the fly with the required metadata that will load the user-defined script for each task.
-Then, Slurm runs to queue the job.
+Then, Slurm queues the job.
 
-  \item \textbf{Slurm} \cite{Jette02slurm:simple} schedules the execution of the job, considering the scheduling policy of the data centre.
+  \item \textbf{Slurm} plans the execution of the job, considering the scheduling policy of the data centre.
   Once the job is ready to be dispatched, i.e., resources are available, and the job priority is the highest, it is started on the supercomputer.
@@
   \item \textbf{Job} runs the user-provided script on one single node assigned by Slurm.
-  The script has set environment variables containing information about the execution (e.g., a task in the workflow) and the Slurm job (e.g., number of nodes).
+  The script set environment variables containing information about the execution (e.g., a task in the workflow) and the Slurm job (e.g., number of nodes).
 
-  \item \textbf{Script} typically calls commands to create a filename with a template provided by the user and feeds this information to the application which, ultimately, defines the storage location.
-  A script typically consists of multiple fine-grained commands and may even call multiple parallel applications.
+  \item \textbf{Script} typically consists of multiple fine-grained commands to create a filename with a template provided by the user.
+The information is fed into the application which, ultimately, defines the storage location.
+%  and may even call multiple parallel applications.
 
   \item \textbf{Application} runs taking the generated filenames that the script has set.
-  The user-specified parallel application may use XIOS \cite{xios} or NetCDF \cite{Jenter92netcdf:a} to perform the I/O.
+  The user-specified parallel application may use XIOS or NetCDF to perform the I/O.
 
 \end{enumerate}
+
+\subsection{Cylc}
+
+\lr{Intro}
+
+Cylc is in charge to execute and monitor workflows in which each step is submitted to the batch scheduler of a data centre.
+ %, and the relevance of heterogeneous infrastructure to run workflows.
+%We focus on high-level considerations to illustrate mandatory concepts.
+
+Consider the Cylc workflow for a toy monthly cycling workflow in \figref{pic/cylc1} \cite{8675433}.
+In this workflow, a warm-cycled atmospheric model (\textbf{model}) simulates the physics from a current state to predict the future after a timestep, for example, six hours.
+This process is repeated in the model to simulate month and years into the future.
+Once the simulation of any month is computed, the data for this month becomes available and can be analysed.
+A user has specified the detailed workflow in a Cylc configuration file (\texttt{conf.rc}).
+In this workflow, the \textbf{model} is followed by postprocessing (\textbf{post}), forecast verification (\textbf{ver}), and product generation (\textbf{prod}) tasks.
+Also, consider, in addition, a task \textbf{check} that compares some verification metric against products from two cycles earlier.\lr{Needed?}
+
+\fig{width=0.9\columnwidth}{pic/cylc1}{Example workflow}
+
+\figref{pic/cylc1} presents the workflow and the dependencies among the tasks, but it is missing the information about I/O and storage.
+Decisions about storage must take into consideration the architecture in which the workflow will run.
+The developers make the configuration about parallelism and where the data is stored in a script file that is executed for each specific task.
 
@@
 
-Usually, the scripts for tasks in climate and weather workflows define how the data is stored on the available storage systems.
-What happens in many current workflows is that they ignore the benefits of using multiple file systems concurrently or exploiting the data locality between tasks by co-locating the tasks.
+Usually, the scripts for tasks in climate and weather workflows define how the data is stored on the available system.
+What happens in many current workflows is that they ignore the benefits of using multiple file systems concurrently and exploiting the data locality between tasks to co-locating them.
 On top of that, in the current state-of-the-art, scientists have to optimise the available storage resources intuitively and insert the information about this decision-making process manually.
 Once we have the complete knowledge about how the workflow is expected to run, we can optimise steps that were before predefined by the user, usually considering only options for feasibility.
-Machine Learning has been proven effective in replacing human-beings, and it may adapt the workflow dynamically in case of unexpected situations (the system is down, nodes are unresponsive, etc.).
 
-Consider that the machine has three file systems: a fast \textbf{scratch} file system on which data may reside only for a week, a slower \textbf{work} file system, and each node has a \textbf{local} file system.
-What happens in many current workflows is that they use the only \textbf{work} and \textbf{scratch}.
+Consider, for instance, one machine with three file systems: a fast \textbf{scratch} file system on which data may reside only for a week, a slower \textbf{work} file system, and each node has a \textbf{local} file system.
+What happens in many current workflows is that they only use \textbf{work} and \textbf{scratch}.
 When a task is set to computing, the corresponding dataset is moved to \textbf{scratch}, processed, and the resulting datasets are transferred back to \textbf{work}.
-If the \textbf{scratch} filesystem achieves its capacity, then the datasets move back to \textbf{work} and keep running the task until it is finished.
+If the \textbf{scratch} filesystem achieves its capacity, the datasets move back to \textbf{work} and keep running the task until it is finished which might be inefficient.
 If the I/O planning is made a priori, this situation can be avoided.
@@
 In the end, this dataset might be an intermediate product that can then be deleted.
-Alternatively (\figref{fig:lifecycle2}, it could be stored on the \textbf{scratch} file system immediately and accessed there.
+Alternatively (\figref{fig:lifecycle2}), the dataset could be stored on the \textbf{scratch} file system immediately and accessed there.
 However, that would require that the last read access happens before files on \textbf{scratch} are automatically removed.
-The last scenario (\figref{fig:lifecycle3}) presents the case where data is created on \textbf{work} by a task.
-Still, then it is copied to a \textbf{local} node to allow read accesses of subsequent tasks which might be beneficial for small random accesses.
+The last scenario (\figref{fig:lifecycle3}) presents the case where the dataset is created on \textbf{work} by a task and it is copied to a \textbf{local} node.
+This local node allows read accesses of subsequent tasks which might be beneficial for small random accesses.
 However, it would require that the subsequent tasks are placed on the same machine where data is now stored.
@@
 
-Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of files on a particular storage system and, usually, moving data between storage by using scripts.
+Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of files on a particular storage system) and, usually, moving data between storage by using scripts.
 One limitation here is that decisions about how data structures are mapped and packaged into files are made once by the producing application, and cannot be changed without manual intervention by a downstream application.
@@
 A policy system (e.g., deployed on a burst buffer) aims to simplify the data movement for the user, but typically migrates objects in the coarse granularity of files.
-However, the semantical information that can be used by a policy system to make decisions is limited, e.g., data location, file extension, age of the file, etc.
+However, the semantical information that can be used by this type of system to make decisions is limited, e.g., data location, file extension, age of the file, etc.
 
@@
 
+Slurm \cite{Jette02slurm:simple}
+
+ESDM \cite{esdm}
+
+Cylc \cite{8675433}
+
+XIOS \cite{xios}
+
+NetCDF \cite{Jenter92netcdf:a}
+
+Compression \cite{TDTSOCAFQC17}
+
 \section{Vision for I/O-Aware Workflows}
@@
 
-There are several approaches to implement the technology for this vision, and changes need to be made to software components to realise it.
-In the next sections, we will discuss a specific design for our transitional roadmap considering climate and weather workflows and the tools scientists from this field already use in their routine research.
-
-Applied scientists should not spend time understanding hardware characteristics, but using their time to develop their work, let the computer do the job the most straightforward way as possible, and just collect and analyse the results.
-However, nowadays, to be able to run a job in a HPC environment efficiently, researchers have to have profound knowledge about their workflow, which is expected, but also about decisions regarding storage, communication, and computing.
-
 Data system do not respect the importance of data depends at all.
 Aspects like costs to reproduce the data (can it be reproduced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow are not considered.
-Value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability.
+Value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability.\lr{kind of lost here too}
+
+There are several approaches to implement the technology for the vision proposed in this work, and changes need to be made to software components to realise it.
+In the next sections, we will discuss a specific design for our transitional roadmap considering climate and weather workflows and tools scientists from this field already use in their routine research.
+Applied scientists should not spend time understanding hardware characteristics, but using their time to develop their work, let the computer do the job as most straightforward as possible, and just collect and analyse the results.
+However, nowadays, to be able to run a job in a HPC environment efficiently, researchers have to develop profound knowledge about their workflow, which is expected, but also about decisions regarding storage, communication, and computing.
 
@@
 Once we have an automated decision about where the job will run and how the storage will be managed, scientists can then run their workflows in any machine without further interactions and even without previous knowledge about the machine architecture.
-
 Our vision for I/O-aware workflows requires two additional information.
@@
 In general, climate and weather workflows allow specifying tasks and the dependencies among them.
-We aim to enable enriching this information with characteristics for input, and output, i.e., the datasets.
+We aim to enable enhancing the current information with characteristics for input, and output, i.e., the datasets.
 An example workflow containing input datasets and (intermediate) products is illustrated in \figref{pic/workflow}.
@@
 This last step represents the typical uncertainty of data reuse, i.e., it is unclear how Product\,3 will be used further.
-In the approach proposed in this work, each task is annotated with the required input datasets and the generated products with their characteristics.
-Each data product is annotated with metadata such as data life cycle information, e.g., the value of data and how long should it be kept.
-
-By monitoring I/O accesses, we can automatically draft such I/O descriptions on behalf of the user to simplify the specification.
-All we need is to run one cycle of the specific workflow, considering that a workflow cycle is executed many times in climate and weather; this is a cheap approach.
-From the recorded I/O accesses, we analyse the I/O patterns and dependencies associated with it to generate the configuration file.
+In the approach proposed in this work, each task is annotated with the required input datasets and the generated products must include metadata such as data life cycle information, the value of data, and how long should it be kept.
 
@@
 
+\lr{Replicate this figure and explain the idea of recomputing of data.}
+
+D1 D2 P1 P2 P3 + arrows
+
+D1 + D2 (small) + P1 + P2 (short) + P3 (huge) => P3 might be recomputed
+
 \subsection{System Information}
 
-While many optimisations are possible once the abstraction is in place, here we focus on utilising a heterogeneous storage environment.
-The optimisations we discuss are related to the life cycle of datasets and the placement of such datasets into specific storage according to system performance characteristics and the workflow specification.
+% Here we focus on utilising a heterogeneous storage environment.
+While many optimisations are possible once an abstraction is in place, the optimisations we discuss here are related to the life cycle of datasets and the placement of such datasets into specific storage according to system performance characteristics and the workflow specification.
+For achieving that, the system information shall comprise of all available storage systems, the system topology, and details of each of the required components.
+Simplified and complex models of the components can be included to approximate performance for specific I/O patterns allowing to predict performance behaviour.
+It is expected that the data centre (or expert users) can create such a configuration file, e.g., by using vendor-provided information or by executing benchmarks.
+With this information, the I/O scheduler can make the placement and migration decisions for individual datasets and manage their relocation during their life cycle, as discussed in \Cref{sec:datamanagement}.
+This separation of concerns allows us to abstract from the workflow what is essential and what a machine should optimise to ensure that the resources available are being used smartly.
+\lr{Check abstract!}
 
-The system information shall comprise of all available storage systems, the system topology, and details of each of its required components.
-This may include simplified or complex models of the components to approximate performance for specific I/O patterns allowing to predict performance behaviour.
-It is expected that the data centre (or expert users) can create such a configuration file, e.g., by using vendor-provided information or by executing benchmarks.
-With this information, the I/O scheduler can make the placement and migration decisions for individual datasets and manage their relocation during their life cycle, as discussed in \figref{sec:datamanagement}.\lr{Check ref.}
-This separation of concerns allows us to abstract from the workflow what is essential and what a machine should optimise to ensure that the resources available are being used smartly.
+\subsection{Data Movement}
 
@@
 
-In this paper, we focus on utilising a heterogeneous storage environment instead of exhaustively discussing the potential that is unleashed with the right level of abstraction in place by researchers that exploit the available opportunities.
+In this paper, we focus on utilising a heterogeneous storage environment instead of exhaustively discussing the potential that is unleashed with the right level of abstraction in place.
+Researchers are welcome to exploit the available opportunities.\lr{Lost.}
 
-Consider the hypothetical case to have just two tasks A and B that need to be repeated many times:
+\lr{Combine next two paragraphs}
+
+Consider the hypothetical case of having just two tasks A and B that need to be repeated many times.
 Datasets generated by Task\,\textbf{A} are used by Task\,\textbf{B} and hence both tasks must have access to the same data in the system.
@@
 Such configuration was done manually and with restricted information about the storage and system architecture.
-It would be interesting to explore storage options for the datasets and, e.g., have different cycles stored placed in different storage systems.
-If the lifetime of such dataset is, for instance, two cycles, alternate the datasets into two scratch file systems is something that would be a simple job an I/O-aware scheduler.
-However, manually, that implies at having at least two scripts\footnote{Complicated scripts would have allowed changing the storage type depending on the cycle.
-Still, it is a significant burden to the user.} for each task with information about the different storage.
+It would be interesting to explore storage options for the datasets and, e.g., have different cycles placed in different storage systems.
+If the lifetime of such dataset is, for instance, two cycles, alternate the datasets into two scratch file systems is something that would be a simple job for an I/O-aware scheduler.
+However, currently, that implies in having at least two scripts\footnote{Complicated scripts would have allowed changing the storage type depending on the cycle.
+Still, it is a significant burden to the user.} for each task with information about the different storage placement.
 
@@
 Assume you have, for instance, five cycles of the model running with all the necessary post-processing steps.
-Typically, the user has written the scripts for each task by defining a filename using the same prefix that is then placed at the same storage type every time.
+Typically, the user has written the scripts for each task by defining a filename using the same prefix that places the dataset at the same storage type every time.
 That happens because all datasets generated by Task \textbf{model} will be used by Task \textbf{post\,prod}, so they should stay on the same node.
@@
 Similarly, the scheduler may decide to couple subsequent steps of a workflow directly without storing intermediate data products on persistent storage.
-Finally, data might be replicated but enables the system to rerun parts of the workflow in case of a data loss.
+Finally, data might be replicated by enabling the system to rerun parts of the workflow in case of a data loss.
 
-Ultimately, by abstracting and automatizing these tasks, we provide a smarter route taking into consideration the architecture available and the way the datasets are stored during the intermediate steps of the workflow.
+Ultimately, by abstracting and automatising the workflow, we provide a smarter route taking into consideration the architecture available and the way the datasets are stored during the intermediate steps of the workflow.
 
-\fig{width=0.8\columnwidth}{pic/cycle4}{Four stages workflow example}
+\fig{width=0.8\columnwidth}{pic/cycle4}{Five stages workflow example}
+
+\lr{Color the pic and change the definition of the other pic.}
 
@@
 
-\begin{itemize}
+\begin{description}
 
-\item Abstraction: The user does not have to worry about the architecture that the workflow will run, removing the specialist from the decision-making in the storage process.
+\item[Abstraction] The user does not have to know the architecture that the workflow will run, removing the specialist from the decision-making process.
 
-\item Optimisation: The workflow will be optimised specifically for the available system infrastructure considering potentially runtime characteristics and storage resilience which will increase performance and system utilisation.
+\item[Optimisation] The workflow will be optimised specifically for the available system infrastructure considering potentially runtime characteristics and storage resilience which will increase performance and system utilisation.
 By knowing the workflow behind the data creation, the system could even trade data security with potential recomputing opportunities to optimise the cost-efficiency.
@@
 
-\item Performance-portability: The user has to specify the I/O requirements only once for the tasks of a specific workflow, and the workflow can now run on any system without user intervention.
+\item[Performance-portability] The user has to specify the I/O requirements only once for the tasks of a specific workflow, and the workflow can now run on any system without user intervention.
 Even more, if the system characteristics change, e.g., it gets upgraded, an additional storage tier becomes available, or if storage degrades, the I/O-aware workflow could automatically adapt and make use of this new environment.
 
-\end{itemize}
+\end{description}
 
@@
 
-This section describes our approach to incrementally extend the current software stack for climate and weather workflows which realises parts of our vision.
+This section describes our first approach to incrementally extend the current software stack for climate and weather workflows which realises parts of our vision.
 To automatically make scheduling decisions, the software stack needs to:
@@
 
-\item Adapts the resulting workflow, individual scripts and application executions to consider the automatic I/O scheduler.
+\item Adapts the resulting workflow, individual scripts and application executions to consider the adaptative I/O scheduler.
 
@@
 
-\subsection{Cylc}
-
-Cylc is in charge to execute and monitor workflows in which each step is submitted to the batch scheduler of a data centre.
- %, and the relevance of heterogeneous infrastructure to run workflows.
-%We focus on high-level considerations to illustrate mandatory concepts.
-
-Consider the Cylc workflow for a toy monthly cycling workflow in \figref{pic/cylc1} \cite{8675433}.
-In this workflow, a warm-cycled atmospheric model (\textbf{model}) simulates the physics from a current state to predict the future after a timestep, for example, after six hours.
-This process is repeated in the model to simulate month and years into the future.
-Once the simulation of any month is computed, the data for this month becomes available and can be analysed.
-A user has specified a detailed workflow in a Cylc configuration file (\texttt{conf.rc}).
-In this workflow, the \textbf{model} is followed by postprocessing (\textbf{post}), forecast verification (\textbf{ver}), and product generation (\textbf{prod}) tasks.
-Also, consider, in addition, a task \textbf{check} that compares some verification metric against products from two cycles earlier.
-
-\fig{width=0.9\columnwidth}{pic/cylc1}{Example workflow}
-
-This workflow presents the dependencies among the tasks, but it is missing the information about I/O and storage.
-Decisions about storage must take into consideration the architecture in which the workflow will run.
-The developers make the configuration about parallelism and where the data is stored in a script file that is executed for each specific task.
+\lr{What comes next...}
 
@@
 \figref{fig:cycle1} introduces one cycle of the workflow in \figref{pic/cylc1} with a simplified notation.
-The arrows represent the tasks dependencies, but there is no information about the data that are generated.
+The arrows represent the tasks dependencies, but there is no information about data that are generated.
 Tasks \textbf{verification} and \textbf{post\,proc} depend on Task \textbf{model}, but does Task \textbf{model} needs to be fully completed to satisfy the dependencies?
@@
 
-The scenario presented in \figref{fig:cycle1} shows the same cycle of \figref{fig:cycle-io-dep}, but  extended information about datasets is added.
+\lr{Think about split the colours.  }
+
+The scenario presented in \figref{fig:cycle-io-dep} shows the same cycle of \figref{fig:cycle1}, but extended information about datasets is added.
 In this picture, we have the tasks dependencies, and we explicitly list the datasets dependencies regarding the tasks that generated them.
@@
 Datasets colours indicate which task will use their information and datasets in yellow represent information that will be used in a different cycle.
+From the diagram we can also notice that Task \textbf{model} will generate one dataset for Task \textbf{post\,proc}, one dataset for Task \textbf{prod\,gen}, two datasets for Task \textbf{verification}, and three datasets for later cycles.
 The tricky part when we consider only tasks and not the datasets in the workflow is that it may lead to the false impression that one task depends only on the immediately previous one.
 Note, for instance, that one dataset produced by Task \textbf{model} is later used in Task \textbf{prod\,gen}.
-That is not an immediate conclusion when we analyse just the arrows for the dependencies.
+That is not an immediate conclusion when we analyse just the arrows for the dependencies among tasks.
 It is clear that Task \textbf{model} is a dependency for Task \textbf{post\,prod}, which is then a dependency for Task \textbf{prod\,gen}.
-Still, the crucial information here is that any dataset currently being used in the current cycle can also be used later.
-From the diagram we can also notice that Task \textbf{model} will generate one dataset for Task \textbf{post\,proc}, one dataset for Task \textbf{prod\,gen}, two datasets for Task \textbf{verification}, and three datasets for later cycles.
+Still, the crucial information here is that any dataset being used in the current cycle can also be used later.
 
-The access to this information before the process starts can be used to optimise where compute jobs are executed and how these datasets will be stored during the workflow.
-This information today is static and manual, leaving opportunities for automatisation and optimisations open.
+The access to this information before the process starts can be used to optimise where compute jobs will be executed and how these datasets will be stored during the workflow.
+This information today is static and manual, leaving open opportunities for automatisation and optimisations.
 The idea here is to embrace the concept that tasks dependencies are really imposed by datasets dependencies.
-Once we can insert the datasets explicitly in the workflow, it is easier to see that many optimisations regarding storage can be made.
+Once we insert the datasets explicitly in the workflow, it is easier to see that many optimisations regarding storage can be made.
 
@@
 
-The user provides information about datasets required for input and the generated output in a separate file similarly to Cylc's workflow configuration file.
-This information contains dataset information for each Cylc task:
+The user now has to provide information about datasets required for input and the generated output for each Cylc task in a separate file similarly to Cylc's workflow configuration file.
 
@@
   \item[Name] The name of the field generated/required, which may include a pattern such as supported by Cylc.
+
   \item[Size] An estimate of the file size.
+
   \item[Lifetime] How long the data must be retained on storage (if at all).
+
   \item[Type] It may specify the type of data, i.e., checkpoint, diagnostics, temporary.
 
+\lr{Sketch a config file based on Cylc}
+
 \end{description}
-
-% Expressions will support generic variables such as the number of nodes a job runs...
-
-% \begin{itemize}
-% \item Task\,1 usually requires ten hours running in a pc with standard configuration\footnote{Intel Core i5, 8 GB RAM, and 500 GB internal storage drive, for instance.}, requires 100 GB of storage that has to be persistent for no more than one week.
-% \item Task\,2 usually requires two hours running in a pc with standard configuration, requires 150 GB of storage that has to be persistent for no more than two weeks week.
-% \item Task\,3 usually requires five hours running in a pc with standard configuration, and it does not require extra storage other than the storage for the final results.
-% \end{itemize}
 
@@
 
+\lr{Insert esdm.conf and explain.}
+
 \subsection{Smarter Scheduling}
@@
 The ESDM Scheduler will make the schedule, and it can be used in different steps of the process.
-Assuming the user can provide extra information in advance, ESDM can interact in the process before the user gives the data to the Cylc tool, using the information to construct an optimised workflow.
+Assuming the user can provide extra information in advance, ESDM can interact in the process before the user gives the data to the Cylc tool, using the gathered information to construct a temptive optimised workflow.
 Once Cylc builds the workflow, ESDM can interact again after the configuration of the scripts for both effective and efficient run of the parallel applications regarding the available storage and architecture.
+
+\lr{Explore potential for migration using IME.}
 
@@
 
-The end-user has now to provide an additional file that covers the I/O information for each task, and slight changes (discussed below) have to be made to the scripts currently being executed.
-Here, we use ESDM \cite{esdm} as the middleware that aims to exploit heterogeneous system environments and Cylc \cite{8675433} as the workflow engine.
-In the following, we describe the modifications we propose in this vision paper in each stage of the software stack.
+The end-user has now to provide an additional file that covers the I/O information for each task, and slightly changes (discussed below) have to be made to the current scripts``.
+Here, we use ESDM as the middleware that aims to exploit heterogeneous system environments and Cylc as the workflow engine.
+In the following, we describe the modifications we propose in this vision paper for each stage of the software stack.
 
@@
 
-  \item \textbf{Scientist} In addition, now the user also has to provide data dependencies and characteristics.
+  \item \textbf{Scientist} In addition, now the user also has to provide data dependencies and characteristics of the datasets and products generated.
 
-  \item \textbf{Cylc}
-  Cylc invokes EIOS to identify potential optimisations in the schedule before generating the Slurm script.
-  EIOS may decide that subsequent jobs shall be placed on the same node or to reorder the execution of some jobs.
-  Also, it may provide information for conducting data migration.
+  \item \textbf{Cylc} EIOS is invoked by Cylc to identify potential optimisations in the schedule before generating the Slurm script.
+  EIOS may decide that subsequent jobs shall be placed on the same node, reorder the execution of some jobs, and provide information for conducting data migration.
 
-  \item \textbf{\color{red}{EIOS}} reads the information about the workflow (original and I/O workflow configuration files) and acts depending on the stage of the execution.
+  \item \textbf{\color{red}{EIOS}} The ESDM I/O Scheduler reads the information about the workflow (original and I/O workflow configuration files) and acts depending on the stage of the execution.
   EIOS consists of several subcomponents:
-
     \begin{itemize}
 
-      \item The high-level scheduler that interfaces with Cylc.
+      \item The high-level scheduler that interfaces with Cylc.\\
 
-      \item A tool that interacts with Cylc and calls job scripts to generate pseudo-file names used by applications with ESDM support.
+      \item A tool based on Cylc to generate pseudo-file names used by applications with ESDM support.\\
 
-      \item A data management service (not shown on the figure) that purges or migrate data according to information from EIOS, e.g., the user life cycle specification.
+      \item A data management service (not shown on the figure) that migrate and purge data according to information from EIOS, e.g., the user life cycle specification.\\
 
     \end{itemize}
-
-  EIOS components use ESDM through an \texttt{esdm.conf} file that contains information about the available technology in the data centre and its I/O characteristics which contains information on how to prioritise I/O targets.
+  EIOS components use ESDM through the \texttt{esdm.conf} file \ref{aa}.
+This file contains information about the available technology in the data centre and its I/O characteristics to make decisions about `how to prioritise I/O targets.
 
   \item \textbf{Slurm}
-  Cylc may now have added a constraint to place a job on the same node as a previous job, and that is now handled by a modified Slurm.
+  Cylc may now have added an optimisation identified by EIOS and that is now handled by a modified Slurm.
   Also, if migrations have to be performed, Slurm will administer them according to the specification in the job script.
@@
   \item \textbf{Job}
-  A job might run on the same node than a previous job to utilise local storage.
+  A job might run on the same node than a previous job to utilise local storage. \lr{+migration}
 
@@
   This filename will encode additional information for ESDM about how to prioritise data placement according to data access.
-  Potentially, EIOS may generate an \texttt{esdm.conf} file for the specific application to adjust the mapping of data to available storage targets.
 
@@
 The changes proposed in this work increase the opportunity for smarter scheduling of computing and storage in heterogeneous storage environments considering characteristics of workflows and data.
+
 Organising the data placement on storage tiers is typically performed manually by the developer/user or via policies, leading to suboptimal optimisations.
+
 Manual and hard-coded workflows cannot handle changes in the environment. On top of that, it is a tedious, non-portable, and error-prone task.
-Any hard-coded decision has significant implications for the performance of both applications and systems because the optimal choices about file size and content may be very different for the data production and any subsequent data analysis phase.
+Any hard-coded decision has significant implications for the performance of both applications and systems because the optimal choices about file size and content may be very different for data production and any subsequent data analysis phase.
 By increasing the abstraction level for scientists, not only tedious manual optimisation could be automatised, but also novel strategies for further optimisations can be harnessed.
@@
 \end{document}
+
+Machine Learning has been proven effective in replacing human-beings, and it may also adapt the workflow dynamically in case of unexpected situations (the system is down, nodes are unresponsive, etc.).
+
+(3.3)
+
+By monitoring I/O accesses, we can automatically draft such I/O descriptions on behalf of the user to simplify the specification.
+All we need is to run one cycle of the specific workflow, considering that a workflow cycle is executed many times in climate and weather; this is a cheap approach.
+From the recorded I/O accesses, we analyse the I/O patterns and dependencies associated with it to generate the configuration file.
+
+\cite{TUIBIHWLSC19}
+
+3.2 end
+
+- Collect all ideas and enumerate them.
+
+- Decide on what we can actually will do.
+
+Potentially, EIOS may generate an \texttt{esdm.conf} file for the specific application to adjust the mapping of data to available storage targets.

### 2020-02-07 paper/paper-old.tex

--- 2020-01-29 paper/paper.tex
@@
   \item \textbf{Cylc}
-  %Cylc parses the suite file, generate dependencies and defines a schedule for the execution.
   Now, Cylc calls EIOS to identify potential optimisations in the schedule.
@@
   \item \textbf{Slurm}
-  %Schedules the execution of the job considering the scheduling policy of the data centre.
   Cylc may now have added a constraint to place a job on the same node as a previous job and that is now handled by a modified Slurm.
@@
   A job might be scheduled on the same nodes than a previous job to utilize local storage.
-  %The user-provided script is executed on a single node assigned to the job.
-  % The script has set environment variables containing information about the Cylc execution (e.g., a task in the workflow) and the Slurm job (e.g., number of nodes).
 
   \item \textbf{Script}
-  %The shell-script executes the sequence of commands.
-  % Typically, to create a filename, the script calls Cylc commands with a template provided by the user and feeds this information to the application.
-  % This ultimately defines the storage location to be used.
-  % A script typically consists of multiple fine-grained commands and may even call multiple parallel applications.
   The way the filenames are generated is now different. A replacement command invokes EIOS to generate a pseudo filename.
@@
   \item \textbf{Application}
-  %The user-specified parallel application runs.
   The application may either use NetCDF with ESDM support or ESDM directly to access datasets.
@@
   ESDM loads the \texttt{esdm.conf} file that contains the information about the available storage backends and their configuration.
-  % Knowing the potential targets, ESDM schedules application-local I/O.
   ESDM extracts the long-term schedule information from the generated filenames and considers it during the I/O scheduling to exploit the data locality between tasks.
